# Data Overview

We compiled a collection of validation datasets across diverse forest conditions for this specific analysis project. These datasets consist of coincident aerial point cloud data (including USGS 3DEP ALS, UAS-lidar, and UAS-SfM) paired with field-measured stem maps or plot inventories where species identities are known.

| Name | State | Forest Type | Point Cloud Data | Field Data |
| :--- | :--- | :--- | :--- | :--- |
| Penobscot Experimental Forest | Maine | northern mixed conifer | ALS | tree locations |
| Manitou Experimental Forest (N1) | Colorado | ponderosa pine | UAS-SfM | tree locations |
| Manitou Experimental Forest (Big Plot) | Colorado | western mixed conifer | UAS-SfM | tree locations |
| CSFS State Forest State Park | Colorado | lodgepole pine | UAS-lidar | plots |
| BLM Dawson Creek | Colorado | pinyon-juniper | UAS-SfM | tree locations |
| Ute Creek | Colorado | | UAS-SfM | plots |
| WFSI Fort Stewart | Georgia | southern yellow pine forest | UAS-lidar | plots |
| WFSI Elgin AFB | Florida | longleaf pine | UAS-lidar | plots |
| Wabikon Lake Forest Dynamics Plot (ForestGEO) | Wisconsin | northern mixed hardwood | ALS | tree locations |
| UC Santa Cruz Forest Ecology Research Plot (ForestGEO) | California | coastal mixed-evergreen & coast redwood | ALS | tree locations |
| Smithsonian Environmental Research Center (ForestGEO) | Maryland | mixed deciduous hardwood | ALS | tree locations |
| Smithsonian Conservation Biology Institute (ForestGEO) | Virginia | mid-atlantic deciduous | ALS | tree locations |

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # the cloud2trees
```

```{r pkg-ld, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Penobscot Experimental Forest (Maine)

Field data from this study site is from: 

* Kenefic, Laura S.; Allogio, Jeanette A.; Gerndt, Kathryn M.; Knapp, Rachel A.; Rogers, Nicole S.; Puhlick, Joshua J. 2025. Overstory tree and regeneration data from the "Silvicultural Effects on Composition, Structure, and Growth" study at Penobscot Experimental Forest. 3rd Edition. Fort Collins, CO: Forest Service Research Data Archive. [https://doi.org/10.2737/RDS-2012-0008-3](https://doi.org/10.2737/RDS-2012-0008-3)

>Data were collected on a series of permanent sample plots distributed across management units receiving various silvicultural treatments. Plot data are available in eight separate data sets. Overstory tree and sapling data include tree species, condition (e.g., merchantability status and cause of mortality), and diameter at breast height of living and dead trees, 1952 to 2023...Tree location data include the distance and azimuth from plot center of a subsample of trees, 2003 to 2023. Height and crown radius data include total height, height to crown base, and crown radius measurements for a subsample of live trees, 2003 to 2020 (heights) and 2003 to 2017 (radii)...Permanent sample plot location data include geospatial coordinates for permanent sample plots, 2018 and 2020...Future inventories are planned for this study indefinitely.

> 1/5th Acre Plots (were used to measure) Trees ≥ 4.5 inches DBH (with a) 52.7-foot plot radius

let's load in the plot data first to orient ourselves

```{r}
usfs_penobscotef_plots <- readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_PlotLocationsData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::filter(
    geodoubt == "N"
    , active == 1
    , !is.na(utm_east)
    , !is.na(utm_north)
  ) %>% 
  dplyr::mutate(
    plot_id = 
      mu %>%
        stringr::str_squish() %>% 
        stringr::str_c("-", as.character(plot))
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("utm_east", "utm_north"), crs = 26919, remove = F) %>% 
  dplyr::mutate(
    plot_area_ac = 1/5
    # 1 ha = 2.47105 acres
    , plot_area_ha = plot_area_ac/2.47105
    , plot_area_m2 = plot_area_ha*10000
    , plot_radius_meters = sqrt(plot_area_m2 / pi)    
  ) %>% 
  sf::st_buffer(.$plot_radius_meters)
# huh?
# usfs_penobscotef_plots %>% dplyr::count(projection)
usfs_penobscotef_plots %>% dplyr::glimpse()
```

we did a lot of data cleaning in there which we won't describe (you can read the code if you're interested). let's look at the plot locations...

```{r}
usfs_penobscotef_plots %>% 
  dplyr::select(plot_id,mu,plot,plot_area_ha) %>% 
  mapview::mapview(zcol = "mu", col.regions = sample(viridis::turbo(100)), layer.name = "plots", legend = F)
```






ForestGEO SCBI

```{r}
boundary_scbi_temp <- sf::st_read("../data/ForestGEO_SCBI/ForestGEO_SCBI_shapefiles/ForestGEO_grid_outline.shp") %>% 
  sf::st_union() 
boundary_scbi_temp %>% 
  mapview::mapview(layer.name = "SCBI")
boundary_scbi_temp %>% sf::st_centroid() %>% sf::st_transform(4326) %>% sf::st_coordinates()
# 38.893480, -78.145380
```






The `cloud2trees` package can be installed by following the directions listed in the [README file on GitHub](https://github.com/georgewoolsey/cloud2trees). If one is still experiencing difficulties installing the package, see the [setup guide](https://georgewoolsey.github.io/cloud2trees/articles/cloud2trees-setup.html) which details how to install the package using a virgin R instance.

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get cloud2trees
remotes::install_github(repo = "georgewoolsey/cloud2trees", upgrade = F)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```
