# Data Overview

```{r, include=F, warning=F, message=F, eval = F}
######################################################################################################################
######################################################################################################################
# https://doi.org/10.5281/zenodo.6637599
# Tallo database to build ITD segmentation ws functions by genus/forest type based on height-crown diameter?
######################################################################################################################
######################################################################################################################
# https://doi.org/10.5281/zenodo.6637599
# we developed the Tallo database, a collection of 498,838 georeferenced and taxonomically 
# standardized records of individual trees for which stem diameter, height and/or crown radius 
# have been measured. These data were collected at 61,856 globally distributed sites, spanning 
# all major forested and non-forested biomes. The majority of trees in the database are identified
# to species (88%), and collectively Tallo includes data for 5163 species distributed 
# across 1453 genera and 187 plant families.

######################################################################################################################
######################################################################################################################
# https://gdam-api-v2-782971006568.us-west1.run.app/docs#/Index/index
# fastfuels api to get species?
######################################################################################################################
######################################################################################################################
###########################################################
###########################################################
# here's the example py from the docs
###########################################################
###########################################################
# import pandas as pd
# import requests

# # 1. Prepare your input data
# input_df = pd.DataFrame({
#     "Lat": [48.1234, 48.1240],
#     "Lon": [-114.5678, -114.5680],
#     "HT": [25.5, 18.2],
#     "DIA": [None, 8.5]  # Missing fields will be imputed
# })

# # 2. Format the payload using orient="split"
# payload = {
#     "trees": input_df.to_dict(orient="split")
# }

# # 3. Send the request
# response = requests.post("https://<API_URL>/v1/predict/batch", json=payload)

# # 4. Reconstruct the output directly back into a Pandas DataFrame
# response_data = response.json()
# predictions_df = pd.DataFrame(**response_data["predictions"])

# print(predictions_df.head())

###########################################################
###########################################################
# here's the start of formatting data to send to api
###########################################################
###########################################################
validation_trees %>% 
  sf::st_transform(crs = sf::st_crs(4326)) %>% 
  dplyr::mutate(
    tree_lon = sf::st_coordinates(.)[,1]
    , tree_lat = sf::st_coordinates(.)[,2]
    , fastfuels_api = paste0(
      # https://gdam-api-v2-782971006568.us-west1.run.app/docs#/Index/index
      # fastfuels api to get species?
      # {
      #   "Lat": 48.1234,
      #   "Lon": -114.5678,
      #   "HT": 25.5
      # }
      "{"
        , "\"Lat\":" , scales::number(tree_lat, accuracy = 0.000001)
        , ",\"Lon\":" , scales::number(tree_lon, accuracy = 0.000001)
        , ",\"HT\":" , scales::number(field_tree_height_m, accuracy = 0.01)
      , "}"
      
    )
  ) %>% 
  dplyr::slice_sample(n=44) %>% 
  dplyr::slice_head(n=7) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(fastfuels_api)

# {"Lat":39.125813,"Lon":-105.115625,"HT":19.05}    # 39.125813,-105.115625
# {"Lat":39.125659,"Lon":-105.113507,"HT":18.90}    # 39.125659,-105.113507
# {"Lat":39.125840,"Lon":-105.112728,"HT":20.73}    # 39.125840,-105.112728
# {"Lat":39.126451,"Lon":-105.115426,"HT":2.44}   # 39.126451,-105.115426
# {"Lat":39.126394,"Lon":-105.115221,"HT":2.13}   # 39.126394,-105.115221
# {"Lat":39.124983,"Lon":-105.115187,"HT":15.54}    # 39.124983,-105.115187
# {"Lat":39.125319,"Lon":-105.115390,"HT":3.35}   # 39.125319,-105.115390


# {"Lat":39.127170,"Lon":-105.115781,"HT":20.12}
# {"Lat":39.127175,"Lon":-105.115772,"HT":10.06}
# {"Lat":39.127174,"Lon":-105.115760,"HT":7.01}"
# {"Lat":39.127124,"Lon":-105.115732,"HT":11.28}
# {"Lat":39.127110,"Lon":-105.115742,"HT":23.47}
# {"Lat":39.127099,"Lon":-105.115714,"HT":11.89}
# {"Lat":39.127073,"Lon":-105.115716,"HT":20.12}
```

We compiled a collection of validation datasets across diverse forest conditions for this specific analysis project. These datasets consist of coincident aerial point cloud data (including USGS 3DEP ALS, UAS-lidar, and UAS-SfM) paired with field-measured stem maps or plot inventories where species identities are known.

| Name | State | Forest Type | Point Cloud Data | Field Data |
| :--- | :--- | :--- | :--- | :--- |
| Penobscot Experimental Forest | Maine | northern mixed conifer | ALS | tree locations |
| Manitou Experimental Forest (N1) | Colorado | ponderosa pine | UAS-SfM | tree locations |
| Manitou Experimental Forest (Big Plot) | Colorado | western mixed conifer | UAS-SfM | tree locations |
| CSFS State Forest State Park | Colorado | lodgepole pine | UAS-lidar | plots |
| BLM Dawson Creek | Colorado | pinyon-juniper | UAS-SfM | tree locations |
| Ute Creek | Colorado | | UAS-SfM | plots |
| WFSI Fort Stewart | Georgia | southern yellow pine forest | UAS-lidar | plots |
| WFSI Elgin AFB | Florida | longleaf pine | UAS-lidar | plots |
| Wabikon Lake Forest Dynamics Plot (ForestGEO) | Wisconsin | northern mixed hardwood | ALS | tree locations |
| UC Santa Cruz Forest Ecology Research Plot (ForestGEO) | California | coastal mixed-evergreen & coast redwood | ALS | tree locations |
| Smithsonian Environmental Research Center (ForestGEO) | Maryland | mixed deciduous hardwood | ALS | tree locations |
| Smithsonian Conservation Biology Institute (ForestGEO) | Virginia | mid-atlantic deciduous | ALS | tree locations |

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # the cloud2trees
```

```{r pkg-ld, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's make a quick data frame of our project data which is organized by site

```{r}
project_data_df <-
  dplyr::tibble(
   data_dir = list.dirs("../data", recursive = F) 
  ) %>% 
  dplyr::mutate(
    data_site_nm = basename(data_dir)
  )
# project_data_df$data_site_nm
# make a list for the CHM, predicted trees, field trees, and plot spatial data we can add to
chm_list <- setNames( vector("list", nrow(project_data_df)), project_data_df$data_site_nm)
# chm_list
# names(chm_list)
pred_trees_list <- chm_list
ref_trees_list <- chm_list
ref_plots_list <- chm_list
```

## Eglin Air Force Base (Florida)

### Field Data

Field data from this study site is from: 

* Andrew Hudak, Seth Bigelow, Nuria Sánchez-López, and Vanessa Niemczyk. 2025. Field measurements of coarse wood, duff, litter, and trees for the Eglin Air Force Base 2023 fuels campaign. Wildland Fire Science Initiative (WFSI) [https://doi.org/10.60594/W4RG6P](https://doi.org/10.60594/W4RG6P)

>Field measurements collected before and after 2023 study fires at Eglin Air Force Base (AFB) to document fuel loads (large diameter fuels, litter depth, duff depth) and trees in a 3-year rough in sandhill longleaf pine, and the effects of study fires on these variables. Fuel loads, which were measured along transects, were co-located with tree observations, which were measured within fixed-radius plots. These field observations were also co-located with three-dimensional fuel and laser scan plots to allow for leveraging of these other data.

>After the study fires in March 2023, trees within 10 m of plot center were identified to functional group (pine, hardwood or species), measured for diameter at breast height with a diameter tape, measured for distance from outer bark to plot center, and measured for azimuth from plot center to tree. Some post-processing was done using lidar scans to distinguish between evergreeen and deciduous hardwoods. Saplings were tallied by ordinal quadrant (up to 5 m from plot center). Tree measurements were not taken in unit L25. In unit G20, plots HH01, HH02, HH03, HH04, HH05 were sampled pre-fire but not post fire. In unit G25, plots HH105, HH95, PP16 were sampled pre-fire but not post fire.

according to *Eglin_2022_Tree_and_Transect_Field_Protocol.pdf*, the post-fire sampling protocol for trees was:

I. Determine if plot radius is 8 or 11.3 meters and record on datasheet
  a. If there are 4 or more trees within an 8-meter radius, then the plot radius is 8-meters.
  b. If there are 0 to 3 trees within an 8-meter radius, then increase the plot radius to 11.3-meters.
II. Sample all trees ≥10 cm DBH within the plot radius and over from plot center 
III. Sample all pine saplings ONLY ≥2.5 to 10 cm DBH by recording:
  a. Radius (4 or 5.6m):
    i. If tree sampling radius was 8-meters, then sapling sampling radius = 4-meters
    ii. If tree sampling radius was 11.3-meters, then sapling sampling radius = 5.6-meters
  b. Quadrant = NE, SE, SW, or NW
  c. Count = How many of the same species in that quadrant
  d. Species = Longleaf pine (Pinus palustris), Loblolly pine (Pinus taeda), Sand pine (Pinus clausa) or TBD

for those interested, pre- and post-burn photographs for each plot were archived here: [https://doi.org/10.60594/W4GW2T](https://doi.org/10.60594/W4GW2T)

```{r}
# https://doi.org/10.60594/W4RG6P
wfsi_eglinafb_trees <-
  readr::read_csv(
    "../data/WFSI_EglinAFB/Eglin_Air_Force_Base_2023_fuels/data/trees.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  # dplyr::count(species)
  dplyr::filter(
    species != ""
    , !is.na(species)
    , !is.na(tree_x)
    , !is.na(tree_y)
  ) %>% 
  # spatial that
  # epsg from "data_dictionary.csv"
  sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = 26916, remove = F) %>% 
  # make a plot_id that is unique for all so we don't have to join by multiple columns
  # and for all sites we'll just use a plot_id
  dplyr::mutate(
    plot_id = stringr::str_c(unit,plot,sep = "_")
  )
# huh?
wfsi_eglinafb_trees %>% dplyr::glimpse()
# wfsi_eglinafb_trees %>% dplyr::select(dbh_cm) %>% summary()
```

let's check those field trees out to orient ourselves to the location

```{r}
wfsi_eglinafb_trees %>% 
  mapview::mapview(
    zcol = "species"
    , layer.name="WFSI Eglin AFB: field trees"
    , popup=F
  )
# wfsi_eglinafb_trees %>%
#   sf::st_union() %>%
#   sf::st_bbox() %>%
#   sf::st_as_sfc() %>%
#   sf::st_centroid() %>%
#   sf::st_transform(crs=4326) %>%
#   sf::st_coordinates()
# # 30.53834,-86.7296
```

what is the count of trees measured pre- and post-burn?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(burn_status)
```

so these are all post-burn samples. what is the distribution of species?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species) %>% 
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

how many plots have trees by unit?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::distinct(unit,plot) %>% 
  dplyr::count(unit)
```

we need to get the plot location points and buffer them by 10 m of plot center which was said to be the field sampling protocol to "*Sample all trees ≥10 cm DBH within the plot radius and over from plot center*" ...clear as mud!

```{r}
# https://doi.org/10.60594/W4T880
wfsi_eglinafb_plots <-
  readr::read_csv(
    "../data/WFSI_EglinAFB/Macroplot_Geographical_Location_Data/data/Geographical_Macroplot_Data_Eglin_2023.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::filter(
    !is.na(latitude_y)
    , !is.na(longitude_x)
    , !is.na(macroplot)
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("longitude_x", "latitude_y"), crs = 4326, remove = F) %>% 
  # converts proj to match trees and horizontal units to m
  sf::st_transform(sf::st_crs(wfsi_eglinafb_trees)) %>% 
  # "After the study fires in March 2023, trees within 10 m of plot center were identified..."
  sf::st_buffer(10) %>% 
  dplyr::mutate(
    plot_area_m2 = sf::st_area(.) %>% as.numeric()
    # this is good data mgmt: make sure to specify unit orientation in the plot data but not the tree data
    , burn_unit = burn_unit %>% stringr::str_remove("[EWSNewsn]$")
    # # another good practice is to format the plot id's to have padding of two zeros in the plot data but only padding of one zero in the tree data
    , macroplot = dplyr::case_when(
      stringr::str_sub(macroplot, 3, 4) == "00" ~ stringr::str_c( stringr::str_sub(macroplot,1,2), "0", stringr::str_sub(macroplot,5) )
      , stringr::str_sub(macroplot, 3, 3) == "0" ~  stringr::str_c( stringr::str_sub(macroplot,1,2), stringr::str_sub(macroplot,4) )
      , T ~ macroplot
    )
  ) %>% 
  dplyr::rename(
    plot = macroplot
    , unit = burn_unit
  )

# wfsi_eglinafb_plots %>% dplyr::glimpse()
# wfsi_eglinafb_plots$plot %>% unique()

# # check for plot and unit non-matches
# wfsi_eglinafb_trees %>%
#   sf::st_drop_geometry() %>%
#   dplyr::distinct(unit,plot) %>%
#   dplyr::anti_join(
#     wfsi_eglinafb_plots %>%
#       sf::st_drop_geometry() %>%
#       dplyr::distinct(unit,plot,plot_area_m2)
#     , by = dplyr::join_by(unit,plot)
#   ) %>%
#   dplyr::glimpse()
  
# lets keep only plots for which tree data exists since we know those were measured post-burn
wfsi_eglinafb_plots <- wfsi_eglinafb_plots %>% 
  dplyr::inner_join(
    wfsi_eglinafb_trees %>%
      sf::st_drop_geometry() %>%
      dplyr::distinct(unit,plot,plot_id)
    , by = dplyr::join_by(unit,plot)
  )
  
# wfsi_eglinafb_plots %>% dplyr::glimpse()

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
ref_plots_list[[nm_temp]] <- wfsi_eglinafb_plots
# ref_plots_list

```

let's quickly check the trees and plots to make sure they align

```{r}
# wfsi_eglinafb_trees %>% dplyr::glimpse()
# plot function
p_fn_temp <- function(nr){
    p <- wfsi_eglinafb_plots %>% dplyr::slice(nr)
    t <- wfsi_eglinafb_trees %>% 
      dplyr::inner_join(
          p %>% 
          dplyr::select(unit,plot) %>%
          sf::st_drop_geometry()
        , by = dplyr::join_by(unit,plot)
      )
    ggplot2::ggplot() + 
    ggplot2::geom_sf(
      data = t
      , mapping = ggplot2::aes(color = dbh_cm)
      , size = 2.5
    ) + 
    ggplot2::geom_sf(
      data = p
      , color = "gray33"
      , fill = NA
      , lwd = 1.5
    ) + 
    ggplot2::geom_sf(
      data = t
      , mapping = ggplot2::aes(color = dbh_cm)
      , size = 2.5
    ) + 
    ggplot2::scale_color_viridis_c(
      option = "mako"
      , limits = c(0, ceiling(max(wfsi_eglinafb_trees$dbh_cm,na.rm=T)*1.01))
      , direction = -1
      , name = "DBH (cm)"
    ) +
    ggplot2::labs(
      subtitle = paste0(p$unit, " | ", p$plot)
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
      plot.subtitle = ggplot2::element_text(size = 8, hjust = 0.5, face = "bold")
    )
  }

pl_temp <- 
  1:nrow(wfsi_eglinafb_plots) %>% 
  sample(12) %>% 
  purrr::map(p_fn_temp)
patchwork::wrap_plots(pl_temp,ncol=4,guides = "collect")
```

spot check of alignment between plot and tree spatial data looks good but we can quickly look at all plots and trees to see if any trees fall outside of the plot boundaries

```{r}
# wfsi_eglinafb_trees %>% dplyr::glimpse()
wfsi_eglinafb_trees <- 
  1:nrow(wfsi_eglinafb_plots) %>% 
  purrr::map(function(nr){
    p <- wfsi_eglinafb_plots %>% dplyr::slice(nr)
    t <- wfsi_eglinafb_trees %>% 
      dplyr::inner_join(
          p %>% 
          dplyr::select(unit,plot) %>%
          sf::st_drop_geometry()
        , by = dplyr::join_by(unit,plot)
      )
    # points outside?
    out <- t %>% sf::st_filter(p, .predicate = sf::st_disjoint)
      
    return(
      t %>% 
        dplyr::left_join(
          out %>% 
            sf::st_drop_geometry() %>% 
            dplyr::select(plot,unit,tree_number) %>% 
            dplyr::mutate(is_out_of_plot = T)
          , by = dplyr::join_by(plot,unit,tree_number)
        ) %>% 
        dplyr::mutate(
          is_out_of_plot = dplyr::coalesce(is_out_of_plot,F)
        )
    )
  }) %>% 
  purrr::list_rbind()

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
ref_trees_list[[nm_temp]] <- wfsi_eglinafb_trees
# ref_trees_list 

# are there any out?
wfsi_eglinafb_trees %>% sf::st_drop_geometry() %>% dplyr::count(is_out_of_plot)
```

uh-oh we have one tree outside of the plot

```{r}
wfsi_eglinafb_plots %>% 
  dplyr::mutate(rn=dplyr::row_number()) %>% 
  dplyr::inner_join(
    wfsi_eglinafb_trees %>% 
      dplyr::filter(is_out_of_plot) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(unit,plot)
    , by = dplyr::join_by(unit,plot) 
  ) %>% 
  dplyr::pull(rn) %>% 
  purrr::map(p_fn_temp)

```

right on the edge, but just outside based on a slight misalignment in GPS locations or in/out determination in the field

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Point Cloud Data

Andrew Hudak, & Benjamin Bright. (2026). *Airborne Lidar Data from the 2023 Department of Defense Wildland Fire Science Initiative Eglin Air Force Base Campaign*. WFSI Data Portal. [doi:10.60594/W4B017](https://doi.org/10.60594/W4B017)

>This dataset contains airborne light detection and ranging (lidar) data across burn units G20, G25, and L25 in Eglin AFB, FL. Kucera International acquired airborne lidar on January 28, 2023 (before prescribed fire), and on April 26, 2023 (after prescribed fire); prescribed fires occurred from March 11 to 19, 2023. Lidar data were acquired to characterize pre- and post-fire vegetation and estimate fire-caused biomass consumption. 

some other relevant notes about the data:

>Kucera International acquired airborne lidar data across burn units G20, G25, and L25 on January 28, 2023 (before prescribed fire), and across burn units G20 and G25 on April 26, 2023 (after prescribed fire); prescribed fires occurred from March 11 to 19, 2023. The lidar sensor, an Optech Galaxy T2000, was mounted on a PA-31 Piper Navajo Chieftain fixed-wing aircraft, which flew at an elevation of 1341 m (4400 ft) above ground level at a speed of about 150 knots, with a planned sidelap of 30%. The Optech Galaxy T2000 scanned with a laser pulse rate of 1600 kHz, a scan rate of 150 Hz, and a scan field of view of 30 degrees, which resulted in swath widths of 716 meters (2350 feet) and an average point density of 15 points per square meter.

>LAZ file horizontal distance units are meters, in the UTM Zone 16N NAD83 (2011) geographic coordinate system (EPSG 6345)

let's check out the raw ALS data

```{r}
wfsi_eglinafb_las_f_temp <- "../data/WFSI_EglinAFB/post_fire_laz_apr2023"
wfsi_eglinafb_las_ctg_temp <- lidR::readLAScatalog(wfsi_eglinafb_las_f_temp)
wfsi_eglinafb_las_ctg_temp
# wfsi_eglinafb_las_ctg_temp@data %>% dplyr::glimpse()
```

let's check out the lidar data coverage compared to the field plot data

```{r}
wfsi_eglinafb_las_ctg_temp@data$geometry %>% 
  sf::st_union() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(data = "ALS data extent") %>% 
mapview::mapview(
 zcol = "data"
  , layer.name="WFSI Eglin AFB: ALS data"
  , popup=F
) +
mapview::mapview(
  wfsi_eglinafb_trees
  , zcol = "species"
  , layer.name="WFSI Eglin AFB: field trees"
  , popup=F
)
```

use `cloud2trees` to process the raw point cloud data to generate a CHM for use in tree detection

```{r}
################################################
# input/output data str
################################################
# outdir
c2t_output_dir_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_dir)
c2t_process_dir_temp <- file.path(c2t_output_dir_temp, "point_cloud_processing_delivery")
# timer file
c2t_tracking_fnm_temp <- file.path(c2t_process_dir_temp, "processed_tracking_data.csv")
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    c2t_process_dir = ifelse(data_dir==c2t_output_dir_temp, c2t_process_dir_temp, "")
    , c2t_tracking_fnm = ifelse(data_dir==c2t_output_dir_temp, c2t_tracking_fnm_temp, "")
  )
# project_data_df %>% print(n=nrow(project_data_df))

################################################
# use point cloud point density
# to automatically set CHM and DTM res
# make a fn to do this based on input LAScatalog
################################################

get_ptcld_density_chm_res_fn <- function(las_ctg, rast_res_buffer = 0.01, min_chm_res_m = 0.15, chm_factor_for_dtm = 2) {
  rast_res_buffer <- dplyr::coalesce(as.numeric(rast_res_buffer),0)
  min_chm_res_m <- dplyr::coalesce(as.numeric(min_chm_res_m),0.15)
  chm_factor_for_dtm <- dplyr::coalesce(as.numeric(chm_factor_for_dtm),2)
  
  if(!inherits(las_ctg, "LAScatalog")){stop("must be a `LAScatalog` read with `lidR::readLAScatalog()`")}
  if(
    !( methods::slotNames(las_ctg) %>% stringr::str_equal("data") %>% any() )
  ){stop("can't find `data` layer in `LAScatalog` read with `lidR::readLAScatalog()`")}
  if(!inherits(las_ctg@data, "sf")){stop("non-sf `data` from `LAScatalog` read with `lidR::readLAScatalog()`")}
  
  # some point clouds have tiles with lots of overlap
  # some point clouds have tiles with no overlap
  
  # union over total area will help us identify overlap
  # high overlap >> union will be much smaller than total area (e.g., uoa ratio < 0.66)
  # low overlap >> union will be similar to total area (e.g., uoa ratio > 0.88)
  
  # union area
  las_uoa_temp <- 
  (
    las_ctg@data %>% 
    dplyr::ungroup() %>% 
    sf::st_union() %>% 
    sf::st_area() %>% 
    as.numeric()
  ) /
  # tot area
  (
    las_ctg@data %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(
      area_m2 = sf::st_area(geometry) %>% as.numeric()
    ) %>% 
    dplyr::summarise(area_m2=sum(area_m2,na.rm=T)) %>% 
    dplyr::pull(area_m2)
  )
  # las_uoa_temp
  # when a lot of overlap, union and calculate density based off union
  if(las_uoa_temp<0.66){
    pts_m2_temp <- 
      sum(las_ctg@data$Number.of.point.records,na.rm=T) /
      ( las_ctg@data$geometry %>% sf::st_union() %>% sf::st_area() %>% as.numeric() )  
  }else{
  # when NOT a lot of overlap, calculate density as-is
    pts_m2_temp <- 
      las_ctg@data %>% 
      dplyr::mutate(
        area_m2 = sf::st_area(geometry) %>% as.numeric()
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::ungroup() %>% 
      dplyr::summarise(
        pts_m2 = sum(Number.of.point.records,na.rm = T)/sum(area_m2,na.rm = T)
      ) %>% 
      dplyr::pull(pts_m2)
  }
  # pts_m2_temp
  # 1/sqrt(pts_m2_temp)
  
  chm_res_m_temp <- max(round(1/sqrt(pts_m2_temp),2) + rast_res_buffer, min_chm_res_m) 
  dtm_res_m_temp <- round(as.numeric(chm_res_m_temp)*chm_factor_for_dtm,2)
  
  return(list(
    pts_m2 = pts_m2_temp
    , chm_res_m = chm_res_m_temp
    , dtm_res_m = dtm_res_m_temp
  ))
}

# get point cloud density and raster resolutions
get_ptcld_density_chm_res_temp <- get_ptcld_density_chm_res_fn(wfsi_eglinafb_las_ctg_temp)
# get_ptcld_density_chm_res_temp

# fnm
chm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("chm_",get_ptcld_density_chm_res_temp$chm_res_m,"m.tif") )
dtm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("dtm_",get_ptcld_density_chm_res_temp$dtm_res_m,"m.tif") )
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    las_pts_m2 = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$pts_m2, as.numeric(NA))
    , chm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$chm_res_m, as.numeric(NA))
    , chm_fnm = ifelse(data_dir==c2t_output_dir_temp, chm_fnm_temp, "")
    , dtm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$dtm_res_m, as.numeric(NA))
    , dtm_fnm = ifelse(data_dir==c2t_output_dir_temp, dtm_fnm_temp, "")
  )

# project_data_df %>% dplyr::filter(data_dir==c2t_output_dir_temp) %>% dplyr::glimpse()

##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists(chm_fnm_temp)
  || !file.exists(dtm_fnm_temp)
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir_temp
    , input_las_dir = wfsi_eglinafb_las_f_temp
    , accuracy_level = 2
    , keep_intrmdt = T
    , dtm_res_m = get_ptcld_density_chm_res_temp$dtm_res_m # (1/sqrt(pts_m2_temp))*2
    , chm_res_m = get_ptcld_density_chm_res_temp$chm_res_m # 1/sqrt(pts_m2_temp)
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm_temp
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( dtm_fnm_temp )
  chm_temp <- terra::rast( chm_fnm_temp )
  
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

we'll set the minimum CHM resolution to 0.25 m to match the `cloud2trees` default. if the automatically-determined CHM resolution based on the point cloud density used in the `cloud2trees::cloud2raster()` was lower than this level, we'll aggregate the raster using `terra::aggregate()` and `terra::resample()` to get the exact target resolution. if the determined CHM resolution was higher than this level due to low point density, we will use the CHM as-is since this is already the finest resolution possible with the point cloud density.

```{r}
min_chm_res <- 0.25
##################################################################
# aggregate to make raster more coarse for tree detection
##################################################################
agg_chm_rast_fn <- function(chm, res, dir) {
  fp <- file.path(dir, paste0("chm_",res,"m.tif"))
  if(terra::res(chm)[1]<res){
    # first, we'll borrow from the `cloud2trees` codebase to get a function to change the resolution of a raster exactly
    ###___________________________________________###
    # adjust the resolution of a raster to be in exactly the target resolution
    ###___________________________________________###
    if(
      !file.exists(fp)
    ){
      agg_chm_rast <- cloud2trees:::adjust_raster_resolution(
        chm
        , target_resolution = res
        , fun = max
        , resample_method = "max"
        , ofile = fp
      )
    }else{
      agg_chm_rast <- terra::rast( fp )
    }
  }else{
    agg_chm_rast <- chm  
  }
  return(agg_chm_rast)
}

```

let's aggregate the CHM for this site and put it into a list of CHM rasters that we can map over in all of the processing

```{r}
agg_chm_rast_temp <- agg_chm_rast_fn(
  chm = cloud2raster_ans$chm_rast
  , res = min_chm_res
  , dir = 
    project_data_df %>% 
    dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
    dplyr::pull(c2t_process_dir)
)
# agg_chm_rast_temp
# agg_chm_rast_temp %>% terra::clamp(lower = 1.37, values = F) %>% terra::plot()

# add to the chm list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
chm_list[[nm_temp]] <- agg_chm_rast_temp
# chm_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Tree Detection

`cloud2trees` implements a CHM-based individual tree detection method accessible via `cloud2trees::raster2trees()`. this method requires a window function used to set the search radius used to find local maxima in the CHM. These local maxima represent individual tree top points and are subsequently used to delineate individual tree crowns via a marker-controlled watershed segmentation

because the objective of this study is to evaluate a species/genera classification method, we just want the best ITD results we can get based on the field reference data. We are not interested in evaluating tree segmentation here, that has already been done across varied input data and forest types.

let's make a function to test many different window functions, compare the segmentation results to the field reference data, then select the best window function based on tree detection accuracy.

```{r}
# test a bunch of different variable window sizes within `lidR::lmf()` 
# that cover a range of possibilities and function shapes including linear, concave up, and constant
get_ws_fn_list <- function(min_ws = 0.5, max_ws = 6.0, n_linear = 40, n_concave = 30, n_cr = 30, max_ht = 45) {
  
  # empty list
  ws_functions <- list()
  
  ###########################################################################
  # linear functions over a range
  ###########################################################################
  # end points for the line
  linear_targets <- base::seq(from = max_ws, to = max_ws*4*10, length.out = n_linear)
  
  n_linear_pad <- ceiling(n_linear*1.5)
  
  for (i in 1:n_linear_pad) {
    target_height <- linear_targets[i]
    slope <- (max_ws - min_ws) / target_height
    
    # Pad names with a zero (01 to 60) for clean sorting
    func_name <- base::paste0("linear_", base::sprintf("%02d", i))
    
    v_min  <- base::as.numeric(min_ws)
    v_max  <- base::as.numeric(max_ws)
    v_slope <- base::as.numeric(slope)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        (v_min + v_slope * x) < v_min ~ v_min,
        (v_min + v_slope * x) > v_max ~ v_max,
        TRUE ~ (v_min + v_slope * x)
      ),
      list(v_min = v_min, v_max = v_max, v_slope = v_slope)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  
  # the flat slopes are too close together...let's remove every other one
  # how many to remove
  num_to_remove <- length(ws_functions) - n_linear
  
  # half
  start_idx <- length(ws_functions) - (num_to_remove - 1) * 2
  
  # remove
  remove_idx <- seq(from = start_idx, to = length(ws_functions), by = 2)
  ws_functions <- ws_functions[-remove_idx]
  names(ws_functions) <- base::paste0("linear_", base::sprintf("%02d", c(1:length(ws_functions)) ))
  
  # ws_functions[1]
  # ws_functions[[1]]
  # class(ws_functions[[1]])
  # ws_functions[[1]](1.5)
  # ws_functions[[1]](0)
  # ws_functions[[1]](18)
  # 
  # ws_functions[[40]](1.5)
  # ws_functions[[40]](0)
  # ws_functions[[40]](18)
  
  ###########################################################################
  # concave down functions over a range
  ###########################################################################
  # total potential growth (or distance) between the floor and the ceiling is: 6.0-0.5=5.5
  # the halfway point of growth is exactly half of that distance: 5.5/2=2.75
  # when the curve reaches its exact halfway point, the resulting window size must be the baseline floor plus that half-distance: 0.5+2.75=3.25
  # in `half_sat`, setting the constant of 1.375 means the function will hit its halfway mark of 3.25 when the tree is only 1.375 meters tall
    # max slope = 4 (for ever 1 m growth, ws increases by 4 m)
    # 4 = (6.0-0.5)/c >> 4 = 5.5/c >> c = 5.5/4 >> c = 1.375
  # how far out to look based on maximum plot/tree height? 45 m tree??
  
  for (i in 1:n_concave) {
    c <- 1.375 # ~dbh height
    max_end <- max_ht # biggest tree everrrrr
    half_sat <- c + (i - 1) * ( (max_end-c) / (n_concave-1))
    
    func_name <- base::paste0("concave_down_", base::sprintf("%02d", i))
    
    v_min  <- base::as.numeric(min_ws)
    v_max  <- base::as.numeric(max_ws)
    v_half <- base::as.numeric(half_sat)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        (v_min + ((v_max - v_min) * x) / (v_half + x)) < v_min ~ v_min,
        (v_min + ((v_max - v_min) * x) / (v_half + x)) > v_max ~ v_max,
        TRUE ~ (v_min + ((v_max - v_min) * x) / (v_half + x))
      ),
      list(v_min = v_min, v_max = v_max, v_half = v_half)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  # ws_functions[length(ws_functions)]
  # ws_functions[length(ws_functions)-1]
  
  ###########################################################################
  # constant functions over a range
  ###########################################################################
  # range
  constant_values <- base::seq(from = min_ws, to = max_ws * 0.95, by = 0.5)
  
  for (i in base::seq_along(constant_values)) {
    const_val <- constant_values[i]
    
    func_name <- base::paste0("constant_", base::sprintf("%02d", i))
    
    v_min <- base::as.numeric(min_ws)
    v_val <- base::as.numeric(const_val)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        TRUE ~ v_val
      ),
      list(v_min = v_min, v_val = v_val)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  # ws_functions[length(ws_functions)]
  # ws_functions[[length(ws_functions)]](0.1)
  # ws_functions[[length(ws_functions)]](4)
  # ws_functions[[length(ws_functions)]](15)
  # ws_functions[length(ws_functions)-1]
  
  
  # ###########################################################################
  # # two-step linear functions over a range
  # ###########################################################################
  # # how many functions to test
  # n_temp <- 16
  # # max tree height in CHM
  # 
  # # Calculate fractional boundary envelopes based on global constraints
  # xb_min <- max_ht * 0.09   # ~5m height
  # xb_max <- max_ht * 0.66   # ~25m height
  # yb_min <- min_ws + (max_ws - min_ws) * 0.36  # ~2.5m window size
  # yb_max <- min_ws + (max_ws - min_ws) * 0.88  # ~4.5m window size
  # 
  # xb_sequence  <- base::seq(from = xb_min, to = xb_max, length.out = n_temp)
  # yb_sequence  <- base::seq(from = yb_min, to = yb_max, length.out = n_temp)
  # cap_sequence <- base::seq(from = 15.0,   to = 40.0,   length.out = n_temp)
  # 
  # for (i in 1:n_temp) {
  #   x_b <- xb_sequence[i]
  #   y_b <- yb_sequence[i]
  #   
  #   slope1 <- (y_b - min_ws) / x_b
  #   
  #   target_cap_x <- x_b + cap_sequence[i]
  #   slope2 <- (max_ws - y_b) / (target_cap_x - x_b)
  #   
  #   func_name <- base::paste0("twostep_linear_", base::sprintf("%02d", i))
  #   
  #   v_min    <- base::as.numeric(min_ws)
  #   v_max    <- base::as.numeric(max_ws)
  #   v_xb     <- base::as.numeric(x_b)
  #   v_yb     <- base::as.numeric(y_b)
  #   v_slope1 <- base::as.numeric(slope1)
  #   v_slope2 <- base::as.numeric(slope2)
  #   
  #   func_body <- base::substitute(
  #     dplyr::case_when(
  #       base::is.na(x) ~ v_min,
  #       x <= v_xb ~ {
  #         val <- v_min + v_slope1 * x
  #         dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
  #       },
  #       x > v_xb ~ {
  #         val <- v_yb + v_slope2 * (x - v_xb)
  #         dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
  #       }
  #     ),
  #     list(v_min = v_min, v_max = v_max, v_xb = v_xb, v_yb = v_yb, v_slope1 = v_slope1, v_slope2 = v_slope2)
  #   )
  #   
  #   ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  # }
  # 
  # 
  # # ws_functions[length(ws_functions)]
  # # ws_functions[[length(ws_functions)]](0.1)
  # # ws_functions[[length(ws_functions)]](4)
  # # ws_functions[[length(ws_functions)]](15)
  # # ws_functions[length(ws_functions)-1]
  
  
  ###########################################################################
  # chapman richards functions over a range
  ###########################################################################
  # Generate parameters that smoothly spread across your MAX_CANOPY_HEIGHT
  # Beta controls horizontal stretching; Gamma controls the sharpness of the initial lag
  beta_sequence  <- base::seq(from = 0.15, to = 0.02, length.out = n_cr)
  gamma_sequence <- base::seq(from = 1.5,  to = 4.0,  length.out = n_cr)
  
  for (i in 1:n_cr) {
    b_param <- beta_sequence[i]
    g_param <- gamma_sequence[i]
    
    func_name <- base::paste0("chapman_richards_", base::sprintf("%02d", i))
    
    v_min   <- base::as.numeric(min_ws)
    v_max   <- base::as.numeric(max_ws)
    v_beta  <- base::as.numeric(b_param)
    v_gamma <- base::as.numeric(g_param)
    
    # Chapman-Richards equation with case_when boundaries
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        TRUE ~ {
          # Core Chapman-Richards Sigmoid Formula
          val <- v_min + (v_max - v_min) * (1 - base::exp(-v_beta * x))^v_gamma
          dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
        }
      ),
      list(v_min = v_min, v_max = v_max, v_beta = v_beta, v_gamma = v_gamma)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  
  return(ws_functions)
}
```

let's take that function for a spin to get a list of window functions for testing and plot the functions over a height range

```{r}
get_ws_fn_list_ans <- get_ws_fn_list(min_ws = 0.5, max_ws = 7.0, n_linear = 30, n_concave = 30, n_cr = 30, max_ht = 45)
# get_ws_fn_list_ans %>% names()
# get_ws_fn_list_ans %>% length()

# cloud2trees::itd_ws_functions() %>% names()
# get_ws_fn_list_ans %>% names()
# 
# cloud2trees::itd_ws_functions()[[1]] %>% class()
# get_ws_fn_list_ans[[1]] %>% class()
# 
# cloud2trees::itd_ws_functions()[[1]](2.22) %>% class()
# get_ws_fn_list_ans[[1]](2.22) %>% class()
# 
# cloud2trees::itd_ws_functions()[[1]](2.22)
# get_ws_fn_list_ans[[1]](2.22)

# plot that
ht_seq_temp <- base::seq(0, 50, by = 0.25)

p_dta_temp <- 
  purrr::map_dfr(
  names(get_ws_fn_list_ans)
  , function(name) {
    tibble::tibble(
      height = ht_seq_temp
      , window_size = get_ws_fn_list_ans[[name]](ht_seq_temp)
      , function_id = name
      , shape_type = dplyr::case_when(
        stringr::str_starts(name,"linear_") ~ "linear"
        , stringr::str_starts(name,"concave_down_") ~ "concave down"
        , stringr::str_starts(name,"twostep_linear_") ~ "two-step linear"
        , stringr::str_starts(name,"constant_") ~ "constant"
        , stringr::str_starts(name,"chapman_richards_") ~ "chapman-richards"
      )
    )
  }
)
# p_dta_temp %>% dplyr::count(shape_type)
# p_dta_temp %>% dplyr::distinct(function_id,shape_type) %>% dplyr::count(shape_type)

p_temp <- 
  ggplot2::ggplot(
    data = p_dta_temp
    , mapping = ggplot2::aes(x = height, y = window_size, group = function_id, color = shape_type)
  ) +
    ggplot2::geom_line(alpha = 0.6, linewidth = 1) +
    ggplot2::labs(
      x = "CHM (m)",
      , y = "ws diameter (m)"
      , color = "ws type"
    ) +
    # ggplot2::facet_wrap(facets = dplyr::vars(shape_type)) +
    ggplot2::scale_color_viridis_d(option = "viridis", direction = -1) +
    ggplot2::scale_y_continuous(limits=c(0,NA), breaks = scales::breaks_extended(n=6)) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
    )
p_temp

```

let's look at the functions by type so they are a little easier to see since that is `r length(get_ws_fn_list_ans)` unique functions

```{r}
p_temp + 
  ggplot2::facet_wrap(facets = dplyr::vars(shape_type))
```

now we need a function to clip the CHM to the plot plus a buffer and run `cloud2trees::raster2trees()` for each window function. the return should be a spatial tree list unique by function.

```{r}
# function to loop over ws function list in raster2trees
ws_testtest_fn <- function(
    chm
    , buffer = 5
    , plot_sf
    , ws_fn_list # a named list of functions ex: cloud2trees::itd_ws_functions()
  ) {
  # min ht for chm so we can test non-na
  min_chm_ht <- 1.37
  # crop chm
  chm_crop <- chm %>% 
    terra::crop(
      plot_sf %>% 
        sf::st_transform(terra::crs(chm)) %>% 
        sf::st_buffer(buffer)
      , mask = T
    ) %>% 
    terra::clamp(lower = min_chm_ht, values = F)
  
  # terra::plot(chm_crop)
  # terra::plot(
  #   plot_sf %>%
  #     sf::st_transform(terra::crs(chm))
  #   , add = T
  #   , border = "cyan"
  #   , col = NA
  # )

  if( !any( terra::global(chm_crop, fun = "anynotNA") ) ){
    return(NULL)
  }
  
  # map over the function list
  ws_tree_list <- 
    names(ws_fn_list) %>%
    purrr::set_names() %>%
    purrr::map(function(nm){
      if (exists("my_temp_dir") && dir.exists(my_temp_dir)) {
        unlink(my_temp_dir, recursive = TRUE, force = TRUE)
      }
      if (exists("my_temp_dir2") && dir.exists(my_temp_dir2)) {
        unlink(my_temp_dir2, recursive = TRUE, force = TRUE)
      }
      warning(paste0("making dir: ",nm,"....",Sys.time()),immediate. = T,call. = F)
      my_temp_dir <- tempfile(pattern = nm)
      dir.create(my_temp_dir, showWarnings = FALSE)
      my_temp_dir2 <- tempfile(pattern = paste0("xx",nm))
      dir.create(my_temp_dir2, showWarnings = FALSE)
      warning(paste0("raster2trees: ",nm,"....",Sys.time()),immediate. = T,call. = F)
      
      safe_raster2trees <- purrr::safely( cloud2trees::raster2trees )
      xxx_raster2trees_ans <- safe_raster2trees(
          chm_rast = chm_crop
          , outfolder = my_temp_dir
          , ws = ws_fn_list[[nm]]
          , min_height = min_chm_ht
          , min_crown_area = 0.5
          , tempdir = my_temp_dir2
        )
      
      if(!is.null(xxx_raster2trees_ans$result)){
        xxx_raster2trees_ans <- xxx_raster2trees_ans$result
        # xxx_raster2trees_ans %>% names()
        return(
          xxx_raster2trees_ans %>% 
            cloud2trees::simplify_multipolygon_crowns()
            # sf::st_drop_geometry() %>% 
            # sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(xxx_raster2trees_ans))
        )
      }else{return(NULL)}
    }) %>% 
    purrr::list_rbind(names_to = "ws_fn_nm")
  
  # ws_tree_list %>% dplyr::glimpse()
  # ws_tree_list %>% sf::st_drop_geometry() %>% dplyr::count(ws_fn_nm)
  # ws_tree_list %>% class()
  
  return(ws_tree_list)
}
```

get the individual tree predictions for each window function for the area within the field reference plots with a buffer

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, "")
  )
if(!file.exists(fnm_temp)){
  wfsi_eglinafb_ws_pred_trees <- 
    1:nrow(wfsi_eglinafb_plots) %>% 
    purrr::map(
      \(x)
      ws_testtest_fn(
        chm = chm_list$WFSI_EglinAFB
        , buffer = 5.5
        , plot_sf = 
          wfsi_eglinafb_plots %>% 
          dplyr::ungroup() %>% 
          dplyr::slice(x) %>%
          sf::st_union()
        # a named list of functions ex: cloud2trees::itd_ws_functions()
        , ws_fn_list = get_ws_fn_list_ans # cloud2trees::itd_ws_functions() # 
      )  
    ) %>% 
    purrr::list_rbind(names_to = "plot_dta_idx")
  
  # add plot id
  wfsi_eglinafb_ws_pred_trees <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::left_join(
      wfsi_eglinafb_plots %>% 
        dplyr::mutate(plot_dta_idx=dplyr::row_number()) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(plot_dta_idx,plot_id)
      , by = "plot_dta_idx"
    )
  # get dbh
  wfsi_eglinafb_ws_pred_trees_dbh_temp <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number()
    ) %>% 
    cloud2trees::trees_dbh()
  # wfsi_eglinafb_ws_pred_trees_dbh_temp %>% dplyr::glimpse()
  # add to trees
  wfsi_eglinafb_ws_pred_trees <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::mutate(
      join_id = dplyr::row_number()
    ) %>% 
    dplyr::left_join(
      wfsi_eglinafb_ws_pred_trees_dbh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          join_id = as.numeric(treeID)
        ) %>% 
        dplyr::select(
          join_id, tidyselect::starts_with("fia_est_dbh"), dbh_cm
        )
      , by = "join_id"
    )
  # wfsi_eglinafb_ws_pred_trees %>% dplyr::glimpse()
  
  # # plot real quick
  # wfsi_eglinafb_ws_pred_trees %>% 
  #   sf::st_drop_geometry() %>% 
  #   dplyr::slice_sample(n = 9999) %>% 
  #   ggplot2::ggplot(mapping = ggplot2::aes(x=tree_height_m,y=dbh_cm)) + 
  #   ggplot2::geom_point() +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_lower),color="gray") +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_upper),color="gray")

  # save
  sf::st_write(wfsi_eglinafb_ws_pred_trees, dsn = fnm_temp, quiet = T, append = F)
}else{
  wfsi_eglinafb_ws_pred_trees <- sf::st_read(fnm_temp, quiet = T)
}
wfsi_eglinafb_ws_pred_trees <- wfsi_eglinafb_ws_pred_trees %>% 
  dplyr::mutate(
    ws_type = 
      ws_fn_nm %>% 
      # removes all numeric digits
      stringr::str_remove_all("[0-9]+") %>% 
      stringr::str_replace_all("_"," ") %>% 
      stringr::str_squish() %>% 
      as.factor() %>% 
      forcats::fct_inorder(ordered = T)
  )
# wfsi_eglinafb_ws_pred_trees %>% dplyr::glimpse()
# wfsi_eglinafb_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_type)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's check out one example field plot and a few different window function tree detection and crown delineation results

```{r, fig.height=10}
# plot
p_temp <- wfsi_eglinafb_plots %>% dplyr::slice(1)
# some trees
t_temp <- 
  wfsi_eglinafb_ws_pred_trees %>% 
  dplyr::filter(plot_id==p_temp$plot_id[1]) %>% 
  dplyr::inner_join(
    wfsi_eglinafb_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm,ws_type) %>% 
      dplyr::slice_sample(n = 3, by = ws_type) %>% 
      dplyr::select(ws_fn_nm)
    , by = "ws_fn_nm"
  )
chm_temp <- chm_list[["WFSI_EglinAFB"]] %>% 
  terra::crop(
    t_temp %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      terra::vect()
    , mask = T
  )
# terra::plot(chm_temp)
plt_temp <-
  chm_temp %>% 
  terra::clamp(lower = 1.37, values = F) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x=x,y=y,fill=f)
    , color = NA
    , lwd = 0
  ) +
  ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent", name = "CHM (m)") +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top")
plt_temp +
  ggplot2::geom_sf(
    data = p_temp %>% sf::st_transform(terra::crs(chm_temp))
    , fill = NA
    , color = "black"
    , lwd = 0.8
  ) + 
  ggplot2::geom_sf(
    data = t_temp
    , mapping = ggplot2::aes(color = ws_type)
    , fill = NA
    , lwd = 0.7
  ) +
  ggplot2::scale_color_viridis_d() +
  ggplot2::facet_wrap(facets = dplyr::vars(ws_fn_nm), ncol = 3) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold")
    , strip.background = ggplot2::element_rect(color = "black")
    , panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  ggplot2::guides(
    color = "none"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now we need a function to perform instance matching of the predictions and reference data. to keep things simple, since all the reference field data includes point locations, we'll perform matching based on a spatial intersection of these points with the predicted crowns. if multiple tree points fall within a single crown, we'll select the tree that is closest in terms of size based on whatever field data is available (height or DBH). One predicted tree will only be matched with one field reference tree and vice-versa.

```{r}
# instance matching function for points reference and polys predictions
# matches a single set of predictions with a single reference set
instance_match_fn <- function(
  pred_crowns
  , ref_points
){
  
  if(nrow(ref_points)==0){return(NULL)}
  if(nrow(pred_crowns)==0 && nrow(ref_points)>0){
    return(
      ref_points %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          ref_match_idx = dplyr::row_number()
          , pred_match_idx = as.numeric(NA)
        ) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(
          match_grp = "omission"
          , match_grp = factor(
            match_grp
            , ordered = T
            , levels = c(
              "true positive"
              , "commission"
              , "omission"
            )
          ) %>% forcats::fct_rev()
        )
    )
  }
  
  # set up matching col based on what data is available
  if(
    any( stringr::str_equal(names(pred_crowns),"tree_height_m") ) &&
    any( stringr::str_equal(names(ref_points),"tree_height_m") )
  ){
    pred_crowns <- pred_crowns %>% dplyr::mutate(xxx_match_col = as.numeric(tree_height_m))
    ref_points <- ref_points %>% dplyr::mutate(xxx_match_col = as.numeric(tree_height_m))
  }else if(
    any( stringr::str_equal(names(pred_crowns),"dbh_cm") ) &&
    any( stringr::str_equal(names(ref_points),"dbh_cm") )
  ){
    pred_crowns <- pred_crowns %>% dplyr::mutate(xxx_match_col = as.numeric(dbh_cm))
    ref_points <- ref_points %>% dplyr::mutate(xxx_match_col = as.numeric(dbh_cm))
  }
  # get rid of data without data
  pred_crowns <- pred_crowns %>% dplyr::filter(!is.na(xxx_match_col), xxx_match_col>0) %>% 
    # make an id
    dplyr::mutate(xxx_id = dplyr::row_number())
  ref_points <- ref_points %>% dplyr::filter(!is.na(xxx_match_col), xxx_match_col>0) %>% 
    # make it the same crs
    sf::st_transform(sf::st_crs(pred_crowns)) %>% 
    # make an id
    dplyr::mutate(xxx_id = dplyr::row_number())
  
  # intersection for tp's
  i_temp <-
    pred_crowns %>% 
    dplyr::rename_with(
      .fn = ~stringr::str_replace(.x, "^xxx_", "xxx_pred_")
      , .cols = tidyselect::starts_with("xxx_")
    ) %>% 
    sf::st_intersection(
      ref_points %>% 
        dplyr::select(xxx_id,xxx_match_col) %>% 
        dplyr::rename_with(
          .fn = ~stringr::str_replace(.x, "^xxx_", "xxx_ref_")
          , .cols = tidyselect::starts_with("xxx_")
        )
    ) %>% 
    sf::st_drop_geometry() %>% 
    # absolute diff between match col
    dplyr::mutate(
      xxx_abs_diff = abs(xxx_pred_match_col - xxx_ref_match_col)
    ) %>% 
    # sorting is key to make the purrr::reduce work how we want it
    dplyr::arrange(
      desc(xxx_pred_match_col)
      , xxx_pred_id
      , xxx_abs_diff
    )
  # i_temp %>% dplyr::glimpse()
  
  # if there even are any matches
  if(nrow(i_temp)>1){
    # ?purrr::reduce
    # purrr::reduce(letters[1:5], paste, sep = ".")
    # 1:3  %>% reduce(`+`)
    match_temp <- purrr::reduce(
      # go through all rows except first since we start there
      .x = c(2:nrow(i_temp))
      # start with the first row
      , .init = i_temp %>% dplyr::slice(1) # = iter_temp
      , .f = function(iter_temp, i){
        
        #ids of curr row
        pred_id  <- i_temp$xxx_pred_id[i]
        ref_id <- i_temp$xxx_ref_id[i]
        
        # either one used?
        pred_used  <- pred_id %in% iter_temp$xxx_pred_id
        ref_used <- ref_id %in% iter_temp$xxx_ref_id
        
        # add if freee
        if(!pred_used && !ref_used) {
          return(
            iter_temp %>% 
              dplyr::bind_rows(
                dplyr::slice(i_temp, i)
              )
          )
        }else{
          # return current state
          return(iter_temp)
        }
    })
    # match_temp %>% dplyr::glimpse()
    # match_temp %>% dplyr::count(xxx_pred_id) %>% dplyr::filter(n>1) %>% nrow()
    # match_temp %>% dplyr::count(xxx_ref_id) %>% dplyr::filter(n>1) %>% nrow()
  }else{
    match_temp <- i_temp
  }
  # match_temp %>% dplyr::glimpse()
  # these are matches
  return_df <- match_temp %>% 
    dplyr::rename(
      ref_match_idx = xxx_ref_id
      , pred_match_idx = xxx_pred_id
    ) %>% 
    dplyr::select(dplyr::all_of(c(
      "ref_match_idx"
      , "pred_match_idx"
    ))) %>% 
    dplyr::mutate(match_grp = "true positive")
  # return_df %>% dplyr::glimpse()
  
  
  return_df <-
    return_df %>% 
    # if reference but no predictions...they are all omissions (FN)
    dplyr::bind_rows(
      ref_points %>% 
        dplyr::filter(
          !(xxx_id %in% match_temp$xxx_ref_id)
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::rename(
          ref_match_idx = xxx_id
        ) %>% 
        dplyr::mutate(pred_match_idx = as.numeric(NA)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(match_grp = "omission")  
    ) %>% 
    # if predictions but no reference...they are all commissions (FP)
    dplyr::bind_rows(
      pred_crowns %>% 
        dplyr::filter(
          !(xxx_id %in% match_temp$xxx_pred_id)
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::rename(
          pred_match_idx = xxx_id
        ) %>% 
        dplyr::mutate(ref_match_idx = as.numeric(NA)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(match_grp = "commission")
    )
    
  # return_df %>% dplyr::glimpse()
  # return_df %>% View()
  # return_df %>% dplyr::count(match_grp)
  
  # make match_grp factor
    return_df <- return_df %>% 
      dplyr::mutate(
        match_grp = factor(
          match_grp
          , ordered = T
          , levels = c(
            "true positive"
            , "commission"
            , "omission"
          )
        ) %>% forcats::fct_rev()
      )
    return(return_df)
}

###################################
### test for one plot
### test for one plot
### test for one plot
###################################
instance_match_fn(
  pred_crowns = 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::filter(
      plot_id == wfsi_eglinafb_plots$plot_id[1]
      , ws_fn_nm==wfsi_eglinafb_ws_pred_trees$ws_fn_nm[1]
    )
  , ref_points = 
    wfsi_eglinafb_trees %>% 
    dplyr::filter(
      plot_id == wfsi_eglinafb_plots$plot_id[1]
    )
) %>% 
dplyr::count(match_grp)
```

remember, we have predictions for many different window functions for each plot. we need a function to iterate over each set of predictions (i.e. window function) for a single plot. this is where we'll filter the predictions to attempt to match the field data collection protocol (e.g. measure trees >= 10 cm DBH)

```{r, fig.height=8.8}
ws_iter_instance_match_fn <- function(
  plot_sf # needs plot_id # should be a single plot
  , ref_pts # needs plot_id
  , pred_polys # needs plot_id
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = NA
){
  if(
    !any( stringr::str_equal(names(plot_sf),"plot_id") )
  ){
    stop("`plot_sf` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(ref_pts),"plot_id") )
  ){
    stop("`ref_pts` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"plot_id") )
  ){
    stop("`pred_polys` needs column named `plot_id`")
  }
  
  # pred polys
  pred_polys <- 
    pred_polys %>% 
    dplyr::filter(
      plot_id == plot_sf$plot_id[1]
    )
  
  # reference points
  ref_pts <- 
    ref_pts %>% 
    dplyr::filter(
      plot_id == plot_sf$plot_id[1]
    )
  
  # filter based on field survey limit
  if(
    !is.na(as.numeric(min_dbh_cm))
  ){
    # do they exist? .... PRED
    if(
      any( stringr::str_equal(names(pred_polys),"fia_est_dbh_cm_upper") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          fia_est_dbh_cm_upper >= as.numeric(min_dbh_cm)
        )
    }else if(
      any( stringr::str_equal(names(pred_polys),"dbh_cm") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          dbh_cm >= as.numeric(min_dbh_cm)
        )
    }else{
      stop("`pred_polys` needs column named `dbh_cm`")
    }
    
    # do they exist? .... REF
    if(
      any( stringr::str_equal(names(ref_pts),"dbh_cm") )
    ){
      ref_pts <- ref_pts %>% 
        dplyr::filter(
          dbh_cm >= as.numeric(min_dbh_cm)
        )
    }else{
      stop("`ref_pts` needs column named `dbh_cm`")
    }
  }else if(
    !is.na(as.numeric(min_ht_m))
  ){
    # do they exist? .... PRED
    if(
      any( stringr::str_equal(names(pred_polys),"tree_height_m") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          tree_height_m >= as.numeric(min_ht_m)
        )
    }else{
      stop("`pred_polys` needs column named `tree_height_m`")
    }
    
    # do they exist? .... REF
    if(
      any( stringr::str_equal(names(ref_pts),"tree_height_m") )
    ){
      ref_pts <- ref_pts %>% 
        dplyr::filter(
          tree_height_m >= as.numeric(min_ht_m)
        )
    }else{
      stop("`ref_pts` needs column named `tree_height_m`")
    }
  }
  
  if(nrow(ref_pts)==0){return(NULL)}
  
  # map over it
  if(
    !any( stringr::str_equal(names(pred_polys),"ws_fn_nm") )
  ){
    stop("`pred_polys` needs column named `ws_fn_nm` ... see ws_testtest_fn() ")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"tree_x") )
  ){
    stop("`pred_polys` needs column named `tree_x` ... see ws_testtest_fn() ")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"tree_y") )
  ){
    stop("`pred_polys` needs column named `tree_y` ... see ws_testtest_fn() ")
  }
  
  # map it
  instance_match_ans <- 
    unique(pred_polys$ws_fn_nm) %>% 
    purrr::set_names() %>%
    purrr::map(function(nm){
      # which preds?
      these_pred_polys <- 
        pred_polys %>% 
        dplyr::filter(
          ws_fn_nm == nm
        )
      # instance match
      im <- 
        instance_match_fn(
          pred_crowns = these_pred_polys
          , ref_points = ref_pts
        )
      # remove commission points outside plot boundary?
      # remove commission points outside plot boundary?
      # get rid of commissions outside of plot
      outs <- 
        these_pred_polys %>% 
        dplyr::mutate(
          pred_match_idx = dplyr::row_number()
        ) %>% 
        sf::st_drop_geometry() %>%
        sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(pred_polys)) %>% 
        sf::st_filter(
          plot_sf %>% sf::st_union() %>% sf::st_transform(sf::st_crs(pred_polys))
          , .predicate = sf::st_disjoint
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(pred_match_idx) %>% 
        dplyr::mutate(is_out = T)
      
      return_df <- 
        im %>% 
        dplyr::left_join(
          outs
          , by = "pred_match_idx"
        ) %>% 
        dplyr::filter(
          !(match_grp == "commission" & dplyr::coalesce(is_out,F)==T)
        ) %>% 
        dplyr::select(-is_out)
      return(return_df)
    }) %>% 
    purrr::list_rbind(names_to = "ws_fn_nm")
  
  
  return(instance_match_ans)
}

###################################
### test for one plot
### test for one plot
### test for one plot
###################################

ws_iter_instance_match_fn(
  # needs plot_id # should be a single plot
  plot_sf = wfsi_eglinafb_plots %>% dplyr::slice(4) 
  # needs plot_id
  , ref_pts = wfsi_eglinafb_trees 
  # needs plot_id
  , pred_polys = wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::inner_join(
      wfsi_eglinafb_ws_pred_trees %>% 
        sf::st_drop_geometry() %>% 
        dplyr::distinct(ws_fn_nm,ws_type) %>% 
        dplyr::slice_sample(n = 3, by = ws_type) %>% 
        dplyr::select(ws_fn_nm)
      , by = "ws_fn_nm"
    )
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = 10
) %>% 

# plot
  dplyr::count(ws_fn_nm,match_grp) %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::mutate(pct = n/sum(n)) %>% 

  ggplot2::ggplot(
    mapping = ggplot2::aes(x = pct, y = ws_fn_nm, fill=match_grp, group=match_grp)
  ) +
  ggplot2::geom_col(
    width = 0.7, alpha=0.95
  ) +
  ggplot2::scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.direction  = "horizontal"
  )

```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now, we need to iterate over all plots at a study site

```{r}
plot_iter_instance_match_fn <- function(
  ref_plots
  , ref_pts
  # needs plot_id
  , pred_polys
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = NA
) {
  if(
    !any( stringr::str_equal(names(ref_plots),"plot_id") )
  ){
    stop("`ref_plots` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(ref_pts),"plot_id") )
  ){
    stop("`ref_pts` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"plot_id") )
  ){
    stop("`pred_polys` needs column named `plot_id`")
  }
  
  ws_iter_instance_match_ans <- 
    unique(ref_plots$plot_id) %>% 
    purrr::set_names() %>% 
    purrr::map(
      \(pid)
      ws_iter_instance_match_fn(
        # needs plot_id # should be a single plot
        plot_sf = ref_plots %>% dplyr::filter(plot_id==pid) 
        # needs plot_id
        , ref_pts = ref_pts 
        # needs plot_id
        , pred_polys = pred_polys
        # minimum tree height
        , min_ht_m = min_ht_m
        # -OR- minimum DBH
        , min_dbh_cm = min_dbh_cm
      )
  ) %>% 
  purrr::list_rbind(names_to = "plot_id")
  
  return(ws_iter_instance_match_ans)
}
# get the ans
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_instance_match.csv")
if(!file.exists(fnm_temp)){
  wfsi_eglinafb_ws_instance_match <- 
    plot_iter_instance_match_fn(
      ref_plots = ref_plots_list[["WFSI_EglinAFB"]]
      , ref_pts = ref_trees_list[["WFSI_EglinAFB"]]
      , pred_polys = wfsi_eglinafb_ws_pred_trees
      , min_ht_m = NA
      , min_dbh_cm = 10
    )
  readr::write_csv(wfsi_eglinafb_ws_instance_match, file = fnm_temp, append = F, progress = F)
}else{
  wfsi_eglinafb_ws_instance_match <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# wfsi_eglinafb_ws_instance_match %>% dplyr::glimpse()
```

lastly, we need a function to aggregate these instance segmentation counts (TP, FP, and FN) to evaluate the overall accuracy of the tree segmentation method. then, we'll select the best window function and apply that across the entire study area.

```{r}
###############################################################
###############################################################
###############################################################
###############################################################
# first function takes df with cols tp_n, fp_n, and fn_n to calculate rates
###############################################################
###############################################################
###############################################################
###############################################################
confusion_matrix_scores_fn <- function(df){
  df %>% 
  dplyr::mutate(
    omission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 0 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 1 # every single actual tree was missed
      , dplyr::coalesce(fn_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fn_n/(tp_n+fn_n)
    ) # False Negative Rate or Miss Rate
    , commission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 0 # if no predictions are made, the model could not have made any commission errors
      , dplyr::coalesce(fp_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fp_n/(tp_n+fp_n)
    ) # False Positive Rate
    , precision = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 1 # if no predictions are made, the model made zero incorrect positive claims
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) > 0 ~ 0
      , T ~ tp_n/(tp_n+fp_n)
    )
    , recall = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 1 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 0 # every single actual tree was missed
      , T ~ tp_n/(tp_n+fn_n)
    )
    , f_score = dplyr::case_when(
      dplyr::coalesce(precision,0) == 0 | dplyr::coalesce(recall,0) == 0 ~ 0
      , T ~ 2 * ( (precision*recall)/(precision+recall) )
    )    
  ) 
}
###############################################################
###############################################################
###############################################################
###############################################################
# aggregate results from ground_truth_prediction_match()
###############################################################
###############################################################
###############################################################
###############################################################
agg_ground_truth_match <- function(ground_truth_prediction_match_ans) {
  if(nrow(ground_truth_prediction_match_ans)==0){return(NULL)}
  if( !(names(ground_truth_prediction_match_ans) %>% stringr::str_equal("match_grp") %>% any()) ){stop("ground_truth_prediction_match_ans must contain `match_grp` column")}
  
  # check for difference columns (contains "_diff") and calc rmse for only those to return a single line df with colums for each diff_rmse
    if(
      (ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(tidyselect::starts_with("diff_") | tidyselect::starts_with("pct_diff_")) %>% 
        ncol() )>0
    ){
      # get rmse and mean difference/error for all columns with "_diff" but not "pct_diff"
      # get mape for all columns with "pct_diff" but not "diff_"
      rmse_df <- 
        ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::ungroup() %>% 
        dplyr::select(
          tidyselect::starts_with("diff_")
          | tidyselect::starts_with("pct_diff_")
        ) %>%
        tidyr::pivot_longer(dplyr::everything(), values_drop_na = T) %>% 
        dplyr::group_by(name) %>% 
        dplyr::summarise(
          sq = sum(value^2, na.rm = T)
          , mean = mean(value, na.rm = T)
          , sumabs = sum(abs(value), na.rm = T)
          , nomiss = sum(!is.na(value))
        ) %>% 
        dplyr::ungroup() %>% 
        dplyr::mutate(
          rmse = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sqrt(sq/nomiss)
          )
          , mape = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sumabs/nomiss
          )
        ) %>% 
        # NA nonsense values
        dplyr::mutate(
          mape = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ mape
            , T ~ as.numeric(NA)
          )
          , rmse = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ rmse
          )
          , mean = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ mean
          )
        ) %>% 
        dplyr::select(name,rmse,mean,mape) %>% 
        tidyr::pivot_wider(
          names_from = name
          , values_from = c(rmse,mean,mape)
          , names_glue = "{name}_{.value}"
        ) %>% 
        # remove columns with NA in all rows
        dplyr::select( dplyr::where( ~!all(is.na(.x)) ) )
      
      if(
        dplyr::coalesce(nrow(rmse_df),0)==0
        || dplyr::coalesce(ncol(rmse_df),0)==0
      ){
        # empty df
        rmse_df <- dplyr::tibble()
      }
    }else{
      # empty df
      rmse_df <- dplyr::tibble()
    }
  
  # count by match group
    agg <-
      ground_truth_prediction_match_ans %>% 
      sf::st_drop_geometry() %>% 
      dplyr::ungroup() %>% 
      dplyr::count(match_grp) %>% 
      dplyr::mutate(
        match_grp = dplyr::recode_values(
          match_grp
          , "true positive"~"tp"
          , "commission"~"fp"
          , "omission"~"fn"
        )
      )
    # true positive, false positive, false negative rates
    return_df <- dplyr::tibble(match_grp = c("tp","fp","fn")) %>% 
      dplyr::left_join(agg, by = "match_grp") %>% 
      dplyr::mutate(dplyr::across(.cols = c(n), .fn = ~dplyr::coalesce(.x,0))) %>% 
      tidyr::pivot_wider(
        names_from = match_grp
        , values_from = c(n)
        , names_glue = "{match_grp}_{.value}"
      )
    # rates, precision, recall, f-score
    return_df <- confusion_matrix_scores_fn(return_df)
    # add rmse
    if(nrow(rmse_df)>0){
      return_df <- return_df %>% dplyr::bind_cols(rmse_df)
    }
  # return
  return(return_df)
}
# agg_ground_truth_match(
#   wfsi_eglinafb_ws_instance_match %>% 
#     dplyr::filter(
#       plot_id==wfsi_eglinafb_ws_instance_match$plot_id[1]
#       , ws_fn_nm==wfsi_eglinafb_ws_instance_match$ws_fn_nm[1]
#     )
# )

# map it over
# ?purrr::pmap
wfsi_eglinafb_agg_ground_truth_match_ans <- 
  wfsi_eglinafb_ws_instance_match %>% 
  dplyr::distinct(plot_id,ws_fn_nm) %>% 
  purrr::pmap(function(plot_id, ws_fn_nm) {
    
    d <- tibble(plot_id = plot_id, ws_fn_nm = ws_fn_nm)
    
    j <- 
      wfsi_eglinafb_ws_instance_match %>% 
      dplyr::inner_join(
        d   
        , by = dplyr::join_by("plot_id","ws_fn_nm")
      )
    
    d %>% 
      dplyr::bind_cols(agg_ground_truth_match(j))
    
  }) %>% 
  list_rbind()
# huh?
# wfsi_eglinafb_agg_ground_truth_match_ans %>% dplyr::glimpse()
```

aggregate the plot level data to the window function to get the answer we've been searching for

```{r}
# wfsi_eglinafb_agg_ground_truth_match_ans %>% 
#   dplyr::count(ws_fn_nm) %>% 
#   dplyr::count(n)
agg_temp <- 
  wfsi_eglinafb_agg_ground_truth_match_ans %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::summarize(
    n_plots = dplyr::n()
    , dplyr::across(
      .cols = c(precision,recall,f_score,tidyselect::ends_with("_rate"))
      , list( 
        mean= ~ mean(.x,na.rm=T)
        , median= ~ median(.x,na.rm=T)
      )
    )
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      .cols = dplyr::where(is.numeric)
      , ~ round(.x,2) %>% as.character() %>% as.numeric()
    )
  ) %>% 
  dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean))

```


```{r, include = F, eval = F}
# xxx_treetops_sf %>% mapview::mapview(zcol="tree_height_m", layer.name="trees")
  # nrow(xxx_treetops_sf)
  xxx_treetops_sf <- 
    sf::st_filter(
      xxx_treetops_sf
      , piles_workflow_ans %>% 
        dplyr::filter(is_in_trtmnt_unit, inrange_th_votes>=5) %>% 
        sf::st_union()
      , .predicate = sf::st_disjoint
    )
  # nrow(xxx_treetops_sf)
  warning(paste0("reference_prediction_match: ",nm,"....",Sys.time()),immediate. = T,call. = F)
  ################### map it
  xxx_gt_pred_match_map_ans <- validation_trees$plot %>% 
    unique() %>% 
    purrr::set_names() %>% 
    purrr::map(function(plt_id) {
      # reference_prediction_match
      rp_match_ans_temp <- reference_prediction_match(
        pred_data = 
          xxx_treetops_sf %>% 
            dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
            sf::st_intersection(
              validation_plots %>% 
                dplyr::filter(plot==plt_id) %>%  
                dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% 
                sf::st_transform(sf::st_crs(xxx_treetops_sf)) %>% 
                sf::st_buffer(.$plot_radius_m) %>% 
                dplyr::select(plot_area_m2)
            ) %>% 
            dplyr::left_join(
              xxx_treetops_sf %>% 
                # clip to study boundary
                sf::st_intersection(
                  validation_plots %>% 
                    dplyr::filter(plot==plt_id) %>%  
                    sf::st_transform(sf::st_crs(xxx_treetops_sf)) %>% 
                    sf::st_buffer(.$plot_radius_m) %>% 
                    dplyr::mutate(is_in_stand = T)
                ) %>% 
                sf::st_drop_geometry() %>% 
                dplyr::select(treeID, is_in_stand)
              , by = "treeID"
            ) %>% 
            dplyr::mutate(
              is_in_stand = dplyr::coalesce(is_in_stand,F)
            )
        , ref_data = 
          validation_trees %>% 
            dplyr::filter(plot==plt_id) %>%
            dplyr::rename(tree_height_m=height_m, dbh_cm=dbh_cm, tree_cbh_m=cbh_m) %>% 
            sf::st_transform(sf::st_crs(xxx_treetops_sf))
        , max_dist_m = 2 # round(median(validation_trees$crown_width_m)/2)
        , max_height_error_m = 1.5 # round(sd(validation_trees$height_m)*1.1,1)
      )
      # # wha?
      # rp_match_ans_temp$ref_data %>% dplyr::glimpse()
      # rp_match_ans_temp$pred_data %>% dplyr::glimpse()
      
      gt_temp <- ground_truth_prediction_match(
        reference_prediction_match_ans = rp_match_ans_temp
        , comp_cols = c("tree_height_m", "dbh_cm") #, "tree_cbh_m")
      )
      
      if(!is.null(rp_match_ans_temp$pred_data) && nrow(rp_match_ans_temp$pred_data)>0){
        gt_temp <- gt_temp %>% 
          # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
          dplyr::left_join(
            rp_match_ans_temp$pred_data %>% 
              sf::st_drop_geometry() %>% 
              dplyr::select(pred_match_idx, is_in_stand)
            , by = "pred_match_idx"
          ) %>% 
          dplyr::filter(
            !(!is_in_stand & match_grp=="commission")
          ) 
      }
      # gt_temp %>% agg_ground_truth_match() %>% dplyr::glimpse()
      return(gt_temp)
  })
  xxx_ground_truth_prediction_match_ans <- 
    xxx_gt_pred_match_map_ans %>%
    purrr::compact() %>% 
    purrr::list_rbind(names_to = "plot")
  ans <- xxx_ground_truth_prediction_match_ans %>% 
    sf::st_drop_geometry() %>% 
    agg_ground_truth_match()
  ans %>% readr::write_csv(file.path("../data/testtest", paste0(nm,".csv")))
  return(ans)
}

# testtest_ans <- 
#   names(ws_functions) %>%
#   # .[2:3] %>% 
#   purrr::set_names() %>% 
#   purrr::map(testtest_fn)
# testtest_ans %>% 
#   purrr::compact() %>% 
#   purrr::list_rbind(names_to = "vws_function_nm") %>% 
#   dplyr::glimpse()

readr::write_csv(file.path("../data/testtest", paste0(nm,".csv")))

p_temp <- 
  testtest_ans %>% 
  purrr::compact() %>% 
  purrr::list_rbind(names_to = "vws_function_nm") %>% 
  ggplot(aes(x=recall,y=precision,color=f_score,label=vws_function_nm)) +
  geom_point() +
  scale_color_viridis_c()
p_temp
plotly::ggplotly(p_temp)
```

for this project, we're going to set a minimum height to qualify as a "tree" which will be applied across study sites. Since many field-measured data do not include height measurements but included diameter limits for measurement, this height threshold will align with measuring only larger trees, ignoring seedlings, saplings, and shrubs.


## CSFS SWE

### Point Cloud Data

UAS-lidar data was collected with a DJI Matrice M350 (DJI, Shenzhen, China) platform equipped with  a Zenmuse L2 lidar sensor.

let's check out the raw ALS data

```{r}
# csfs_swe
csfs_swe_las_f_temp <- "../data/csfs_swe/las"
csfs_swe_las_ctg_temp <- lidR::readLAScatalog(csfs_swe_las_f_temp)
csfs_swe_las_ctg_temp
# csfs_swe_las_ctg_temp@data %>% dplyr::glimpse()
```

let's check out the lidar data coverage compared to the field plot data

```{r}
csfs_swe_las_ctg_temp@data$geometry %>% 
  sf::st_union() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(data = "UAS-lidar data extent") %>% 
mapview::mapview(
 zcol = "data"
  , layer.name="CSFS SWE: UAS-lidar data"
  , popup=F
)
# mapview::mapview(
#   csfs_swe_trees
#   , zcol = "species"
#   , layer.name="WFSI Eglin AFB: field trees"
#   , popup=F
# )
# 40.59503,-105.9903
```

use `cloud2trees` to process the raw point cloud data to generate a CHM for use in tree detection

```{r}
################################################
# input/output data str
################################################
# outdir
c2t_output_dir_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="csfs_swe") %>% 
  dplyr::pull(data_dir)
c2t_process_dir_temp <- file.path(c2t_output_dir_temp, "point_cloud_processing_delivery")
# timer file
c2t_tracking_fnm_temp <- file.path(c2t_process_dir_temp, "processed_tracking_data.csv")
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    c2t_process_dir = ifelse(data_dir==c2t_output_dir_temp, c2t_process_dir_temp, c2t_process_dir)
    , c2t_tracking_fnm = ifelse(data_dir==c2t_output_dir_temp, c2t_tracking_fnm_temp, c2t_tracking_fnm)
  )
# project_data_df %>% print(n=nrow(project_data_df))

################################################
# use point cloud point density
# to automatically set CHM and DTM res
# make a fn to do this based on input LAScatalog
################################################

# get point cloud density and raster resolutions
get_ptcld_density_chm_res_temp <- get_ptcld_density_chm_res_fn(csfs_swe_las_ctg_temp)
# get_ptcld_density_chm_res_temp

# fnm
chm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("chm_",get_ptcld_density_chm_res_temp$chm_res_m,"m.tif") )
dtm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("dtm_",get_ptcld_density_chm_res_temp$dtm_res_m,"m.tif") )
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    , las_pts_m2 = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$pts_m2, las_pts_m2)
    , chm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$chm_res_m, chm_res_m)
    , chm_fnm = ifelse(data_dir==c2t_output_dir_temp, chm_fnm_temp, chm_fnm)
    , dtm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$dtm_res_m, dtm_res_m)
    , dtm_fnm = ifelse(data_dir==c2t_output_dir_temp, dtm_fnm_temp, dtm_fnm)
  )

# project_data_df %>% dplyr::filter(data_dir==c2t_output_dir_temp) %>% dplyr::glimpse()
# project_data_df %>% print(n=nrow(project_data_df))

##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists(chm_fnm_temp)
  || !file.exists(dtm_fnm_temp)
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir_temp
    , input_las_dir = csfs_swe_las_f_temp
    , accuracy_level = 2
    , keep_intrmdt = T
    , dtm_res_m = get_ptcld_density_chm_res_temp$dtm_res_m # (1/sqrt(pts_m2_temp))*2
    , chm_res_m = get_ptcld_density_chm_res_temp$chm_res_m # 1/sqrt(pts_m2_temp)
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm_temp
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( dtm_fnm_temp )
  chm_temp <- terra::rast( chm_fnm_temp )
  
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```


## Penobscot Experimental Forest (Maine)

### Field Data

Field data from this study site is from: 

* Kenefic, Laura S.; Allogio, Jeanette A.; Gerndt, Kathryn M.; Knapp, Rachel A.; Rogers, Nicole S.; Puhlick, Joshua J. 2025. Overstory tree and regeneration data from the "Silvicultural Effects on Composition, Structure, and Growth" study at Penobscot Experimental Forest. 3rd Edition. Fort Collins, CO: Forest Service Research Data Archive. [https://doi.org/10.2737/RDS-2012-0008-3](https://doi.org/10.2737/RDS-2012-0008-3)

>Data were collected on a series of permanent sample plots distributed across management units receiving various silvicultural treatments. Plot data are available in eight separate data sets. Overstory tree and sapling data include tree species, condition (e.g., merchantability status and cause of mortality), and diameter at breast height of living and dead trees, 1952 to 2023...Tree location data include the distance and azimuth from plot center of a subsample of trees, 2003 to 2023. Height and crown radius data include total height, height to crown base, and crown radius measurements for a subsample of live trees, 2003 to 2020 (heights) and 2003 to 2017 (radii)...Permanent sample plot location data include geospatial coordinates for permanent sample plots, 2018 and 2020...Future inventories are planned for this study indefinitely.

> 1/5th Acre Plots (were used to measure) Trees ≥ 4.5 inches DBH (with a) 52.7-foot plot radius

let's load in the plot data first to orient ourselves

```{r}
usfs_penobscotef_plots <- readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_PlotLocationsData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::filter(
    geodoubt == "N"
    , active == 1
    , !is.na(utm_east)
    , !is.na(utm_north)
  ) %>% 
  dplyr::mutate(
    plot_id = 
      mu %>%
        stringr::str_squish() %>% 
        stringr::str_c("-", as.character(plot))
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("utm_east", "utm_north"), crs = 26919, remove = F) %>% 
  dplyr::mutate(
    plot_area_ac = 1/5
    # 1 ha = 2.47105 acres
    , plot_area_ha = plot_area_ac/2.47105
    , plot_area_m2 = plot_area_ha*10000
    , plot_radius_meters = sqrt(plot_area_m2 / pi)
  ) %>% 
  sf::st_buffer(.$plot_radius_meters)
# huh?
# usfs_penobscotef_plots %>% dplyr::count(projection)
usfs_penobscotef_plots %>% dplyr::glimpse()
# usfs_penobscotef_plots %>% sf::st_drop_geometry() %>% dplyr::count(mu,plot) %>% dplyr::filter(n>1)
```

we did a lot of data cleaning in there which we won't describe (you can read the code if you're interested). let's look at the plot locations...

load in the trees

```{r}
usfs_penobscotef_trees <- 
  readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_TreeLocationData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::left_join(
    usfs_penobscotef_plots %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(mu,plot,plot_utm_north=utm_north,plot_utm_east=utm_east)
    , by = dplyr::join_by(mu,plot)
  ) %>% 
  # convert from dist and azimuth to coords
  dplyr::mutate(
    # dist to m
    dist_m = dist/3.280839895
    # convert azimuth from degrees to radians
    , az_rad = az * pi / 180
    , tree_utm_east = plot_utm_east + (dist_m * sin(az_rad)),
    , tree_utm_north = plot_utm_north + (dist_m * cos(az_rad))
  ) %>% 
  # filter
  dplyr::filter(
    !is.na(tree_utm_east)
    , !is.na(tree_utm_north)
  ) %>% 
  # what if a tree has multiple locations? ... take the most recent
  dplyr::arrange(mu,plot,treenum,desc(year),desc(month),desc(inv)) %>% 
  dplyr::group_by(mu,plot,treenum) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(mu,plot,treenum,tidyselect::starts_with("tree_utm_")) %>% 
  # spatial that
  sf::st_as_sf(coords = c("tree_utm_east", "tree_utm_north"), crs = sf::st_crs(usfs_penobscotef_plots), remove = F)

# read and attach tree information
usfs_penobscotef_trees <- 
  usfs_penobscotef_trees %>% 
  dplyr::inner_join(
    readr::read_csv(
        "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_TreeData.csv"
        , progress = F
        , show_col_types = F
      ) %>% 
      dplyr::rename_with(~tolower(.x) %>% make.names())
    , by = dplyr::join_by(mu,plot,treenum)
    , relationship = "one-to-many"
  )
# huh?
usfs_penobscotef_trees %>% dplyr::glimpse()
# only keep plots we have trees for
usfs_penobscotef_plots <- usfs_penobscotef_plots %>% 
  dplyr::inner_join(
    usfs_penobscotef_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(mu,plot)
    , by = dplyr::join_by(mu,plot)
  )
```

let's check those field plots out to orient ourselves to the location

```{r}
usfs_penobscotef_plots %>% 
  mapview::mapview(
    layer.name="USFS Penobscot EF: field plots"
    , popup=F
  )
# usfs_penobscotef_plots %>% 
#   sf::st_union() %>% 
#   sf::st_bbox() %>% 
#   sf::st_as_sfc() %>% 
#   sf::st_centroid() %>% 
#   sf::st_transform(crs=4326) %>% 
#   sf::st_coordinates()
# 44.85681,-68.62513
```






ForestGEO SCBI

```{r}
boundary_scbi_temp <- sf::st_read("../data/ForestGEO_SCBI/ForestGEO_SCBI_shapefiles/ForestGEO_grid_outline.shp") %>% 
  sf::st_union() 
boundary_scbi_temp %>% 
  mapview::mapview(layer.name = "SCBI")
boundary_scbi_temp %>% sf::st_centroid() %>% sf::st_transform(4326) %>% sf::st_coordinates()
# 38.893480, -78.145380
```






The `cloud2trees` package can be installed by following the directions listed in the [README file on GitHub](https://github.com/georgewoolsey/cloud2trees). If one is still experiencing difficulties installing the package, see the [setup guide](https://georgewoolsey.github.io/cloud2trees/articles/cloud2trees-setup.html) which details how to install the package using a virgin R instance.

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get cloud2trees
remotes::install_github(repo = "georgewoolsey/cloud2trees", upgrade = F)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```
