# Data and Process Overview

We compiled a collection of validation datasets across diverse forest conditions for this specific analysis project. These datasets consist of coincident aerial point cloud data (including USGS 3DEP ALS, UAS-lidar, and UAS-SfM) paired with field-measured stem maps or plot inventories where species identities are known.

| Name | State | Forest Type | Point Cloud Data | Field Data |
| :--- | :--- | :--- | :--- | :--- |
| Penobscot Experimental Forest | Maine | northern mixed conifer | ALS | tree locations |
| Manitou Experimental Forest (N1) | Colorado | ponderosa pine | UAS-SfM | tree locations |
| Manitou Experimental Forest (Big Plot) | Colorado | western mixed conifer | UAS-SfM | tree locations |
| CSFS State Forest State Park | Colorado | lodgepole pine | UAS-lidar | plots |
| BLM Dawson Creek | Colorado | pinyon-juniper | UAS-SfM | tree locations |
| Ute Creek | Colorado | | UAS-SfM | plots |
| WFSI Fort Stewart | Georgia | southern yellow pine forest | UAS-lidar | plots |
| WFSI Elgin AFB | Florida | longleaf pine | UAS-lidar | plots |
| Wabikon Lake Forest Dynamics Plot (ForestGEO) | Wisconsin | northern mixed hardwood | ALS | tree locations |
| UC Santa Cruz Forest Ecology Research Plot (ForestGEO) | California | coastal mixed-evergreen & coast redwood | ALS | tree locations |
| Smithsonian Environmental Research Center (ForestGEO) | Maryland | mixed deciduous hardwood | ALS | tree locations |
| Smithsonian Conservation Biology Institute (ForestGEO) | Virginia | mid-atlantic deciduous | ALS | tree locations |

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # the cloud2trees
```

```{r pkg-ld, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's make a quick data frame of our project data which is organized by site

```{r}
project_data_df <-
  dplyr::tibble(
   data_dir = list.dirs("../data", recursive = F) 
  ) %>% 
  dplyr::mutate(
    data_site_nm = basename(data_dir)
  )
# project_data_df$data_site_nm
# make a list for the CHM, predicted trees, field trees, and plot spatial data we can add to
chm_list <- setNames( vector("list", nrow(project_data_df)), project_data_df$data_site_nm)
# chm_list
# names(chm_list)
pred_trees_list <- chm_list
ref_trees_list <- chm_list
ref_plots_list <- chm_list
```

We're going to use this first study site and associated data to demonstrate the entire analysis process which includes:

* [Reference Field Data](#01_fld_dta): load and attempt to standardize the field reference spatial tree inventory and polygon plot data
* [Point Cloud Data](#02_ptcld_dta): process the raw point cloud data to generate the CHM used for individual tree detection
* [Individual Tree Detection](#03_itd): Use the CHM for individual tree detection. This includes many steps.
  - Utilize multiple window functions for ITD
  - Evaluate detection accuracy with segmentation matching of the predicted crowns with the field reference points
  - Select the best performing window function with the ITD result used for the remaining species method validation
* [Species/Genera Prediction](#04_species): Predict species for the point cloud-generated tree list
* [Tree-level Species/Genera Evaluation]: Evaluate the tree-level accuracy of the proposed classification method
* [Site-level Species/Genera Evaluation]: Evaluate the site-level accuracy of the proposed classification method based on proportional distributions

## Eglin Air Force Base (Florida)

### Field Data{#01_fld_dta}

Field data from this study site is from: 

* Andrew Hudak, Seth Bigelow, Nuria Sánchez-López, and Vanessa Niemczyk. 2025. Field measurements of coarse wood, duff, litter, and trees for the Eglin Air Force Base 2023 fuels campaign. Wildland Fire Science Initiative (WFSI) [https://doi.org/10.60594/W4RG6P](https://doi.org/10.60594/W4RG6P)

>Field measurements collected before and after 2023 study fires at Eglin Air Force Base (AFB) to document fuel loads (large diameter fuels, litter depth, duff depth) and trees in a 3-year rough in sandhill longleaf pine, and the effects of study fires on these variables. Fuel loads, which were measured along transects, were co-located with tree observations, which were measured within fixed-radius plots. These field observations were also co-located with three-dimensional fuel and laser scan plots to allow for leveraging of these other data.

>After the study fires in March 2023, trees within 10 m of plot center were identified to functional group (pine, hardwood or species), measured for diameter at breast height with a diameter tape, measured for distance from outer bark to plot center, and measured for azimuth from plot center to tree. Some post-processing was done using lidar scans to distinguish between evergreeen and deciduous hardwoods. Saplings were tallied by ordinal quadrant (up to 5 m from plot center). Tree measurements were not taken in unit L25. In unit G20, plots HH01, HH02, HH03, HH04, HH05 were sampled pre-fire but not post fire. In unit G25, plots HH105, HH95, PP16 were sampled pre-fire but not post fire.

according to *Eglin_2022_Tree_and_Transect_Field_Protocol.pdf*, the post-fire sampling protocol for trees was:

I. Determine if plot radius is 8 or 11.3 meters and record on datasheet
  a. If there are 4 or more trees within an 8-meter radius, then the plot radius is 8-meters.
  b. If there are 0 to 3 trees within an 8-meter radius, then increase the plot radius to 11.3-meters.
II. Sample all trees ≥10 cm DBH within the plot radius and over from plot center 
III. Sample all pine saplings ONLY ≥2.5 to 10 cm DBH by recording:
  a. Radius (4 or 5.6m):
    i. If tree sampling radius was 8-meters, then sapling sampling radius = 4-meters
    ii. If tree sampling radius was 11.3-meters, then sapling sampling radius = 5.6-meters
  b. Quadrant = NE, SE, SW, or NW
  c. Count = How many of the same species in that quadrant
  d. Species = Longleaf pine (Pinus palustris), Loblolly pine (Pinus taeda), Sand pine (Pinus clausa) or TBD

for those interested, pre- and post-burn photographs for each plot were archived here: [https://doi.org/10.60594/W4GW2T](https://doi.org/10.60594/W4GW2T)

```{r}
# https://doi.org/10.60594/W4RG6P
wfsi_eglinafb_trees <-
  readr::read_csv(
    "../data/WFSI_EglinAFB/Eglin_Air_Force_Base_2023_fuels/data/trees.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  # dplyr::count(species)
  dplyr::filter(
    species != ""
    , !is.na(species)
    , !is.na(tree_x)
    , !is.na(tree_y)
  ) %>% 
  # spatial that
  # epsg from "data_dictionary.csv"
  sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = 26916, remove = F) %>% 
  # make a plot_id that is unique for all so we don't have to join by multiple columns
  # and for all sites we'll just use a plot_id
  dplyr::mutate(
    plot_id = stringr::str_c(unit,plot,sep = "_")
    # also, each tree needs a treeID
    , treeID = dplyr::row_number()
  )
# huh?
wfsi_eglinafb_trees %>% dplyr::glimpse()
# wfsi_eglinafb_trees %>% dplyr::select(dbh_cm) %>% summary()
```

let's check those field trees out to orient ourselves to the location

```{r}
wfsi_eglinafb_trees %>% 
  mapview::mapview(
    zcol = "species"
    , layer.name="WFSI Eglin AFB: field trees"
    , popup=F
  )
# wfsi_eglinafb_trees %>%
#   sf::st_union() %>%
#   sf::st_bbox() %>%
#   sf::st_as_sfc() %>%
#   sf::st_centroid() %>%
#   sf::st_transform(crs=4326) %>%
#   sf::st_coordinates()
# # 30.53834,-86.7296
```

what is the count of trees measured pre- and post-burn?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(burn_status)
```

so these are all post-burn samples. what is the distribution of species?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species) %>% 
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

how many plots have trees by unit?

```{r}
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::distinct(unit,plot) %>% 
  dplyr::count(unit)
```

we need to get the plot location points and buffer them by 10 m of plot center which was said to be the field sampling protocol to "*Sample all trees ≥10 cm DBH within the plot radius and over from plot center*" ...clear as mud!

```{r}
# https://doi.org/10.60594/W4T880
wfsi_eglinafb_plots <-
  readr::read_csv(
    "../data/WFSI_EglinAFB/Macroplot_Geographical_Location_Data/data/Geographical_Macroplot_Data_Eglin_2023.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::filter(
    !is.na(latitude_y)
    , !is.na(longitude_x)
    , !is.na(macroplot)
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("longitude_x", "latitude_y"), crs = 4326, remove = F) %>% 
  # converts proj to match trees and horizontal units to m
  sf::st_transform(sf::st_crs(wfsi_eglinafb_trees)) %>% 
  # "After the study fires in March 2023, trees within 10 m of plot center were identified..."
  sf::st_buffer(10) %>% 
  dplyr::mutate(
    plot_area_m2 = sf::st_area(.) %>% as.numeric()
    # this is good data mgmt: make sure to specify unit orientation in the plot data but not the tree data
    , burn_unit = burn_unit %>% stringr::str_remove("[EWSNewsn]$")
    # # another good practice is to format the plot id's to have padding of two zeros in the plot data but only padding of one zero in the tree data
    , macroplot = dplyr::case_when(
      stringr::str_sub(macroplot, 3, 4) == "00" ~ stringr::str_c( stringr::str_sub(macroplot,1,2), "0", stringr::str_sub(macroplot,5) )
      , stringr::str_sub(macroplot, 3, 3) == "0" ~  stringr::str_c( stringr::str_sub(macroplot,1,2), stringr::str_sub(macroplot,4) )
      , T ~ macroplot
    )
  ) %>% 
  dplyr::rename(
    plot = macroplot
    , unit = burn_unit
  )

# wfsi_eglinafb_plots %>% dplyr::glimpse()
# wfsi_eglinafb_plots$plot %>% unique()

# # check for plot and unit non-matches
# wfsi_eglinafb_trees %>%
#   sf::st_drop_geometry() %>%
#   dplyr::distinct(unit,plot) %>%
#   dplyr::anti_join(
#     wfsi_eglinafb_plots %>%
#       sf::st_drop_geometry() %>%
#       dplyr::distinct(unit,plot,plot_area_m2)
#     , by = dplyr::join_by(unit,plot)
#   ) %>%
#   dplyr::glimpse()
  
# lets keep only plots for which tree data exists since we know those were measured post-burn
wfsi_eglinafb_plots <- wfsi_eglinafb_plots %>% 
  dplyr::inner_join(
    wfsi_eglinafb_trees %>%
      sf::st_drop_geometry() %>%
      dplyr::distinct(unit,plot,plot_id)
    , by = dplyr::join_by(unit,plot)
  )
  
# wfsi_eglinafb_plots %>% dplyr::glimpse()

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
ref_plots_list[[nm_temp]] <- wfsi_eglinafb_plots
# ref_plots_list

```

let's quickly check the trees and plots to make sure they align

```{r}
# wfsi_eglinafb_trees %>% dplyr::glimpse()
# plot function
p_fn_temp <- function(nr){
    p <- wfsi_eglinafb_plots %>% dplyr::slice(nr)
    t <- wfsi_eglinafb_trees %>% 
      dplyr::inner_join(
          p %>% 
          dplyr::select(unit,plot) %>%
          sf::st_drop_geometry()
        , by = dplyr::join_by(unit,plot)
      )
    ggplot2::ggplot() + 
    ggplot2::geom_sf(
      data = t
      , mapping = ggplot2::aes(color = dbh_cm)
      , size = 2.5
    ) + 
    ggplot2::geom_sf(
      data = p
      , color = "gray33"
      , fill = NA
      , lwd = 1.5
    ) + 
    ggplot2::geom_sf(
      data = t
      , mapping = ggplot2::aes(color = dbh_cm)
      , size = 2.5
    ) + 
    ggplot2::scale_color_viridis_c(
      option = "mako"
      , limits = c(0, ceiling(max(wfsi_eglinafb_trees$dbh_cm,na.rm=T)*1.01))
      , direction = -1
      , name = "DBH (cm)"
    ) +
    ggplot2::labs(
      subtitle = paste0(p$unit, " | ", p$plot)
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
      plot.subtitle = ggplot2::element_text(size = 8, hjust = 0.5, face = "bold")
    )
  }

pl_temp <- 
  1:nrow(wfsi_eglinafb_plots) %>% 
  sample(12) %>% 
  purrr::map(p_fn_temp)
patchwork::wrap_plots(pl_temp,ncol=4,guides = "collect")
```

spot check of alignment between plot and tree spatial data looks good but we can quickly look at all plots and trees to see if any trees fall outside of the plot boundaries

```{r}
# wfsi_eglinafb_trees %>% dplyr::glimpse()
wfsi_eglinafb_trees <- 
  1:nrow(wfsi_eglinafb_plots) %>% 
  purrr::map(function(nr){
    p <- wfsi_eglinafb_plots %>% dplyr::slice(nr)
    t <- wfsi_eglinafb_trees %>% 
      dplyr::inner_join(
          p %>% 
          dplyr::select(unit,plot) %>%
          sf::st_drop_geometry()
        , by = dplyr::join_by(unit,plot)
      )
    # points outside?
    out <- t %>% sf::st_filter(p, .predicate = sf::st_disjoint)
      
    return(
      t %>% 
        dplyr::left_join(
          out %>% 
            sf::st_drop_geometry() %>% 
            dplyr::select(plot,unit,tree_number) %>% 
            dplyr::mutate(is_out_of_plot = T)
          , by = dplyr::join_by(plot,unit,tree_number)
        ) %>% 
        dplyr::mutate(
          is_out_of_plot = dplyr::coalesce(is_out_of_plot,F)
        )
    )
  }) %>% 
  purrr::list_rbind()

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
ref_trees_list[[nm_temp]] <- wfsi_eglinafb_trees
# ref_trees_list 

# are there any out?
wfsi_eglinafb_trees %>% sf::st_drop_geometry() %>% dplyr::count(is_out_of_plot)
```

uh-oh we have one tree outside of the plot

```{r}
wfsi_eglinafb_plots %>% 
  dplyr::mutate(rn=dplyr::row_number()) %>% 
  dplyr::inner_join(
    wfsi_eglinafb_trees %>% 
      dplyr::filter(is_out_of_plot) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(unit,plot)
    , by = dplyr::join_by(unit,plot) 
  ) %>% 
  dplyr::pull(rn) %>% 
  purrr::map(p_fn_temp)

```

right on the edge, but just outside based on a slight misalignment in GPS locations or in/out determination in the field

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Point Cloud Data{#02_ptcld_dta}

Andrew Hudak, & Benjamin Bright. (2026). *Airborne Lidar Data from the 2023 Department of Defense Wildland Fire Science Initiative Eglin Air Force Base Campaign*. WFSI Data Portal. [doi:10.60594/W4B017](https://doi.org/10.60594/W4B017)

>This dataset contains airborne light detection and ranging (lidar) data across burn units G20, G25, and L25 in Eglin AFB, FL. Kucera International acquired airborne lidar on January 28, 2023 (before prescribed fire), and on April 26, 2023 (after prescribed fire); prescribed fires occurred from March 11 to 19, 2023. Lidar data were acquired to characterize pre- and post-fire vegetation and estimate fire-caused biomass consumption. 

some other relevant notes about the data:

>Kucera International acquired airborne lidar data across burn units G20, G25, and L25 on January 28, 2023 (before prescribed fire), and across burn units G20 and G25 on April 26, 2023 (after prescribed fire); prescribed fires occurred from March 11 to 19, 2023. The lidar sensor, an Optech Galaxy T2000, was mounted on a PA-31 Piper Navajo Chieftain fixed-wing aircraft, which flew at an elevation of 1341 m (4400 ft) above ground level at a speed of about 150 knots, with a planned sidelap of 30%. The Optech Galaxy T2000 scanned with a laser pulse rate of 1600 kHz, a scan rate of 150 Hz, and a scan field of view of 30 degrees, which resulted in swath widths of 716 meters (2350 feet) and an average point density of 15 points per square meter.

>LAZ file horizontal distance units are meters, in the UTM Zone 16N NAD83 (2011) geographic coordinate system (EPSG 6345)

let's check out the raw ALS data

```{r}
wfsi_eglinafb_las_f_temp <- "../data/WFSI_EglinAFB/post_fire_laz_apr2023"
wfsi_eglinafb_las_ctg_temp <- lidR::readLAScatalog(wfsi_eglinafb_las_f_temp)
wfsi_eglinafb_las_ctg_temp
# wfsi_eglinafb_las_ctg_temp@data %>% dplyr::glimpse()
```

let's check out the lidar data coverage compared to the field plot data

```{r}
wfsi_eglinafb_las_ctg_temp@data$geometry %>% 
  sf::st_union() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(data = "ALS data extent") %>% 
mapview::mapview(
 zcol = "data"
  , layer.name="WFSI Eglin AFB: ALS data"
  , popup=F
) +
mapview::mapview(
  wfsi_eglinafb_trees
  , zcol = "species"
  , layer.name="WFSI Eglin AFB: field trees"
  , popup=F
)
```

use `cloud2trees` to process the raw point cloud data to generate a CHM for use in tree detection

```{r}
################################################
# input/output data str
################################################
# outdir
c2t_output_dir_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_dir)
c2t_process_dir_temp <- file.path(c2t_output_dir_temp, "point_cloud_processing_delivery")
# timer file
c2t_tracking_fnm_temp <- file.path(c2t_process_dir_temp, "processed_tracking_data.csv")
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    c2t_process_dir = ifelse(data_dir==c2t_output_dir_temp, c2t_process_dir_temp, "")
    , c2t_tracking_fnm = ifelse(data_dir==c2t_output_dir_temp, c2t_tracking_fnm_temp, "")
  )
# project_data_df %>% print(n=nrow(project_data_df))

################################################
# use point cloud point density
# to automatically set CHM and DTM res
# make a fn to do this based on input LAScatalog
################################################

get_ptcld_density_chm_res_fn <- function(las_ctg, rast_res_buffer = 0.01, min_chm_res_m = 0.15, chm_factor_for_dtm = 2) {
  rast_res_buffer <- dplyr::coalesce(as.numeric(rast_res_buffer),0)
  min_chm_res_m <- dplyr::coalesce(as.numeric(min_chm_res_m),0.15)
  chm_factor_for_dtm <- dplyr::coalesce(as.numeric(chm_factor_for_dtm),2)
  
  if(!inherits(las_ctg, "LAScatalog")){stop("must be a `LAScatalog` read with `lidR::readLAScatalog()`")}
  if(
    !( methods::slotNames(las_ctg) %>% stringr::str_equal("data") %>% any() )
  ){stop("can't find `data` layer in `LAScatalog` read with `lidR::readLAScatalog()`")}
  if(!inherits(las_ctg@data, "sf")){stop("non-sf `data` from `LAScatalog` read with `lidR::readLAScatalog()`")}
  
  # some point clouds have tiles with lots of overlap
  # some point clouds have tiles with no overlap
  
  # union over total area will help us identify overlap
  # high overlap >> union will be much smaller than total area (e.g., uoa ratio < 0.66)
  # low overlap >> union will be similar to total area (e.g., uoa ratio > 0.88)
  
  # union area
  las_uoa_temp <- 
  (
    las_ctg@data %>% 
    dplyr::ungroup() %>% 
    sf::st_union() %>% 
    sf::st_area() %>% 
    as.numeric()
  ) /
  # tot area
  (
    las_ctg@data %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(
      area_m2 = sf::st_area(geometry) %>% as.numeric()
    ) %>% 
    dplyr::summarise(area_m2=sum(area_m2,na.rm=T)) %>% 
    dplyr::pull(area_m2)
  )
  # las_uoa_temp
  # when a lot of overlap, union and calculate density based off union
  if(las_uoa_temp<0.66){
    pts_m2_temp <- 
      sum(las_ctg@data$Number.of.point.records,na.rm=T) /
      ( las_ctg@data$geometry %>% sf::st_union() %>% sf::st_area() %>% as.numeric() )  
  }else{
  # when NOT a lot of overlap, calculate density as-is
    pts_m2_temp <- 
      las_ctg@data %>% 
      dplyr::mutate(
        area_m2 = sf::st_area(geometry) %>% as.numeric()
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::ungroup() %>% 
      dplyr::summarise(
        pts_m2 = sum(Number.of.point.records,na.rm = T)/sum(area_m2,na.rm = T)
      ) %>% 
      dplyr::pull(pts_m2)
  }
  # pts_m2_temp
  # 1/sqrt(pts_m2_temp)
  
  chm_res_m_temp <- max(round(1/sqrt(pts_m2_temp),2) + rast_res_buffer, min_chm_res_m) 
  dtm_res_m_temp <- round(as.numeric(chm_res_m_temp)*chm_factor_for_dtm,2)
  
  return(list(
    pts_m2 = pts_m2_temp
    , chm_res_m = chm_res_m_temp
    , dtm_res_m = dtm_res_m_temp
  ))
}

# get point cloud density and raster resolutions
get_ptcld_density_chm_res_temp <- get_ptcld_density_chm_res_fn(wfsi_eglinafb_las_ctg_temp)
# get_ptcld_density_chm_res_temp

# fnm
chm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("chm_",get_ptcld_density_chm_res_temp$chm_res_m,"m.tif") )
dtm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("dtm_",get_ptcld_density_chm_res_temp$dtm_res_m,"m.tif") )
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    las_pts_m2 = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$pts_m2, as.numeric(NA))
    , chm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$chm_res_m, as.numeric(NA))
    , chm_fnm = ifelse(data_dir==c2t_output_dir_temp, chm_fnm_temp, "")
    , dtm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$dtm_res_m, as.numeric(NA))
    , dtm_fnm = ifelse(data_dir==c2t_output_dir_temp, dtm_fnm_temp, "")
  )

# project_data_df %>% dplyr::filter(data_dir==c2t_output_dir_temp) %>% dplyr::glimpse()

##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists(chm_fnm_temp)
  || !file.exists(dtm_fnm_temp)
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir_temp
    , input_las_dir = wfsi_eglinafb_las_f_temp
    , accuracy_level = 2
    , keep_intrmdt = T
    , dtm_res_m = get_ptcld_density_chm_res_temp$dtm_res_m # (1/sqrt(pts_m2_temp))*2
    , chm_res_m = get_ptcld_density_chm_res_temp$chm_res_m # 1/sqrt(pts_m2_temp)
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm_temp
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( dtm_fnm_temp )
  chm_temp <- terra::rast( chm_fnm_temp )
  
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

we'll set the minimum CHM resolution to 0.25 m to match the `cloud2trees` default. if the automatically-determined CHM resolution based on the point cloud density used in the `cloud2trees::cloud2raster()` was lower than this level, we'll aggregate the raster using `terra::aggregate()` and `terra::resample()` to get the exact target resolution. if the determined CHM resolution was higher than this level due to low point density, we will use the CHM as-is since this is already the finest resolution possible with the point cloud density.

```{r}
min_chm_res <- 0.25
##################################################################
# aggregate to make raster more coarse for tree detection
##################################################################
agg_chm_rast_fn <- function(chm, res, dir) {
  fp <- file.path(dir, paste0("chm_",res,"m.tif"))
  if(terra::res(chm)[1]<res){
    # first, we'll borrow from the `cloud2trees` codebase to get a function to change the resolution of a raster exactly
    ###___________________________________________###
    # adjust the resolution of a raster to be in exactly the target resolution
    ###___________________________________________###
    if(
      !file.exists(fp)
    ){
      agg_chm_rast <- cloud2trees:::adjust_raster_resolution(
        chm
        , target_resolution = res
        , fun = max
        , resample_method = "max"
        , ofile = fp
      )
    }else{
      agg_chm_rast <- terra::rast( fp )
    }
  }else{
    agg_chm_rast <- chm  
  }
  return(agg_chm_rast)
}

```

let's aggregate the CHM for this site and put it into a list of CHM rasters that we can map over in all of the processing

```{r}
agg_chm_rast_temp <- agg_chm_rast_fn(
  chm = cloud2raster_ans$chm_rast
  , res = min_chm_res
  , dir = 
    project_data_df %>% 
    dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
    dplyr::pull(c2t_process_dir)
)
# agg_chm_rast_temp
# agg_chm_rast_temp %>% terra::clamp(lower = 1.37, values = F) %>% terra::plot()

# add to the chm list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
chm_list[[nm_temp]] <- agg_chm_rast_temp
# chm_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Tree Detection{#03_itd}

`cloud2trees` implements a CHM-based individual tree detection method accessible via `cloud2trees::raster2trees()`. this method requires a window function used to set the search radius used to find local maxima in the CHM. These local maxima represent individual tree top points and are subsequently used to delineate individual tree crowns via a marker-controlled watershed segmentation

because the objective of this study is to evaluate a species/genera classification method, we just want the best ITD results we can get based on the field reference data. We are not interested in evaluating tree segmentation here, that has already been done across varied input data and forest types.

let's make a function to test many different window functions, compare the segmentation results to the field reference data, then select the best window function based on tree detection accuracy.

```{r}
# test a bunch of different variable window sizes within `lidR::lmf()` 
# that cover a range of possibilities and function shapes including linear, concave up, and constant
get_ws_fn_list <- function(min_ws = 0.5, max_ws = 6.0, n_linear = 40, n_concave = 30, n_cr = 30, max_ht = 45) {
  
  # empty list
  ws_functions <- list()
  
  ###########################################################################
  # linear functions over a range
  ###########################################################################
  # end points for the line
  linear_targets <- base::seq(from = max_ws, to = max_ws*4*10, length.out = n_linear)
  
  n_linear_pad <- ceiling(n_linear*1.5)
  
  for (i in 1:n_linear_pad) {
    target_height <- linear_targets[i]
    slope <- (max_ws - min_ws) / target_height
    
    # Pad names with a zero (01 to 60) for clean sorting
    func_name <- base::paste0("linear_", base::sprintf("%02d", i))
    
    v_min  <- base::as.numeric(min_ws)
    v_max  <- base::as.numeric(max_ws)
    v_slope <- base::as.numeric(slope)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        (v_min + v_slope * x) < v_min ~ v_min,
        (v_min + v_slope * x) > v_max ~ v_max,
        TRUE ~ (v_min + v_slope * x)
      ),
      list(v_min = v_min, v_max = v_max, v_slope = v_slope)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  
  # the flat slopes are too close together...let's remove every other one
  # how many to remove
  num_to_remove <- length(ws_functions) - n_linear
  
  # half
  start_idx <- length(ws_functions) - (num_to_remove - 1) * 2
  
  # remove
  remove_idx <- seq(from = start_idx, to = length(ws_functions), by = 2)
  ws_functions <- ws_functions[-remove_idx]
  names(ws_functions) <- base::paste0("linear_", base::sprintf("%02d", c(1:length(ws_functions)) ))
  
  # ws_functions[1]
  # ws_functions[[1]]
  # class(ws_functions[[1]])
  # ws_functions[[1]](1.5)
  # ws_functions[[1]](0)
  # ws_functions[[1]](18)
  # 
  # ws_functions[[40]](1.5)
  # ws_functions[[40]](0)
  # ws_functions[[40]](18)
  
  ###########################################################################
  # concave down functions over a range
  ###########################################################################
  # total potential growth (or distance) between the floor and the ceiling is: 6.0-0.5=5.5
  # the halfway point of growth is exactly half of that distance: 5.5/2=2.75
  # when the curve reaches its exact halfway point, the resulting window size must be the baseline floor plus that half-distance: 0.5+2.75=3.25
  # in `half_sat`, setting the constant of 1.375 means the function will hit its halfway mark of 3.25 when the tree is only 1.375 meters tall
    # max slope = 4 (for ever 1 m growth, ws increases by 4 m)
    # 4 = (6.0-0.5)/c >> 4 = 5.5/c >> c = 5.5/4 >> c = 1.375
  # how far out to look based on maximum plot/tree height? 45 m tree??
  
  for (i in 1:n_concave) {
    c <- 1.375 # ~dbh height
    max_end <- max_ht # biggest tree everrrrr
    half_sat <- c + (i - 1) * ( (max_end-c) / (n_concave-1))
    
    func_name <- base::paste0("concave_down_", base::sprintf("%02d", i))
    
    v_min  <- base::as.numeric(min_ws)
    v_max  <- base::as.numeric(max_ws)
    v_half <- base::as.numeric(half_sat)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        (v_min + ((v_max - v_min) * x) / (v_half + x)) < v_min ~ v_min,
        (v_min + ((v_max - v_min) * x) / (v_half + x)) > v_max ~ v_max,
        TRUE ~ (v_min + ((v_max - v_min) * x) / (v_half + x))
      ),
      list(v_min = v_min, v_max = v_max, v_half = v_half)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  # ws_functions[length(ws_functions)]
  # ws_functions[length(ws_functions)-1]
  
  ###########################################################################
  # constant functions over a range
  ###########################################################################
  # range
  constant_values <- base::seq(from = min_ws, to = max_ws * 0.95, by = 0.5)
  
  for (i in base::seq_along(constant_values)) {
    const_val <- constant_values[i]
    
    func_name <- base::paste0("constant_", base::sprintf("%02d", i))
    
    v_min <- base::as.numeric(min_ws)
    v_val <- base::as.numeric(const_val)
    
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        TRUE ~ v_val
      ),
      list(v_min = v_min, v_val = v_val)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  # ws_functions[length(ws_functions)]
  # ws_functions[[length(ws_functions)]](0.1)
  # ws_functions[[length(ws_functions)]](4)
  # ws_functions[[length(ws_functions)]](15)
  # ws_functions[length(ws_functions)-1]
  
  
  # ###########################################################################
  # # two-step linear functions over a range
  # ###########################################################################
  # # how many functions to test
  # n_temp <- 16
  # # max tree height in CHM
  # 
  # # Calculate fractional boundary envelopes based on global constraints
  # xb_min <- max_ht * 0.09   # ~5m height
  # xb_max <- max_ht * 0.66   # ~25m height
  # yb_min <- min_ws + (max_ws - min_ws) * 0.36  # ~2.5m window size
  # yb_max <- min_ws + (max_ws - min_ws) * 0.88  # ~4.5m window size
  # 
  # xb_sequence  <- base::seq(from = xb_min, to = xb_max, length.out = n_temp)
  # yb_sequence  <- base::seq(from = yb_min, to = yb_max, length.out = n_temp)
  # cap_sequence <- base::seq(from = 15.0,   to = 40.0,   length.out = n_temp)
  # 
  # for (i in 1:n_temp) {
  #   x_b <- xb_sequence[i]
  #   y_b <- yb_sequence[i]
  #   
  #   slope1 <- (y_b - min_ws) / x_b
  #   
  #   target_cap_x <- x_b + cap_sequence[i]
  #   slope2 <- (max_ws - y_b) / (target_cap_x - x_b)
  #   
  #   func_name <- base::paste0("twostep_linear_", base::sprintf("%02d", i))
  #   
  #   v_min    <- base::as.numeric(min_ws)
  #   v_max    <- base::as.numeric(max_ws)
  #   v_xb     <- base::as.numeric(x_b)
  #   v_yb     <- base::as.numeric(y_b)
  #   v_slope1 <- base::as.numeric(slope1)
  #   v_slope2 <- base::as.numeric(slope2)
  #   
  #   func_body <- base::substitute(
  #     dplyr::case_when(
  #       base::is.na(x) ~ v_min,
  #       x <= v_xb ~ {
  #         val <- v_min + v_slope1 * x
  #         dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
  #       },
  #       x > v_xb ~ {
  #         val <- v_yb + v_slope2 * (x - v_xb)
  #         dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
  #       }
  #     ),
  #     list(v_min = v_min, v_max = v_max, v_xb = v_xb, v_yb = v_yb, v_slope1 = v_slope1, v_slope2 = v_slope2)
  #   )
  #   
  #   ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  # }
  # 
  # 
  # # ws_functions[length(ws_functions)]
  # # ws_functions[[length(ws_functions)]](0.1)
  # # ws_functions[[length(ws_functions)]](4)
  # # ws_functions[[length(ws_functions)]](15)
  # # ws_functions[length(ws_functions)-1]
  
  
  ###########################################################################
  # chapman richards functions over a range
  ###########################################################################
  # Generate parameters that smoothly spread across your MAX_CANOPY_HEIGHT
  # Beta controls horizontal stretching; Gamma controls the sharpness of the initial lag
  beta_sequence  <- base::seq(from = 0.15, to = 0.02, length.out = n_cr)
  gamma_sequence <- base::seq(from = 1.5,  to = 4.0,  length.out = n_cr)
  
  for (i in 1:n_cr) {
    b_param <- beta_sequence[i]
    g_param <- gamma_sequence[i]
    
    func_name <- base::paste0("chapman_richards_", base::sprintf("%02d", i))
    
    v_min   <- base::as.numeric(min_ws)
    v_max   <- base::as.numeric(max_ws)
    v_beta  <- base::as.numeric(b_param)
    v_gamma <- base::as.numeric(g_param)
    
    # Chapman-Richards equation with case_when boundaries
    func_body <- base::substitute(
      dplyr::case_when(
        base::is.na(x) ~ v_min,
        TRUE ~ {
          # Core Chapman-Richards Sigmoid Formula
          val <- v_min + (v_max - v_min) * (1 - base::exp(-v_beta * x))^v_gamma
          dplyr::case_when(val < v_min ~ v_min, val > v_max ~ v_max, TRUE ~ val)
        }
      ),
      list(v_min = v_min, v_max = v_max, v_beta = v_beta, v_gamma = v_gamma)
    )
    
    ws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
  }
  
  return(ws_functions)
}
```

let's take that function for a spin to get a list of window functions for testing and plot the functions over a height range

```{r}
get_ws_fn_list_ans <- get_ws_fn_list(min_ws = 0.5, max_ws = 7.0, n_linear = 30, n_concave = 30, n_cr = 30, max_ht = 45)
# get_ws_fn_list_ans %>% names()
# get_ws_fn_list_ans %>% length()

# cloud2trees::itd_ws_functions() %>% names()
# get_ws_fn_list_ans %>% names()
# 
# cloud2trees::itd_ws_functions()[[1]] %>% class()
# get_ws_fn_list_ans[[1]] %>% class()
# 
# cloud2trees::itd_ws_functions()[[1]](2.22) %>% class()
# get_ws_fn_list_ans[[1]](2.22) %>% class()
# 
# cloud2trees::itd_ws_functions()[[1]](2.22)
# get_ws_fn_list_ans[[1]](2.22)

# plot that
ht_seq_temp <- base::seq(0, 50, by = 0.25)

p_dta_temp <- 
  purrr::map_dfr(
  names(get_ws_fn_list_ans)
  , function(name) {
    tibble::tibble(
      height = ht_seq_temp
      , window_size = get_ws_fn_list_ans[[name]](ht_seq_temp)
      , function_id = name
      , shape_type = dplyr::case_when(
        stringr::str_starts(name,"linear_") ~ "linear"
        , stringr::str_starts(name,"concave_down_") ~ "concave down"
        , stringr::str_starts(name,"twostep_linear_") ~ "two-step linear"
        , stringr::str_starts(name,"constant_") ~ "constant"
        , stringr::str_starts(name,"chapman_richards_") ~ "chapman-richards"
      )
    )
  }
)
# p_dta_temp %>% dplyr::count(shape_type)
# p_dta_temp %>% dplyr::distinct(function_id,shape_type) %>% dplyr::count(shape_type)

p_temp <- 
  ggplot2::ggplot(
    data = p_dta_temp
    , mapping = ggplot2::aes(x = height, y = window_size, group = function_id, color = shape_type)
  ) +
    ggplot2::geom_line(alpha = 0.6, linewidth = 1) +
    ggplot2::labs(
      x = "CHM (m)",
      , y = "ws diameter (m)"
      , color = "ws type"
    ) +
    # ggplot2::facet_wrap(facets = dplyr::vars(shape_type)) +
    ggplot2::scale_color_viridis_d(option = "viridis", direction = -1) +
    ggplot2::scale_y_continuous(limits=c(0,NA), breaks = scales::breaks_extended(n=6)) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
    )
p_temp

```

let's look at the functions by type so they are a little easier to see since that is `r length(get_ws_fn_list_ans)` unique functions

```{r}
p_temp + 
  ggplot2::facet_wrap(facets = dplyr::vars(shape_type))
```

now we need a function to clip the CHM to the plot plus a buffer and run `cloud2trees::raster2trees()` for each window function. the return should be a spatial tree list unique by function.

```{r}
# function to loop over ws function list in raster2trees
ws_testtest_fn <- function(
    chm
    , buffer = 5
    , plot_sf
    , ws_fn_list # a named list of functions ex: cloud2trees::itd_ws_functions()
  ) {
  # min ht for chm so we can test non-na
  min_chm_ht <- 1.37
  # crop chm
  chm_crop <- chm %>% 
    terra::crop(
      plot_sf %>% 
        sf::st_transform(terra::crs(chm)) %>% 
        sf::st_buffer(buffer)
      , mask = T
    ) %>% 
    terra::clamp(lower = min_chm_ht, values = F)
  
  # terra::plot(chm_crop)
  # terra::plot(
  #   plot_sf %>%
  #     sf::st_transform(terra::crs(chm))
  #   , add = T
  #   , border = "cyan"
  #   , col = NA
  # )

  if( !any( terra::global(chm_crop, fun = "anynotNA") ) ){
    return(NULL)
  }
  
  # map over the function list
  ws_tree_list <- 
    names(ws_fn_list) %>%
    purrr::set_names() %>%
    purrr::map(function(nm){
      if (exists("my_temp_dir") && dir.exists(my_temp_dir)) {
        unlink(my_temp_dir, recursive = TRUE, force = TRUE)
      }
      if (exists("my_temp_dir2") && dir.exists(my_temp_dir2)) {
        unlink(my_temp_dir2, recursive = TRUE, force = TRUE)
      }
      warning(paste0("making dir: ",nm,"....",Sys.time()),immediate. = T,call. = F)
      my_temp_dir <- tempfile(pattern = nm)
      dir.create(my_temp_dir, showWarnings = FALSE)
      my_temp_dir2 <- tempfile(pattern = paste0("xx",nm))
      dir.create(my_temp_dir2, showWarnings = FALSE)
      warning(paste0("raster2trees: ",nm,"....",Sys.time()),immediate. = T,call. = F)
      
      safe_raster2trees <- purrr::safely( cloud2trees::raster2trees )
      xxx_raster2trees_ans <- safe_raster2trees(
          chm_rast = chm_crop
          , outfolder = my_temp_dir
          , ws = ws_fn_list[[nm]]
          , min_height = min_chm_ht
          , min_crown_area = 0.5
          , tempdir = my_temp_dir2
        )
      
      if(!is.null(xxx_raster2trees_ans$result)){
        xxx_raster2trees_ans <- xxx_raster2trees_ans$result
        # xxx_raster2trees_ans %>% names()
        return(
          xxx_raster2trees_ans %>% 
            cloud2trees::simplify_multipolygon_crowns()
            # sf::st_drop_geometry() %>% 
            # sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(xxx_raster2trees_ans))
        )
      }else{return(NULL)}
    }) %>% 
    purrr::list_rbind(names_to = "ws_fn_nm")
  
  # ws_tree_list %>% dplyr::glimpse()
  # ws_tree_list %>% sf::st_drop_geometry() %>% dplyr::count(ws_fn_nm)
  # ws_tree_list %>% class()
  
  return(ws_tree_list)
}
```

get the individual tree predictions for each window function for the area within the field reference plots with a buffer

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, "")
  )
if(!file.exists(fnm_temp)){
  wfsi_eglinafb_ws_pred_trees <- 
    1:nrow(wfsi_eglinafb_plots) %>% 
    purrr::map(
      \(x)
      ws_testtest_fn(
        chm = chm_list$WFSI_EglinAFB
        , buffer = 5.5
        , plot_sf = 
          wfsi_eglinafb_plots %>% 
          dplyr::ungroup() %>% 
          dplyr::slice(x) %>%
          sf::st_union()
        # a named list of functions ex: cloud2trees::itd_ws_functions()
        , ws_fn_list = get_ws_fn_list_ans # cloud2trees::itd_ws_functions() # 
      )  
    ) %>% 
    purrr::list_rbind(names_to = "plot_dta_idx")
  
  # add plot id
  wfsi_eglinafb_ws_pred_trees <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::left_join(
      wfsi_eglinafb_plots %>% 
        dplyr::mutate(plot_dta_idx=dplyr::row_number()) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(plot_dta_idx,plot_id)
      , by = "plot_dta_idx"
    )
  # get dbh
  wfsi_eglinafb_ws_pred_trees_dbh_temp <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number()
    ) %>% 
    cloud2trees::trees_dbh()
  # wfsi_eglinafb_ws_pred_trees_dbh_temp %>% dplyr::glimpse()
  # add to trees
  wfsi_eglinafb_ws_pred_trees <- 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::mutate(
      join_id = dplyr::row_number()
    ) %>% 
    dplyr::left_join(
      wfsi_eglinafb_ws_pred_trees_dbh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          join_id = as.numeric(treeID)
        ) %>% 
        dplyr::select(
          join_id, tidyselect::starts_with("fia_est_dbh"), dbh_cm
        )
      , by = "join_id"
    )
  # wfsi_eglinafb_ws_pred_trees %>% dplyr::glimpse()
  
  # # plot real quick
  # wfsi_eglinafb_ws_pred_trees %>% 
  #   sf::st_drop_geometry() %>% 
  #   dplyr::slice_sample(n = 9999) %>% 
  #   ggplot2::ggplot(mapping = ggplot2::aes(x=tree_height_m,y=dbh_cm)) + 
  #   ggplot2::geom_point() +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_lower),color="gray") +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_upper),color="gray")

  # save
  sf::st_write(wfsi_eglinafb_ws_pred_trees, dsn = fnm_temp, quiet = T, append = F)
}else{
  wfsi_eglinafb_ws_pred_trees <- sf::st_read(fnm_temp, quiet = T)
}
wfsi_eglinafb_ws_pred_trees <- wfsi_eglinafb_ws_pred_trees %>% 
  dplyr::mutate(
    ws_type = 
      ws_fn_nm %>% 
      # removes all numeric digits
      stringr::str_remove_all("[0-9]+") %>% 
      stringr::str_replace_all("_"," ") %>% 
      stringr::str_squish() %>% 
      as.factor() %>% 
      forcats::fct_inorder(ordered = T)
  )
# wfsi_eglinafb_ws_pred_trees %>% dplyr::glimpse()
# wfsi_eglinafb_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_type)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's check out one example field plot and a few different window function tree detection and crown delineation results

```{r, fig.height=10}
# plot
p_temp <- wfsi_eglinafb_plots %>% dplyr::slice(1)
# some trees
t_temp <- 
  wfsi_eglinafb_ws_pred_trees %>% 
  dplyr::filter(plot_id==p_temp$plot_id[1]) %>% 
  dplyr::inner_join(
    wfsi_eglinafb_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm,ws_type) %>% 
      dplyr::slice_sample(n = 3, by = ws_type) %>% 
      dplyr::select(ws_fn_nm)
    , by = "ws_fn_nm"
  )
chm_temp <- chm_list[["WFSI_EglinAFB"]] %>% 
  terra::crop(
    t_temp %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      terra::vect()
    , mask = T
  )
# terra::plot(chm_temp)
plt_temp <-
  chm_temp %>% 
  terra::clamp(lower = 1.37, values = F) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x=x,y=y,fill=f)
    , color = NA
    , lwd = 0
  ) +
  ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent", name = "CHM (m)") +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top")
plt_temp +
  ggplot2::geom_sf(
    data = p_temp %>% sf::st_transform(terra::crs(chm_temp))
    , fill = NA
    , color = "black"
    , lwd = 0.8
  ) + 
  ggplot2::geom_sf(
    data = t_temp
    , mapping = ggplot2::aes(color = ws_type)
    , fill = NA
    , lwd = 0.7
  ) +
  ggplot2::scale_color_viridis_d() +
  ggplot2::facet_wrap(facets = dplyr::vars(ws_fn_nm), ncol = 3) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold")
    , strip.background = ggplot2::element_rect(color = "black")
    , panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  ggplot2::guides(
    color = "none"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now we need a function to perform instance matching of the predictions and reference data. to keep things simple, since all the reference field data includes point locations, we'll perform matching based on a spatial intersection of these points with the predicted crowns. if multiple tree points fall within a single crown, we'll select the tree that is closest in terms of size based on whatever field data is available (height or DBH). One predicted tree will only be matched with one field reference tree and vice-versa.

```{r}
# instance matching function for points reference and polys predictions
# matches a single set of predictions with a single reference set
instance_match_fn <- function(
  pred_crowns
  , ref_points
){
  # check
  if(
    !any( stringr::str_equal(names(pred_crowns),"treeID") ) ||
    !any( stringr::str_equal(names(ref_points),"treeID") )
  ){
    stop("both reference and predicted data must have `treeID` column")
  }
  
  if(nrow(ref_points)==0){return(NULL)}
  if(nrow(pred_crowns)==0 && nrow(ref_points)>0){
    return(
      ref_points %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          ref_match_idx = dplyr::row_number()
          , pred_match_idx = as.numeric(NA)
        ) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(
          match_grp = "omission"
          , match_grp = factor(
            match_grp
            , ordered = T
            , levels = c(
              "true positive"
              , "commission"
              , "omission"
            )
          ) %>% forcats::fct_rev()
        )
    )
  }
  
  # set up matching col based on what data is available
  if(
    any( stringr::str_equal(names(pred_crowns),"tree_height_m") ) &&
    any( stringr::str_equal(names(ref_points),"tree_height_m") )
  ){
    pred_crowns <- pred_crowns %>% dplyr::mutate(xxx_match_col = as.numeric(tree_height_m))
    ref_points <- ref_points %>% dplyr::mutate(xxx_match_col = as.numeric(tree_height_m))
  }else if(
    any( stringr::str_equal(names(pred_crowns),"dbh_cm") ) &&
    any( stringr::str_equal(names(ref_points),"dbh_cm") )
  ){
    pred_crowns <- pred_crowns %>% dplyr::mutate(xxx_match_col = as.numeric(dbh_cm))
    ref_points <- ref_points %>% dplyr::mutate(xxx_match_col = as.numeric(dbh_cm))
  }else{
    stop("both reference and predicted data must have `tree_height_m` or `dbh_cm` column")
  }
  # get rid of data without data
  pred_crowns <- pred_crowns %>% dplyr::filter(!is.na(xxx_match_col), xxx_match_col>0) %>% 
    # make an id
    dplyr::mutate(xxx_id = dplyr::row_number())
  ref_points <- ref_points %>% dplyr::filter(!is.na(xxx_match_col), xxx_match_col>0) %>% 
    # make it the same crs
    sf::st_transform(sf::st_crs(pred_crowns)) %>% 
    # make an id
    dplyr::mutate(xxx_id = dplyr::row_number())
  
  # intersection for tp's
  i_temp <-
    pred_crowns %>% 
    dplyr::rename_with(
      .fn = ~stringr::str_replace(.x, "^xxx_", "xxx_pred_")
      , .cols = tidyselect::starts_with("xxx_")
    ) %>% 
    sf::st_intersection(
      ref_points %>% 
        dplyr::select(xxx_id,xxx_match_col) %>% 
        dplyr::rename_with(
          .fn = ~stringr::str_replace(.x, "^xxx_", "xxx_ref_")
          , .cols = tidyselect::starts_with("xxx_")
        )
    ) %>% 
    sf::st_drop_geometry() %>% 
    # absolute diff between match col
    dplyr::mutate(
      xxx_abs_diff = abs(xxx_pred_match_col - xxx_ref_match_col)
    ) %>% 
    # sorting is key to make the purrr::reduce work how we want it
    dplyr::arrange(
      desc(xxx_pred_match_col)
      , xxx_pred_id
      , xxx_abs_diff
    )
  # i_temp %>% dplyr::glimpse()
  
  # if there even are any matches
  if(nrow(i_temp)>1){
    # ?purrr::reduce
    # purrr::reduce(letters[1:5], paste, sep = ".")
    # 1:3  %>% reduce(`+`)
    match_temp <- purrr::reduce(
      # go through all rows except first since we start there
      .x = c(2:nrow(i_temp))
      # start with the first row
      , .init = i_temp %>% dplyr::slice(1) # = iter_temp
      , .f = function(iter_temp, i){
        
        #ids of curr row
        pred_id  <- i_temp$xxx_pred_id[i]
        ref_id <- i_temp$xxx_ref_id[i]
        
        # either one used?
        pred_used  <- pred_id %in% iter_temp$xxx_pred_id
        ref_used <- ref_id %in% iter_temp$xxx_ref_id
        
        # add if freee
        if(!pred_used && !ref_used) {
          return(
            iter_temp %>% 
              dplyr::bind_rows(
                dplyr::slice(i_temp, i)
              )
          )
        }else{
          # return current state
          return(iter_temp)
        }
    })
    # match_temp %>% dplyr::glimpse()
    # match_temp %>% dplyr::count(xxx_pred_id) %>% dplyr::filter(n>1) %>% nrow()
    # match_temp %>% dplyr::count(xxx_ref_id) %>% dplyr::filter(n>1) %>% nrow()
  }else{
    match_temp <- i_temp
  }
  # match_temp %>% dplyr::glimpse()
  # these are matches
  return_df <- match_temp %>% 
    dplyr::rename(
      ref_match_idx = xxx_ref_id
      , pred_match_idx = xxx_pred_id
    ) %>% 
    dplyr::select(dplyr::all_of(c(
      "ref_match_idx"
      , "pred_match_idx"
    ))) %>% 
    dplyr::mutate(match_grp = "true positive")
  # return_df %>% dplyr::glimpse()

  return_df <-
    return_df %>% 
    # if reference but no predictions...they are all omissions (FN)
    dplyr::bind_rows(
      ref_points %>% 
        dplyr::filter(
          !(xxx_id %in% match_temp$xxx_ref_id)
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::rename(
          ref_match_idx = xxx_id
        ) %>% 
        dplyr::mutate(pred_match_idx = as.numeric(NA)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(match_grp = "omission")  
    ) %>% 
    # if predictions but no reference...they are all commissions (FP)
    dplyr::bind_rows(
      pred_crowns %>% 
        dplyr::filter(
          !(xxx_id %in% match_temp$xxx_pred_id)
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::rename(
          pred_match_idx = xxx_id
        ) %>% 
        dplyr::mutate(ref_match_idx = as.numeric(NA)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
        ))) %>% 
        dplyr::mutate(match_grp = "commission")
    ) %>% 
    # add on treeID
    dplyr::left_join(
      ref_points %>% 
        sf::st_drop_geometry() %>%
        dplyr::select(xxx_id, treeID) %>% 
        dplyr::rename(
          ref_match_idx = xxx_id
          , ref_treeID = treeID
        )
      , by = "ref_match_idx"
    ) %>% 
    # add on treeID
    dplyr::left_join(
      pred_crowns %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(xxx_id, treeID) %>% 
        dplyr::rename(
          pred_match_idx = xxx_id
          , pred_treeID = treeID
        )
      , by = "pred_match_idx"
    )
    
  # return_df %>% dplyr::glimpse()
  # return_df %>% View()
  # return_df %>% dplyr::count(match_grp)
  
  # make match_grp factor
    return_df <- return_df %>% 
      dplyr::mutate(
        match_grp = factor(
          match_grp
          , ordered = T
          , levels = c(
            "true positive"
            , "commission"
            , "omission"
          )
        ) %>% forcats::fct_rev()
      )
    return(return_df)
}

###################################
### test for one plot
### test for one plot
### test for one plot
###################################
instance_match_fn(
  pred_crowns = 
    wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::filter(
      plot_id == wfsi_eglinafb_plots$plot_id[1]
      , ws_fn_nm==wfsi_eglinafb_ws_pred_trees$ws_fn_nm[1]
    )
  , ref_points = 
    wfsi_eglinafb_trees %>% 
    dplyr::filter(
      plot_id == wfsi_eglinafb_plots$plot_id[1]
    )
) %>% 
# dplyr::glimpse()
dplyr::count(match_grp)
```

remember, we have predictions for many different window functions for each plot. we need a function to iterate over each set of predictions (i.e. window function) for a single plot. this is where we'll filter the predictions to attempt to match the field data collection protocol (e.g. measure trees >= 10 cm DBH)

```{r}
ws_iter_instance_match_fn <- function(
  plot_sf # needs plot_id # should be a single plot
  , ref_pts # needs plot_id
  , pred_polys # needs plot_id
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = NA
){
  if(
    !any( stringr::str_equal(names(plot_sf),"plot_id") )
  ){
    stop("`plot_sf` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(ref_pts),"plot_id") )
  ){
    stop("`ref_pts` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"plot_id") )
  ){
    stop("`pred_polys` needs column named `plot_id`")
  }
  
  # pred polys
  pred_polys <- 
    pred_polys %>% 
    dplyr::filter(
      plot_id == plot_sf$plot_id[1]
    )
  
  # reference points
  ref_pts <- 
    ref_pts %>% 
    dplyr::filter(
      plot_id == plot_sf$plot_id[1]
    )
  
  # filter based on field survey limit
  if(
    !is.na(as.numeric(min_dbh_cm))
  ){
    # do they exist? .... PRED
    if(
      any( stringr::str_equal(names(pred_polys),"fia_est_dbh_cm_upper") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          dplyr::coalesce(fia_est_dbh_cm_upper,0) >= as.numeric(min_dbh_cm)
        )
    }else if(
      any( stringr::str_equal(names(pred_polys),"dbh_cm") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          dplyr::coalesce(dbh_cm,0) >= as.numeric(min_dbh_cm)
        )
    }else{
      stop("`pred_polys` needs column named `dbh_cm`")
    }
    
    # do they exist? .... REF
    if(
      any( stringr::str_equal(names(ref_pts),"dbh_cm") )
    ){
      ref_pts <- ref_pts %>% 
        dplyr::filter(
          dplyr::coalesce(dbh_cm,0) >= as.numeric(min_dbh_cm)
        )
    }else{
      stop("`ref_pts` needs column named `dbh_cm`")
    }
  }else if(
    !is.na(as.numeric(min_ht_m))
  ){
    # do they exist? .... PRED
    if(
      any( stringr::str_equal(names(pred_polys),"tree_height_m") )
    ){
      pred_polys <- pred_polys %>% 
        dplyr::filter(
          dplyr::coalesce(tree_height_m,0) >= as.numeric(min_ht_m)
        )
    }else{
      stop("`pred_polys` needs column named `tree_height_m`")
    }
    
    # do they exist? .... REF
    if(
      any( stringr::str_equal(names(ref_pts),"tree_height_m") )
    ){
      ref_pts <- ref_pts %>% 
        dplyr::filter(
          dplyr::coalesce(tree_height_m,0) >= as.numeric(min_ht_m)
        )
    }else{
      stop("`ref_pts` needs column named `tree_height_m`")
    }
  }
  
  if(nrow(ref_pts)==0){return(NULL)}
  
  # map over it
  if(
    !any( stringr::str_equal(names(pred_polys),"ws_fn_nm") )
  ){
    stop("`pred_polys` needs column named `ws_fn_nm` ... see ws_testtest_fn() ")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"tree_x") )
  ){
    stop("`pred_polys` needs column named `tree_x` ... see ws_testtest_fn() ")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"tree_y") )
  ){
    stop("`pred_polys` needs column named `tree_y` ... see ws_testtest_fn() ")
  }
  
  # map it
  instance_match_ans <- 
    unique(pred_polys$ws_fn_nm) %>% 
    purrr::set_names() %>%
    purrr::map(function(nm){
      # which preds?
      these_pred_polys <- 
        pred_polys %>% 
        dplyr::filter(
          ws_fn_nm == nm
        )
      # instance match
      im <- 
        instance_match_fn(
          pred_crowns = these_pred_polys
          , ref_points = ref_pts
        )
      # remove commission points outside plot boundary?
      # remove commission points outside plot boundary?
      # get rid of commissions outside of plot
      outs <- 
        these_pred_polys %>% 
        dplyr::mutate(
          pred_match_idx = dplyr::row_number()
        ) %>% 
        sf::st_drop_geometry() %>%
        sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(pred_polys)) %>% 
        sf::st_filter(
          plot_sf %>% sf::st_union() %>% sf::st_transform(sf::st_crs(pred_polys))
          , .predicate = sf::st_disjoint
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(pred_match_idx) %>% 
        dplyr::mutate(is_out = T)
      
      return_df <- 
        im %>% 
        dplyr::left_join(
          outs
          , by = "pred_match_idx"
        ) %>% 
        dplyr::filter(
          !(match_grp == "commission" & dplyr::coalesce(is_out,F)==T)
        ) %>% 
        dplyr::select(-is_out)
      return(return_df)
    }) %>% 
    purrr::list_rbind(names_to = "ws_fn_nm")
  
  
  return(instance_match_ans)
}
```

let's test that for a single plot

```{r, fig.height=8.8}
###################################
### test for one plot
### test for one plot
### test for one plot
###################################

ws_iter_instance_match_fn(
  # needs plot_id # should be a single plot
  plot_sf = wfsi_eglinafb_plots %>% dplyr::slice(4) 
  # needs plot_id
  , ref_pts = wfsi_eglinafb_trees 
  # needs plot_id
  , pred_polys = wfsi_eglinafb_ws_pred_trees %>% 
    dplyr::inner_join(
      wfsi_eglinafb_ws_pred_trees %>% 
        sf::st_drop_geometry() %>% 
        dplyr::distinct(ws_fn_nm,ws_type) %>% 
        dplyr::slice_sample(n = 3, by = ws_type) %>% 
        dplyr::select(ws_fn_nm)
      , by = "ws_fn_nm"
    )
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = 10
) %>% 

# plot
  dplyr::count(ws_fn_nm,match_grp) %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::mutate(pct = n/sum(n)) %>% 

  ggplot2::ggplot(
    mapping = ggplot2::aes(x = pct, y = ws_fn_nm, fill=match_grp, group=match_grp)
  ) +
  ggplot2::geom_col(
    width = 0.7, alpha=0.95
  ) +
  ggplot2::scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.direction  = "horizontal"
  )

```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now, we need to iterate over all plots at a study site

```{r}
plot_iter_instance_match_fn <- function(
  ref_plots
  , ref_pts
  # needs plot_id
  , pred_polys
  # minimum tree height
  , min_ht_m = NA
  # -OR- minimum DBH
  , min_dbh_cm = NA
) {
  if(
    !any( stringr::str_equal(names(ref_plots),"plot_id") )
  ){
    stop("`ref_plots` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(ref_pts),"plot_id") )
  ){
    stop("`ref_pts` needs column named `plot_id`")
  }
  if(
    !any( stringr::str_equal(names(pred_polys),"plot_id") )
  ){
    stop("`pred_polys` needs column named `plot_id`")
  }
  
  ws_iter_instance_match_ans <- 
    unique(ref_plots$plot_id) %>% 
    purrr::set_names() %>% 
    purrr::map(
      \(pid)
      ws_iter_instance_match_fn(
        # needs plot_id # should be a single plot
        plot_sf = ref_plots %>% dplyr::filter(plot_id==pid) 
        # needs plot_id
        , ref_pts = ref_pts 
        # needs plot_id
        , pred_polys = pred_polys
        # minimum tree height
        , min_ht_m = min_ht_m
        # -OR- minimum DBH
        , min_dbh_cm = min_dbh_cm
      )
  ) %>% 
  purrr::list_rbind(names_to = "plot_id")
  
  return(ws_iter_instance_match_ans)
}
# get the ans
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_instance_match.csv")
if(!file.exists(fnm_temp)){
  wfsi_eglinafb_ws_instance_match <- 
    plot_iter_instance_match_fn(
      ref_plots = ref_plots_list[["WFSI_EglinAFB"]]
      , ref_pts = ref_trees_list[["WFSI_EglinAFB"]]
      , pred_polys = wfsi_eglinafb_ws_pred_trees
      , min_ht_m = NA
      , min_dbh_cm = 10
    )
  readr::write_csv(wfsi_eglinafb_ws_instance_match, file = fnm_temp, append = F, progress = F)
}else{
  wfsi_eglinafb_ws_instance_match <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# wfsi_eglinafb_ws_instance_match %>% dplyr::glimpse()
```

lastly, we need a function to aggregate these instance segmentation counts (TP, FP, and FN) to evaluate the overall accuracy of the tree segmentation method. then, we'll select the best window function and apply that across the entire study area.

```{r}
###############################################################
###############################################################
###############################################################
###############################################################
# first function takes df with cols tp_n, fp_n, and fn_n to calculate rates
###############################################################
###############################################################
###############################################################
###############################################################
confusion_matrix_scores_fn <- function(df){
  df %>% 
  dplyr::mutate(
    omission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 0 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 1 # every single actual tree was missed
      , dplyr::coalesce(fn_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fn_n/(tp_n+fn_n)
    ) # False Negative Rate or Miss Rate
    , commission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 0 # if no predictions are made, the model could not have made any commission errors
      , dplyr::coalesce(fp_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fp_n/(tp_n+fp_n)
    ) # False Positive Rate
    , precision = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 1 # if no predictions are made, the model made zero incorrect positive claims
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) > 0 ~ 0
      , T ~ tp_n/(tp_n+fp_n)
    )
    , recall = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 1 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 0 # every single actual tree was missed
      , T ~ tp_n/(tp_n+fn_n)
    )
    , f_score = dplyr::case_when(
      dplyr::coalesce(precision,0) == 0 | dplyr::coalesce(recall,0) == 0 ~ 0
      , T ~ 2 * ( (precision*recall)/(precision+recall) )
    )    
  ) 
}
###############################################################
###############################################################
###############################################################
###############################################################
# aggregate results from ground_truth_prediction_match()
###############################################################
###############################################################
###############################################################
###############################################################
agg_ground_truth_match <- function(ground_truth_prediction_match_ans) {
  if(nrow(ground_truth_prediction_match_ans)==0){return(NULL)}
  if( !(names(ground_truth_prediction_match_ans) %>% stringr::str_equal("match_grp") %>% any()) ){stop("ground_truth_prediction_match_ans must contain `match_grp` column")}
  
  # check for difference columns (contains "_diff") and calc rmse for only those to return a single line df with colums for each diff_rmse
    if(
      (ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(tidyselect::starts_with("diff_") | tidyselect::starts_with("pct_diff_")) %>% 
        ncol() )>0
    ){
      # get rmse and mean difference/error for all columns with "_diff" but not "pct_diff"
      # get mape for all columns with "pct_diff" but not "diff_"
      rmse_df <- 
        ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::ungroup() %>% 
        dplyr::select(
          tidyselect::starts_with("diff_")
          | tidyselect::starts_with("pct_diff_")
        ) %>%
        tidyr::pivot_longer(dplyr::everything(), values_drop_na = T) %>% 
        dplyr::group_by(name) %>% 
        dplyr::summarise(
          sq = sum(value^2, na.rm = T)
          , mean = mean(value, na.rm = T)
          , sumabs = sum(abs(value), na.rm = T)
          , nomiss = sum(!is.na(value))
        ) %>% 
        dplyr::ungroup() %>% 
        dplyr::mutate(
          rmse = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sqrt(sq/nomiss)
          )
          , mape = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sumabs/nomiss
          )
        ) %>% 
        # NA nonsense values
        dplyr::mutate(
          mape = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ mape
            , T ~ as.numeric(NA)
          )
          , rmse = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ rmse
          )
          , mean = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ mean
          )
        ) %>% 
        dplyr::select(name,rmse,mean,mape) %>% 
        tidyr::pivot_wider(
          names_from = name
          , values_from = c(rmse,mean,mape)
          , names_glue = "{name}_{.value}"
        ) %>% 
        # remove columns with NA in all rows
        dplyr::select( dplyr::where( ~!all(is.na(.x)) ) )
      
      if(
        dplyr::coalesce(nrow(rmse_df),0)==0
        || dplyr::coalesce(ncol(rmse_df),0)==0
      ){
        # empty df
        rmse_df <- dplyr::tibble()
      }
    }else{
      # empty df
      rmse_df <- dplyr::tibble()
    }
  
  # count by match group
    agg <-
      ground_truth_prediction_match_ans %>% 
      sf::st_drop_geometry() %>% 
      dplyr::ungroup() %>% 
      dplyr::count(match_grp) %>% 
      dplyr::mutate(
        match_grp = dplyr::recode_values(
          match_grp
          , "true positive"~"tp"
          , "commission"~"fp"
          , "omission"~"fn"
        )
      )
    # true positive, false positive, false negative rates
    return_df <- dplyr::tibble(match_grp = c("tp","fp","fn")) %>% 
      dplyr::left_join(agg, by = "match_grp") %>% 
      dplyr::mutate(dplyr::across(.cols = c(n), .fn = ~dplyr::coalesce(.x,0))) %>% 
      tidyr::pivot_wider(
        names_from = match_grp
        , values_from = c(n)
        , names_glue = "{match_grp}_{.value}"
      )
    # rates, precision, recall, f-score
    return_df <- confusion_matrix_scores_fn(return_df)
    # add rmse
    if(nrow(rmse_df)>0){
      return_df <- return_df %>% dplyr::bind_cols(rmse_df)
    }
  # return
  return(return_df)
}
# agg_ground_truth_match(
#   wfsi_eglinafb_ws_instance_match %>% 
#     dplyr::filter(
#       plot_id==wfsi_eglinafb_ws_instance_match$plot_id[1]
#       , ws_fn_nm==wfsi_eglinafb_ws_instance_match$ws_fn_nm[1]
#     )
# )

# map it over

# get the ans
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"agg_ground_truth_match_ans.csv")
if(!file.exists(fnm_temp)){
  # ?purrr::pmap
  # A data frame is an important special case of .l. It will cause .f to be called once for each row.
  wfsi_eglinafb_agg_ground_truth_match_ans <- 
    wfsi_eglinafb_ws_instance_match %>% 
    dplyr::distinct(plot_id,ws_fn_nm) %>% 
    purrr::pmap(function(plot_id, ws_fn_nm) {
      # the unique row data
      d <- dplyr::tibble(plot_id = plot_id, ws_fn_nm = ws_fn_nm)
      # filter full set based on the data
      j <- 
        wfsi_eglinafb_ws_instance_match %>% 
        dplyr::inner_join(
          d   
          , by = dplyr::join_by("plot_id","ws_fn_nm")
        )
      # agg_ground_truth_match
      d %>% 
        dplyr::bind_cols(agg_ground_truth_match(j))
      
    }) %>% 
    list_rbind()
  readr::write_csv(wfsi_eglinafb_agg_ground_truth_match_ans, file = fnm_temp, append = F, progress = F)
}else{
  wfsi_eglinafb_agg_ground_truth_match_ans <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# wfsi_eglinafb_agg_ground_truth_match_ans %>% dplyr::glimpse()
```

aggregate the plot level data to the window function to get the answer we've been searching for

```{r}
# wfsi_eglinafb_agg_ground_truth_match_ans %>% 
#   dplyr::count(ws_fn_nm) %>% 
#   dplyr::count(n)
agg_temp <- 
  wfsi_eglinafb_agg_ground_truth_match_ans %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::summarize(
    n_plots = dplyr::n()
    , dplyr::across(
      .cols = c(precision,recall,f_score,tidyselect::ends_with("_rate"))
      , list( 
        mean= ~ mean(.x,na.rm=T)
        , median= ~ median(.x,na.rm=T)
      )
    )
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::left_join(
    wfsi_eglinafb_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::mutate(
    f_score_rank = dplyr::min_rank(-f_score_mean)
    , ws_fn_nm_fact = forcats::fct_reorder(ws_fn_nm, f_score_rank)
    # , dplyr::across(
    #   .cols = dplyr::where(is.numeric)
    #   , ~ round(.x,2) %>% as.character() %>% as.numeric()
    # )
  ) %>% 
  dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean))
# agg_temp %>% dplyr::glimpse()
```

let's check out the plot-level distribution of F-score for the top window functions

```{r}
wfsi_eglinafb_agg_ground_truth_match_ans %>% 
  # add agg data
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::filter(f_score_rank<=12) %>% 
      dplyr::select(ws_fn_nm, ws_fn_nm_fact, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::arrange(ws_fn_nm_fact) %>% 
  dplyr::mutate(ws_fn_nm_fact=forcats::fct_rev(ws_fn_nm_fact)) %>% 
  # dplyr::pull(ws_fn_nm) %>% levels()
  # View()
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = f_score, y = ws_fn_nm_fact, fill=ws_type)
  ) +
  ggplot2::geom_boxplot(
    width = 0.4
    , alpha=0.9
    , outliers = F
    , color = "black"
  ) +
  ggplot2::geom_point(color = "gray77") +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(x="F-score", y="") +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , legend.direction  = "horizontal"
  )
```

we can print the raw window function definition for those that achieved the highest F-score

```{r}
agg_temp %>% 
  dplyr::filter(f_score_rank<=5) %>% 
  dplyr::arrange(f_score_rank) %>% 
  dplyr::pull(ws_fn_nm) %>% 
  purrr::set_names() %>% 
  # purrr::map(\(nm) purrr::pluck(get_ws_fn_list_ans, nm))
  purrr::map(function(nm){
    fn <- purrr::pluck(get_ws_fn_list_ans, nm)
    dplyr::tibble(
      ws_fn_nm = nm
      , def = paste(format(fn), collapse = "\n")
    )
  }) %>%
  purrr::list_rbind() %>% 
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::select(ws_fn_nm, f_score_mean, recall_mean, precision_mean) %>% 
      dplyr::mutate(
        dplyr::across(
          dplyr::where(is.numeric)
          , ~ scales::percent(.x,accuracy = 0.1)
        )
      )
  ) %>% 
  dplyr::relocate(def,.after = dplyr::everything()) %>% 
  kableExtra::kbl(
    caption = "WFSI Eglin Air Force Base top ITD window functions"
    , col.names = c(
      "function<br>name"
      , "F-score<br>(mean)"
      , "recall<br>(mean)"
      , "precision<br>(mean)"
      , "function<br>definition"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  # make r line breaks (\n) and use monospace font for code
  kableExtra::column_spec(5, monospace = T, extra_css = "white-space: pre;")
```

nice, we'll select the best non-constant window function for use in evaluation of the species accuracy. if we were to use the constant functions, even if they are the best performing, we would likely be overfitting to this specific data set. in addition to being more in-line with real-world tree height-to-crown relationships, the variable window functions are likely more broadly applicable

```{r}
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    best_ws_fn_nm = ifelse(
      tolower(data_site_nm)=="wfsi_eglinafb"
      , agg_temp %>% 
        dplyr::filter(ws_type!="constant") %>% 
        dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean)) %>% 
        dplyr::slice(1) %>% 
        dplyr::pull(ws_fn_nm)
      , ""
    )
  )
# project_data_df %>% dplyr::glimpse()

# let's also save the plot-level ITD predictions for this ws
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
  dplyr::pull(data_site_nm)
pred_trees_list[[nm_temp]] <- 
  wfsi_eglinafb_ws_pred_trees %>% 
  # which ws
  dplyr::filter(
    ws_fn_nm %in% (
      project_data_df %>% 
      dplyr::filter(tolower(data_site_nm)=="wfsi_eglinafb") %>% 
      dplyr::pull(best_ws_fn_nm)
    )
  ) %>% 
  # filter for trees used in instance matching 
    # which included thresholds used for measuring field trees 
    # and plot spatial alignment
  dplyr::inner_join(
    wfsi_eglinafb_ws_instance_match %>% 
      dplyr::select(plot_id,ws_fn_nm,pred_treeID) %>% 
      dplyr::filter(!is.na(pred_treeID)) %>% 
      dplyr::rename(treeID=pred_treeID)
    , by = dplyr::join_by(plot_id,ws_fn_nm,treeID)
  )
# pred_trees_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(list = ls()[grep("^wfsi",ls())])
remove(cloud2raster_ans)
gc()
```

### Species/Genera Prediction Method{#04_species}

for some context of the primary objective of this analysis, here is what we said in the `cloud2trees` manuscript (a draft of the manuscript which may be different than what was actually published): 

>However, taxonomic differentiation using airborne lidar alone remains one of the most technically challenging aspects of forest inventory workflows (White et al., 2016, 2025; Fassnacht et al., 2024). In practice, genus-level distinction is often more realistic when relying on remote sensing and supplemental spatial data ([Vauhkonen et al., 2014b](https://doi.org/10.1007/978-94-017-8663-8_7)). As a practical, computationally efficient approach, future research could evaluate predicting tree-level species or genus for potential integration into cloud2trees by leveraging supplemental spatial datasets. Testing could evaluate either using tree lists from FIA plots linked through a spatial overlay with TreeMap (Houtman et al., 2025) to train a taxonomic classification model based on structural and competition metrics shared between FIA and cloud2trees tree lists, or by proportionally assigning species using BigMap species-level proportional biomass raster data (Wilson et al., 2012, 2018). Incorporating finer-scale taxonomic classification than the currently implemented FIA forest type groups into the cloud2trees tree list will enable the integration of detailed biomass equations such as the National Scale Volume Biomass system (NSVB; Westfall et al., 2024) and other allometries that require species- or genera-level information (e.g., Réjou‐Méchain et al., 2017; Gonzalez‐Akre et al., 2022).

`cloud2trees` does not currently include functionality to attach species to it's tree list. there are a few paths forward for performing species-specific workflows with the `cloud2trees` tree list:

the proposed method which we are evaluating presently, will predict tree-level species using the TreeMap imputed FIA plot tree list to build a species classification model based on height and tree location to attach tree species to our tree list. The model will inherently account for the proportional distribution of different species. 

early testing of the proposed method used a  bayesian hierarchical model grouped by imputed FIA plot identifier that allows the intercept and the slope of height to vary by plot took quite a while to run for a relatively small area (~20-30 mins). as such, we are going to use a random forest (RF) classifier to enable more computationally efficient classification. `cloud2trees` already implements random forest models to model missing HMD and CBH values that were not successfully extracted from the point cloud and uses a model averaging method that subsamples the training data (to reduce model fit times), estimates the model several times with different subsamples, then model averages the predictions. Because the species is not a metric variable that can be averaged, we'll use the same method but employ a model majority voting system instead. we can do this with some of the `cloud2trees` built-in functionality.

ultimately, if the proposed species/genera classification method achieves decent accuracy results, it could be integrated into the `cloud2trees` framework as `cloud2trees::trees_species()`, for example. this function would take an input spatial tree list and attach the predicted species/genera

#### Training Data

We'll start by reviewing the tree lists we'll used for training data which come from the TreeMap FIA imputed plots

load in the TreeMap data and clip to our trees as in `cloud2trees::trees_dbh()`

```{r}
# let's make a function to standardize this process since we'll be using it in more than one `cloud2trees::trees_*` function now
treemap_data_loader <- function(sf_poly, input_treemap_dir = NULL) {
  ####################################################################
  # check external data
  ####################################################################
    # find external data
    find_ext_data_ans <- cloud2trees:::find_ext_data(
      input_treemap_dir = input_treemap_dir
    )
    # if can't find external treemap data
    if(is.null(find_ext_data_ans$treemap_dir)){
      stop(paste0(
        "Treemap data has not been downloaded to package contents. Use `get_treemap()` first."
        , "\nIf you supplied a value to the `input_treemap_dir` parameter check that directory for data."
      ))
    }

    #treemap files
    treemap_data_finder_ans <- cloud2trees:::treemap_data_finder(find_ext_data_ans$treemap_dir)
    # treemap_data_finder_ans$which_treemap
    # treemap_data_finder_ans$treemap_trees
    # treemap_data_finder_ans$treemap_rast
  ####################################################################
  # check polygons
  ####################################################################
    if(inherits(sf_poly, "sfc")){sf_poly <- sf::st_as_sf(sf_poly)}
    if(!inherits(sf_poly, "sf")){stop("must include `sf` data object in 'sf_poly'")}
    if( !all(sf::st_is(sf_poly, type = c("POLYGON", "MULTIPOLYGON"))) ){
      stop(paste0(
        "`sf_poly` data must be an `sf` class object with POLYGON geometry (see [sf::st_geometry_type()])"
      ))
    }
    sf_poly <- sf_poly %>% dplyr::ungroup() %>% sf::st_union()
    
  ####################################################################
  # read in treemap data
  ####################################################################
    # read in treemap data
    # read in treemap (no memory is taken)
    treemap_rast <- terra::rast(treemap_data_finder_ans$treemap_rast)

    # check study boundary against the raster
    # the resulting matrix will have a true value where an intersection exists
    intersection_result <- terra::relate(
      x = sf_poly %>%
        sf::st_transform(terra::crs(treemap_rast)) %>%
        terra::vect()
      , y = treemap_rast
      , relation = "intersects"
    )

    if(!any(intersection_result)) {
      stop(paste0(
        "The search area does not overlap with an area within CONUS."
        , "\n.... If provided data in `study_boundary` ensure that it overlaps with CONUS"
        , "\n.... If did not provide data in `study_boundary` ensure that the `tree_list` data overlaps with CONUS"
      ))
    }
    
    # calculate buffer based on the expected window size used for any terra::focal operations
    ws_temp <- 5
    buffer_temp <- ((ws_temp-1)/2)*terra::res(treemap_rast)[1]
    
    ### filter treemap based on las...rast now in memory
    treemap_rast <- treemap_rast %>%
      terra::crop(
        sf_poly %>%
          sf::st_transform(terra::crs(treemap_rast)) %>%
          sf::st_buffer(buffer_temp) %>% 
          terra::vect()
      ) %>%
      terra::subset(1)

    # ggplot(treemap_rast %>% as.data.frame(xy=T) %>% rename(f=3)) +
    #   geom_tile(aes(x=x,y=y,fill=as.factor(f))) +
    #   scale_fill_viridis_d(option = "turbo") +
    #   theme_light() + theme(legend.position = "none")

    # check for all NA's
    notna_cells <- any( terra::global(treemap_rast, fun = "anynotNA") )
    if( !notna_cells ){
      stop(paste0(
        "The search area does not overlap with a forested area within CONUS. Cannot estimate DBH."
        , "\n ... try expanding the `boundary_buffer` ?"
      ))
    }
    
    ### get weights for weighting each tree in the population models
    # treemap id = tm_id for linking to tabular data
    tm_id_weight_temp <- treemap_rast %>% ## works
      terra::values() %>%
      table() %>%
      dplyr::as_tibble() %>%
      dplyr::rename(tm_id=1,tree_weight=n) %>%
      dplyr::mutate(tm_id = cloud2trees:::as_character_safe(tm_id))
    # tm_id_weight_temp
    
    ############################################################################
    ### get a stability weight of the imputed plot id's
    ############################################################################
    
    # when we consider the tree composition of an entire forest stand we expect TreeMap 
    # to generally get the main species and forest type correct 
    # but for each 30m raster cell (i.e. the plot location that we get the X and Y coordinates for the tree list used for training from) 
    # we expect it to be less accurate with a cell potentially being a completely incorrect representation of the 
    # actual ground setting. As such, we we’ll weight the individual tree records based on the stability of 
    # the imputed plot id’s looking over a window surrounding the focal cell. Specifically, we’ll calculate 
    # the proportion of surrounding cells (5x5) that share the same imputed plot id as the focal cell 
    # and use this weight as a measure of TreeMap’s predicted stability to place less weight 
    # on plots that occur infrequently and more weight on plots that occur frequently.
    treemap_id_rast <- 
      treemap_rast %>% 
      terra::catalyze() %>% # to numeric
      terra::subset(1)
    
    treemap_prop_rast_temp <- 
      treemap_id_rast %>% 
      terra::focal(
        # # 5x5 = 2.25 ha; 3x3 = 0.81 ha
        w = ws_temp
        , fun = match_focal_cells_fn # add to cloud2trees:::
        # , na.poicy = "omit"
        , na.rm = T
      ) %>% 
      setNames("stability_weight")
    # treemap_prop_rast_temp %>% terra::hist()
    
    ############################################################################
    ### get the TreeMap FIA tree list for only the plots included
    ############################################################################
    if(treemap_data_finder_ans$which_treemap==2022){
      treemap_cols <- c(
        "TM_ID"
        , "PLT_CN"
        , "SPECIES_SYMBOL"
        , "STATUSCD"
        , "DIA"
        , "HT"
        , "CR" # crown ratio
      )
    }else if(treemap_data_finder_ans$which_treemap==2016){
      treemap_cols <- c(
        "tm_id"
        , "CN"
        , "SPECIES_SYMBOL"
        , "STATUSCD"
        , "DIA"
        , "HT"
        , "CR" # crown ratio
      )
    }else{
      stop("unknown TreeMap vintage")
    }
    ### read it
    # treemap_data_finder_ans$treemap_trees
    treemap_trees_df <-
      readr::read_csv(
        treemap_data_finder_ans$treemap_trees
        , col_select = treemap_cols
        , progress = F
        , show_col_types = F
      ) %>%
      dplyr::rename_with(tolower)
    # rename cn to fit with original table str
    if(treemap_data_finder_ans$which_treemap==2022){
      treemap_trees_df <- treemap_trees_df %>% dplyr::rename(cn = plt_cn)
    }
    # clean it
    treemap_trees_df <-
      treemap_trees_df %>%
      dplyr::mutate(
        cn = cloud2trees:::as_character_safe(cn)
        , tm_id = cloud2trees:::as_character_safe(tm_id)
        , cr = ifelse(cr>100|cr<0,NA,cr)*0.01
      ) %>%
      dplyr::inner_join(
        tm_id_weight_temp
        , by = dplyr::join_by("tm_id")
      ) %>%
      dplyr::filter(
        # keep live trees only: 1=live;2=dead
        statuscd == 1
        & !is.na(dia)
        & !is.na(ht)
        & !is.na(tree_weight)
      ) %>%
      dplyr::mutate(
        dbh_cm = dia*2.54
        , tree_height_m = ht*0.3048
        , tm_id = as.factor(tm_id)
        , species_symbol = factor(species_symbol)
        # # normalize ht
        # , ht_wt_mean = sum(tree_height_m * tree_weight) / sum(tree_weight)
        # , ht_wt_sd = sqrt( sum(tree_weight * (tree_height_m - ht_wt_mean)^2 ) / sum(tree_weight) )
        # , ht_z = (tree_height_m - ht_wt_mean) / ht_wt_sd
      ) %>%
      dplyr::select(-c(statuscd,dia,ht,cr)) # maybe we can use cr in a future version with and NSUR approach
    
    # treemap_trees_df %>% dplyr::glimpse()
    
    # check FIA model data
    if(dplyr::coalesce(nrow(treemap_trees_df),0)==0){
      stop(paste0(
        "No trees in Treemap data. Ensure that "
        , "\nIf provided data in `study_boundary` ensure that it is located in the continental US"
        , "\nIf did not provide data in `study_boundary` ensure that the `tree_list` data has is located in the continental US"
      ))
    }
    
    # we need the spatial information from the raster...
    ############################################################################
    ### attach spatial information from raster to trees
    ############################################################################
    
    # create the tree list with trees assigned the X and Y of the raster cell and 
    # for imputed plot id's the specific tree will appear multiple times based on 
    # the number of times the plot is replicated within the area of interest but 
    # each tree will have a different X,Y and stability weight based on the surrounding imputed plots 
    
    treemap_trees_spatial_df <-
      treemap_id_rast %>% 
      terra::as.data.frame(xy = T, cells = T) %>% 
      dplyr::rename(tm_id=4) %>% 
      # add on weights based on moving window stability of imputed plot
      dplyr::left_join(
        treemap_prop_rast_temp %>% 
          terra::as.data.frame(xy = F, cells = T)
        , by = "cell"
      ) %>% 
      dplyr::mutate(
        tm_id = cloud2trees:::as_character_safe(tm_id)
        # clean stability_weight
        , stability_weight = ifelse(stability_weight>1,1,stability_weight) %>% dplyr::coalesce(0.001)
      ) %>% 
      # dplyr::glimpse()
      # add on the tree data in which a row is unique by tm_id and the tree (which doesn't have an id so we'll fake one)
      dplyr::inner_join(
        treemap_trees_df %>% 
          dplyr::ungroup() %>% 
          dplyr::mutate(
            tree_id = dplyr::row_number()
            , tm_id = cloud2trees:::as_character_safe(tm_id)
          )
        , by = "tm_id"
        , relationship = "many-to-many"
      ) %>% 
      # a row is now unique by x,y,tree_id resulting in the number of rows per tree_id = tm_id_weight_temp$tree_weight
      dplyr::mutate(
        # to factor
        # tm_id = forcats::fct_relevel(tm_id, levels(treemap_trees_df$tm_id) ) ##!!!!! do we need this?
        species_symbol = factor(species_symbol)
        # slightly jitter the tree x,y so that RF doesn't lock-in on grid pattern from raster
        # jitter within the 30m cell +/- 15 m
        , x = x + stats::runif(
          dplyr::n()
          , min = -(terra::res(treemap_rast)[1]/2)
          , max = (terra::res(treemap_rast)[1]/2)
        )
        , y = y + stats::runif(
          dplyr::n()
          , min = -(terra::res(treemap_rast)[1]/2)
          , max = (terra::res(treemap_rast)[1]/2)
        )
      )
    
    # dplyr::glimpse(treemap_trees_spatial_df)
    # treemap_trees_spatial_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) + ggplot2::geom_point() + ggplot2::theme_void()
    # treemap_trees_spatial_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=stability_weight)) + ggplot2::geom_point() + ggplot2::theme_void()
    # treemap_trees_spatial_df %>% 
    #   ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=factor(cell))) + 
    #   ggplot2::geom_point() + 
    #   ggplot2::scale_color_manual(values = sample(
    #       viridis::turbo(n=length(unique(treemap_trees_spatial_df$cell)))
    #     )
    #   ) + 
    #   ggplot2::theme_void() + 
    #   ggplot2::theme(legend.position = "none")
    
    return(list(
      treemap_trees_df = treemap_trees_df
      , treemap_rast = treemap_rast
      , treemap_trees_spatial_df = treemap_trees_spatial_df
      , treemap_prop_rast = treemap_prop_rast_temp
    ))
}

# add to cloud2trees:::
# focal matrix for a x by x window
# focal_kernel_temp <- matrix(1, nrow = 3, ncol = 3)
# focal_kernel_temp
# length(focal_kernel_temp)*terra::res(treemap_rast) %>% prod() %>% `/`(10000)
# 5x5 = 2.25 ha; 3x3 = 0.81 ha
# custom window fn to get prop of surrounding cells matching focal value
# ?terra::focal
match_focal_cells_fn <- function(v, ...) {
  if(all(is.na(v))){ return(NA) }
  
  # center index of the window vector
  center_idx <- (length(v) + 1) / 2
  focal_val  <- v[center_idx]
  
  if(is.na(focal_val)){ return(NA) }
  
  # count how many cells in the window match the focal cell
  n_matches <- sum(v %in% focal_val, ...)
  prop_matches <- n_matches / length(v)
  
  return(prop_matches)
}


```

if we are going to treat the FIA measured trees as individual observations to use for training data without using the imputed plot ID from TreeMap and weight by the number of times that specific id is repeated in the area of interest, then we need to get a tree list with the trees replicated based on the number of times the plot id is replicated. in addition, the TreeMap data product is a raster of imputed plots for a given area at 30m resolution and we should not assume that the tree list of that "guessed" plot represent the actual trees present at that location with complete accuracy. Instead, we expect the TreeMap imputed plots to be more accurate as we look over a broader area with accuracy decreasing at smaller scales. That is, when we consider the tree composition of an entire forest stand we expect TreeMap to generally get the main species and forest type correct but for each 30m raster cell (i.e. the plot location that we get the X and Y coordinates for the tree list used for training from) we expect it to be less accurate with a cell potentially being a completely incorrect representation of the actual ground setting. As such, we we'll weight the individual tree records based on the stability of the imputed plot id's looking over a window surrounding the focal cell. Specifically, we'll calculate the proportion of surrounding cells (5x5) that share the same imputed plot id as the focal cell and use this weight as a measure of TreeMap's predicted stability to place less weight on plots that occur infrequently and more weight on plots that occur frequently.

```{r}
# test
treemap_data_loader_ans <-
  pred_trees_list$WFSI_EglinAFB %>% 
  dplyr::ungroup() %>% 
  sf::st_union() %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  treemap_data_loader()

treemap_data_loader_ans$treemap_rast %>% terra::plot()
treemap_data_loader_ans$treemap_prop_rast %>% terra::plot()
treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  dplyr::slice_sample(n=11111) %>% 
  ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) + 
  ggplot2::geom_point() + 
  ggplot2::scale_color_viridis_d() + 
  ggplot2::theme_void()

treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species_symbol) %>% 
  dplyr::mutate(
    pct=n/sum(n)
    , pct = scales::percent(pct,accuracy=0.1)
  ) %>% 
  dplyr::arrange(desc(n))
```

what do these rasters look like?

```{r, fig.height=8.8}
xxxxxxxxx

terra::plot(
  c(
    treemap_rast %>%
      terra::catalyze() %>%
      terra::subset(1) %>%
      terra::as.factor()
    , treemap_prop_rast_temp 
      # `*`(100) %>%
      # round() %>%
      # terra::as.factor()
  )
  , nc = 1
  , main = c("Imputed Plot ID", "Plot Stability %")
  , axes = F
  # , col = c(viridis::viridis(n=100),viridis::magma(n=100))
  , fun = function() {
   # This code executes on top of every single panel automatically
   lines(
     stand_boundary %>% sf::st_transform(terra::crs(treemap_rast)) %>% terra::vect()
     , col = "red", lwd = 2
    )
 }
)
```

create the tree list with trees assigned the X and Y of the raster cell and for imputed plot id's the specific tree will appear multiple times based on the number of times the plot is replicated within the area of interest but each tree will have a different X,Y and stability weight based on the surrounding imputed plots 

```{r}
# add to tree list
treemap_tree_list_df <-
  treemap_rast %>% 
  terra::catalyze() %>% 
  terra::subset(1) %>% 
  terra::as.data.frame(xy = T, cells = T) %>% 
  dplyr::rename(tm_id=4) %>% 
  # add on weights based on moving window stability of imputed plot
  dplyr::left_join(
    treemap_prop_rast_temp %>% 
      terra::as.data.frame(xy = F, cells = T)
    , by = "cell"
  ) %>% 
  dplyr::mutate(
    tm_id = cloud2trees:::as_character_safe(tm_id)
    # clean stability_weight
    , stability_weight = ifelse(stability_weight>1,1,stability_weight) %>% dplyr::coalesce(0.01)
  ) %>% 
  # dplyr::glimpse()
  # add on the tree data in which a row is unique by tm_id and the tree (which doesn't have an id so we'll fake one)
  dplyr::inner_join(
    treemap_trees_df %>% 
      dplyr::ungroup() %>% 
      dplyr::mutate(
        tree_id = dplyr::row_number()
        , tm_id = as.character(tm_id)
      )
    , by = "tm_id"
    , relationship = "many-to-many"
  ) %>% 
  # a row is now unique by x,y,tree_id resulting in the number of rows per tree_id = tree_weight
  dplyr::mutate(
    # to factor
    tm_id = forcats::fct_relevel(tm_id, levels(treemap_trees_df$tm_id) )
    , species_symbol = factor(species_symbol)
    # slightly jitter the tree x,y so that RF doesn't lock-in on grid pattern from raster
    # jitter within the 30m cell +/- 15 m
    , x = x + stats::runif(
      dplyr::n()
      , min = -(terra::res(treemap_rast)[1]/2)
      , max = (terra::res(treemap_rast)[1]/2)
    )
    , y = y + stats::runif(
      dplyr::n()
      , min = -(terra::res(treemap_rast)[1]/2)
      , max = (terra::res(treemap_rast)[1]/2)
    )
  )
# dplyr::glimpse(treemap_tree_list_df)
# treemap_tree_list_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) + ggplot2::geom_point() + ggplot2::theme_void()
# treemap_tree_list_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=stability_weight)) + ggplot2::geom_point() + ggplot2::theme_void()
# treemap_tree_list_df %>% 
#   ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=factor(cell))) + 
#   ggplot2::geom_point() + 
#   ggplot2::scale_color_manual(values = sample(
#       viridis::turbo(n=length(unique(treemap_tree_list_df$cell)))
#     )
#   ) + 
#   ggplot2::theme_void() + 
#   ggplot2::theme(legend.position = "none")

```

what's in the FIA tree list using the TreeMap imputed plots filtered for our area of interest?

```{r}
treemap_tree_list_df %>% dplyr::glimpse()
```

this is essentially the same thing as `regional_dbh_height_model_training_data.csv` that we reviewed above, but we added the spatial x and y coordinates

before we build the model we need to scale the x, y, and height data but we need to be sure to scale both the training and data to be predicted using the same scale so we use data on the same scale across both data. Centering and scaling the x, y, and height variables puts the center of the area of interest (weighted by number of trees) at the "zero" point and resolves the sign (i.e. +/-) conflicts of the coordinate system (i.e. UTMs) by making the signs represent relative direction (north vs south) rather than arbitrary coordinate origins. we also need to convert the tree species in the training data to factor from character data type.

```{r}
# scale fn
my_scale_fn <- function(x, ref_mean, ref_sd) {
  (x - ref_mean) / ref_sd
}

# ref values from training data
train_stats_temp <- list(
  x_mean = mean(treemap_tree_list_df$x, na.rm = T)
  , x_sd   = sd(treemap_tree_list_df$x, na.rm = T)
  , y_mean = mean(treemap_tree_list_df$y, na.rm = T)
  , y_sd   = sd(treemap_tree_list_df$y, na.rm = T)
  , h_mean = mean(treemap_tree_list_df$tree_height_m, na.rm = T)
  , h_sd   = sd(treemap_tree_list_df$tree_height_m, na.rm = T)
)
# train_stats_temp

# apply to training data
treemap_tree_list_df <- 
  treemap_tree_list_df %>% 
  dplyr::mutate(
    x_z = my_scale_fn(x, train_stats_temp$x_mean, train_stats_temp$x_sd)
    , y_z = my_scale_fn(y, train_stats_temp$y_mean, train_stats_temp$y_sd)
    , ht_z = my_scale_fn(tree_height_m, train_stats_temp$h_mean, train_stats_temp$h_sd)
  )
# treemap_tree_list_df %>% dplyr::glimpse()
# treemap_tree_list_df %>% dplyr::select(x_z,y_z,ht_z) %>% summary()
# apply to prediction data
# !!! use the train_stats_temp here, NOT the prediction data mean/sd
# predicted_trees %>% purrr::list_rbind(names_to = "folder") %>% dplyr::glimpse()

predicted_trees <-
  predicted_trees %>% 
  purrr::map(function(x){
    orig_crs <- sf::st_crs(x)
    # z val of x,y,ht
    dta_temp <- 
      x %>% 
      sf::st_transform(terra::crs(treemap_rast)) %>% 
      dplyr::mutate(
        x = sf::st_coordinates(.)[,1]
        , y = sf::st_coordinates(.)[,2]
        , x_z = my_scale_fn(x, train_stats_temp$x_mean, train_stats_temp$x_sd)
        , y_z = my_scale_fn(y, train_stats_temp$y_mean, train_stats_temp$y_sd)
        , ht_z = my_scale_fn(tree_height_m, train_stats_temp$h_mean, train_stats_temp$h_sd)
      )
    # add tm_id
    tmids_temp <- 
      terra::extract(
        x = treemap_rast %>% 
          terra::catalyze() %>% 
          terra::subset(1)
        , y = dta_temp %>% 
          terra::vect()
      ) %>% 
      dplyr::rename_with(tolower) %>% 
      dplyr::mutate(
        tm_id = cloud2trees:::as_character_safe(tm_id) %>% 
          forcats::as_factor() %>% 
          forcats::fct_expand( levels(treemap_trees_df$tm_id) ) %>% 
          forcats::fct_relevel( levels(treemap_trees_df$tm_id) )
      )
    dta_temp$tm_id <- tmids_temp$tm_id
    # orig_crs
    dta_temp <- dta_temp %>% sf::st_transform(orig_crs)
    return(dta_temp)
  })
# predicted_trees %>% purrr::list_rbind(names_to = "folder") %>% dplyr::glimpse()
# predicted_trees[[1]] %>% ggplot2::ggplot() + ggplot2::geom_sf(mapping = ggplot2::aes(color=tm_id))

# align the factors
validation_trees <- 
  validation_trees %>% 
  dplyr::mutate(
    species_symbol = 
      species_symbol %>% 
      forcats::as_factor() %>% 
      forcats::fct_expand( levels(treemap_trees_df$species_symbol) ) %>% 
      forcats::fct_relevel( levels(treemap_trees_df$species_symbol) )
  )

pal_species <- 
  forcats::lvls_union(list(
    validation_trees$species_symbol
    , treemap_trees_df$species_symbol
  )) %>% 
  length() %>% 
  # RColorBrewer::brewer.pal(name = "Set2") %>% 
  viridis::turbo() %>% 
  setNames(
    forcats::lvls_union(list(
      validation_trees$species_symbol
      , treemap_trees_df$species_symbol
    )) %>% 
    as.character()
  )
# scales::show_col(pal_species)
```

let's quickly look at the training data species over the independent variables of height and x, y

```{r,eval=T, fig.height = 10.5, fig.width = 8.7}
# x,y
plt1_temp <-
  treemap_tree_list_df %>% 
  dplyr::slice_sample(n = 1111) %>% 
  dplyr::mutate(flab = "Y coord vs X coord") %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = x_z, y = y_z, fill = species_symbol, color = species_symbol, label = species_symbol)
  ) +
  ggplot2::geom_hline(yintercept = 0, color = "black") + 
  ggplot2::geom_vline(xintercept = 0, color = "black") + 
  # ggplot2::geom_text(size = 3) + 
  ggrepel::geom_text_repel(
    size = 1.8
    , segment.color = NA
    , box.padding = 0.1
    , point.padding = 0
    , force = 0.5
    , max.overlaps = Inf
  ) +
  ggplot2::facet_wrap(facets = dplyr::vars(flab)) +
  # ggplot2::scale_color_discrete(palette = "Set2") + 
  ggplot2::scale_color_manual(values = pal_species) +
  ggplot2::coord_equal() + 
  ggplot2::labs(
    x = "x", y = "y"
  ) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(color = "black", size = 9, face = "bold")
    , strip.background = ggplot2::element_rect(fill = "gray88", color = "gray88")
  )
# plt1_temp

# ht,x
plt2_temp <-
  treemap_tree_list_df %>% 
  dplyr::slice_sample(n = 1111) %>% 
  dplyr::mutate(flab = "Height vs X coord") %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = x_z, y = ht_z, fill = species_symbol, color = species_symbol, label = species_symbol)
  ) +
  ggplot2::geom_hline(yintercept = 0, color = "black") + 
  ggplot2::geom_vline(xintercept = 0, color = "black") + 
  # ggplot2::geom_text(size = 1.8) + 
  ggrepel::geom_text_repel(
    size = 1.8
    , segment.color = NA
    , box.padding = 0.1
    , point.padding = 0
    , force = 0.5
    , max.overlaps = Inf
  ) +
  ggplot2::facet_wrap(facets = dplyr::vars(flab)) +
  # ggplot2::scale_color_discrete(palette = "Set2") + 
  ggplot2::scale_color_manual(values = pal_species) +
  ggplot2::coord_equal() + 
  ggplot2::labs(
    x = "x", y = "tree height (z)"
  ) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(color = "black", size = 9, face = "bold")
    , strip.background = ggplot2::element_rect(fill = "gray88", color = "gray88")
  )
# plt2_temp

# ht,y
plt3_temp <-
  treemap_tree_list_df %>% 
  dplyr::slice_sample(n = 1111) %>% 
  dplyr::mutate(flab = "Height vs Y coord") %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = y_z, y = ht_z, fill = species_symbol, color = species_symbol, label = species_symbol)
  ) +
  ggplot2::geom_hline(yintercept = 0, color = "black") + 
  ggplot2::geom_vline(xintercept = 0, color = "black") + 
  # ggplot2::geom_text(size = 1.8) + 
  ggrepel::geom_text_repel(
    size = 1.8
    , segment.color = NA
    , box.padding = 0.1
    , point.padding = 0
    , force = 0.5
    , max.overlaps = Inf
  ) +
  ggplot2::facet_wrap(facets = dplyr::vars(flab)) +
  # ggplot2::scale_color_discrete(palette = "Set2") + 
  ggplot2::scale_color_manual(values = pal_species) +
  ggplot2::coord_equal() + 
  ggplot2::labs(
    x = "y", y = "tree height (z)"
  ) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(color = "black", size = 9, face = "bold")
    , strip.background = ggplot2::element_rect(fill = "gray88", color = "gray88")
  )
# plt3_temp
patchwork::wrap_plots(list(plt2_temp,plt3_temp,plt1_temp), ncol = 2)
# ggplot2::ggsave(filename = "../data/spec_predictors_ex.jpg", height = 10.5, width = 9, dpi = "print")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```


### RF Predict Species

implement that random forest model majority voting for species with the objective to reduce estimation time while improving or maintaining the species composition accuracy compared to the field data

```{r, results=F, message=F, warning=F}
# treemap_tree_list_df %>% dplyr::glimpse()
# set max FIA trees to use in each subsample
mod_n_subsample_temp <- min(nrow(treemap_tree_list_df)-1,22222)
# number of models to estimate with this subsample size
# go for at least 50% of training data sampled
# it's just that the individual model fits will be smaller
ntimes_temp <-
  ((nrow(treemap_tree_list_df)*0.5)/mod_n_subsample_temp) %>%
  ceiling() %>%
  max(6) %>%
  min(66)
# ntimes_temp
st_temp <- Sys.time()
# estimate

# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg
# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg
# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg
# rf_spec_mod <- cloud2trees:::rf_subsample_and_model_n_times(
#   predictors = treemap_tree_list_df %>% dplyr::select(x,y,tree_height_m)
#   , response = treemap_tree_list_df$species_symbol
#   , mod_n_subsample = mod_n_subsample_temp
#   , mod_n_times = ntimes_temp
# )
# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg
# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg
# #### need to update cloud2trees:::rf_subsample_and_model_n_times to include case.weights arg

# ?randomForest::tuneRF
optimal_mtry_temp <- cloud2trees:::rf_tune_subsample(
  predictors = treemap_tree_list_df %>% dplyr::select(x,y,tree_height_m)
  , response = treemap_tree_list_df$species_symbol
)
ntimes_temp <- 1
mod_n_subsample_temp <- nrow(treemap_tree_list_df)
rf_spec_mod <- list(
  randomForest::randomForest(
    x = treemap_tree_list_df %>% dplyr::select(x,y,tree_height_m)
    , y = treemap_tree_list_df$species_symbol
    , case.weights = treemap_tree_list_df$stability_weight
    , mtry = optimal_mtry_temp
    , na.action = na.omit
    , ntree = 500
  )
)
# randomForest::varImpPlot(rf_spec_mod[[1]])
# length(rf_spec_mod)
# timer
mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
# mins_temp
```

processing time was much better and took only `r scales::comma(mins_temp, accuracy = 0.01)` minutes (`r scales::comma(mins_temp*60, accuracy = 0.01)` seconds) to estimate `r ntimes_temp` RF models using `r scales::comma(mod_n_subsample_temp)` training records (`r scales::percent(mod_n_subsample_temp/nrow(treemap_tree_list_df), accuracy = 0.01)` of the training data) from which will use majority voting to determine the predicted class

this RF approach has potential for cloud2trees implementation if the species composition accuracy is acceptable. what is acceptable? that's going to take much more testing than is done here...

attach RF species predictions back to our tree list for comparison against the reference data

```{r, results=F, message=F, warning=F}
# function to vote for class based on many models
rf_model_vote_predictions <- function(mod_list, predict_df){
  # get model predictions for each model
  predicted_temp <-
    1:length(mod_list) %>%
      purrr::map(function(x){
        predict(
          mod_list[[x]]
          , predict_df
        ) %>%
        dplyr::as_tibble() %>%
        dplyr::pull(1) %>%
        dplyr::as_tibble() %>%
        dplyr::rename(value=1) %>%
        dplyr::mutate(id=dplyr::row_number())
      }) %>%
      dplyr::bind_rows()
  # predicted_temp %>% dplyr::glimpse()

  # model vote
  predicted_temp <- predicted_temp %>%
    dplyr::count(id,value) %>% 
    dplyr::group_by(id) %>% 
    dplyr::slice_max(order_by = n, n = 1, with_ties = F) %>% 
    dplyr::ungroup() %>%
    dplyr::arrange(id) %>%
    dplyr::rename(predicted=value) %>% 
    dplyr::select(predicted)
  # return
  return(predicted_temp)
}

# predict it, the species symb
# predicted_trees[[1]] %>% dplyr::glimpse()
# predicted_trees[[1]] %>% 
#   ggplot2::ggplot() + 
#   ggplot2::geom_sf(mapping = ggplot2::aes(color=species_symbol)) +
#   ggplot2::scale_color_manual(values = pal_species)
# names(predicted_trees)
predicted_trees <-
  predicted_trees %>% 
  purrr::map(function(x){
    # get predicted values
    preds_temp <- rf_model_vote_predictions(
      mod_list = rf_spec_mod
      , predict_df = 
        x %>% 
        sf::st_drop_geometry() %>%
        dplyr::select(x,y,tree_height_m)
    )

    # preds back to tree list
    dta_temp <-
      x %>% 
      dplyr::mutate(
        species_symbol = 
          preds_temp$predicted
          # # rf does the factoring already
          # factor(
          #   levels = rf_spec_mod[[1]]$classes
          #   , labels = rf_spec_mod[[1]]$classes
          # )
      )
    return(dta_temp)
  })
# predicted_trees %>% purrr::list_rbind(names_to = "folder") %>% dplyr::glimpse()
# predicted_trees[[1]] %>%
#   ggplot2::ggplot() +
#   ggplot2::geom_sf(mapping = ggplot2::aes(color=species_symbol)) +
#   ggplot2::scale_color_manual(values = pal_species)
# predicted_trees[[1]] %>% sf::st_drop_geometry() %>% dplyr::count(species_symbol)
```

let's map predicted species in space for each flight

```{r, fig.height = 9.8, fig.width = 6.5}
predicted_trees %>%
  purrr::list_rbind(names_to = "dataset_factor") %>%
  dplyr::mutate(
    dataset_factor = ordered(dataset_factor, levels(tracking_df$dataset_factor))
  ) %>% 
  # sf::st_drop_geometry() %>% dplyr::count(folder) %>% 
  dplyr::inner_join(
    tracking_df %>% 
      dplyr::select(-c(path))
    , by = "dataset_factor"
  ) %>%
  dplyr::arrange(dataset_factor, desc(species_symbol)) %>% 
  # plot it
  # plt
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = stand_boundary
      , fill = NA, color = "blue"
    ) +
    ggplot2::geom_sf(
      mapping = ggplot2::aes(geometry = geom, color = species_symbol)
      , size = 0.6, alpha = 0.88
    ) +
    ggplot2::scale_color_manual(values = pal_species, drop = F) +
    ggplot2::facet_wrap(
      facets = dplyr::vars(dataset_factor)
      , ncol = 2
    ) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
      , legend.title = ggplot2::element_blank()
      , legend.key = ggplot2::element_blank()
      , strip.text = ggplot2::element_text(face = "bold", color = "black", margin = ggplot2::margin(t = 4, b = 4))
      , strip.background = ggplot2::element_rect(fill = "gray88", color = "gray88")
      # , panel.background = ggplot2::element_rect(fill = NA, color = "black")
      # , panel.spacing = ggplot2::unit(1,"lines")
      , panel.grid.major = ggplot2::element_blank()
      , panel.grid.minor = ggplot2::element_blank()
      , axis.text = ggplot2::element_blank()
      , axis.title = ggplot2::element_blank()
      , axis.ticks = ggplot2::element_blank()
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
      , fill = "none"
    )
```

#### RF Predicted Species Composition Accuracy

perform same prediction accuracy as above for comparison

for this demonstration, we'll limit our evaluation to a single flight since all flights used the same species prediction model and we are evaluating the method moreso than the flight settings when it comes to the species prediction. since we have data across the different flight settings we'll just use the flight that resulted in the highest F-score noting that the flights were very similar in their detection accuracy

```{r}
species_comp_temp <- 
  forcats::lvls_union(list(
    validation_trees$species_symbol
    , predicted_trees %>% purrr::list_rbind() %>% dplyr::pull(species_symbol)
  )) %>% 
  dplyr::as_tibble() %>% 
  dplyr::rename(species_symbol=value) %>% 
  # count reference trees
  dplyr::left_join(
    validation_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::count(species_symbol) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        ref_n=n
        , ref_pct=pct
      )
    , by = "species_symbol"
  ) %>% 
  # count pred trees
  dplyr::left_join(
    predicted_trees %>% 
      purrr::pluck(
        agg_ground_truth_match_ans %>% 
          dplyr::slice_max(n=1, order_by = f_score, with_ties = F) %>% 
          dplyr::pull(dataset_factor)    
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::count(species_symbol) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        pred_n=n
        , pred_pct=pct
      )
    , by = "species_symbol"
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric)
      , ~ dplyr::coalesce(.x,0)
    )
    # calc intersecting shared area per species (Schoener's Overlap index component)
    , overlap_pct = pmin(pred_pct, ref_pct)
    , sq_pct_error = (pred_pct - ref_pct)^2
  )
# species_comp_temp  
```

table the predicted versus reference counts by species

```{r}
species_comp_temp %>% 
  dplyr::mutate(
    # # one column for diff/pct
    pred_pct = paste0(scales::comma(pred_n,accuracy=1), "<br />(", scales::percent(pred_pct,accuracy=0.1), ")")
    , ref_pct = paste0(scales::comma(ref_n,accuracy=1), "<br />(", scales::percent(ref_pct,accuracy=0.1), ")")
  ) %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
    , names_to = "dta"
    , values_to = "prop"
  ) %>% 
  dplyr::mutate(
    dta = ifelse(dta == "pred_pct", "predicted", "field")
  ) %>% 
  tidyr::pivot_wider(
    names_from = species_symbol
    , values_from = prop
  ) %>% 
  dplyr::rename(`data`=dta) %>% 
  kableExtra::kbl(
    caption = paste0(
      "Species Proportional Comparision"
      , "<br />Total Overlap Index: "
      , sum(species_comp_temp$overlap_pct) %>% scales::percent(accuracy = 0.1 )
      # , "<br />Proportion RMSE: "
      # , sqrt(mean(species_comp_temp$sq_pct_error)) %>% scales::percent(accuracy = 0.1 )
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(
    c(" " = 1, "Species Tree Counts and Proportions" = nrow(species_comp_temp))
  )
```

looks pretty similar to the bayesian model demonstrated above with significant time savings

plot the proportion of total by species

```{r}
species_comp_temp %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "field")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_manual(values = pal_species) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

check out the site-wide dissimilarity/similarity between the predicted and reference species compositions

```{r}
library(vegan)
# vegan needs a matrix :\
bc_dissim_temp <-
  species_comp_temp %>% 
  dplyr::select(species_symbol, pred_n, ref_n) %>% 
  tidyr::pivot_longer(cols = c(pred_n, ref_n), names_to = "which_dta", values_to = "n") %>% 
  tidyr::pivot_wider(names_from = species_symbol, values_from = n) %>% 
  dplyr::mutate(which_dta = ifelse(which_dta == "pred_n", "predicted", "field")) %>% 
  tibble::column_to_rownames(var = "which_dta") %>% 
  # relative props
  vegan::decostand(method = "total") %>% 
  # dissimilarity
  vegan::vegdist(method = "bray") %>% 
  as.numeric()
# sim
bc_sim_temp <- 1 - bc_dissim_temp
# table
dplyr::tibble(
    metric = c(
      "Bray-Curtis Dissimilarity Index"
      , "Bray-Curtis Similarity"
      , "Schoener's D (total overlap) Index"
    )
    , value = c(
      bc_dissim_temp
      , bc_sim_temp
      , sum(species_comp_temp$overlap_pct) 
    )
    , huh = c(
      "proportion of relative species profile mis-matched"
      , "proportion of total distribution matching perfectly"
      , "equal to Bray-Curtis similarity for proportions"
    )
  ) %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.01)
  )) %>% 
  kableExtra::kbl(
    caption = "Predicted vs Reference Profile Similarity Metrics"
    , col.names = c("","value","interpretation")
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::column_spec(1:2, bold = TRUE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(bc_dissim_temp, bc_sim_temp)
gc()
```

lastly, let's quantify the compositional dissimilarity between the reference and predicted tree species assemblages by height class

we'll group our trees into quartile height classes based on the reference tree heights

```{r}
ht_qtr_temp <- quantile(validation_trees$field_tree_height_m, probs = c(0.25, 0.5, 0.75))
# reference trees
valht_temp <- 
  validation_trees %>% 
  dplyr::mutate(
    tree_height_m_bin = 
      cut(
        field_tree_height_m
        , breaks = c(0,ht_qtr_temp,Inf)
      ) %>% 
      make_bins_pretty(suffix = "m") %>% 
      forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(my_tree_ht_m,"-")) )
    , tree_height_m_bin = fct_relabel(tree_height_m_bin, ~ {
        levels_list <- levels(tree_height_m_bin)
        nums <- match(.x, levels_list)
        paste0("Q",nums, ": ", .x)
      })
  ) %>% 
  sf::st_drop_geometry() %>%
  # dplyr::group_by(tree_height_m_bin) %>% 
  # dplyr::summarise(
  #   dplyr::across(field_tree_height_m, .fns = list(min,max))
  # )
  # species count
  dplyr::count(tree_height_m_bin,species_symbol)

# data with row unique by species and height bin
species_comp_temp <- 
  tidyr::crossing(
    species_comp_temp %>% dplyr::distinct(species_symbol)
    , valht_temp %>% dplyr::distinct(tree_height_m_bin)
  ) %>% 
  # count reference trees
  dplyr::left_join(
    valht_temp %>% 
      dplyr::group_by(tree_height_m_bin) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        ref_n=n
        , ref_pct=pct
      ) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(species_symbol,tree_height_m_bin)
  ) %>% 
  # count pred trees
  dplyr::left_join(
    predicted_trees %>% 
      purrr::pluck(
        agg_ground_truth_match_ans %>% 
          dplyr::slice_max(n=1, order_by = f_score, with_ties = F) %>% 
          dplyr::pull(dataset_factor)    
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::mutate(
        tree_height_m_bin = 
          cut(
            tree_height_m
            , breaks = c(0,ht_qtr_temp,Inf)
          ) %>% 
          make_bins_pretty(suffix = "m") %>% 
          forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(my_tree_ht_m,"-")) )
        , tree_height_m_bin = fct_relabel(tree_height_m_bin, ~ {
            levels_list <- levels(tree_height_m_bin)
            nums <- match(.x, levels_list)
            paste0("Q",nums, ": ", .x)
          })
      ) %>% 
      dplyr::count(tree_height_m_bin,species_symbol) %>% 
      dplyr::group_by(tree_height_m_bin) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        pred_n=n
        , pred_pct=pct
      ) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(species_symbol,tree_height_m_bin)
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric)
      , ~ dplyr::coalesce(.x,0)
    )
  ) %>% 
  dplyr::filter(ref_n+pred_n>0)
```

plot the proportion of total

```{r, fig.height=9}
species_comp_temp %>% 
  dplyr::select(tree_height_m_bin, species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "field")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::facet_wrap(
    facets = dplyr::vars(tree_height_m_bin)
    , ncol = 2
    , axes = "all" 
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_manual(values = pal_species) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison\nby tree height quartile"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , axis.text = ggplot2::element_text(size = 8)
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
  )
```

most of the misclassification occurs in the smaller tree classes (i.e. below the median tree height) which have less influence on stand-level metrics like basal area and total volume

let's get the similarity/dissimilarity values by height bin to quantify the proportion of the total predicted trees that were assigned the incorrect species relative to the reference (i.e. field-measured) community profile

```{r}
bc_df_temp <- 
  species_comp_temp$tree_height_m_bin %>% 
  unique() %>% 
  purrr::map(function(x){
    # vegan needs a matrix :\
    bc_dissim_temp <-
      species_comp_temp %>% 
      dplyr::filter(tree_height_m_bin==x) %>% 
      dplyr::select(species_symbol, pred_n, ref_n) %>% 
      tidyr::pivot_longer(cols = c(pred_n, ref_n), names_to = "which_dta", values_to = "n") %>% 
      tidyr::pivot_wider(names_from = species_symbol, values_from = n) %>% 
      dplyr::mutate(which_dta = ifelse(which_dta == "pred_n", "predicted", "field")) %>% 
      tibble::column_to_rownames(var = "which_dta") %>% 
      # relative props
      vegan::decostand(method = "total") %>% 
      # dissimilarity
      vegan::vegdist(method = "bray") %>% 
      as.numeric()
    # sim
    bc_sim_temp <- 1 - bc_dissim_temp
    dplyr::tibble(
      tree_height_m_bin = x
      , dissim = bc_dissim_temp
      , sim = bc_sim_temp
    ) 
  }) %>% 
  purrr::list_rbind() %>% 
  dplyr::mutate(
    tree_height_m_bin = tree_height_m_bin %>% 
      forcats::as_factor() %>% 
      forcats::fct_expand( levels(species_comp_temp$tree_height_m_bin) ) %>% 
      forcats::fct_relevel( levels(species_comp_temp$tree_height_m_bin) )
  ) %>% 
  dplyr::arrange(tree_height_m_bin)
  
# table
bc_df_temp %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.01)
  )) %>% 
  kableExtra::kbl(
    caption = "Predicted vs Reference Profile Similarity Metrics<br />by tree height quartile"
    , col.names = c(
      "Ht. Quartile"
      , "Bray-Curtis Dissimilarity Index"
      , "Bray-Curtis Similarity"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

let's visualize that

```{r}
bc_df_temp %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.1)
    , .names = "{.col}_lab"
  )) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = tree_height_m_bin, group = 1)
  ) +
  ggplot2::geom_line(mapping = ggplot2::aes(y = dissim, color = "dissimilarity"), lwd = 2) +
  ggplot2::geom_line(mapping = ggplot2::aes(y = sim, color = "similarity"), lwd = 2) +
  ggplot2::geom_point(mapping = ggplot2::aes(y = dissim, color = "dissimilarity"), size = 4) +
  ggplot2::geom_point(mapping = ggplot2::aes(y = sim, color = "similarity"), size = 4) +
  ggrepel::geom_text_repel(mapping = ggplot2::aes(y = dissim, label = dissim_lab), size = 2.5) +
  ggrepel::geom_text_repel(mapping = ggplot2::aes(y = sim, label = sim_lab), size = 2.5) +
  ggplot2::scale_color_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
  ggplot2::scale_y_continuous(
    limits = c(0,1)
    , breaks = seq(0,1,by=0.2)
    , labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.02))
  ) +
  ggplot2::labs(
    x = "tree height quartile", y = "", color = ""
    , subtitle = "Predicted vs Reference Species Composition Similarity Metrics"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "top"
    , axis.text.x = ggplot2::element_text(size = 11, face = "bold")
    , axis.title.x = ggplot2::element_text(size = 9)
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
    , fill = "none"
  )
```

neat, a different trend than the bayesian estimation approach with the first and second height quartile species similarity scores swapped

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

#### RF Predicted Species Volume Comparison

let's update the volume estimates based on these new species predictions

```{r}
# ground_truth_prediction_match_ans[[1]] %>% dplyr::glimpse()
# validation_trees %>% dplyr::glimpse()
# validation_trees %>% dplyr::glimpse()

ground_truth_prediction_match_ans_updt <-
  tracking_df$dataset_factor %>% 
  purrr::map(function(x){
    # add on to precitions
    ground_truth_prediction_match_ans <- 
      ground_truth_prediction_match_ans[[x]] %>% 
      # add predicted species
      dplyr::left_join(
        predicted_trees[[x]] %>% 
          sf::st_drop_geometry() %>% 
          dplyr::select(treeID, species_symbol) %>% 
          dplyr::rename(
            pred_tree_id=treeID
            , pred_species_symbol = species_symbol
          )
        , by = "pred_tree_id"
      ) %>% 
      # add reference species
      dplyr::left_join(
        validation_trees %>% 
          sf::st_drop_geometry() %>% 
          dplyr::select(field_tree_id, species_symbol,tidyselect::contains("_volume_")) %>% 
          dplyr::rename(
            ref_tree_id=field_tree_id
            , ref_species_symbol = species_symbol
          ) %>% 
          dplyr::rename_with(
            .cols = c(tidyselect::starts_with("grs_volume_"),tidyselect::starts_with("net_volume_"))
            , ~ paste0("ref_",.x,recycle0 = T)
          )
        , by = "ref_tree_id"
      ) 
    # add volume
    ground_truth_prediction_match_ans <- 
      ground_truth_prediction_match_ans %>% 
      # increments for dbh,ht
      dplyr::mutate(
        dia = floor(pred_dbh_cm*0.394) %>% as.integer()
        , ht = floor(pred_tree_height_m*3.281) %>% as.integer()
        , species_symbol = as.character(pred_species_symbol)
      ) %>% 
      # add on preds
      dplyr::left_join(
        vol_lookup_pred %>% dplyr::mutate(species_symbol = as.character(species_symbol))
        , dplyr::join_by(species_symbol,ht,dia)
        , relationship = "many-to-one"
      ) %>% 
      # add on fia
      dplyr::left_join(
        vol_lookup %>% dplyr::select(species_symbol,ht,dia,tidyselect::contains("_volume_")) %>% 
           dplyr::mutate(species_symbol = as.character(species_symbol))
        , dplyr::join_by(species_symbol,ht,dia)
        , relationship = "many-to-one"
      ) %>% 
      # fill in missing fia volume
      dplyr::mutate(
        net_volume_ft3 = ifelse(is.na(net_volume_ft3), pred_net_volume_ft3, net_volume_ft3) %>% 
          # any species not in the FIA data are set to zero (probs not timber...JUSC2,QUGA)
          dplyr::coalesce(0) # smallest of trees
        , grs_volume_ft3 = ifelse(is.na(grs_volume_ft3), pred_grs_volume_ft3, grs_volume_ft3) %>% 
          # any species not in the FIA data are set to zero (probs not timber...JUSC2,QUGA)
          dplyr::coalesce(0) # smallest of trees
        , net_volume_m3 = ifelse(is.na(net_volume_m3), pred_net_volume_m3, net_volume_m3) %>% 
          # any species not in the FIA data are set to zero (probs not timber...JUSC2,QUGA)
          dplyr::coalesce(0) # smallest of trees
        , grs_volume_m3 = ifelse(is.na(grs_volume_m3), pred_grs_volume_m3, grs_volume_m3) %>% 
          # any species not in the FIA data are set to zero (probs not timber...JUSC2,QUGA)
          dplyr::coalesce(0) # smallest of trees
      ) %>% 
      dplyr::select(
        -tidyselect::starts_with("pred_net_volume")
        ,-tidyselect::starts_with("pred_grs_volume")
        , -c(dia,ht,species_symbol)
      ) %>% 
      dplyr::rename_with(
        .cols = c(tidyselect::starts_with("grs_volume_"),tidyselect::starts_with("net_volume_"))
        , ~ paste0("pred_",.x,recycle0 = T)
      )
    return(ground_truth_prediction_match_ans)
  })
names(ground_truth_prediction_match_ans_updt) <- tracking_df$dataset_factor

# ground_truth_prediction_match_ans_updt[[1]] %>% dplyr::glimpse()
# ground_truth_prediction_match_ans_updt[[1]] %>% 
#   sf::st_drop_geometry() %>% 
#   dplyr::count(ref_species_symbol,pred_species_symbol)
#   dplyr::glimpse()
# ground_truth_prediction_match_ans_updt %>%
#   purrr::list_rbind(names_to = "folder") %>%
#   sf::st_drop_geometry() %>% 
#   dplyr::filter(match_grp %in% c("true positive","commission")) %>% 
#   dplyr::select(
#     folder
#     , pred_match_idx
#     , c(tidyselect::starts_with("pred_") & tidyselect::contains("_volume_"))
#   ) %>% 
#   dplyr::group_by(folder) %>% 
#   dplyr::summarize(
#     dplyr::across(
#       .cols = tidyselect::contains("_volume_")
#       , sum
#     )
#     # , n_trees = dplyr::n()
#   ) %>% 
#   dplyr::ungroup()


# agg_ground_truth_match_ans %>% dplyr::glimpse()
agg_ground_truth_match_ans <-
  agg_ground_truth_match_ans %>% 
  # get rid of columns we'll create
  dplyr::select( -dplyr::any_of(c(
    "hey_xxxxxxxxxx"
    , "pred_net_volume_ft3"
    , "pred_grs_volume_ft3"
    , "pred_net_volume_m3"
    , "pred_grs_volume_m3" 
  ))) %>% 
  dplyr::left_join(
    ground_truth_prediction_match_ans_updt %>%
      purrr::list_rbind(names_to = "dataset_factor") %>%
      dplyr::mutate(
        dataset_factor = ordered(dataset_factor, levels(tracking_df$dataset_factor))
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::filter(match_grp %in% c("true positive","commission")) %>% 
      dplyr::select(
        dataset_factor
        , pred_match_idx
        , c(tidyselect::starts_with("pred_") & tidyselect::contains("_volume_"))
      ) %>% 
      dplyr::group_by(dataset_factor) %>% 
      dplyr::summarize(
        dplyr::across(
          .cols = tidyselect::contains("_volume_")
          , sum
        )
        # , n_trees = dplyr::n()
      ) %>% 
      dplyr::ungroup()
    , by = "dataset_factor"
  ) %>% 
  dplyr::mutate(
    dataset_factor = ordered(dataset_factor, levels(tracking_df$dataset_factor))
  ) %>% 
  # diff cols
  dplyr::mutate(
    # 'diff_' columns are calculated as the predicted value minus the actual value
    diff_net_volume_m3 = pred_net_volume_m3-ref_net_volume_m3
    , diff_grs_volume_m3 = pred_grs_volume_m3-ref_grs_volume_m3
    # 'abs_diff_' columns are calculated as the predicted value minus the actual value
    , abs_diff_net_volume_m3 = abs(diff_net_volume_m3)
    , abs_diff_grs_volume_m3 = abs(diff_grs_volume_m3)
    # 'pct_diff_' columns are calculated as the predicted value minus the actual value divided by the actual value
    , pct_diff_net_volume_m3 = (pred_net_volume_m3-ref_net_volume_m3)/ref_net_volume_m3
    , pct_diff_grs_volume_m3 = (pred_grs_volume_m3-ref_grs_volume_m3)/ref_grs_volume_m3
  )
# agg_ground_truth_match_ans %>% dplyr::glimpse()
```

table that

```{r}
agg_ground_truth_match_ans %>% 
  dplyr::mutate(
    ref_trees = tp_n+fn_n
    , pred_trees = tp_n+fp_n
  ) %>% 
  dplyr::select(
    dataset_factor, tidyselect::starts_with("flight_")
    # , site_area_m2
    # detection
    , ref_trees
    , pred_trees
    # quantification
    , tidyselect::ends_with("_volume_m3")
  ) %>% 
  dplyr::select(!tidyselect::starts_with("abs_diff_")) %>% 
  dplyr::mutate(
    dplyr::across(
      tidyselect::starts_with("flight_")
      , as.factor
    )
    , dplyr::across(
      .cols = c(
        tidyselect::starts_with("pct_diff_")
        , tidyselect::ends_with("_rate")
      )
      , .fn = ~ scales::percent(.x, accuracy = .1)
    )
    , dplyr::across(
      .cols = c(tidyselect::ends_with("_trees"))
      , .fn = ~ scales::comma(.x, accuracy = 1)
    )
    , dplyr::across(
      .cols = dplyr::where(is.numeric)
      , .fn = ~ scales::comma(.x, accuracy = 0.1)
    )
    # # one column for diff/pct
    , diff_net_volume_m3 = paste0(diff_net_volume_m3, "<br>(", pct_diff_net_volume_m3, ")")
    , diff_grs_volume_m3 = paste0(diff_grs_volume_m3, "<br>(", pct_diff_grs_volume_m3, ")")
  ) %>% 
  dplyr::select(
    !tidyselect::starts_with("pct_diff_")
  ) %>% 
  dplyr::arrange(dataset_factor) %>% 
  dplyr::select(
    -c(
      dataset_factor
    )
  ) %>% 
  dplyr::select(
    tidyselect::starts_with("flight_")
    , tidyselect::ends_with("_trees")
    , tidyselect::contains("grs_volume")
    , tidyselect::contains("net_volume")
  ) %>% 
  # dplyr::glimpse()
  kableExtra::kbl(
    caption = "Volume: Stand-Level Accuracy"
    , col.names = c(
      "altitude (m)", "speed (m/s)"
      , "reference", "predicted"
      , rep(c("reference","predicted","difference"), times = 2)
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 11.5) %>% 
  kableExtra::add_header_above(
    c(
      " "=2
      , "# Trees" = 2
      # , "Area" = 3
      , "Gross<br />Volume m<sup>3</sup>" = 3
      , "Net<br />Volume m<sup>3</sup>" = 3
    )
    , escape = F
  ) %>% 
  kableExtra::column_spec( c(2,seq(4,10,by=3) ), border_right = TRUE, include_thead = TRUE) %>% 
  kableExtra::collapse_rows(columns = c(1), valign = "top")
```

plot

```{r, fig.height=8}
agg_ground_truth_match_ans %>% 
  dplyr::select(
    dataset_factor
    , pct_diff_net_volume_m3
    , pct_diff_grs_volume_m3
  ) %>% 
  dplyr::mutate(
    tot = -1* (abs(pct_diff_net_volume_m3) + abs(pct_diff_grs_volume_m3))
  ) %>% 
  tidyr::pivot_longer(cols = -c(dataset_factor,tot)) %>% 
  dplyr::mutate(
    name = dplyr::recode_values(
      name
      , "pct_diff_net_volume_m3" ~ latex2exp::TeX("Net volume $m^{3}$", output = "character")
      , "pct_diff_grs_volume_m3" ~ latex2exp::TeX("Gross volume $m^{3}$", output = "character")
    )
    , value_lab = scales::percent(value,accuracy=0.1)
  ) %>% 

ggplot2::ggplot(
  mapping = ggplot2::aes(
    y = reorder(dataset_factor, tot)
    , x = value
    , group = name
  )
) +
  ggplot2::geom_vline(xintercept = 0, color = "gray95", lwd = 1.5, alpha = 0.7) +
  ggplot2::geom_vline(xintercept = c(0.05,-0.05), color = "gray77", lwd = 0.5, alpha = 0.7) +
  ggplot2::geom_vline(xintercept = c(0.1,-0.1), color = "gray55", lwd = 0.5, alpha = 0.7) +
  ggplot2::geom_vline(xintercept = c(0.15,-0.15), color = "gray33", lwd = 0.5, alpha = 0.7) +
  ggplot2::geom_vline(xintercept = c(0.2,-0.2), color = "gray11", lwd = 0.5) +
  ggplot2::geom_segment(
    mapping = ggplot2::aes(yend = dataset_factor, xend = 0,color = value > 0)
    # , color = "gray44"
  ) +
  ggplot2::geom_point(
    mapping = ggplot2::aes(color = value > 0)
    , size = 3
  ) +
  # lhs labs
  ggplot2::geom_text(
    data = function(x){dplyr::filter(x,value<0)}
    , mapping = ggplot2::aes(label = value_lab, fontface = "bold")
    # , color = "black"
    , size = 3
    , nudge_x = -0.006
    , hjust = 1
  ) +
  # rhs labs
  ggplot2::geom_text(
    data = function(x){dplyr::filter(x,value>=0)}
    , mapping = ggplot2::aes(label = value_lab, fontface = "bold")
    # , color = "black"
    , size = 3
    , nudge_x = 0.006
    , hjust = 0
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(name)
    # , labeller = ggplot2::label_parsed
    , labeller = ggplot2::as_labeller(
      function(x) {
        paste0("atop(", x, ", 'percent error')")
      }
      , default = ggplot2::label_parsed
    )
  ) +
  ggplot2::scale_color_manual(values = c("firebrick","navy")) +
  ggplot2::scale_x_continuous(
    limits = c(-0.2,0.2)
    , labels = scales::percent
    # , breaks = scales::breaks_extended(n=10)
    # , breaks = function(x) seq(min(x), max(x), by = 0.05)
  ) +
  ggplot2::labs(
    y = "" # "comparison"
    , x = "percent error"
    , subtitle = "positive = UAS-measured higher than field-measured\nnegative = UAS-measured lower than field-measured"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , axis.text.y = ggplot2::element_text(size = 9, face = "bold")
    # , plot.subtitle = ggplot2::element_text(hjust = 1)
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , panel.grid.major.x = element_blank()
    , panel.grid.minor.x = element_blank()
  )
```

these volume errors are slightly larger than the volume errors based on when species was randomly assigned to trees to [proportionally match the distribution](#spec_prop_alloc) of tree species. the reason for this increase in error is that when species were proportionally, randomly assigned there were a greater number of non-timber species (i.e. JUSC2, POTR5, and QUGA) which had the effect of removing those trees from the stand-level volume estimates. because our estimated DBH values tended to be larger than the field-measured values (~`r scales::comma(mean(agg_ground_truth_match_ans$diff_dbh_cm_mean), accuracy = 0.01)` cm mean error) and our stand-level TPA were highly accurate (~ `r scales::percent(mean(agg_ground_truth_match_ans$pct_diff_trees_per_ha), accuracy = 0.1)` error), our predicted stand-level basal area was overestimated by `r scales::percent(fivenum(agg_ground_truth_match_ans$pct_diff_basal_area_m2_per_ha)[c(1,5)], accuracy = 0.1) %>% paste0(collapse = " to ")` across the flights. The overestimation of stand-level gross volume by `r scales::percent(fivenum(agg_ground_truth_match_ans$pct_diff_grs_volume_m3)[c(1,5)], accuracy = 0.1) %>% paste0(collapse = " to ")` mirrors this trend, especially when trees aren't (improperly) removed from the volume summation due to being classified as a non-timber species as in the random, proportional assignment.

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

To potentially improve DBH estimates we could explore linking species and DBH is a sequential model approach. That is, we could run the species model (using RF, probably), then in the DBH model we would allow the coefficients to vary by predicted species

```{r, include=F, eval=FALSE}
# estimate species and DBH simultaneously with brms::mvbf()
# the "link" between the two models is the `tm_id` in the grouping term
# this type of model will estimate whether a plot that is "above average" for a certain species might also be "above average" for the asymptotic diameter in the Chapman-richards growth curve

# define the species formula
bf_species <- brms::bf(
  species_symbol | weights(tree_weight) ~ 1 + tree_height_m + (1 + tree_height_m | p | tm_id)
) + brms::categorical(link = "logit")

# define the non-linear dbh formula
# note the use of | p | to link the random effects to the species model
bf_dbh <- brms::bf(
  dbh_cm | weights(tree_weight) ~ asym * (1 - exp(-k * tree_height_m))^p
  , asym ~ 1 + (1 | p | tm_id)
  , k ~ 1 + (1 | p | tm_id)
  , p ~ 1 + (1 | p | tm_id)
  , nl = TRUE
) + brms::lognormal()

st_temp <- Sys.time()
# combine into a multivariate model
joint_model <- brms::brm(
  formula = bf_species + bf_dbh
  , data = treemap_trees_df
  , prior = c(
    brms::prior(normal(60, 20), nlpar = "asym", resp = "dbhcm", lb = 0)
    , brms::prior(normal(0.05, 0.02), nlpar = "k", resp = "dbhcm", lb = 0)
    , brms::prior(normal(2, 0.5), nlpar = "p", resp = "dbhcm", lb = 0)
  )
  , iter = 6000
  , warmup = 3000 
  , chains = 4
  , cores = lasR::half_cores()
  , file = paste0("../data/", "dbh_species_model")
)
# timer
(mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric())

summary(joint_model)
```

```{r, include=F, eval=FALSE}
# another option for linking species and DBH is a sequential model approach
# that is, run the species model then the dbh model with the asym and k coefficients
# allowed to vary by predicted species:
# Non-linear DBH model where asym and k vary by species
# p is kept as a global parameter to simplify estimation
dbh_species_model <- brms::brm(
  formula = brms::bf(
    dbh_cm | weights(tree_weight) ~ asym * (1 - exp(-k * tree_height_m))^p,
    asym ~ 0 + species_symbol + (1 | tm_id),
    k ~ 0 + species_symbol + (1 | tm_id),
    p ~ 1,
    nl = TRUE
  ),
  data = treemap_trees_df,
  prior = c(
    brms::prior(normal(60, 20), nlpar = "asym", lb = 0),
    brms::prior(normal(0.05, 0.02), nlpar = "k", lb = 0),
    brms::prior(normal(2, 0.5), nlpar = "p", lb = 0)
  ),
  family = brms::lognormal(),
  iter = 4000, 
  warmup = 2000, 
  chains = 4, 
  cores = 6,
  backend = "cmdstanr"
)

# if the brms species classifier is taking too long to run for package implementation
# use a random forest classifier

# Train the Random Forest
# We use the weights to account for TreeMap imputation frequency
species_rf <- randomForest::randomForest(
  formula = species_symbol ~ tree_height_m + x + y,
  data = treemap_trees_df,
  weights = treemap_trees_df$tree_weight,
  ntree = 500,
  mtry = 2,
  importance = TRUE
)

# Predict species for the target tree list
# We assign the discrete label directly in a tidy pipe
target_tree_list <- target_tree_list |>
  dplyr::mutate(
    species_symbol = stats::predict(species_rf, newdata = target_tree_list)
  )
# Fit the non-linear model with species as a covariate
dbh_model <- brms::brm(
  formula = brms::bf(
    dbh_cm | weights(tree_weight) ~ asym * (1 - exp(-k * tree_height_m))^p,
    asym ~ 0 + species_symbol + (1 | tm_id),
    k ~ 0 + species_symbol + (1 | tm_id),
    p ~ 1,
    nl = TRUE
  ),
  data = treemap_trees_df,
  prior = c(
    brms::prior(normal(60, 20), nlpar = "asym", lb = 0),
    brms::prior(normal(0.05, 0.02), nlpar = "k", lb = 0),
    brms::prior(normal(2, 0.5), nlpar = "p", lb = 0)
  ),
  family = brms::lognormal(),
  iter = 4000, 
  warmup = 2000, 
  chains = 4, 
  cores = 6
)
# Generate DBH predictions
dbh_preds <- brms::posterior_epred(
  object = dbh_model,
  newdata = target_tree_list,
  ndraws = 100, # Use enough draws to capture the mean
  re_formula = NULL # Keep the plot-level spatial variation
)

# Attach the mean prediction to the list
target_tree_list <- target_tree_list |>
  dplyr::mutate(
    predicted_dbh_cm = base::colMeans(dbh_preds)
  ) 
```


#### Field Data

let's check out the reference data species distribution

let's map the reference species in space

```{r}
# plt
validation_trees %>% 
  dplyr::arrange(desc(species_symbol)) %>% 
  dplyr::mutate(dataset_factor="Reference Trees") %>% 
  # plot it
  # plt
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = stand_boundary
      , fill = NA, color = "blue"
    ) +
    ggplot2::geom_sf(
      mapping = ggplot2::aes(color = species_symbol)
      , size = 1.2, alpha = 0.88
    ) +
    ggplot2::scale_color_manual(values = pal_species) +
    ggplot2::facet_wrap(
      facets = dplyr::vars(dataset_factor)
      , ncol = 3
    ) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
      , legend.title = ggplot2::element_blank()
      , legend.key = ggplot2::element_blank()
      , strip.text = ggplot2::element_text(face = "bold", color = "black", margin = ggplot2::margin(t = 4, b = 4))
      , strip.background = ggplot2::element_rect(fill = "gray88", color = "gray88")
      # , panel.background = ggplot2::element_rect(fill = NA, color = "black")
      # , panel.spacing = ggplot2::unit(1,"lines")
      , panel.grid.major = ggplot2::element_blank()
      , panel.grid.minor = ggplot2::element_blank()
      , axis.text = ggplot2::element_blank()
      , axis.title = ggplot2::element_blank()
      , axis.ticks = ggplot2::element_blank()
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
      , fill = "none"
    )
```

and table that

```{r}
dta_temp <- validation_trees %>%
  sf::st_drop_geometry() %>% 
  dplyr::mutate(dataset_factor="Reference Trees") %>% 
  dplyr::select(
    dataset_factor
    , species_symbol
  ) %>% 
  dplyr::count(dplyr::pick(dplyr::everything())) %>% 
  dplyr::group_by(dataset_factor) %>% 
  dplyr::mutate(pct=n/sum(n,na.rm = T)) %>% 
  dplyr::ungroup() 
# plt
dta_temp %>% 
  dplyr::mutate(
    # # one column for diff/pct
    n = paste0(scales::comma(n,accuracy=1), "<br />(", scales::percent(pct,accuracy=0.1), ")")
  ) %>% 
  dplyr::select(-c(dataset_factor,pct)) %>% 
  tidyr::pivot_wider(values_from = n, names_from = species_symbol) %>% 
  # dplyr::glimpse()
  kableExtra::kbl(
    caption = "Reference Species"
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling()
  
```

pretty simple species mix...

plot the proportional distribution

```{r}
dta_temp %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    lab = ifelse(
      pct>0.0005
      , paste0(
        scales::percent(pct,accuracy=1) 
        ,"\n("
        , scales::comma(n,accuracy=1) 
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct, x = species_symbol)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = species_symbol)
    , width = 0.6
    , color = NA, alpha = 0.8
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab)
    , color = "black", size = 2.5, vjust = -0.2
  ) +
  ggplot2::scale_fill_manual(values = pal_species) +
  # ggplot2::scale_color_manual(values = c("black","white"), guide = "none") +
  ggplot2::scale_x_discrete(drop = F) +
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.15))
  ) +
  ggplot2::labs(
    x = "", y = "", fill = ""
    , subtitle = "Reference Species"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 7, angle = 45, hjust = 1, vjust = 1)
    , axis.text.y = ggplot2::element_blank()
    , axis.ticks.y = ggplot2::element_blank()
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

#### Predicted Species Composition Accuracy

accuracy metrics for a species classification model are calculated using a confusion matrix comparing the predicted and reference data

for this demonstration, we'll limit our evaluation to a single flight since all flights used the same species prediction model and we are evaluating the method moreso than the flight settings when it comes to the species prediction. since we have data across the different flight settings we'll just use the flight that resulted in the highest F-score noting that the flights were very similar in their detection accuracy

```{r}
species_comp_temp <-
  forcats::lvls_union(list(
    validation_trees$species_symbol
    , predicted_trees %>% purrr::list_rbind() %>% dplyr::pull(species_symbol)
  )) %>% 
  dplyr::as_tibble() %>% 
  dplyr::rename(species_symbol=value) %>% 
  # count reference trees
  dplyr::left_join(
    validation_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::count(species_symbol) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        ref_n=n
        , ref_pct=pct
      )
    , by = "species_symbol"
  ) %>% 
  # count pred trees
  dplyr::left_join(
    predicted_trees %>% 
      purrr::pluck(
        agg_ground_truth_match_ans %>% 
          dplyr::slice_max(n=1, order_by = f_score, with_ties = F) %>% 
          dplyr::pull(dataset_factor)    
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::count(species_symbol) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        pred_n=n
        , pred_pct=pct
      )
    , by = "species_symbol"
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric)
      , ~ dplyr::coalesce(.x,0)
    )
    # calc intersecting shared area per species (Schoener's Overlap index component)
    , overlap_pct = pmin(pred_pct, ref_pct)
    , sq_pct_error = (pred_pct - ref_pct)^2
  )
# species_comp_temp  
```

let's start by plotting the predicted versus reference counts by species

```{r}
species_comp_temp %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x=ref_n,y=pred_n,color=species_symbol,label=species_symbol,fontface="bold")
  ) +
  ggplot2::geom_abline() +
  ggplot2::geom_point(size = 3, alpha = 0.88) +
  ggrepel::geom_text_repel(
    size = 4
    , segment.color = NA
    , box.padding = 0.1
    , point.padding = 0
    , force = 0.5
    # , max.overlaps = Inf
  ) +
  ggplot2::scale_color_manual(values = pal_species) +
  ggplot2::scale_x_continuous(
    limits = c(0,max(species_comp_temp$pred_n,species_comp_temp$ref_n))
    , labels = scales::comma
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0,max(species_comp_temp$pred_n,species_comp_temp$ref_n))
    , labels = scales::comma
  ) +
  ggplot2::labs(
    x = "# field trees"
    , y = "# predicted trees"
    , subtitle = "Species Count Equality"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
      legend.position = "top"
      , legend.title = ggplot2::element_blank()
      , legend.key = ggplot2::element_blank()
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
      , fill = "none"
    )  
```

table it

```{r}
species_comp_temp %>% 
  dplyr::mutate(
    # # one column for diff/pct
    pred_pct = paste0(scales::comma(pred_n,accuracy=1), "<br />(", scales::percent(pred_pct,accuracy=0.1), ")")
    , ref_pct = paste0(scales::comma(ref_n,accuracy=1), "<br />(", scales::percent(ref_pct,accuracy=0.1), ")")
  ) %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
    , names_to = "dta"
    , values_to = "prop"
  ) %>% 
  dplyr::mutate(
    dta = ifelse(dta == "pred_pct", "predicted", "field")
  ) %>% 
  tidyr::pivot_wider(
    names_from = species_symbol
    , values_from = prop
  ) %>% 
  dplyr::rename(`data`=dta) %>% 
  kableExtra::kbl(
    caption = paste0(
      "Species Proportional Comparision"
      , "<br />Total Overlap Index: "
      , sum(species_comp_temp$overlap_pct) %>% scales::percent(accuracy = 0.1 )
      # , "<br />Proportion RMSE: "
      # , sqrt(mean(species_comp_temp$sq_pct_error)) %>% scales::percent(accuracy = 0.1 )
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(
    c(" " = 1, "Species Tree Counts and Proportions" = nrow(species_comp_temp))
  )

```

in the table above, we calculated the total overlap index of `r sum(species_comp_temp$overlap_pct) %>% scales::percent(accuracy = 0.1 )` which is sometimes referred to as Schoener's D index where a value of 1 means complete overlap (identical), while 0 means no overlap. we'll explore this concept of similarity/dissimilarity between the predicted and reference species compositions in more detail below

we can plot the predicted vs reference proportional distribution by species

```{r}
species_comp_temp %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(name = ifelse(name == "pred_pct", "predicted", "field")) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = species_symbol, y = value, fill = name, group = name)
  ) +
  ggplot2::geom_area(position = "identity", alpha = 0.4) +
  ggplot2::geom_line(
    mapping = ggplot2::aes(color = name)
    , linewidth = 1
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  harrypotter::scale_color_hp_d(option = "hermionegranger") +
  harrypotter::scale_fill_hp_d(option = "hermionegranger") +
  ggplot2::labs(
    subtitle = "Relative Species Abundance"
    , x = ""
    , y = "relative proportion of total"
    , fill = "", color = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

and the proportion of total

```{r}
species_comp_temp %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "field")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_manual(values = pal_species) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

While not specifically designed for evaluating the accuracy of a proposed prediction method, the `vegan` [package](https://vegandevs.github.io/vegan/reference/vegdist.html) provides utilities for comparing species assemblages using community ecology methods. Let's use the `vegan::vegdist()` function to compute the Bray-Curtis index which can be used to quantify the compositional dissimilarity between two different samples. This index is bounded between 0 and 1, where 0 = same composition and 1 = maximally dissimilar (see [here](https://peat-clark.github.io/BIO381/veganTutorial.html)). For our purposes, dissimilarity quantifies the proportion of the total predicted trees that were assigned the incorrect species relative to the reference (i.e. field-measured) community profile.

Because Bray-Curtis is a **dissimilarity** measure and Schoener's D (total overlap) index which we looked at above is a **similarity** measure, they can be interpreted as: 

$$\textrm{Bray-Curtis}_{dissimilarity} \approx 1-\textrm{Schoener}_{similarity}$$

we'll start by looking at the site-wide dissimilarity between the predicted and reference species compositions

```{r}
library(vegan)
# vegan needs a matrix :\
bc_dissim_temp <-
  species_comp_temp %>% 
  dplyr::select(species_symbol, pred_n, ref_n) %>% 
  tidyr::pivot_longer(cols = c(pred_n, ref_n), names_to = "which_dta", values_to = "n") %>% 
  tidyr::pivot_wider(names_from = species_symbol, values_from = n) %>% 
  dplyr::mutate(which_dta = ifelse(which_dta == "pred_n", "predicted", "field")) %>% 
  tibble::column_to_rownames(var = "which_dta") %>% 
  # relative props
  vegan::decostand(method = "total") %>% 
  # dissimilarity
  vegan::vegdist(method = "bray") %>% 
  as.numeric()
# sim
bc_sim_temp <- 1 - bc_dissim_temp
# table
dplyr::tibble(
    metric = c(
      "Bray-Curtis Dissimilarity Index"
      , "Bray-Curtis Similarity"
      , "Schoener's D (total overlap) Index"
    )
    , value = c(
      bc_dissim_temp
      , bc_sim_temp
      , sum(species_comp_temp$overlap_pct) 
    )
    , huh = c(
      "proportion of relative species profile mis-matched"
      , "proportion of total distribution matching perfectly"
      , "equal to Bray-Curtis similarity for proportions"
    )
  ) %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.01)
  )) %>% 
  kableExtra::kbl(
    caption = "Predicted vs Reference Profile Similarity Metrics"
    , col.names = c("","value","interpretation")
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::column_spec(1:2, bold = TRUE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(bc_dissim_temp, bc_sim_temp)
gc()
```

lastly, let's quantify the compositional dissimilarity between the reference and predicted tree species assemblages by height class

we'll group our trees into quartile height classes based on the reference tree heights

```{r}
ht_qtr_temp <- quantile(validation_trees$field_tree_height_m, probs = c(0.25, 0.5, 0.75))
# reference trees
valht_temp <- 
  validation_trees %>% 
  dplyr::mutate(
    tree_height_m_bin = 
      cut(
        field_tree_height_m
        , breaks = c(0,ht_qtr_temp,Inf)
      ) %>% 
      make_bins_pretty(suffix = "m") %>% 
      forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(my_tree_ht_m,"-")) )
    , tree_height_m_bin = fct_relabel(tree_height_m_bin, ~ {
        levels_list <- levels(tree_height_m_bin)
        nums <- match(.x, levels_list)
        paste0("Q",nums, ": ", .x)
      })
  ) %>% 
  sf::st_drop_geometry() %>%
  # dplyr::group_by(tree_height_m_bin) %>% 
  # dplyr::summarise(
  #   dplyr::across(field_tree_height_m, .fns = list(min,max))
  # )
  # species count
  dplyr::count(tree_height_m_bin,species_symbol)

# data with row unique by species and height bin
species_comp_temp <- 
  tidyr::crossing(
    species_comp_temp %>% dplyr::distinct(species_symbol)
    , valht_temp %>% dplyr::distinct(tree_height_m_bin)
  ) %>% 
  # count reference trees
  dplyr::left_join(
    valht_temp %>% 
      dplyr::group_by(tree_height_m_bin) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        ref_n=n
        , ref_pct=pct
      ) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(species_symbol,tree_height_m_bin)
  ) %>% 
  # count pred trees
  dplyr::left_join(
    predicted_trees %>% 
      purrr::pluck(
        agg_ground_truth_match_ans %>% 
          dplyr::slice_max(n=1, order_by = f_score, with_ties = F) %>% 
          dplyr::pull(dataset_factor)    
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::mutate(
        tree_height_m_bin = 
          cut(
            tree_height_m
            , breaks = c(0,ht_qtr_temp,Inf)
          ) %>% 
          make_bins_pretty(suffix = "m") %>% 
          forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(my_tree_ht_m,"-")) )
        , tree_height_m_bin = fct_relabel(tree_height_m_bin, ~ {
            levels_list <- levels(tree_height_m_bin)
            nums <- match(.x, levels_list)
            paste0("Q",nums, ": ", .x)
          })
      ) %>% 
      dplyr::count(tree_height_m_bin,species_symbol) %>% 
      dplyr::group_by(tree_height_m_bin) %>% 
      dplyr::mutate(pct=n/sum(n)) %>% 
      dplyr::rename(
        pred_n=n
        , pred_pct=pct
      ) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(species_symbol,tree_height_m_bin)
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric)
      , ~ dplyr::coalesce(.x,0)
    )
  ) %>% 
  dplyr::filter(ref_n+pred_n>0)
```

plot the count of trees by species and height bin

```{r, fig.height=9}
species_comp_temp %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x=ref_n,y=pred_n,color=species_symbol,label=species_symbol,fontface="bold")
  ) +
  ggplot2::geom_abline() +
  ggplot2::geom_point(size = 3, alpha = 0.88) +
  ggrepel::geom_text_repel(
    size = 2.5
    , segment.color = NA
    , box.padding = 0.1
    , point.padding = 0
    , force = 0.5
    # , max.overlaps = Inf
  ) +
  ggplot2::facet_wrap(
    facets = dplyr::vars(tree_height_m_bin)
    , ncol = 2
    , axes = "all" 
  ) +
  ggplot2::scale_color_manual(values = pal_species) +
  ggplot2::scale_x_continuous(
    limits = c(0,max(species_comp_temp$pred_n,species_comp_temp$ref_n))
    , labels = scales::comma
    , breaks = scales::breaks_extended(n=7)
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0,max(species_comp_temp$pred_n,species_comp_temp$ref_n))
    , labels = scales::comma
    , breaks = scales::breaks_extended(n=7)
  ) +
  ggplot2::labs(
    x = "# field trees"
    , y = "# predicted trees"
    , subtitle = "Species Count Equality\nby tree height quartile"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.title = ggplot2::element_blank()
    , legend.key = ggplot2::element_blank()
    , axis.text = ggplot2::element_text(size = 8)
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
    , fill = "none"
  )
```

table it by height bin

```{r}
species_comp_temp %>% 
  dplyr::mutate(
    # # one column for diff/pct
    pred_pct = paste0(scales::comma(pred_n,accuracy=1), "<br />(", scales::percent(pred_pct,accuracy=0.1), ")")
    , ref_pct = paste0(scales::comma(ref_n,accuracy=1), "<br />(", scales::percent(ref_pct,accuracy=0.1), ")")
  ) %>% 
  dplyr::select(tree_height_m_bin, species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
    , names_to = "dta"
    , values_to = "prop"
  ) %>% 
  dplyr::mutate(
    dta = ifelse(dta == "pred_pct", "predicted", "field")
  ) %>% 
  tidyr::pivot_wider(
    names_from = species_symbol
    , values_from = prop
  ) %>% 
  dplyr::arrange(tree_height_m_bin) %>% 
  dplyr::rename(`Ht. Quartile`=tree_height_m_bin, `data`=dta) %>% 
  kableExtra::kbl(
    caption = paste0(
      "Species Proportional Comparision<br />by tree height quartile"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(
    c(" " = 2, "Species Tree Counts and Proportions by Height" =length(unique(species_comp_temp$species_symbol)) )
  ) %>% 
  kableExtra::column_spec(c(2), border_right = TRUE, include_thead = TRUE) %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

interesting

plot the proportion of total

```{r, fig.height=9}
species_comp_temp %>% 
  dplyr::select(tree_height_m_bin, species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "field")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::facet_wrap(
    facets = dplyr::vars(tree_height_m_bin)
    , ncol = 2
    , axes = "all" 
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_manual(values = pal_species) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison\nby tree height quartile"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , axis.text = ggplot2::element_text(size = 8)
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
  )
```

most of the misclassification occurs in the smaller tree classes which have less influence on stand-level metrics like basal area and total volume

let's get the similarity/dissimilarity values by height bin to quantify the proportion of the total predicted trees that were assigned the incorrect species relative to the reference (i.e. field-measured) community profile

```{r}
bc_df_temp <- 
  species_comp_temp$tree_height_m_bin %>% 
  unique() %>% 
  purrr::map(function(x){
    # vegan needs a matrix :\
    bc_dissim_temp <-
      species_comp_temp %>% 
      dplyr::filter(tree_height_m_bin==x) %>% 
      dplyr::select(species_symbol, pred_n, ref_n) %>% 
      tidyr::pivot_longer(cols = c(pred_n, ref_n), names_to = "which_dta", values_to = "n") %>% 
      tidyr::pivot_wider(names_from = species_symbol, values_from = n) %>% 
      dplyr::mutate(which_dta = ifelse(which_dta == "pred_n", "predicted", "field")) %>% 
      tibble::column_to_rownames(var = "which_dta") %>% 
      # relative props
      vegan::decostand(method = "total") %>% 
      # dissimilarity
      vegan::vegdist(method = "bray") %>% 
      as.numeric()
    # sim
    bc_sim_temp <- 1 - bc_dissim_temp
    dplyr::tibble(
      tree_height_m_bin = x
      , dissim = bc_dissim_temp
      , sim = bc_sim_temp
    ) 
  }) %>% 
  purrr::list_rbind() %>% 
  dplyr::mutate(
    tree_height_m_bin = tree_height_m_bin %>% 
      forcats::as_factor() %>% 
      forcats::fct_expand( levels(species_comp_temp$tree_height_m_bin) ) %>% 
      forcats::fct_relevel( levels(species_comp_temp$tree_height_m_bin) )
  ) %>% 
  dplyr::arrange(tree_height_m_bin)
  
# table
bc_df_temp %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.01)
  )) %>% 
  kableExtra::kbl(
    caption = "Predicted vs Reference Profile Similarity Metrics<br />by tree height quartile"
    , col.names = c(
      "Ht. Quartile"
      , "Bray-Curtis Dissimilarity Index"
      , "Bray-Curtis Similarity"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

let's visualize that

```{r}
bc_df_temp %>% 
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric)
    , ~scales::percent(.x,accuracy=0.1)
    , .names = "{.col}_lab"
  )) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = tree_height_m_bin, group = 1)
  ) +
  ggplot2::geom_line(mapping = ggplot2::aes(y = dissim, color = "dissimilarity"), lwd = 2) +
  ggplot2::geom_line(mapping = ggplot2::aes(y = sim, color = "similarity"), lwd = 2) +
  ggplot2::geom_point(mapping = ggplot2::aes(y = dissim, color = "dissimilarity"), size = 4) +
  ggplot2::geom_point(mapping = ggplot2::aes(y = sim, color = "similarity"), size = 4) +
  ggrepel::geom_text_repel(mapping = ggplot2::aes(y = dissim, label = dissim_lab), size = 2.5) +
  ggrepel::geom_text_repel(mapping = ggplot2::aes(y = sim, label = sim_lab), size = 2.5) +
  ggplot2::scale_color_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
  ggplot2::scale_y_continuous(
    limits = c(0,1)
    , breaks = seq(0,1,by=0.2)
    , labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.02))
  ) +
  ggplot2::labs(
    x = "tree height quartile", y = "", color = ""
    , subtitle = "Predicted vs Reference Species Composition Similarity Metrics"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "top"
    , axis.text.x = ggplot2::element_text(size = 11, face = "bold")
    , axis.title.x = ggplot2::element_text(size = 9)
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
    , fill = "none"
  )
```

neat 

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

