# Data Overview

```{r, include=F, warning=F, message=F, eval = F}
######################################################################################################################
######################################################################################################################
# https://doi.org/10.5281/zenodo.6637599
# Tallo database to build ITD segmentation ws functions by genus/forest type based on height-crown diameter?
######################################################################################################################
######################################################################################################################
# https://doi.org/10.5281/zenodo.6637599
# we developed the Tallo database, a collection of 498,838 georeferenced and taxonomically 
# standardized records of individual trees for which stem diameter, height and/or crown radius 
# have been measured. These data were collected at 61,856 globally distributed sites, spanning 
# all major forested and non-forested biomes. The majority of trees in the database are identified
# to species (88%), and collectively Tallo includes data for 5163 species distributed 
# across 1453 genera and 187 plant families.

######################################################################################################################
######################################################################################################################
# https://gdam-api-v2-782971006568.us-west1.run.app/docs#/Index/index
# fastfuels api to get species?
######################################################################################################################
######################################################################################################################
###########################################################
###########################################################
# here's the example py from the docs
###########################################################
###########################################################
# import pandas as pd
# import requests

# # 1. Prepare your input data
# input_df = pd.DataFrame({
#     "Lat": [48.1234, 48.1240],
#     "Lon": [-114.5678, -114.5680],
#     "HT": [25.5, 18.2],
#     "DIA": [None, 8.5]  # Missing fields will be imputed
# })

# # 2. Format the payload using orient="split"
# payload = {
#     "trees": input_df.to_dict(orient="split")
# }

# # 3. Send the request
# response = requests.post("https://<API_URL>/v1/predict/batch", json=payload)

# # 4. Reconstruct the output directly back into a Pandas DataFrame
# response_data = response.json()
# predictions_df = pd.DataFrame(**response_data["predictions"])

# print(predictions_df.head())

###########################################################
###########################################################
# here's the start of formatting data to send to api
###########################################################
###########################################################
validation_trees %>% 
  sf::st_transform(crs = sf::st_crs(4326)) %>% 
  dplyr::mutate(
    tree_lon = sf::st_coordinates(.)[,1]
    , tree_lat = sf::st_coordinates(.)[,2]
    , fastfuels_api = paste0(
      # https://gdam-api-v2-782971006568.us-west1.run.app/docs#/Index/index
      # fastfuels api to get species?
      # {
      #   "Lat": 48.1234,
      #   "Lon": -114.5678,
      #   "HT": 25.5
      # }
      "{"
        , "\"Lat\":" , scales::number(tree_lat, accuracy = 0.000001)
        , ",\"Lon\":" , scales::number(tree_lon, accuracy = 0.000001)
        , ",\"HT\":" , scales::number(field_tree_height_m, accuracy = 0.01)
      , "}"
      
    )
  ) %>% 
  dplyr::slice_sample(n=44) %>% 
  dplyr::slice_head(n=7) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(fastfuels_api)

# {"Lat":39.125813,"Lon":-105.115625,"HT":19.05}    # 39.125813,-105.115625
# {"Lat":39.125659,"Lon":-105.113507,"HT":18.90}    # 39.125659,-105.113507
# {"Lat":39.125840,"Lon":-105.112728,"HT":20.73}    # 39.125840,-105.112728
# {"Lat":39.126451,"Lon":-105.115426,"HT":2.44}   # 39.126451,-105.115426
# {"Lat":39.126394,"Lon":-105.115221,"HT":2.13}   # 39.126394,-105.115221
# {"Lat":39.124983,"Lon":-105.115187,"HT":15.54}    # 39.124983,-105.115187
# {"Lat":39.125319,"Lon":-105.115390,"HT":3.35}   # 39.125319,-105.115390


# {"Lat":39.127170,"Lon":-105.115781,"HT":20.12}
# {"Lat":39.127175,"Lon":-105.115772,"HT":10.06}
# {"Lat":39.127174,"Lon":-105.115760,"HT":7.01}"
# {"Lat":39.127124,"Lon":-105.115732,"HT":11.28}
# {"Lat":39.127110,"Lon":-105.115742,"HT":23.47}
# {"Lat":39.127099,"Lon":-105.115714,"HT":11.89}
# {"Lat":39.127073,"Lon":-105.115716,"HT":20.12}
```

We compiled a collection of validation datasets across diverse forest conditions for this specific analysis project. These datasets consist of coincident aerial point cloud data (including USGS 3DEP ALS, UAS-lidar, and UAS-SfM) paired with field-measured stem maps or plot inventories where species identities are known.

| Name | State | Forest Type | Point Cloud Data | Field Data |
| :--- | :--- | :--- | :--- | :--- |
| Penobscot Experimental Forest | Maine | northern mixed conifer | ALS | tree locations |
| Manitou Experimental Forest (N1) | Colorado | ponderosa pine | UAS-SfM | tree locations |
| Manitou Experimental Forest (Big Plot) | Colorado | western mixed conifer | UAS-SfM | tree locations |
| CSFS State Forest State Park | Colorado | lodgepole pine | UAS-lidar | plots |
| BLM Dawson Creek | Colorado | pinyon-juniper | UAS-SfM | tree locations |
| Ute Creek | Colorado | | UAS-SfM | plots |
| WFSI Fort Stewart | Georgia | southern yellow pine forest | UAS-lidar | plots |
| WFSI Elgin AFB | Florida | longleaf pine | UAS-lidar | plots |
| Wabikon Lake Forest Dynamics Plot (ForestGEO) | Wisconsin | northern mixed hardwood | ALS | tree locations |
| UC Santa Cruz Forest Ecology Research Plot (ForestGEO) | California | coastal mixed-evergreen & coast redwood | ALS | tree locations |
| Smithsonian Environmental Research Center (ForestGEO) | Maryland | mixed deciduous hardwood | ALS | tree locations |
| Smithsonian Conservation Biology Institute (ForestGEO) | Virginia | mid-atlantic deciduous | ALS | tree locations |

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # the cloud2trees
```

```{r pkg-ld, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Eglin Air Force Base (Florida)

Field data from this study site is from: 

* Andrew Hudak, Seth Bigelow, Nuria Sánchez-López, and Vanessa Niemczyk. 2025. Field measurements of coarse wood, duff, litter, and trees for the Eglin Air Force Base 2023 fuels campaign. Wildland Fire Science Initiative (WFSI) [https://doi.org/10.60594/W4RG6P](https://doi.org/10.60594/W4RG6P)

>Field measurements collected before and after 2023 study fires at Eglin Air Force Base (AFB) to document fuel loads (large diameter fuels, litter depth, duff depth) and trees in a 3-year rough in sandhill longleaf pine, and the effects of study fires on these variables. Fuel loads, which were measured along transects, were co-located with tree observations, which were measured within fixed-radius plots. These field observations were also co-located with three-dimensional fuel and laser scan plots to allow for leveraging of these other data.

>After the study fires in March 2023, trees within 10 m of plot center were identified to functional group (pine, hardwood or species), measured for diameter at breast height with a diameter tape, measured for distance from outer bark to plot center, and measured for azimuth from plot center to tree. Some post-processing was done using lidar scans to distinguish between evergreeen and deciduous hardwoods. Saplings were tallied by ordinal quadrant (up to 5 m from plot center). Tree measurements were not taken in unit L25. In unit G20, plots HH01, HH02, HH03, HH04, HH05 were sampled pre-fire but not post fire. In unit G25, plots HH105, HH95, PP16 were sampled pre-fire but not post fire.

the post-fire sampling protocol for trees was:

I. Determine if plot radius is 8 or 11.3 meters and record on datasheet
  a. If there are 4 or more trees within an 8-meter radius, then the plot radius is 8-meters.
  b. If there are 0 to 3 trees within an 8-meter radius, then increase the plot radius to 11.3-meters.
II. Sample all trees >10 cm DBH within the plot radius and over from plot center 
III. Sample all pine saplings ONLY >2.5 to 10 cm DBH by recording:
  a. Radius (4 or 5.6m):
    i. If tree sampling radius was 8-meters, then sapling sampling radius = 4-meters
    ii. If tree sampling radius was 11.3-meters, then sapling sampling radius = 5.6-meters
  b. Quadrant = NE, SE, SW, or NW
  c. Count = How many of the same species in that quadrant
  d. Species = Longleaf pine (Pinus palustris), Loblolly pine (Pinus taeda), Sand pine (Pinus clausa) or TBD


```{r}
wfsi_eglinafb_trees <-
  readr::read_csv(
    "../data/WFSI_EglinAFB/Eglin_Air_Force_Base_2023_fuels/data/trees.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  # dplyr::count(species)
  dplyr::filter(
    species != ""
    , !is.na(species)
    , !is.na(tree_x)
    , !is.na(tree_y)
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = 26916, remove = F)
# huh?
wfsi_eglinafb_trees %>% dplyr::glimpse()
```

check those field trees out in space

```{r}
# option to put satellite imagery as base layer of mapview maps
mapview::mapviewOptions(
  homebutton = FALSE
  , basemaps = c("Esri.WorldImagery","OpenStreetMap")
)
wfsi_eglinafb_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(unit,species)

wfsi_eglinafb_trees %>% 
  mapview::mapview(zcol = "species", layer.name="field trees") +

  mapview::mapview(zcol = "species", layer.name="field trees")


```



## Penobscot Experimental Forest (Maine)

Field data from this study site is from: 

* Kenefic, Laura S.; Allogio, Jeanette A.; Gerndt, Kathryn M.; Knapp, Rachel A.; Rogers, Nicole S.; Puhlick, Joshua J. 2025. Overstory tree and regeneration data from the "Silvicultural Effects on Composition, Structure, and Growth" study at Penobscot Experimental Forest. 3rd Edition. Fort Collins, CO: Forest Service Research Data Archive. [https://doi.org/10.2737/RDS-2012-0008-3](https://doi.org/10.2737/RDS-2012-0008-3)

>Data were collected on a series of permanent sample plots distributed across management units receiving various silvicultural treatments. Plot data are available in eight separate data sets. Overstory tree and sapling data include tree species, condition (e.g., merchantability status and cause of mortality), and diameter at breast height of living and dead trees, 1952 to 2023...Tree location data include the distance and azimuth from plot center of a subsample of trees, 2003 to 2023. Height and crown radius data include total height, height to crown base, and crown radius measurements for a subsample of live trees, 2003 to 2020 (heights) and 2003 to 2017 (radii)...Permanent sample plot location data include geospatial coordinates for permanent sample plots, 2018 and 2020...Future inventories are planned for this study indefinitely.

> 1/5th Acre Plots (were used to measure) Trees ≥ 4.5 inches DBH (with a) 52.7-foot plot radius

let's load in the plot data first to orient ourselves

```{r}
usfs_penobscotef_plots <- readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_PlotLocationsData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(~tolower(.x) %>% make.names()) %>% 
  dplyr::filter(
    geodoubt == "N"
    , active == 1
    , !is.na(utm_east)
    , !is.na(utm_north)
  ) %>% 
  dplyr::mutate(
    plot_id = 
      mu %>%
        stringr::str_squish() %>% 
        stringr::str_c("-", as.character(plot))
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("utm_east", "utm_north"), crs = 26919, remove = F) %>% 
  dplyr::mutate(
    plot_area_ac = 1/5
    # 1 ha = 2.47105 acres
    , plot_area_ha = plot_area_ac/2.47105
    , plot_area_m2 = plot_area_ha*10000
    , plot_radius_meters = sqrt(plot_area_m2 / pi)    
  ) %>% 
  sf::st_buffer(.$plot_radius_meters)
# huh?
# usfs_penobscotef_plots %>% dplyr::count(projection)
usfs_penobscotef_plots %>% dplyr::glimpse()
```

we did a lot of data cleaning in there which we won't describe (you can read the code if you're interested). let's look at the plot locations...

```{r}
usfs_penobscotef_plots %>% 
  dplyr::select(plot_id,mu,plot,plot_area_ha) %>% 
  mapview::mapview(zcol = "mu", col.regions = sample(viridis::turbo(100)), layer.name = "plots", legend = F)
```






ForestGEO SCBI

```{r}
boundary_scbi_temp <- sf::st_read("../data/ForestGEO_SCBI/ForestGEO_SCBI_shapefiles/ForestGEO_grid_outline.shp") %>% 
  sf::st_union() 
boundary_scbi_temp %>% 
  mapview::mapview(layer.name = "SCBI")
boundary_scbi_temp %>% sf::st_centroid() %>% sf::st_transform(4326) %>% sf::st_coordinates()
# 38.893480, -78.145380
```






The `cloud2trees` package can be installed by following the directions listed in the [README file on GitHub](https://github.com/georgewoolsey/cloud2trees). If one is still experiencing difficulties installing the package, see the [setup guide](https://georgewoolsey.github.io/cloud2trees/articles/cloud2trees-setup.html) which details how to install the package using a virgin R instance.

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get cloud2trees
remotes::install_github(repo = "georgewoolsey/cloud2trees", upgrade = F)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```
