# USFS Manitou Experimental Forest (N1)

Now that we have reviewed the process and evaluation method and set up functions to standardize and streamline our data processing, let's do it for another site

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Field Data{#01_fld_dta}

Field data from this study site is from: a stem map including all trees with a height of at least 2 m.

the sampling protocol for trees was: measure all trees at least 2 m in height for height, DBH, and species and record distance and azimuth from gridded plot centers over square study area.

we'll keep trees over 

```{r}
usfs_manitouef_n1_trees <-
  # we have used this data before
  readxl::read_excel("c:/data/usfs/uas_lidar_n1/data/N1_2024_Field_Trees.xlsx") %>% 
  dplyr::rename_with(
    ~ tolower(.x) %>% 
      stringr::str_squish() %>% 
      make.names() %>% 
      stringr::str_replace_all( "\\.{2,}", ".") %>% 
      stringr::str_remove( "\\.$") %>% 
      stringr::str_replace_all( "\\.", "_")
  ) %>% 
  sf::st_as_sf(
    coords = c("easting", "northing")
    , crs = 26913
    , remove=F
  ) %>% 
  dplyr::rename(
    # field_dbh_cm = dbh_cm
    # , field_tree_height_m = ht_m
    dbh_in = dbh_24
    , tree_height_ft = ht_24
  ) %>% 
  # dplyr::select(dbh_in, tree_height_ft) %>% summary()
  dplyr::mutate(
    dbh_cm = dbh_in/0.394
    , tree_height_m = tree_height_ft/3.281
  ) %>% 
  # dplyr::select(dbh_cm, tree_height_m) %>% summary()
  sf::st_set_geometry("geometry") %>% 
  dplyr::mutate(
    tree_x = sf::st_coordinates(geometry)[,1] #lon
    , tree_y = sf::st_coordinates(geometry)[,2] #lat
  ) %>% 
  # make a plot_id that is unique for all so we don't have to join by multiple columns
  # and for all sites we'll just use a plot_id
  dplyr::mutate(
    # plot_id = xxxxxx
    # also, each tree needs a treeID
    treeID = dplyr::row_number() %>% as.character()
    # also, we need a species_symbol to match with the FIADatabase
    , species_symbol = toupper(spp) %>% stringr::str_remove_all("\\s")
  ) %>% 
  # clean
  dplyr::filter(
    !is.na(dbh_cm)
    & !is.na(tree_height_m)
    & sf::st_is_valid(geometry)
    & tree_height_m >= 2
    & !is.na(species_symbol)
    & species_symbol!=""
  )
# huh?
# usfs_manitouef_n1_trees %>% dplyr::glimpse()
# usfs_manitouef_n1_trees %>% dplyr::select(dbh_cm) %>% summary()
```

let's check those field trees out to orient ourselves to the location

```{r}
usfs_manitouef_n1_trees %>% 
  mapview::mapview(
    zcol = "species_symbol"
    , layer.name="USFS Manitou EF (N1): field trees"
    , popup=F
  )
# usfs_manitouef_n1_trees %>%
#   sf::st_union() %>%
#   sf::st_bbox() %>%
#   sf::st_as_sfc() %>%
#   sf::st_centroid() %>%
#   sf::st_transform(crs=4326) %>%
#   sf::st_coordinates()
# # 30.53834,-86.7296
```

what is the distribution of species?

```{r}
usfs_manitouef_n1_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species_symbol) %>% 
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

let's load in the singular plot location (it's just one large study site that we'll call a plot for this analysis)

```{r}
usfs_manitouef_n1_plots <-
   sf::st_read(
    "c:/data/usfs/uas_lidar_n1/data/N1_Boundary/N1_Boundary.shp"
    # "c:/data/usfs/uas_sfm_tree_detection/data/field_validation/field_data_Boundary.shp"
    , quiet = T
  ) %>% 
  dplyr::rename_with(tolower) %>% 
  dplyr::filter(
    name=="N1"
    # site=="N1"
  ) %>% 
  # converts proj to match trees and horizontal units to m
  sf::st_transform(sf::st_crs(usfs_manitouef_n1_trees)) %>% 
  dplyr::mutate(
    plot_area_m2 = sf::st_area(.) %>% as.numeric()
    , plot_id = name
  )

# lets keep only trees within the plot boundar
usfs_manitouef_n1_trees <- 
  usfs_manitouef_n1_trees %>% 
  sf::st_intersection(
    usfs_manitouef_n1_plots %>% dplyr::select(plot_id)
  )
  
# usfs_manitouef_n1_trees %>% dplyr::glimpse()

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(data_site_nm)
ref_plots_list[[nm_temp]] <- usfs_manitouef_n1_plots
ref_trees_list[[nm_temp]] <- usfs_manitouef_n1_trees
# ref_plots_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Point Cloud Data{#02_ptcld_dta}

data was collected over the study site via a UAS-lidar flight at an altitude of 122 m (400 ft) and speed of 8.9 meters per second (20 miles per hour)

let's check out the raw point cloud data

```{r}
usfs_manitouef_n1_las_f_temp <- "c:/data/usfs/uas_lidar_n1/data/N1_400AGL_20MPH_TFOFF/"
usfs_manitouef_n1_las_ctg_temp <- lidR::readLAScatalog(usfs_manitouef_n1_las_f_temp)
usfs_manitouef_n1_las_ctg_temp
# usfs_manitouef_n1_las_ctg_temp@data %>% dplyr::glimpse()
```

let's check out the lidar data coverage compared to the field plot data

```{r}
usfs_manitouef_n1_las_ctg_temp@data$geometry %>% 
  sf::st_union() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(data = "UAS-lidar data extent") %>% 
mapview::mapview(
 zcol = "data"
  , layer.name="USFS Manitou EF (N1): UAS-lidar data"
  , popup=F
) +
mapview::mapview(
  usfs_manitouef_n1_trees
  , zcol = "species_symbol"
  , layer.name="USFS Manitou EF (N1): field trees"
  , popup=F
)
```

use `cloud2trees` to process the raw point cloud data to generate a CHM for use in tree detection

```{r}
################################################
# input/output data str
################################################
# outdir
c2t_output_dir_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(data_dir)
c2t_process_dir_temp <- file.path(c2t_output_dir_temp, "point_cloud_processing_delivery")
# timer file
c2t_tracking_fnm_temp <- file.path(c2t_process_dir_temp, "processed_tracking_data.csv")
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    c2t_process_dir = ifelse(data_dir==c2t_output_dir_temp, c2t_process_dir_temp, c2t_process_dir)
    , c2t_tracking_fnm = ifelse(data_dir==c2t_output_dir_temp, c2t_tracking_fnm_temp, c2t_tracking_fnm)
  )
# project_data_df %>% print(n=nrow(project_data_df))

# get point cloud density and raster resolutions
get_ptcld_density_chm_res_temp <- get_ptcld_density_chm_res_fn(usfs_manitouef_n1_las_ctg_temp)
# get_ptcld_density_chm_res_temp

# fnm
chm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("chm_",get_ptcld_density_chm_res_temp$chm_res_m,"m.tif") )
dtm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("dtm_",get_ptcld_density_chm_res_temp$dtm_res_m,"m.tif") )
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    las_pts_m2 = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$pts_m2, las_pts_m2) # las_pts_m2
    , chm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$chm_res_m, chm_res_m) # chm_res_m
    , chm_fnm = ifelse(data_dir==c2t_output_dir_temp, chm_fnm_temp, chm_fnm) # chm_fnm
    , dtm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$dtm_res_m, dtm_res_m) # dtm_res_m
    , dtm_fnm = ifelse(data_dir==c2t_output_dir_temp, dtm_fnm_temp, dtm_fnm) # dtm_fnm
  )

# project_data_df %>% dplyr::filter(data_dir==c2t_output_dir_temp) %>% dplyr::glimpse()

##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists(chm_fnm_temp)
  || !file.exists(dtm_fnm_temp)
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir_temp
    , input_las_dir = usfs_manitouef_n1_las_f_temp
    , accuracy_level = 2
    , keep_intrmdt = F
    , dtm_res_m = get_ptcld_density_chm_res_temp$dtm_res_m # (1/sqrt(pts_m2_temp))*2
    , chm_res_m = get_ptcld_density_chm_res_temp$chm_res_m # 1/sqrt(pts_m2_temp)
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm_temp
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( dtm_fnm_temp )
  chm_temp <- terra::rast( chm_fnm_temp )
  
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

we'll set the minimum CHM resolution to `r min_chm_res` m to match the `cloud2trees` default. if the automatically-determined CHM resolution based on the point cloud density used in the `cloud2trees::cloud2raster()` was lower than this level, we'll aggregate the raster using `terra::aggregate()` and `terra::resample()` to get the exact target resolution. if the determined CHM resolution was higher than this level due to low point density, we will use the CHM as-is since this is already the finest resolution possible with the point cloud density.

let's aggregate the CHM for this site and put it into a list of CHM rasters that we can map over in all of the processing

```{r}
agg_chm_rast_temp <- agg_chm_rast_fn(
  chm = cloud2raster_ans$chm_rast
  , res = min_chm_res
  , dir = 
    project_data_df %>% 
    dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
    dplyr::pull(c2t_process_dir)
)
# agg_chm_rast_temp
# agg_chm_rast_temp %>% terra::clamp(lower = 1.37, values = F) %>% terra::plot()

# add to the chm list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(data_site_nm)
chm_list[[nm_temp]] <- agg_chm_rast_temp
# chm_list
```

let's check out the CHM

```{r}
# plot to check out the fine-resolution CHM raster
chm_list$USFS_ManitouEF_N1 %>% 
  terra::plot(
    col = viridis::plasma(n=100)
    , axes=F
    , plg = list(
      title = "CHM (m)"
      , cex = 0.8
      , title.cex = 0.9
      , shrink = 0.6 
    )
  )
```

and convert the reference trees and plot data to the point cloud data projection

```{r}
ref_plots_list$USFS_ManitouEF_N1 <- ref_plots_list$USFS_ManitouEF_N1 %>% sf::st_transform(terra::crs(chm_list$USFS_ManitouEF_N1))
ref_trees_list$USFS_ManitouEF_N1 <- ref_trees_list$USFS_ManitouEF_N1 %>% 
  sf::st_transform(terra::crs(chm_list$USFS_ManitouEF_N1)) %>% 
  dplyr::mutate(
    tree_x = sf::st_coordinates(geometry)[,1] #lon
    , tree_y = sf::st_coordinates(geometry)[,2] #lat
  )
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(usfs_manitouef_n1_plots, usfs_manitouef_n1_trees)
gc()
```

### Tree Detection{#03_itd}

`cloud2trees` implements a CHM-based individual tree detection method accessible via `cloud2trees::raster2trees()`. this method requires a window function used to set the search radius used to find local maxima in the CHM. These local maxima represent individual tree top points and are subsequently used to delineate individual tree crowns via a marker-controlled watershed segmentation

because the objective of this study is to evaluate a species/genera classification method, we just want the best ITD results we can get based on the field reference data. We are not interested in evaluating tree segmentation here, that has already been done across varied input data and forest types.

use our function to clip the CHM to the plot plus a buffer and run `cloud2trees::raster2trees()` for each window function. the return should be a spatial tree list unique by function.

get the individual tree predictions for each window function for the area within the field reference plots with a buffer

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, ws_pred_trees_fnm)
  )

if(!file.exists(fnm_temp)){
  usfs_manitouef_n1_ws_pred_trees <-
    1:nrow(ref_plots_list$USFS_ManitouEF_N1) %>%
    purrr::map(
      \(x)
      ws_testtest_fn(
        chm = chm_list$USFS_ManitouEF_N1
        , buffer = (competition_radius_m+3) # since we only have one plot we're just going to go for it
        , plot_sf =
          ref_plots_list$USFS_ManitouEF_N1 %>%
          dplyr::ungroup() %>%
          dplyr::slice(x) %>%
          sf::st_union()
        # a named list of functions ex: cloud2trees::itd_ws_functions()
        , ws_fn_list = get_ws_fn_list_ans # cloud2trees::itd_ws_functions() #
      )
    ) %>%
    purrr::list_rbind(names_to = "plot_dta_idx")
  
  # add plot id
  usfs_manitouef_n1_ws_pred_trees <- 
    usfs_manitouef_n1_ws_pred_trees %>% 
    dplyr::left_join(
      ref_plots_list$USFS_ManitouEF_N1 %>% 
        dplyr::mutate(plot_dta_idx=dplyr::row_number()) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(plot_dta_idx,plot_id)
      , by = "plot_dta_idx"
    ) %>% 
    sf::st_set_geometry("geometry")
  
  # usfs_manitouef_n1_ws_pred_trees %>% sf::st_crs()
  # usfs_manitouef_n1_ws_pred_trees %>% dplyr::glimpse()
  # usfs_manitouef_n1_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_fn_nm)
  
  # get dbh
  usfs_manitouef_n1_ws_pred_trees_dbh_temp <- 
    usfs_manitouef_n1_ws_pred_trees %>% 
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number() %>% as.character()
    ) %>% 
    cloud2trees::trees_dbh()
  # usfs_manitouef_n1_ws_pred_trees_dbh_temp %>% dplyr::glimpse()
  # add to trees
  usfs_manitouef_n1_ws_pred_trees <-
    usfs_manitouef_n1_ws_pred_trees %>% 
    dplyr::mutate(
      join_id = dplyr::row_number() %>% as.character()
    ) %>% 
    dplyr::left_join(
      usfs_manitouef_n1_ws_pred_trees_dbh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          join_id = as.character(treeID)
        ) %>% 
        dplyr::select(
          join_id, tidyselect::starts_with("fia_est_dbh"), dbh_cm
        )
      , by = "join_id"
    )
  # usfs_manitouef_n1_ws_pred_trees %>% dplyr::glimpse()
  
  # # plot real quick
  # usfs_manitouef_n1_ws_pred_trees %>% 
  #   sf::st_drop_geometry() %>% 
  #   dplyr::slice_sample(n = 9999) %>% 
  #   ggplot2::ggplot(mapping = ggplot2::aes(x=tree_height_m,y=dbh_cm)) + 
  #   ggplot2::geom_point() +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_lower),color="gray") +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_upper),color="gray")

  # save
  sf::st_write(usfs_manitouef_n1_ws_pred_trees, dsn = fnm_temp, quiet = T, append = F)
}else{
  usfs_manitouef_n1_ws_pred_trees <- sf::st_read(fnm_temp, quiet = T)
}
usfs_manitouef_n1_ws_pred_trees <- usfs_manitouef_n1_ws_pred_trees %>% 
  dplyr::mutate(
    ws_type = 
      ws_fn_nm %>% 
      # removes all numeric digits
      stringr::str_remove_all("[0-9]+") %>% 
      stringr::str_replace_all("_"," ") %>% 
      stringr::str_squish() %>% 
      as.factor() %>% 
      forcats::fct_inorder(ordered = T)
    , treeID = as.character(treeID)
  )
# usfs_manitouef_n1_ws_pred_trees %>% dplyr::glimpse()
# usfs_manitouef_n1_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_type)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's check out one example field plot and a few different window function tree detection and crown delineation results

```{r, fig.height=10}
# plot
p_temp <- ref_plots_list$USFS_ManitouEF_N1 %>% dplyr::slice(1)

if(as.numeric(sf::st_area(p_temp))>2000){
  p_temp <- p_temp %>% 
    sf::st_centroid() %>% 
    sf::st_buffer(sqrt(2000)/2) %>% 
    sf::st_bbox() %>% 
    sf::st_as_sfc() %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(plot_id = p_temp$plot_id[1])
}
# some trees
t_temp <- 
  usfs_manitouef_n1_ws_pred_trees %>% 
  dplyr::filter(plot_id==p_temp$plot_id[1]) %>% 
  sf::st_intersection(p_temp) %>% 
  dplyr::inner_join(
    usfs_manitouef_n1_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm,ws_type) %>% 
      dplyr::slice_sample(n = 3, by = ws_type) %>% 
      dplyr::select(ws_fn_nm)
    , by = "ws_fn_nm"
  )
chm_temp <- chm_list$USFS_ManitouEF_N1 %>% 
  terra::crop(
    t_temp %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      terra::vect()
    , mask = T
  )
# terra::plot(chm_temp)
plt_temp <-
  chm_temp %>% 
  terra::clamp(lower = 1.37, values = F) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x=x,y=y,fill=f)
    , color = NA
    , lwd = 0
  ) +
  ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent", name = "CHM (m)") +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top")
plt_temp +
  ggplot2::geom_sf(
    data = p_temp %>% sf::st_transform(terra::crs(chm_temp))
    , fill = NA
    , color = "black"
    , lwd = 0.8
  ) + 
  ggplot2::geom_sf(
    data = t_temp
    , mapping = ggplot2::aes(color = ws_type)
    , fill = NA
    , lwd = 0.7
  ) +
  ggplot2::scale_color_viridis_d() +
  ggplot2::facet_wrap(facets = dplyr::vars(ws_fn_nm), ncol = 3) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold")
    , strip.background = ggplot2::element_rect(color = "black")
    , panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  ggplot2::guides(
    color = "none"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now we need a function to perform instance matching of the predictions and reference data. to keep things simple, since all the reference field data includes point locations, we'll perform matching based on a spatial intersection of these points with the predicted tree crowns. if multiple tree points fall within a single crown, we'll select the tree that is closest in terms of size based on whatever field data is available (height or DBH). One predicted tree will only be matched with one field reference tree and vice-versa.

```{r}
##################################################################
##################################################################
##################################################################
# plot_iter_instance_match_fn
# loop over plots and ws functions for instance match
##################################################################
##################################################################
##################################################################
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_instance_match.csv")
if(!file.exists(fnm_temp)){
  usfs_manitouef_n1_ws_instance_match <- 
    plot_iter_instance_match_fn(
      ref_plots = ref_plots_list$USFS_ManitouEF_N1
      , ref_pts = ref_trees_list$USFS_ManitouEF_N1
      , pred_polys = usfs_manitouef_n1_ws_pred_trees
      , min_ht_m = 2
      , min_dbh_cm = NA
    )
  readr::write_csv(usfs_manitouef_n1_ws_instance_match, file = fnm_temp, append = F, progress = F)
}else{
  usfs_manitouef_n1_ws_instance_match <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# usfs_manitouef_n1_ws_instance_match %>% dplyr::glimpse()
##################################################################
##################################################################
##################################################################
# agg_ground_truth_match
# loop over plots and ws functions for accuracy eval
##################################################################
##################################################################
##################################################################
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"agg_ground_truth_match_ans.csv")
if(!file.exists(fnm_temp)){
  # ?purrr::pmap
  # A data frame is an important special case of .l. It will cause .f to be called once for each row.
  usfs_manitouef_n1_agg_ground_truth_match_ans <- 
    usfs_manitouef_n1_ws_instance_match %>% 
    dplyr::distinct(plot_id,ws_fn_nm) %>% 
    purrr::pmap(function(plot_id, ws_fn_nm) {
      # the unique row data
      d <- dplyr::tibble(plot_id = plot_id, ws_fn_nm = ws_fn_nm)
      # filter full set based on the data
      j <- 
        usfs_manitouef_n1_ws_instance_match %>% 
        dplyr::inner_join(
          d   
          , by = dplyr::join_by("plot_id","ws_fn_nm")
        )
      # agg_ground_truth_match
      d %>% 
        dplyr::bind_cols(agg_ground_truth_match(j))
      
    }) %>% 
    list_rbind()
  readr::write_csv(usfs_manitouef_n1_agg_ground_truth_match_ans, file = fnm_temp, append = F, progress = F)
}else{
  usfs_manitouef_n1_agg_ground_truth_match_ans <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# wfsi_eglinafb_agg_ground_truth_match_ans %>% dplyr::glimpse()
```

aggregate the plot level data to the window function to get the answer we've been searching for

```{r}
# usfs_manitouef_n1_agg_ground_truth_match_ans %>% 
#   dplyr::count(ws_fn_nm) %>% 
#   dplyr::count(n)
agg_temp <- 
  usfs_manitouef_n1_agg_ground_truth_match_ans %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::summarize(
    n_plots = dplyr::n()
    , dplyr::across(
      .cols = c(precision,recall,f_score,tidyselect::ends_with("_rate"))
      , list( 
        mean= ~ mean(.x,na.rm=T)
        , median= ~ median(.x,na.rm=T)
      )
    )
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::left_join(
    usfs_manitouef_n1_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::mutate(
    f_score_rank = dplyr::min_rank(-f_score_mean)
    , ws_fn_nm_fact = forcats::fct_reorder(ws_fn_nm, f_score_rank)
    # , dplyr::across(
    #   .cols = dplyr::where(is.numeric)
    #   , ~ round(.x,2) %>% as.character() %>% as.numeric()
    # )
  ) %>% 
  dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean))
# agg_temp %>% dplyr::glimpse()
```

let's check out the plot-level distribution of F-score for the top window functions (remember this particular data has one "plot", a stem map)

```{r}
usfs_manitouef_n1_agg_ground_truth_match_ans %>% 
  # add agg data
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::filter(f_score_rank<=12) %>% 
      dplyr::select(ws_fn_nm, ws_fn_nm_fact, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::arrange(ws_fn_nm_fact) %>% 
  dplyr::mutate(ws_fn_nm_fact=forcats::fct_rev(ws_fn_nm_fact)) %>% 
  # dplyr::pull(ws_fn_nm) %>% levels()
  # View()
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = f_score, y = ws_fn_nm_fact, fill=ws_type)
  ) +
  ggplot2::geom_boxplot(
    width = 0.4
    , alpha=0.9
    , outliers = F
    , color = "black"
  ) +
  ggplot2::geom_point(color = "gray77") +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(x="F-score", y="") +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , legend.direction  = "horizontal"
  )
```

we can print the raw window function definition for those that achieved the highest F-score

```{r}
agg_temp %>% 
  dplyr::filter(f_score_rank<=7) %>% 
  dplyr::arrange(f_score_rank) %>% 
  dplyr::pull(ws_fn_nm) %>% 
  purrr::set_names() %>% 
  # purrr::map(\(nm) purrr::pluck(get_ws_fn_list_ans, nm))
  purrr::map(function(nm){
    fn <- purrr::pluck(get_ws_fn_list_ans, nm)
    dplyr::tibble(
      ws_fn_nm = nm
      , def = paste(format(fn), collapse = "\n")
    )
  }) %>%
  purrr::list_rbind() %>% 
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::select(ws_fn_nm, f_score_mean, recall_mean, precision_mean) %>% 
      dplyr::mutate(
        dplyr::across(
          dplyr::where(is.numeric)
          , ~ scales::percent(.x,accuracy = 0.1)
        )
      )
  ) %>% 
  dplyr::relocate(def,.after = dplyr::everything()) %>% 
  kableExtra::kbl(
    caption = "WFSI Eglin Air Force Base top ITD window functions"
    , col.names = c(
      "function<br>name"
      , "F-score<br>(mean)"
      , "recall<br>(mean)"
      , "precision<br>(mean)"
      , "function<br>definition"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling(font_size = 12, bootstrap_options = c("responsive", "condensed")) %>% # 
  # make r line breaks (\n) and use monospace font for code
  kableExtra::column_spec(5, monospace = T, extra_css = "white-space: pre;")
```

nice, we'll select the best non-constant window function for use in evaluation of the species accuracy. if we were to use the constant functions, even if they are the best performing, we would likely be overfitting to this specific data set. in addition to being more in-line with real-world tree height-to-crown relationships, the variable window functions are likely more broadly applicable

```{r}
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    best_ws_fn_nm = ifelse(
      tolower(data_site_nm)=="usfs_manitouef_n1"
      , agg_temp %>% 
        dplyr::filter(ws_type!="constant") %>% 
        dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean)) %>% 
        dplyr::slice(1) %>% 
        dplyr::pull(ws_fn_nm)
      , best_ws_fn_nm
    )
  )
# project_data_df %>% dplyr::glimpse()
```

full CHM data processing

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"best_ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    best_ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, best_ws_pred_trees_fnm)
  )

###############################################################################################
###############################################################################################
### raster2trees that best ws fn
###############################################################################################
###############################################################################################
if(!file.exists(fnm_temp)){
  
  ws_temp <- project_data_df %>% 
    dplyr::filter(c2t_process_dir==c2t_process_dir_temp) %>% 
    dplyr::slice(1) %>% 
    dplyr::pull(best_ws_fn_nm)
  
  pred_trees_temp <- 
    usfs_manitouef_n1_ws_pred_trees %>% 
    dplyr::filter(ws_fn_nm==ws_temp) %>% 
    # this is because we did tree detection by plot
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number() %>% as.character()
    )
  
  # get rh
  pred_trees_temp_rh_temp <- 
    # have to do this by plot so that all trees are not included
    unique(pred_trees_temp$plot_dta_idx) %>% 
    purrr::map(
      \(idx)
      pred_trees_temp %>% 
        dplyr::filter(plot_dta_idx==idx) %>% 
        get_relative_height_fn(radius = competition_radius_m) 
    ) %>% 
    purrr::list_rbind()
  # pred_trees_temp_rh_temp %>% dplyr::glimpse()
  # pred_trees_temp_rh_temp %>% 
  #   ggplot2::ggplot() + 
  #   ggplot2::geom_point(ggplot2::aes(x = tree_height_m, y=rel_tree_height_m)) + 
  #   ggplot2::facet_wrap(facets=dplyr::vars(plot_id))
  
  # add to trees
  pred_trees_temp <- 
    pred_trees_temp %>% 
    dplyr::left_join(
      pred_trees_temp_rh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(
          plot_dta_idx,treeID,rel_tree_height_m
        )
      , by = dplyr::join_by(plot_dta_idx,treeID)
    )
  # pred_trees_temp %>% dplyr::glimpse()
  
  # save
  sf::st_write(pred_trees_temp, dsn = fnm_temp, quiet = T, append = F)
}else{
  pred_trees_temp <- sf::st_read(fnm_temp, quiet = T)
}
# pred_trees_temp %>% dplyr::glimpse()

###############################################################################################
###############################################################################################
### now we need to do the matching 
# again to only keep trees in plots for individual tree species comparisons
###############################################################################################
###############################################################################################

im_temp <- 
  plot_iter_instance_match_fn(
    ref_plots = ref_plots_list$USFS_ManitouEF_N1
    , ref_pts = ref_trees_list$USFS_ManitouEF_N1
    , pred_polys = pred_trees_temp
    , min_ht_m = 2
    , min_dbh_cm = NA
  )

# im_temp %>% dplyr::glimpse()

# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
  dplyr::pull(data_site_nm)
pred_ref_tree_match_list[[nm_temp]] <- im_temp

###############################################################################################
###############################################################################################
### save to the pred trees
###############################################################################################
###############################################################################################
# let's also save the plot-level ITD predictions for this ws

pred_trees_list[[nm_temp]] <-
  pred_trees_temp %>% 
  # which ws
  dplyr::filter(
    ws_fn_nm %in% (
      project_data_df %>% 
      dplyr::filter(tolower(data_site_nm)=="usfs_manitouef_n1") %>% 
      dplyr::pull(best_ws_fn_nm)
    )
  ) %>% 
  # filter for trees used in instance matching 
    # which included thresholds used for measuring field trees 
    # and plot spatial alignment
  dplyr::inner_join(
    pred_ref_tree_match_list[[nm_temp]] %>% 
      dplyr::select(plot_id,ws_fn_nm,pred_treeID) %>% 
      dplyr::filter(!is.na(pred_treeID)) %>% 
      dplyr::rename(treeID=pred_treeID)
    , by = dplyr::join_by(plot_id,ws_fn_nm,treeID)
  )
# pred_trees_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(list = ls()[grep("^usfs_manitou",ls())])
remove(cloud2raster_ans)
gc()
```

### Species/Genus Prediction Method{#04_species}

for some context of the primary objective of this analysis, here is what we said in the `cloud2trees` manuscript (a draft of the manuscript which may be different than what was actually published): 

>However, taxonomic differentiation using airborne lidar alone remains one of the most technically challenging aspects of forest inventory workflows (White et al., 2016, 2025; Fassnacht et al., 2024). In practice, genus-level distinction is often more realistic when relying on remote sensing and supplemental spatial data ([Vauhkonen et al., 2014b](https://doi.org/10.1007/978-94-017-8663-8_7)). As a practical, computationally efficient approach, future research could evaluate predicting tree-level species or genus for potential integration into cloud2trees by leveraging supplemental spatial datasets. Testing could evaluate either using tree lists from FIA plots linked through a spatial overlay with TreeMap (Houtman et al., 2025) to train a taxonomic classification model based on structural and competition metrics shared between FIA and cloud2trees tree lists, or by proportionally assigning species using BigMap species-level proportional biomass raster data (Wilson et al., 2012, 2018). Incorporating finer-scale taxonomic classification than the currently implemented FIA forest type groups into the cloud2trees tree list will enable the integration of detailed biomass equations such as the National Scale Volume Biomass system (NSVB; Westfall et al., 2024) and other allometries that require species- or genera-level information (e.g., Réjou‐Méchain et al., 2017; Gonzalez‐Akre et al., 2022).

`cloud2trees` does not currently include functionality to attach species to it's tree list. there are a few paths forward for performing species-specific workflows with the `cloud2trees` tree list:

the proposed method which we are evaluating presently, will predict tree-level species using the TreeMap imputed FIA plot tree list to build a species classification model based on height and tree location to attach tree species to our tree list. The model will inherently account for the proportional distribution of different species. 

early testing of the proposed method used a  bayesian hierarchical model grouped by imputed FIA plot identifier that allows the intercept and the slope of height to vary by plot took quite a while to run for a relatively small area (~20-30 mins). as such, we are going to use a random forest (RF) classifier to enable more computationally efficient classification. `cloud2trees` already implements random forest models to model missing HMD and CBH values that were not successfully extracted from the point cloud and uses a model averaging method that subsamples the training data (to reduce model fit times), estimates the model several times with different subsamples, then model averages the predictions. Because the species is not a metric variable that can be averaged, we'll use the same method but employ a model majority voting system instead. we can do this with some of the `cloud2trees` built-in functionality.

ultimately, if the proposed species/genera classification method achieves decent accuracy results, it could be integrated into the `cloud2trees` framework as `cloud2trees::trees_species()`, for example. this function would take an input spatial tree list and attach the predicted species/genera

#### Training Data

load in the TreeMap data and clip to our trees as in `cloud2trees::trees_dbh()`

```{r}
# let's make a function to standardize this process since we'll be using it in more than one `cloud2trees::trees_*` function now
treemap_data_loader <- function(sf_poly, input_treemap_dir = NULL) {
  ####################################################################
  # check external data
  ####################################################################
    # find external data
    find_ext_data_ans <- cloud2trees:::find_ext_data(
      input_treemap_dir = input_treemap_dir
    )
    # if can't find external treemap data
    if(is.null(find_ext_data_ans$treemap_dir)){
      stop(paste0(
        "Treemap data has not been downloaded to package contents. Use `get_treemap()` first."
        , "\nIf you supplied a value to the `input_treemap_dir` parameter check that directory for data."
      ))
    }

    #treemap files
    treemap_data_finder_ans <- cloud2trees:::treemap_data_finder(find_ext_data_ans$treemap_dir)
    # treemap_data_finder_ans$which_treemap
    # treemap_data_finder_ans$treemap_trees
    # treemap_data_finder_ans$treemap_rast
  ####################################################################
  # check polygons
  ####################################################################
    if(inherits(sf_poly, "sfc")){sf_poly <- sf::st_as_sf(sf_poly)}
    if(!inherits(sf_poly, "sf")){stop("must include `sf` data object in 'sf_poly'")}
    if( !all(sf::st_is(sf_poly, type = c("POLYGON", "MULTIPOLYGON"))) ){
      stop(paste0(
        "`sf_poly` data must be an `sf` class object with POLYGON geometry (see [sf::st_geometry_type()])"
      ))
    }
    sf_poly <- sf_poly %>% dplyr::ungroup() %>% sf::st_union()
    
  ####################################################################
  # read in treemap data
  ####################################################################
    # read in treemap data
    # read in treemap (no memory is taken)
    treemap_rast <- terra::rast(treemap_data_finder_ans$treemap_rast)

    # check study boundary against the raster
    # the resulting matrix will have a true value where an intersection exists
    intersection_result <- terra::relate(
      x = sf_poly %>%
        sf::st_transform(terra::crs(treemap_rast)) %>%
        terra::vect()
      , y = treemap_rast
      , relation = "intersects"
    )

    if(!any(intersection_result)) {
      stop(paste0(
        "The search area does not overlap with an area within CONUS."
        , "\n.... If provided data in `study_boundary` ensure that it overlaps with CONUS"
        , "\n.... If did not provide data in `study_boundary` ensure that the `tree_list` data overlaps with CONUS"
      ))
    }
    
    # calculate buffer based on the expected window size used for any terra::focal operations
    ws_temp <- 5
    buffer_temp <- ((ws_temp-1)/2)*terra::res(treemap_rast)[1]
    
    ### filter treemap based on las...rast now in memory
    treemap_rast <- treemap_rast %>%
      terra::crop(
        sf_poly %>%
          sf::st_transform(terra::crs(treemap_rast)) %>%
          sf::st_buffer(buffer_temp) %>% 
          terra::vect()
      ) %>%
      terra::subset(1)

    # ggplot(treemap_rast %>% as.data.frame(xy=T) %>% rename(f=3)) +
    #   geom_tile(aes(x=x,y=y,fill=as.factor(f))) +
    #   scale_fill_viridis_d(option = "turbo") +
    #   theme_light() + theme(legend.position = "none")

    # check for all NA's
    notna_cells <- any( terra::global(treemap_rast, fun = "anynotNA") )
    if( !notna_cells ){
      stop(paste0(
        "The search area does not overlap with a forested area within CONUS. Cannot estimate DBH."
        , "\n ... try expanding the `boundary_buffer` ?"
      ))
    }
    
    ### get weights for weighting each tree in the population models
    # treemap id = tm_id for linking to tabular data
    tm_id_weight_temp <- treemap_rast %>% ## works
      terra::values() %>%
      table() %>%
      dplyr::as_tibble() %>%
      dplyr::rename(tm_id=1,tree_weight=n) %>%
      dplyr::mutate(tm_id = cloud2trees:::as_character_safe(tm_id))
    # tm_id_weight_temp
    
    ############################################################################
    ### get a stability weight of the imputed plot id's
    ############################################################################
    
    # when we consider the tree composition of an entire forest stand we expect TreeMap 
    # to generally get the main species and forest type correct 
    # but for each 30m raster cell (i.e. the plot location that we get the X and Y coordinates for the tree list used for training from) 
    # we expect it to be less accurate with a cell potentially being a completely incorrect representation of the 
    # actual ground setting. As such, we we’ll weight the individual tree records based on the stability of 
    # the imputed plot id’s looking over a window surrounding the focal cell. Specifically, we’ll calculate 
    # the proportion of surrounding cells (5x5) that share the same imputed plot id as the focal cell 
    # and use this weight as a measure of TreeMap’s predicted stability to place less weight 
    # on plots that occur infrequently and more weight on plots that occur frequently.
    treemap_id_rast <- 
      treemap_rast %>% 
      terra::catalyze() %>% # to numeric
      terra::subset(1)
    
    treemap_prop_rast_temp <- 
      treemap_id_rast %>% 
      terra::focal(
        # # 5x5 = 2.25 ha; 3x3 = 0.81 ha
        w = ws_temp
        , fun = match_focal_cells_fn # add to cloud2trees:::
        # , na.poicy = "omit"
        , na.rm = T
      ) %>% 
      setNames("stability_weight")
    # treemap_prop_rast_temp %>% terra::hist()
    
    ############################################################################
    ### get the TreeMap FIA tree list for only the plots included
    ############################################################################
    if(treemap_data_finder_ans$which_treemap==2022){
      treemap_cols <- c(
        "TM_ID"
        , "PLT_CN"
        , "SPECIES_SYMBOL"
        , "STATUSCD"
        , "DIA"
        , "HT"
        , "CR" # crown ratio
      )
    }else if(treemap_data_finder_ans$which_treemap==2016){
      treemap_cols <- c(
        "tm_id"
        , "CN"
        , "SPECIES_SYMBOL"
        , "STATUSCD"
        , "DIA"
        , "HT"
        , "CR" # crown ratio
      )
    }else{
      stop("unknown TreeMap vintage")
    }
    ### read it
    # treemap_data_finder_ans$treemap_trees
    treemap_trees_df <-
      readr::read_csv(
        treemap_data_finder_ans$treemap_trees
        , col_select = treemap_cols
        , progress = F
        , show_col_types = F
      ) %>%
      dplyr::rename_with(tolower)
    # rename cn to fit with original table str
    if(treemap_data_finder_ans$which_treemap==2022){
      treemap_trees_df <- treemap_trees_df %>% dplyr::rename(cn = plt_cn)
    }
    # clean it
    treemap_trees_df <-
      treemap_trees_df %>%
      dplyr::mutate(
        cn = cloud2trees:::as_character_safe(cn)
        , tm_id = cloud2trees:::as_character_safe(tm_id)
        , cr = ifelse(cr>100|cr<0,NA,cr)*0.01
      ) %>%
      dplyr::inner_join(
        tm_id_weight_temp
        , by = dplyr::join_by("tm_id")
      ) %>%
      dplyr::filter(
        # keep live trees only: 1=live;2=dead
        statuscd == 1
        & !is.na(dia)
        & !is.na(ht)
        & !is.na(tree_weight)
      ) %>%
      dplyr::mutate(
        dbh_cm = dia*2.54
        , tree_height_m = ht*0.3048
        , tm_id = as.factor(tm_id)
        , species_symbol = factor(species_symbol)
        # # normalize ht
        # , ht_wt_mean = sum(tree_height_m * tree_weight) / sum(tree_weight)
        # , ht_wt_sd = sqrt( sum(tree_weight * (tree_height_m - ht_wt_mean)^2 ) / sum(tree_weight) )
        # , ht_z = (tree_height_m - ht_wt_mean) / ht_wt_sd
      ) %>%
      dplyr::select(-c(statuscd,dia,ht,cr)) # maybe we can use cr in a future version with and NSUR approach
    
    # treemap_trees_df %>% dplyr::glimpse()
    
    # check FIA model data
    if(dplyr::coalesce(nrow(treemap_trees_df),0)==0){
      stop(paste0(
        "No trees in Treemap data. Ensure that "
        , "\nIf provided data in `study_boundary` ensure that it is located in the continental US"
        , "\nIf did not provide data in `study_boundary` ensure that the `tree_list` data has is located in the continental US"
      ))
    }
    
    # we need the spatial information from the raster...
    ############################################################################
    ### attach spatial information from raster to trees
    ############################################################################
    
    # create the tree list with trees assigned the X and Y of the raster cell and 
    # for imputed plot id's the specific tree will appear multiple times based on 
    # the number of times the plot is replicated within the area of interest but 
    # each tree will have a different X,Y and stability weight based on the surrounding imputed plots 
    
    treemap_trees_spatial_df <-
      treemap_id_rast %>% 
      terra::as.data.frame(xy = T, cells = T) %>% 
      dplyr::rename(tm_id=4) %>% 
      # add on weights based on moving window stability of imputed plot
      dplyr::left_join(
        treemap_prop_rast_temp %>% 
          terra::as.data.frame(xy = F, cells = T)
        , by = "cell"
      ) %>% 
      dplyr::mutate(
        tm_id = cloud2trees:::as_character_safe(tm_id)
        # clean stability_weight
        , stability_weight = ifelse(stability_weight>1,1,stability_weight) %>% dplyr::coalesce(0.001)
      ) %>% 
      # dplyr::glimpse()
      # add on the tree data in which a row is unique by tm_id and the tree (which doesn't have an id so we'll fake one)
      dplyr::inner_join(
        treemap_trees_df %>% 
          dplyr::ungroup() %>% 
          dplyr::mutate(
            tree_id = dplyr::row_number()
            , tm_id = cloud2trees:::as_character_safe(tm_id)
          )
        , by = "tm_id"
        , relationship = "many-to-many"
      ) %>% 
      # a row is now unique by x,y,tree_id resulting in the number of rows per tree_id = tm_id_weight_temp$tree_weight
      dplyr::mutate(
        # to factor
        # tm_id = forcats::fct_relevel(tm_id, levels(treemap_trees_df$tm_id) ) ##!!!!! do we need this?
        species_symbol = factor(species_symbol)
        # slightly jitter the tree x,y so that RF doesn't lock-in on grid pattern from raster
        # jitter within the 30m cell +/- 15 m
        , x = x + stats::runif(
          dplyr::n()
          , min = -(terra::res(treemap_rast)[1]/2)
          , max = (terra::res(treemap_rast)[1]/2)
        )
        , y = y + stats::runif(
          dplyr::n()
          , min = -(terra::res(treemap_rast)[1]/2)
          , max = (terra::res(treemap_rast)[1]/2)
        )
      )
    
    ############################################################################
    ### add max tree height in moving window to create relative height measurement
    ############################################################################
    # a template raster
    rast_temp <- terra::subset(treemap_rast,1)
    # replace cell values with maximum height from tree data
    values(rast_temp) <-
      dplyr::tibble(
        cell = c(1:terra::ncell(rast_temp))
      ) %>% 
      # left join to leave all NA as NA
      dplyr::left_join(
        treemap_trees_spatial_df %>% 
          sf::st_drop_geometry() %>% 
          dplyr::group_by(cell) %>% 
          dplyr::summarise(tree_height_m = max(tree_height_m,na.rm=F)) %>% 
          dplyr::arrange(cell)  
        , by = "cell"
      ) %>% 
      dplyr::pull(tree_height_m)
    # rast_temp %>% terra::plot()
      
    # moving window with window size of 3
    # rast_temp %>% terra::focal(w = 3, fun=max, na.rm=T) %>% terra::plot()
    max_ht_df <- 
      rast_temp %>% 
      terra::focal(w = 3, fun=max, na.rm=T) %>% 
      terra::as.data.frame(cells = T) %>% 
      dplyr::rename(max_tree_height_m = 2)
    
    # add max ht to spatial tree data
    treemap_trees_spatial_df <- 
      treemap_trees_spatial_df %>% 
      dplyr::left_join(max_ht_df, by = "cell") %>% 
      dplyr::mutate(
        rel_tree_height_m = tree_height_m/max_tree_height_m
        , rel_tree_height_m = ifelse(rel_tree_height_m>1,1,rel_tree_height_m)
      ) %>% 
      # make spatial
      sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(treemap_rast), remove = F)
    
    # dplyr::glimpse(treemap_trees_spatial_df)
    # treemap_trees_spatial_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) + ggplot2::geom_point() + ggplot2::theme_void()
    # treemap_trees_spatial_df %>% ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=stability_weight)) + ggplot2::geom_point() + ggplot2::theme_void()
    # treemap_trees_spatial_df %>% 
    #   ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=factor(cell))) + 
    #   ggplot2::geom_point() + 
    #   ggplot2::scale_color_manual(values = sample(
    #       viridis::turbo(n=length(unique(treemap_trees_spatial_df$cell)))
    #     )
    #   ) + 
    #   ggplot2::theme_void() + 
    #   ggplot2::theme(legend.position = "none")
    
    return(list(
      treemap_trees_df = treemap_trees_df
      , treemap_rast = treemap_rast
      , treemap_trees_spatial_df = treemap_trees_spatial_df
      , treemap_prop_rast = treemap_prop_rast_temp
    ))
}

# add to cloud2trees:::
# focal matrix for a x by x window
# focal_kernel_temp <- matrix(1, nrow = 3, ncol = 3)
# focal_kernel_temp
# length(focal_kernel_temp)*terra::res(treemap_rast) %>% prod() %>% `/`(10000)
# 5x5 = 2.25 ha; 3x3 = 0.81 ha
# custom window fn to get prop of surrounding cells matching focal value
# ?terra::focal
match_focal_cells_fn <- function(v, ...) {
  if(all(is.na(v))){ return(NA) }
  
  # center index of the window vector
  center_idx <- (length(v) + 1) / 2
  focal_val  <- v[center_idx]
  
  if(is.na(focal_val)){ return(NA) }
  
  # count how many cells in the window match the focal cell
  n_matches <- sum(v %in% focal_val, ...)
  prop_matches <- n_matches / length(v)
  
  return(prop_matches)
}


```

We did a few things in that function to return: 

* `treemap_trees_df` : a list of trees associated with a TreeMap imputed plot. the column `tree_weight` is based off of a count of how many times a imputed plot id is replicated across the area of interest. the `regional_dbh_height_model_training_data.csv` file written to the point cloud processing directory when we execute cloud2trees::cloud2trees(..., estimate_tree_dbh = T) matches this data.
* `treemap_rast` : The TreeMap raster data cropped to the area of interest with a buffer
* `treemap_trees_spatial_df` : a *spatial* list of trees associated with a TreeMap imputed plot. A row is now unique by `x`, `y`, `tree_id` resulting in the number of rows per tree determined by the `tree_weight`.
* `treemap_prop_rast` : A raster of the TreeMap imputed plot stability. We expect the TreeMap imputed plots to be more accurate as we look over a broader area with accuracy decreasing at smaller scales. That is, when we consider the tree composition of an entire forest stand we expect TreeMap to generally get the main species and forest type correct but for each 30m raster cell (i.e. the plot location that we get the X and Y coordinates for the tree list used for training from) we expect it to be less accurate with a cell potentially being a completely incorrect representation of the actual ground setting. As such, we we weight the individual tree records based on the stability of the imputed plot id's looking over a window surrounding the focal cell (`stability_weight` in `treemap_trees_spatial_df`). Specifically, we calculate the proportion of surrounding cells (5x5) that share the same imputed plot id as the focal cell and use this weight as a measure of TreeMap's predicted stability to place less weight on plots that occur infrequently and more weight on plots that occur frequently.

```{r}
# test to get the treemap data for our aoi
treemap_data_loader_ans <-
  pred_trees_list$USFS_ManitouEF_N1 %>% 
  dplyr::ungroup() %>% 
  sf::st_union() %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  treemap_data_loader()
```

Check out the TreeMap raster of forest types for our area of interest defined using the spatial extent of the predicted trees

```{r}
treemap_data_loader_ans$treemap_rast %>% terra::plot(axes=F)
```

here is the stability of those imputed TreeMap FIA plot id's

```{r}
treemap_data_loader_ans$treemap_prop_rast %>% terra::plot(col = viridis::mako(n=100), axes=F)
```

lets' review the spatial tree list we'll use for training data which comes from the TreeMap FIA imputed plots

what's in the FIA tree list using the TreeMap imputed plots filtered for our area of interest?

```{r}
treemap_data_loader_ans$treemap_trees_spatial_df %>% dplyr::glimpse()
```

species distribution?

```{r}
treemap_data_loader_ans$treemap_trees_spatial_df %>%
  sf::st_drop_geometry() %>%
  dplyr::count(species_symbol) %>%
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(
    percent=n/sum(n)
    , percent = scales::percent(percent,accuracy=0.1)
    , n = scales::comma(n,accuracy = 1)
  ) %>%
  kableExtra::kbl(
    caption = "TreeMap Species Distribution over Study Area"
  ) %>% 
  kableExtra::kable_styling()
```

here is the spatial distribution of the species for a sample of the trees

```{r}
samp_temp <- treemap_data_loader_ans$treemap_trees_spatial_df %>%
  dplyr::slice_sample(n=2222)
samp_temp %>%
  ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::scale_color_viridis_d(option = "turbo") +
  ggplot2::theme_void()
```

that is difficult to read, let's actually print the species symbol

```{r}
samp_temp %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = x, y = y, fill = species_symbol, color = species_symbol, label = species_symbol)
  ) +
  # ggplot2::geom_hline(yintercept = median(treemap_data_loader_ans$treemap_trees_spatial_df$y) , color = "black") + 
  # ggplot2::geom_vline(xintercept = median(treemap_data_loader_ans$treemap_trees_spatial_df$x), color = "black") + 
  ggplot2::geom_text(size = 2.5) +
  ggplot2::scale_color_viridis_d(option = "turbo") +
  ggplot2::labs(
    x = "x", y = "y"
  ) + 
  ggplot2::theme_void() + 
  ggplot2::theme(
    legend.position = "none"
  )
```

let's check out the `stability_weight` distribution which is unique for each raster cell in the data extent

```{r}
stab_temp <- 
  # stability_weight should be unique by cell
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::distinct(cell,stability_weight) %>% 
  # add on forest type
  dplyr::left_join(
    treemap_data_loader_ans$treemap_rast %>% 
      terra::as.data.frame(xy=T, cells = T)  %>% 
      dplyr::select(cell, ForTypName) %>% 
      dplyr::rename(f_type=ForTypName)
  ) %>% 
  dplyr::mutate(
    f_type = f_type %>% 
      forcats::fct_drop() %>% 
      forcats::fct_reorder(stability_weight,.fun = "median") %>% 
      as.ordered()
  )
stab_temp %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = f_type, x = stability_weight, fill = f_type)
  ) +
  ggplot2::geom_boxplot(
    width = 0.4
    , alpha=0.9
    , outliers = F
    , color = "black"
  ) +
  ggplot2::scale_fill_viridis_d(option = "magma",direction=-1,begin=0.1,end=0.5) +
  ggplot2::labs(y="") +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , legend.direction  = "horizontal"
  )
```

this essentially means that TreeMap is fairly confident that the study area includes `r paste0(levels(forcats::fct_rev(stab_temp$f_type))[1:2], collapse = " and ")`

lastly, we'll look at the relative height data for a sample of the data

```{r}
focal_cell_temp <- 
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::filter(
    stability_weight>=
      quantile(treemap_data_loader_ans$treemap_trees_spatial_df$stability_weight,probs=0.9)
  ) %>% 
  dplyr::count(cell) %>% 
  dplyr::slice_max(order_by = n, n = 10, with_ties = F) %>% 
  dplyr::slice_sample(n=1) %>% 
  dplyr::pull(cell)

aoi_temp <- 
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  dplyr::filter(cell == focal_cell_temp) %>% 
  sf::st_union() %>% 
  sf::st_centroid() %>% 
  sf::st_buffer(terra::res(treemap_data_loader_ans$treemap_rast)[1]) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc()

treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_intersection(aoi_temp) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(x,y,tree_height_m,rel_tree_height_m) %>% 
  # tidyr::pivot_longer(cols = -c(x,y)) %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(x=x,y=y)
  ) + 
  ggplot2::geom_point(mapping = ggplot2::aes(color = rel_tree_height_m), size = 4) +
  ggplot2::scale_color_distiller(palette = "Greys", direction = 1) +
  ggnewscale::new_scale_color() +
  ggplot2::geom_point(mapping = ggplot2::aes(color = tree_height_m), size = 1.8) +
  ggplot2::scale_color_distiller(palette = "Oranges", direction = 1) +
  ggplot2::theme_void()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Predict Species/Genus Application

let's implement a random forest model majority voting for species with the objective to reduce estimation time (compared to a Bayesian Softmax model)

tree species will be our dependent variable and the most simple model will use tree height and tree location (X and Y coordinates) as independent variables. The training FIA tree list data (identified from the TreeMap imputed plots of the area of interest) includes tree height and we'll use the centroid of the TreeMap 30m raster cell to represent the individual tree X and Y location (i.e. each tree within the same cell will have the same coordinate). Considering the combined tree height and spatial location via a full interaction model (i.e. `height*x*y`) allows the model to capture how the species-height relationship varies across space.

we'll first update the `cloud2trees` internal function `rf_subsample_and_model_n_times` to allow for categorical response data

```{r}
####################################!!!!!!!!!!!!!! this is utils_rf.R in cloud2trees code
####################################!!!!!!!!!!!!!! this is utils_rf.R in cloud2trees code
####################################!!!!!!!!!!!!!! this is utils_rf.R in cloud2trees code
rf_tune_subsample <- function(
  predictors
  , response
  , threshold = 14444
  , n_subsamples = 4
  , ntree_try = 44
  , step_factor = 1
  , improve = 0.03
) {
  # set upper limit for ntreeTry
  max_ntreeTry <- 122
  # count the obs
  n_obs <- nrow(predictors)
  # set up list to store values from subsample runs
  optimal_mtry_values <- numeric(as.numeric(n_subsamples))
  # quiet the thing
  quiet_tuneRF <- purrr::quietly(randomForest::tuneRF)
  # if we don't need to subsample
  if(
    n_obs<=threshold
  ){
    # get rid of missing factor levels if not in there
    # results in error: "Can't have empty classes in y." from randomForest::tuneRF
    if(inherits(response,"factor")){
      response <- forcats::fct_drop(response)  
    }
    
    # Tune on the full data if below threshold
    tune_result <- quiet_tuneRF(
      # randomForest::tuneRF(
      y = response
      , x = predictors
      , ntreeTry = max_ntreeTry # this is high for data with less obs
      , stepFactor = step_factor
      , improve = improve
      , plot = F
      , trace = F
    )
    # just get the result
    tune_result <- tune_result$result
    # Extract the optimal mtry value
    optimal_mtry <- tune_result %>%
      dplyr::as_tibble() %>%
      dplyr::filter(OOBError==min(OOBError)) %>%
      dplyr::filter(dplyr::row_number() == 1) %>%
      dplyr::pull(mtry) %>%
      as.numeric()
  }else{
    # number for subsample
    sample_obs <- as.numeric(threshold)
    # run tuning on subsamples
    for (i in 1:n_subsamples) {
      # subsample
      subsample_indices <- sample(x = 1:n_obs, size = sample_obs, replace = F)
      subsample_predictors <- predictors[subsample_indices, ]
      subsample_response <- response[subsample_indices]
      
      # get rid of missing factor levels if not in there
      # results in error: "Can't have empty classes in y." from randomForest::tuneRF
      if(inherits(subsample_response,"factor")){
        subsample_response <- forcats::fct_drop(subsample_response)  
      }
      
      # tune on subsample
      tune_result <- quiet_tuneRF(
        # randomForest::tuneRF(
        y = subsample_response
        , x = subsample_predictors
        , ntreeTry = min(as.numeric(ntree_try),max_ntreeTry)
        , stepFactor = step_factor
        , improve = improve
        , plot = F
        , trace = F
      )
      # just get the result
      tune_result <- tune_result$result
      # Extract the optimal mtry value
      optimal_mtry_values[i] <- tune_result %>%
        dplyr::as_tibble() %>%
        dplyr::filter(OOBError==min(OOBError)) %>%
        dplyr::filter(dplyr::row_number() == 1) %>%
        dplyr::pull(mtry) %>%
        as.numeric()
    }
    # Find most frequent optimal mtry
    optimal_mtry <- table(optimal_mtry_values) %>%
      sort(decreasing = TRUE) %>%
      .[1] %>%
      names() %>%
      as.numeric()
  }
  # ensure that the mtry value is not greater than the number of predictors
  optimal_mtry <- min(optimal_mtry, ncol(predictors))
  # return
  return(optimal_mtry)
}

################################################################################################################################################
# function to combine tuning and modelling
################################################################################################################################################
rf_tune_model <- function(
  predictors
  , response
  , tune_threshold = 14444
  , tune_n_subsamples = 4
  , tune_ntree_try = 44
  , tune_step_factor = 1
  , tune_improve = 0.03
  , ntree = 500
  , weights = NULL
) {
  ### tuning RF model
  # implements steps to mitigate very long run-times when tuning random forests models
  optimal_mtry <- rf_tune_subsample(
    predictors = predictors
    , response = response
    , threshold = tune_threshold
    , n_subsamples = tune_n_subsamples
    , ntree_try = tune_ntree_try
    , step_factor = tune_step_factor
    , improve = tune_improve
  )

  ### Run a randomForest model
  
  # get rid of missing factor levels if not in there
    # results in error: "Can't have empty classes in y." from randomForest::tuneRF
    if(inherits(response,"factor")){
      response <- forcats::fct_drop(response)  
    }
  
  # quiet this
  quiet_rf <- purrr::quietly(randomForest::randomForest)
  # run it
  mod <- quiet_rf(
    y = response
    , x = predictors
    , mtry = optimal_mtry
    , na.action = na.omit
    , ntree = ntree
    # # weights = A vector of length same as y that are positive weights (i.e. >0 for all obs)
    #   # used only in sampling data to grow each tree (not used in any other calculation)
    , weights = weights
  )

  # just get the result
  mod <- mod$result
  return(mod)
}
################################################################################################################################################
# function to subsample, tune, and model
################################################################################################################################################
rf_subsample_and_model <- function(
  predictors
  , response
  , tune_threshold = 14444
  , tune_n_subsamples = 4
  , tune_ntree_try = 44
  , tune_step_factor = 1
  , tune_improve = 0.03
  , mod_n_subsample = 20000
  , ntree = 500
  , weights = NULL
) {
  if(length(response)<=mod_n_subsample){
    mod <- rf_tune_model(
        predictors = predictors
        , response = response
        , tune_threshold = tune_threshold
        , tune_n_subsamples = tune_n_subsamples
        , tune_ntree_try = tune_ntree_try
        , tune_step_factor = tune_step_factor
        , tune_improve = tune_improve
        , ntree = ntree
        , weights = weights
      )
  }else{
    subsample_indices <- sample(x = 1:length(response), size = mod_n_subsample, replace = F)
    subsample_y <- response[subsample_indices]
    subsample_x <- predictors[subsample_indices,]
    # weights
    if(!is.null(weights) && !any(is.na(weights))){
      subsample_w <- weights[subsample_indices]
    }else{
      subsample_w <- NULL
    }
    
    mod <- rf_tune_model(
        predictors = subsample_x
        , response = subsample_y
        , tune_threshold = tune_threshold
        , tune_n_subsamples = tune_n_subsamples
        , tune_ntree_try = tune_ntree_try
        , tune_step_factor = tune_step_factor
        , tune_improve = tune_improve
        , ntree = ntree
        , weights = subsample_w
      )
  }
  return(mod)
}
################################################################################################################################################
# function to subsample, tune, and model N TIMES
# iterates the subsampling and model fitting process n_iterations times
# stores the predictions from each model in a list
################################################################################################################################################
rf_subsample_and_model_n_times <- function(
  predictors
  , response
  , tune_threshold = 14444
  , tune_n_subsamples = 4
  , tune_ntree_try = 44
  , tune_step_factor = 1
  , tune_improve = 0.03
  , mod_n_subsample = 20000
  , mod_n_times = 3
  , ntree = 500
  , weights = NULL
) {
  # set up blank list
  mod_list <- list()
  if(length(response)<=mod_n_subsample){
    mod_list[[1]] <- rf_subsample_and_model(
        predictors = predictors
        , response = response
        , tune_threshold = tune_threshold
        , tune_n_subsamples = tune_n_subsamples
        , tune_ntree_try = tune_ntree_try
        , tune_step_factor = tune_step_factor
        , tune_improve = tune_improve
        , mod_n_subsample = mod_n_subsample
        , ntree = ntree
        , weights = weights
      )
  }else{
    for (i in 1:mod_n_times) {
      mod_list[[i]] <- rf_subsample_and_model(
          predictors = predictors
          , response = response
          , tune_threshold = tune_threshold
          , tune_n_subsamples = tune_n_subsamples
          , tune_ntree_try = tune_ntree_try
          , tune_step_factor = tune_step_factor
          , tune_improve = tune_improve
          , mod_n_subsample = mod_n_subsample
          , ntree = ntree
          , weights = weights
        )
    }
  }
  return(mod_list)
}
################################################################################################################################################
# function to model average predictions using model list
# Model averaging can improve the robustness and accuracy of random forest models, especially when dealing with large datasets
# predict_df must have predictor (x) columns used to develop model
################################################################################################################################################
rf_model_avg_predictions <- function(mod_list, predict_df, levels = NULL){
  mod_type_temp <- mod_list[[1]]$type
  if(mod_type_temp == "classification") {
    if(is.null(levels)){stop("this is a RF classifier and you need to provide `levels` based on training data")}
    predicted_temp <- rf_model_vote_predictions(
        mod_list = mod_list
        , predict_df = predict_df
        , levels = levels
      )
  }else{
    # if (mod_type_temp == "regression") {
    # get model predictions for each model
    predicted_temp <-
      1:length(mod_list) %>%
        purrr::map(function(x){
          predict(
            mod_list[[x]]
            , predict_df
          ) %>%
          dplyr::as_tibble() %>%
          dplyr::pull(1) %>%
          dplyr::as_tibble() %>%
          dplyr::rename(value=1) %>%
          dplyr::mutate(id=dplyr::row_number())
        }) %>%
        dplyr::bind_rows()
    # model average
    predicted_temp <- predicted_temp %>%
      dplyr::group_by(id) %>%
      dplyr::summarise(
        predicted = mean(value, na.rm = T)
      ) %>%
      dplyr::ungroup() %>%
      dplyr::arrange(id) %>%
      dplyr::select(predicted)
  }
  
  # return
  return(predicted_temp)
}
################################################################################################################################################
# !!!!! ADDED for factor dependent variable
# function to model average predictions using model list
# Model averaging can improve the robustness and accuracy of random forest models, especially when dealing with large datasets
# predict_df must have predictor (x) columns used to develop model
################################################################################################################################################
# function to vote for class based on many models
rf_model_vote_predictions <- function(mod_list, predict_df, levels=NULL){
  if(is.null(levels)){stop("this is a RF classifier and you need to provide `levels` based on training data")}
  # get model predictions for each model
  predicted_temp <-
    1:length(mod_list) %>%
      purrr::map(function(x){
        predict(
          mod_list[[x]]
          , predict_df
        ) %>%
        dplyr::as_tibble() %>%
        dplyr::pull(1) %>%
        dplyr::as_tibble() %>%
        dplyr::rename(value=1) %>%
        # !!!!! need to restore the original factor levels impacted when we did `forcats::fct_drop(response)`
        dplyr::mutate(value = factor(value, levels = levels)) %>% 
        # dplyr::pull(value) %>% levels()
        dplyr::mutate(id=dplyr::row_number())
      }) %>%
      dplyr::bind_rows()
  # predicted_temp %>% dplyr::glimpse()
  
  # model vote
  predicted_temp <-
    predicted_temp %>%
    dplyr::count(id,value) %>% 
    # dplyr::arrange(id,desc(n)) %>% print(n=33)
    dplyr::group_by(id) %>% 
    dplyr::slice_max(order_by = n, n = 1, with_ties = F) %>% 
    dplyr::ungroup() %>%
    dplyr::arrange(id) %>%
    dplyr::rename(predicted=value) %>% 
    dplyr::select(predicted)
  # return
  return(predicted_temp)
}

```

let's estimate the RF using data subsampling and model averaging for a computationally efficient way to get estimates over broad areas with potentially large tree lists. Model averaging can improve the robustness and accuracy of random forest models, especially when dealing with large datasets.

```{r}
# treemap_tree_list_df %>% dplyr::glimpse()
# make a function to model and get the predictions attached to the tree list
get_species_symbol_predictions <- name <- function(treemap_data_loader_ans, predict_df) {
  
  # set max FIA trees to use in each subsample
  mod_n_subsample_temp <- min(nrow(treemap_data_loader_ans$treemap_trees_spatial_df)-1,22222)
  # number of models to estimate with this subsample size
  # go for at least 50% of training data sampled
  # it's just that the individual model fits will be smaller
  ntimes_temp <-
    ((nrow(treemap_data_loader_ans$treemap_trees_spatial_df)*0.5)/mod_n_subsample_temp) %>%
    ceiling() %>%
    max(6) %>%
    min(66)
  # ntimes_temp
  
  # estimate
  rf_spec_mod <- rf_subsample_and_model_n_times(
    predictors = treemap_data_loader_ans$treemap_trees_spatial_df %>% sf::st_drop_geometry() %>% dplyr::select(x,y,tree_height_m,rel_tree_height_m)
    # predictors = treemap_data_loader_ans$treemap_trees_spatial_df %>% sf::st_drop_geometry() %>% dplyr::select(x,y,tree_height_m)
    , response = treemap_data_loader_ans$treemap_trees_spatial_df$species_symbol
    , weights = treemap_data_loader_ans$treemap_trees_spatial_df$stability_weight
    , mod_n_subsample = mod_n_subsample_temp
    , mod_n_times = ntimes_temp
  )
  
  # model the missing values
  # vote for class based on many models
  predicted_temp <- rf_model_avg_predictions(
    mod_list = rf_spec_mod
    , predict_df = 
      predict_df %>% 
        sf::st_drop_geometry() %>%
        sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(predict_df), remove = F) %>% 
        sf::st_transform(terra::crs(treemap_data_loader_ans$treemap_rast)) %>% 
        dplyr::mutate(
          x = sf::st_coordinates(.)[,1]
          , y = sf::st_coordinates(.)[,2]
        ) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(x,y,tree_height_m,rel_tree_height_m)
        # dplyr::select(x,y,tree_height_m)
    , levels = levels(treemap_data_loader_ans$treemap_trees_spatial_df$species_symbol)
  )
  # predicted_temp %>% dplyr::glimpse()
  # predicted_temp %>% dplyr::count(predicted) %>% dplyr::arrange(desc(n)) %>% dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
  
  # attach back to data
  predict_df <- predict_df %>%
    dplyr::mutate(
      predicted_species_symbol = predicted_temp$predicted
      , predicted_species_symbol = as.character(predicted_species_symbol)
    )
  
  return(predict_df)
}

# do it with our data
pred_trees_list$USFS_ManitouEF_N1 <- 
  get_species_symbol_predictions(
    treemap_data_loader_ans = treemap_data_loader_ans
    , predict_df = pred_trees_list$USFS_ManitouEF_N1
  ) %>% 
  # for this project we'll just rename to maintain consistent across data sources
  dplyr::rename(species_symbol=predicted_species_symbol)

# pred_trees_list$USFS_ManitouEF_N1 %>%
#   sf::st_drop_geometry() %>%
#   dplyr::count(species_symbol) %>%
#   dplyr::arrange(desc(n)) %>%
#   dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

let's map predicted crowns in space colored by species for a single plot while also showing the field validation data species and the true positive matches (squares)

```{r, fig.height = 9.8, fig.width = 6.5}
pred_trees_list$USFS_ManitouEF_N1 %>%
  dplyr::inner_join(
    ref_plots_list$USFS_ManitouEF_N1 %>% 
      dplyr::slice(4) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(plot_id)
    , by = "plot_id"
  ) %>% 
  dplyr::arrange(desc(species_symbol)) %>% 
  # plot it
  # plt
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = ref_plots_list$USFS_ManitouEF_N1 %>% dplyr::slice(4)
      , fill = NA, color = "blue"
    ) +
    ggplot2::geom_sf(
      mapping = ggplot2::aes(fill = species_symbol)
      , size = 0.6, alpha = 0.88
    ) +
    ggplot2::geom_sf(
      data = 
        ref_trees_list$USFS_ManitouEF_N1 %>%
        dplyr::inner_join(
          ref_plots_list$USFS_ManitouEF_N1 %>% 
            dplyr::slice(4) %>% 
            sf::st_drop_geometry() %>% 
            dplyr::select(plot_id)
          , by = "plot_id"
        ) %>% 
        dplyr::inner_join(
          pred_ref_tree_match_list$USFS_ManitouEF_N1 %>% 
            dplyr::filter(match_grp=="true positive") %>% 
            dplyr::select(ref_treeID)
          , by = dplyr::join_by(treeID==ref_treeID)
        )
      , size = 7, shape = 15, color = "gray88"
    ) +
    ggplot2::geom_sf(
      data = 
        ref_trees_list$USFS_ManitouEF_N1 %>%
        dplyr::inner_join(
          ref_plots_list$USFS_ManitouEF_N1 %>% 
            dplyr::slice(4) %>% 
            sf::st_drop_geometry() %>% 
            dplyr::select(plot_id)
          , by = "plot_id"
        )
      , mapping = ggplot2::aes(color = species)
      , size = 3
    ) +
    ggplot2::scale_fill_viridis_d(option = "turbo", begin = 0.3, end = 0.8) +
    ggplot2::scale_color_viridis_d(option = "magma", begin = 0.3, end = 0.8) +
    ggplot2::labs(color="reference", fill="predicted") +
    ggplot2::theme_void() +
    ggplot2::theme(
      legend.position = "top"
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1))
      , color = ggplot2::guide_legend(override.aes = list(shape = 15, linetype = 0, size = 6, alpha = 1, fill = NA))
    )


```

that looks decent. however, we need to standardize the species code across data sources. what even is the difference between "PIPA2" and "PIPA"?

let's get the species code lookup table from the [FIA DataMart](https://research.fs.usda.gov/products/dataandtools/fia-datamart) to standardize the species code, species symbol, common name, and scientific genus and species

```{r}
# holy smokes, no one knows what these numeric species codes are
  # ....get reference species so a human can understand these
url_temp <- "https://apps.fs.usda.gov/fia/datamart/CSV/FIADB_REFERENCE.zip"
zip_temp <- "FIADB_REFERENCE.zip"
# dir_temp <- file.path("../data", str_remove(basename(zip_temp), "\\.[^.]+$"))
dir_temp <- file.path("c:/data/usfs/uas_lidar_n1/data", str_remove(basename(zip_temp), "\\.[^.]+$"))
# dir.exists(dir_temp)
# dl if needed
if (!dir.exists(dir_temp)){
  download.file(url = url_temp, destfile = zip_temp, mode = "wb")
  unzip(zipfile = zip_temp, exdir = dir_temp)
  file.remove(zip_temp)
}
# read
fia_species_lookup <- readr::read_csv(file = file.path(dir_temp, "REF_SPECIES.csv"), progress = F, show_col_types = F) %>% 
  dplyr::rename_with(.cols = dplyr::everything(), tolower) %>% 
  dplyr::select(spcd, species_symbol, common_name, genus, species)
# huh
fia_species_lookup %>% dplyr::glimpse()
```

what is the correct "PIPA2"?

```{r}
fia_species_lookup %>% 
  dplyr::filter(tolower(species_symbol) %>% stringr::str_starts("pipa"))
```

Let's try to make a function that automatically maps a field-recorded species symbol which is not present in the FIA reference species table to the correct symbol. we'll use the TreeMap data of the area to make the mapping. This will be especially important when we move into more diverse forest types where manually mapping the species codes would be very cumbersome.

```{r}
clean_ref_species_fn <- function(
  df_to_clean
  , my_treemap_data_loader_ans
) {

  # species probs
  species_probs <-
    my_treemap_data_loader_ans$treemap_trees_df %>% 
    dplyr::mutate(species_symbol = as.character(species_symbol)) %>% 
    dplyr::group_by(species_symbol) %>% 
    dplyr::summarise(n = sum(tree_weight)) %>% 
    # just the first 4 characters to get match
    dplyr::mutate(species_symbol4 = stringr::str_sub(species_symbol,1,4)) %>% 
    dplyr::arrange(species_symbol4,desc(n),species_symbol) %>% 
    dplyr::group_by(species_symbol4) %>% 
    dplyr::mutate(is_prob_species = dplyr::row_number()==1) %>% #ties go to alphabetical by species_symbol
    dplyr::ungroup()
  
  # match with df
  ret_df <- 
    df_to_clean %>% 
    dplyr::left_join(
      species_probs %>% 
        dplyr::select(species_symbol) %>% 
        dplyr::mutate(xxx_is_match = T)
      , by = "species_symbol"
    ) %>% 
    # try it with 4
    dplyr::left_join(
      species_probs %>% 
        dplyr::filter(is_prob_species) %>% 
        dplyr::select(species_symbol,species_symbol4) %>% 
        dplyr::mutate(xxx_is_match4 = T) %>% 
        dplyr::rename(
          xxx_new_species_symbol = species_symbol
          , species_symbol = species_symbol4
        )
      , by = "species_symbol"
    ) %>% 
    # update species_symbol
    dplyr::mutate(
      xxx_is_match = dplyr::coalesce(xxx_is_match,F)
      , xxx_is_match4 = dplyr::coalesce(xxx_is_match4,F)
      , species_symbol = dplyr::case_when(
        xxx_is_match ~ species_symbol
        , xxx_is_match4 ~ xxx_new_species_symbol
        , T ~ species_symbol
      )
    ) %>% 
    dplyr::select(-c(xxx_is_match,xxx_is_match4,xxx_new_species_symbol))
  
  # some still don't match FIAdb
  # just take the first one from the fiadb if not found based on the 4 characters
  ret_df %>% 
    dplyr::left_join(
      fia_species_lookup %>% 
        dplyr::select(species_symbol) %>% 
        dplyr::mutate(species_symbol4 = stringr::str_sub(species_symbol,1,4)) %>% 
        dplyr::slice_head(n = 1, by = species_symbol4) %>% 
        dplyr::mutate(xxx_is_match4 = T) %>% 
        dplyr::rename(
          xxx_new_species_symbol = species_symbol
          , species_symbol = species_symbol4
        )
      , by = "species_symbol"
    ) %>% 
    # update species_symbol
    dplyr::mutate(
      xxx_is_match4 = dplyr::coalesce(xxx_is_match4,F)
      , species_symbol = dplyr::case_when(
        xxx_is_match4 ~ xxx_new_species_symbol
        , T ~ species_symbol
      )
    ) %>% 
    dplyr::select(-c(xxx_is_match4,xxx_new_species_symbol))

}
# update the reference tree list
ref_trees_list$USFS_ManitouEF_N1 <- 
  ref_trees_list$USFS_ManitouEF_N1 %>% 
  clean_ref_species_fn(
    my_treemap_data_loader_ans = treemap_data_loader_ans
  )
```

what is the distribution of the cleaned reference field data species?

```{r}
ref_trees_list$USFS_ManitouEF_N1 %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species_symbol)
```

eh. there is one genus only observation and many "HDWD" observations (which i assume is "hard wood"). let's just use the most common of the same genus

```{r}
ref_trees_list$USFS_ManitouEF_N1 <- ref_trees_list$USFS_ManitouEF_N1 %>% 
  dplyr::mutate(
    species_symbol = dplyr::case_when(
      species_symbol %in% c("QU","HDWD") ~ "QULA2"
      , T ~ species_symbol
    )
  )
```

### Species/Genus Accuracy Evaluation

#### Tree-Level Accuracy{#05_tree_acc}

Given that our analysis is focused on evaluating the proposed taxonomic classification algorithm (rather than the individual tree detection method), we'll evaluate the taxonomic classification accuracy conditionally for successfully detected tree crowns (true positive matches). This conditional evaluation will allow us to isolate taxonomic model performance from tree detection errors (reviewed above [here](#03_itd)). In practice, the ITD window function would be tuned for the specific site and data being analyzed, but we implemented an automated ITD window selection process here since we will be evaluating many different sites and forest types.

we're going to use the tree-level prediction and reference data for all TP (true positive) predictions to perform confusion matrix accuracy evaluation of the species predictions. 

using the confusion matrix evaluation we'll calculate the following metrics: 

* **Overall Accuracy**: Measures the global percentage of all correctly classified trees out of the entire dataset of true positive (TP) observations. Provides an overall summary metric but can mask performance on uncommon species.
* **User's Accuracy** (precision): Measures the reliability of specific predicted species. For example, if the method predicts a tree is a ponderosa pine, how often it is actually true.
* **Producer's Accuracy** (recall): Measures how well the model successfully aligns with what was measured on the ground. For example, out of all real ponderosa pine trees measured in the field, what percentage did the method correctly identify.

$$\text{Overall Accuracy} = \frac{\text{Number of Correctly Classified Species Observations}}{\text{Total Number of Successfully Detected Trees (TPs)}}$$



$$\text{User's Accuracy (Species } i\text{)} = \frac{\text{Number of Correctly Classified Observations for Species } i}{\text{Total Number of Predicted Observations for Species } i}$$


$$\text{Producer's Accuracy (Species } i\text{)} = \frac{\text{Number of Correctly Classified Observations for Species } i}{\text{Total Number of Reference Observations for Species } i}$$
```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's make a function to get the confusion matrix and those accuracy metrics

```{r}
# fia_species_lookup %>% dplyr::glimpse()
# function to do confusion matrix and tree-level accuracy metric calcs
confusion_matrix_tree_class_fn <- function(pred_ref_match, pred_trees, ref_trees, use_genus = F, use_only_tp = T) {

  # add species to matches
  w_species_temp <- 
    pred_ref_match %>% 
    ################## predicted ##################
    dplyr::left_join(
      pred_trees %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(treeID, species_symbol) %>% 
        # add on species info
        dplyr::left_join(
          fia_species_lookup %>% 
            dplyr::select(-spcd) %>% 
            dplyr::rename_with(
              .cols = -c(species_symbol)
              , .fn = ~paste0("species_",.x,recycle0 = T)
            )
          , by = "species_symbol"
        ) %>% 
        dplyr::rename_with(
          .cols = tidyselect::starts_with("species_")
          , .fn = ~paste0("pred_",.x,recycle0 = T)
        )
      , by = dplyr::join_by(pred_treeID==treeID)
    ) %>% 
    ################## predicted ##################
    ################## reference ##################
    dplyr::left_join(
      ref_trees %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(treeID, species_symbol) %>% 
        # add on species info
        dplyr::left_join(
          fia_species_lookup %>% 
            dplyr::select(-spcd) %>% 
            dplyr::rename_with(
              .cols = -c(species_symbol)
              , .fn = ~paste0("species_",.x,recycle0 = T)
            )
          , by = "species_symbol"
        ) %>% 
        dplyr::rename_with(
          .cols = tidyselect::starts_with("species_")
          , .fn = ~paste0("ref_",.x,recycle0 = T)
        )
      , by = dplyr::join_by(ref_treeID==treeID)
    ) %>% 
    ################## reference ##################
    # clean up missing
    dplyr::mutate(
      dplyr::across(
        tidyselect::starts_with("pred_species_") & !c("pred_species_symbol")
        , .fns = ~dplyr::case_when(
          is.na(pred_species_symbol) ~ "NONE"
          , is.na(.x) ~ "UNKNOWN"
          , T ~ .x
        )
      )
      , dplyr::across(
        tidyselect::starts_with("ref_species_") & !c("ref_species_symbol")
        , .fns = ~dplyr::case_when(
          is.na(ref_species_symbol) ~ "NONE"
          , is.na(.x) ~ "UNKNOWN"
          , T ~ .x
        )
      )
      , pred_species_symbol = dplyr::case_when(
          is.na(pred_species_symbol) ~ "NONE"
          , is.na(pred_species_symbol) ~ "UNKNOWN"
          , T ~ pred_species_symbol
        )
      , ref_species_symbol = dplyr::case_when(
          is.na(ref_species_symbol) ~ "NONE"
          , is.na(ref_species_symbol) ~ "UNKNOWN"
          , T ~ ref_species_symbol
        )
    )
  
  if(use_only_tp){
    w_species_temp <- w_species_temp %>% dplyr::filter(match_grp == "true positive")
  }
  
  # w_species_temp %>% dplyr::glimpse()
  # w_species_temp %>% dplyr::count(pred_species_genus,ref_species_genus)
  
  if(use_genus){
    w_species_temp <- w_species_temp %>% 
      dplyr::mutate(
        xxx_pred = pred_species_symbol
        , xxx_ref = ref_species_symbol
        , pred_species_symbol = pred_species_genus
        , ref_species_symbol = ref_species_genus
      )
  }
    
  
  
  ##############################################
  # pivot for confusion matrix
  ##############################################
  
  cm_df_temp <-
    w_species_temp %>% 
    dplyr::count(pred_species_symbol,ref_species_symbol) %>% 
    # make sure matrix diagonal is available
    dplyr::full_join(
      w_species_temp %>% 
        dplyr::distinct(ref_species_symbol) %>% 
        dplyr::mutate(pred_species_symbol=ref_species_symbol)
      , by = dplyr::join_by(ref_species_symbol,pred_species_symbol)
    ) %>% 
    dplyr::mutate(n=dplyr::coalesce(n,0)) %>% 
    tidyr::pivot_wider(
      names_from = pred_species_symbol
      , values_from = n
      , values_fill = 0
    ) %>% 
    # again make sure matrix diagonal is available
    dplyr::bind_rows(
      dplyr::tibble(
        ref_species_symbol = 
          base::setdiff(
            unique(w_species_temp$pred_species_symbol)
            , unique(w_species_temp$ref_species_symbol)
          )
      )
    ) %>% 
    dplyr::mutate(
      dplyr::across(
        .cols = dplyr::where(is.numeric)
        , .fns = ~dplyr::coalesce(.x,0)
      )
    ) %>% 
    dplyr::mutate(sorter = ref_species_symbol=="NONE") %>% 
    dplyr::arrange(sorter, ref_species_symbol) %>%
    dplyr::select(-sorter)
  # rearrange columns for matrix diagonal
  cm_df_temp <- 
    cm_df_temp %>% 
    dplyr::select(
      ref_species_symbol
      , dplyr::all_of(
        sort(unique(cm_df_temp$ref_species_symbol))
      )
      , base::setdiff(names(cm_df_temp %>% dplyr::select(-ref_species_symbol)),unique(cm_df_temp$ref_species_symbol))
    )
  
  if(any(stringr::str_equal(names(cm_df_temp), "UNKNOWN"))){
    cm_df_temp <- cm_df_temp %>% dplyr::relocate(UNKNOWN, .after = dplyr::last_col())
  }
  if(any(stringr::str_equal(names(cm_df_temp), "NONE"))){
    cm_df_temp <- cm_df_temp %>% dplyr::relocate(NONE, .after = dplyr::last_col())
  }
  
  # cm_df_temp %>% dplyr::glimpse()
  # cm_df_temp
  
  ##############################################
  # calc accuracy metrics
  ##############################################
  
  #######################
  # overall (OA)
  #######################
  
  # all reference and prediction records
  overall_n_temp <- nrow(w_species_temp) %>% dplyr::coalesce(0)
  # matching but excludes NONE and NONE
  correct_n_temp <- 
    w_species_temp %>% 
    dplyr::filter(
      pred_species_symbol == ref_species_symbol
      , pred_species_symbol != "NONE"
    ) %>% 
    nrow() %>% 
    dplyr::coalesce(0)
  
  overall_accuracy_temp <- ifelse(
    overall_n_temp == 0, NA
    , (correct_n_temp / overall_n_temp)
  )
  # overall_accuracy_temp
  
  #######################
  # UA and PA
  #######################
  # combine total counts
  tp_by_species_temp <- 
    w_species_temp %>% 
    dplyr::filter(
      pred_species_symbol == ref_species_symbol
      , pred_species_symbol != "NONE"
    ) %>% 
    dplyr::count(species_symbol = pred_species_symbol, name = "correct_n")
  
  # tot pred by species (denominator for UA)
  pred_tot_temp <- 
    w_species_temp %>% 
    dplyr::filter(pred_species_symbol != "NONE") %>% 
    dplyr::count(species_symbol = pred_species_symbol, name = "tot_pred_n")
  
  # tot ref by species (denominator for PA)
  ref_tot_temp <- 
    w_species_temp %>% 
    dplyr::filter(ref_species_symbol != "NONE") %>% 
    dplyr::count(species_symbol = ref_species_symbol, name = "tot_ref_n")
  
  # bring together
  species_accuracy_df_temp <-
    pred_tot_temp %>% 
    # full_join makes all combination
    dplyr::full_join(ref_tot_temp, by = "species_symbol") %>% 
    dplyr::left_join(tp_by_species_temp, by = "species_symbol") %>% 
    # clean up
    dplyr::mutate(
      dplyr::across(
        c(correct_n, tot_pred_n, tot_ref_n)
        , ~dplyr::coalesce(.x, 0)
      )
      # handle 0 in denominator
      , users_accuracy = dplyr::case_when(
        tot_pred_n == 0 & tot_ref_n == 0 ~ 1
        , tot_pred_n == 0 ~ as.numeric(NA)
        , T ~ correct_n / tot_pred_n
      )
      , producers_accuracy = dplyr::case_when(
        tot_pred_n == 0 & tot_ref_n == 0 ~ 1
        , tot_ref_n == 0 ~ as.numeric(NA)
        , T ~ correct_n / tot_ref_n
      )
    )
  
  species_accuracy_df_temp
  # cm_df_temp
  
  species_accuracy_df_lab_temp <- species_accuracy_df_temp %>% 
    dplyr::mutate(
      dplyr::across(
        tidyselect::ends_with("_n")
        , ~ scales::comma(.x,accuracy=1)
      )
      , dplyr::across(
        tidyselect::ends_with("_accuracy")
        , ~ scales::percent(.x,accuracy=.1)
      )
    )
  # species_accuracy_df_lab_temp
  
  # make a kableExtra table
  kbl_table_df_temp <-
    cm_df_temp %>% 
    dplyr::mutate(
      dplyr::across(
        dplyr::where(is.numeric)
        , ~ scales::comma(.x,accuracy=1)
      )
    ) %>% 
    dplyr::left_join(
      species_accuracy_df_lab_temp %>% 
        dplyr::select(species_symbol,tot_ref_n,producers_accuracy) %>% 
        dplyr::bind_rows(
          dplyr::tibble(
            species_symbol = "NONE"
            , tot_ref_n = scales::comma(
              (w_species_temp %>% dplyr::filter(ref_species_symbol=="NONE") %>% nrow())
              , accuracy=1
            )
          )
        )
      , by = dplyr::join_by(ref_species_symbol==species_symbol)
    ) %>% 
    # add on column totals
    dplyr::bind_rows(
      cm_df_temp %>% 
        dplyr::distinct(ref_species_symbol) %>% 
        dplyr::left_join(
          species_accuracy_df_lab_temp %>% 
            dplyr::select(species_symbol,tot_pred_n)
          , by = dplyr::join_by(ref_species_symbol==species_symbol)
        ) %>% 
        tidyr::pivot_wider(names_from = ref_species_symbol, values_from = c(tot_pred_n)) %>% 
        dplyr::mutate(ref_species_symbol = "Total Predicted") %>% 
        dplyr::relocate(ref_species_symbol) %>% 
        dplyr::mutate(
          NONE = scales::comma(
              (w_species_temp %>% dplyr::filter(pred_species_symbol=="NONE") %>% nrow())
              , accuracy=1
            )
          # diagonal totals
          , tot_ref_n = scales::comma(
              (w_species_temp %>% nrow())
              , accuracy=1
            )
        )
    ) %>% 
    # add on PA
    dplyr::bind_rows(
      cm_df_temp %>% 
        dplyr::distinct(ref_species_symbol) %>% 
        dplyr::left_join(
          species_accuracy_df_lab_temp %>% 
            dplyr::select(species_symbol,users_accuracy)
          , by = dplyr::join_by(ref_species_symbol==species_symbol)
        ) %>% 
        tidyr::pivot_wider(names_from = ref_species_symbol, values_from = c(users_accuracy)) %>% 
        dplyr::mutate(ref_species_symbol = "User's Accuracy") %>% 
        dplyr::relocate(ref_species_symbol) %>% 
        # add diagonals and overall accuracy
        dplyr::mutate(
          tot_ref_n = as.character(NA)
          , producers_accuracy = scales::number(overall_accuracy_temp[1]*100, accuracy = 0.1, prefix = "<b>OA: ", suffix = "%</b>")
        )
    ) %>% 
    dplyr::rename(
      `Total Reference` = tot_ref_n
      , `Producer's Accuracy` = producers_accuracy
    ) %>% 
    dplyr::mutate(
      grp = ifelse(
        dplyr::row_number()<=(dplyr::n()-2)
        , "Reference"
        , ""
      )
    ) %>% 
    dplyr::relocate(grp)
  
  # kbl_table_df_temp
  
  if(use_only_tp){
    kbl_table_df_temp <- kbl_table_df_temp %>% dplyr::select(-NONE) # none is only for FP and FN
  }
  
  
  ## table that
  kbl_html_table_temp <- 
    kbl_table_df_temp %>% 
    kableExtra::kbl(
      col.names = c(
        "."
        , ""
        , names(kbl_table_df_temp)[3:ncol(kbl_table_df_temp)]
      )
      , escape = F
    ) %>% 
    kableExtra::kable_styling(font_size = 11.5) %>% 
    # header
    kableExtra::add_header_above(
      c(
        " " = 2
        , "Predicted" = (ncol(kbl_table_df_temp)-4)
        , " " = 2
      )
    ) %>% 
    # font
    kableExtra::column_spec(1:2, bold = T) %>% 
    kableExtra::column_spec(
      1
      , width = "4px"
      , extra_css = "transform: rotate(-90deg);padding-left: 0.5px; padding-right: 0.5px;"
    ) %>% 
    # borders
    kableExtra::row_spec(
      c(nrow(kbl_table_df_temp) - 2, nrow(kbl_table_df_temp) - 1, nrow(kbl_table_df_temp))
      # , border = T, 
      , extra_css = "border-bottom: 3px solid black !important;"
    ) %>%
    kableExtra::column_spec(
      c(ncol(kbl_table_df_temp) - 2, ncol(kbl_table_df_temp) - 1) #, ncol(kbl_table_df))
      , border_right = T
      , include_thead = T
      , extra_css = "border-right: 3px solid black !important;"
    ) %>% 
    # ref lab
    kableExtra::collapse_rows(columns = 1, valign = "middle")
  
  ##########
  if(use_genus){
    w_species_temp <- 
      w_species_temp %>% 
      dplyr::mutate(
        pred_species_symbol = xxx_pred
        , ref_species_symbol = xxx_ref
      ) %>% 
      dplyr::select(-c(xxx_ref,xxx_pred))
  }
  return(list(
    pred_ref_match_w_species = w_species_temp
    , overall_accuracy = overall_accuracy_temp
    , species_accuracy_df = species_accuracy_df_temp
    , kbl_html_table = kbl_html_table_temp
  ))
}

# test that
confusion_matrix_species_temp <- confusion_matrix_tree_class_fn(
  pred_ref_match = pred_ref_tree_match_list$USFS_ManitouEF_N1
  , pred_trees = pred_trees_list$USFS_ManitouEF_N1
  , ref_trees = ref_trees_list$USFS_ManitouEF_N1
  , use_genus = F
)
confusion_matrix_genus_temp <- confusion_matrix_tree_class_fn(
  pred_ref_match = pred_ref_tree_match_list$USFS_ManitouEF_N1
  , pred_trees = pred_trees_list$USFS_ManitouEF_N1
  , ref_trees = ref_trees_list$USFS_ManitouEF_N1
  , use_genus = T
)
```

let's check out the confusion matrix at the *species* level

```{r, results='asis'}
confusion_matrix_species_temp$kbl_html_table
```

and the confusion matrix at the *genus* level

```{r, results='asis'}
confusion_matrix_genus_temp$kbl_html_table
```

#### Site-Level Accuracy{#06_site_acc}

we shift away from individual item matching and use the Bray-Curtis Dissimilarity Index. This metric compares the proportional compositions of each species predicted by the model against the reference proportions within a defined area or site.

Instead of evaluating trees one by one, this accuracy evaluation aggregates the species counts or proportions across an entire plot or forest stand. The Bray-Curtis dissimilarity index is used to compare the predicted species composition against the observed field composition. The index returns a value between 0 and 1, where 0 indicates identical species compositions and 1 indicates completely non-overlapping communities. Because Bray-Curtis is a dissimilarity measure, it can be interpreted as: 

$$\textrm{Bray-Curtis}_{dissimilarity} \approx 1-\textrm{Bray-Curtis}_{similarity}$$

The Bray-Curtis index is a widely trusted metric in [community ecology](https://vegandevs.github.io/vegan/reference/vegdist.html). Dissimilarity/similarity metrics tell us how well the proposed method captures the overall ecological community structure. Even if individual tree predictions are misaligned, the proposed method may still capture the correct stand-level abundance proportions, which can still be valuable for management.

*What it means for this evaluation*: Even if individual-level classification accuracy is low, the model may still achieve acceptable aggregate composition accuracy. 
*Real-World Ecological Application*: This aggregate evaluation is valuable for large-scale forest structure measurement including: timber management, carbon and fuel assessments, and landscape ecology patterns. If a forestry manager only needs to know the total volume, fuel loading, or biomass ratios of species across a broad area or management unit to write a management plan, they don't care if individual tree predictions are slightly off. They care about aggregate community structure which makes Bray-Curtis (and other similiarity/dissimilarity metrics) more relevant than overall, producer's, and user's accuracy.


let's make a function to use the full prediction and reference data irrespective of instance matching status to calculate the aggregated species composition metrics

```{r}
similarity_tree_class_fn <- function(
  pred_trees
  , ref_trees
  , use_genus = F
) {
  

  # join species lookup and compare distributions
  species_comp_temp <-
    dplyr::tibble(
      species_symbol = c(ref_trees$species_symbol, pred_trees$species_symbol) %>% unique()
    ) %>% 
    # join on species data
    # add on species info
    dplyr::left_join(
      fia_species_lookup %>% 
        dplyr::select(-spcd) %>% 
        dplyr::rename_with(
          .cols = -c(species_symbol)
          , .fn = ~paste0("species_",.x,recycle0 = T)
        )
        , by = "species_symbol"
      ) %>% 
      # clean up missing
      dplyr::mutate(
        dplyr::across(
          tidyselect::starts_with("species_") & !c("species_symbol")
          , .fns = ~dplyr::case_when(
            is.na(species_symbol) ~ "NONE"
            , is.na(.x) ~ "UNKNOWN"
            , T ~ .x
          )
        )
        , species_symbol = dplyr::case_when(
          is.na(species_symbol) ~ "NONE"
          , is.na(species_symbol) ~ "UNKNOWN"
          , T ~ species_symbol
        )
    ) %>% 
    # count reference trees
    dplyr::left_join(
      ref_trees %>% 
        sf::st_drop_geometry() %>%
        dplyr::ungroup() %>% 
        dplyr::count(species_symbol) %>% 
        dplyr::mutate(pct=n/sum(n)) %>% 
        dplyr::rename(
          ref_n=n
          , ref_pct=pct
        )
      , by = "species_symbol"
    ) %>% 
    # count pred trees
    dplyr::left_join(
      pred_trees %>% 
        sf::st_drop_geometry() %>%
        dplyr::ungroup() %>% 
        dplyr::count(species_symbol) %>% 
        dplyr::mutate(pct=n/sum(n)) %>% 
        dplyr::rename(
          pred_n=n
          , pred_pct=pct
        )
      , by = "species_symbol"
    ) %>% 
    dplyr::mutate(
      dplyr::across(
        dplyr::where(is.numeric)
        , ~ dplyr::coalesce(.x,0)
      )
    )
  # species_comp_temp
  
  if(use_genus){
    species_comp_temp <- 
      species_comp_temp %>% 
      dplyr::group_by(species_genus) %>% 
      dplyr::summarise(
        dplyr::across(
          dplyr::where(is.numeric)
          , ~ sum(.x,na.rm=T)
        ) 
      ) %>% 
      dplyr::ungroup()
  }
  # species_comp_temp
  
  # calc intersecting shared area per species (Schoener's Overlap index component)
  species_comp_temp <- 
    species_comp_temp %>% 
    dplyr::mutate(
      # for similarity index : Schoener's D (total overlap) Index
      overlap_pct = pmin(pred_pct, ref_pct)
      # for rmse : sqrt(mean(species_comp_temp$sq_pct_error))
      , sq_pct_error = (pred_pct - ref_pct)^2
    )
  
  # species_comp_temp
  
  if(use_genus){
    kbl_df_temp <- species_comp_temp %>% dplyr::rename(species_symbol=species_genus)
    cap_temp <- "Taxonomic Genus Proportional Comparision"
    head_temp <- "Genus Proportions and Tree Counts"
  }else{
    kbl_df_temp <- species_comp_temp
    cap_temp <- "Taxonomic Species Proportional Comparision"
    head_temp <- "Species Proportions and Tree Counts"
  }
  # table the predicted versus reference counts by species
  kbl_html_table_temp <-
    kbl_df_temp %>% 
    dplyr::mutate(diff_pct = pred_pct-ref_pct) %>% 
    dplyr::mutate(
      # # one column for diff/pct
      diff_pct = scales::percent(diff_pct,accuracy=0.1)
      , pred_pct = paste0(scales::percent(pred_pct,accuracy=0.1), "<br />(", scales::comma(pred_n,accuracy=1), ")")
      , ref_pct = paste0(scales::percent(ref_pct,accuracy=0.1), "<br />(", scales::comma(ref_n,accuracy=1), ")")
    ) %>% 
    dplyr::select(species_symbol, pred_pct, ref_pct, diff_pct) %>% 
    tidyr::pivot_longer(
      cols = c(pred_pct, ref_pct, diff_pct)
      , names_to = "dta"
      , values_to = "prop"
    ) %>% 
    dplyr::mutate(
      dta = dplyr::recode_values(
        dta
        , "pred_pct" ~ "predicted"
        , "ref_pct" ~ "reference"
        , "diff_pct" ~ "difference<br />(predicted-reference)"
      )
    ) %>% 
    tidyr::pivot_wider(
      names_from = species_symbol
      , values_from = prop
    ) %>% 
    dplyr::arrange(ifelse(stringr::str_starts(dta,"difference"),1,0), desc(dta)) %>% 
    dplyr::rename(`data`=dta) %>% 
    kableExtra::kbl(
      caption = paste0(
        cap_temp # "Taxonomic Proportional Comparision"
        # , "<br />Total Overlap Index: "
        # , sum(species_comp_temp$overlap_pct) %>% scales::percent(accuracy = 0.1 )
        # , "<br />Proportion RMSE: "
        # , sqrt(mean(species_comp_temp$sq_pct_error)) %>% scales::percent(accuracy = 0.1 )
      )
      , escape = F
    ) %>% 
    kableExtra::kable_styling(font_size = 11.5) %>% 
    kableExtra::add_header_above(
      # named numeric vector req'd for dynamic label
      setNames( c(1,nrow(species_comp_temp)), c(" ", head_temp) )
    ) %>% 
    kableExtra::row_spec(
      c(2,3)
      # , border = T, 
      , extra_css = "border-bottom: 3px solid black !important;"
    )
  
  ## dissimilarity index
  # library(vegan)
  
  # format data and pass to vegan::vegdist for bray-curtis dissimilarity
  # ?vegan::vegdist
  # vegan needs a matrix :\
  bc_dissim_temp <-
    kbl_df_temp %>% 
    # relative props
    dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
    tidyr::pivot_longer(cols = c(pred_pct, ref_pct), names_to = "which_dta", values_to = "n") %>% 
    tidyr::pivot_wider(names_from = species_symbol, values_from = n) %>% 
    dplyr::mutate(which_dta = ifelse(which_dta == "pred_pct", "predicted", "reference")) %>% 
    tibble::column_to_rownames(var = "which_dta") %>% 
    # rowSums() # should sum to 1 because proportions
    # # vegan::decostand(method = "total") %>% # if using counts, transform to propotions
    # dissimilarity
    vegan::vegdist(method = "bray") %>% 
    as.numeric()
  # sim
  bc_sim_temp <- 1 - bc_dissim_temp
  # table
  sim_table_temp <- 
    dplyr::tibble(
      metric = c(
        "Bray-Curtis Dissimilarity Index"
        , "Bray-Curtis Similarity"
        # , "Schoener's D (total overlap) Index"
        , "Proportional RMSE"
      )
      , value = c(
        bc_dissim_temp
        , bc_sim_temp
        # , sum(species_comp_temp$overlap_pct) 
        , sqrt(mean(species_comp_temp$sq_pct_error,na.rm=T))
      )
      , metric_desc = c(
        paste0("proportion of relative ", ifelse(use_genus,"genus","species")," distribution mis-matched")
        , "proportion of total distribution matching perfectly"
        # , "equal to Bray-Curtis similarity for proportions"
        , paste0("average percentage deviation error margin per ", ifelse(use_genus,"genus","species"))
      )
    )
  kbl_html_sim_table_temp <- 
    sim_table_temp %>% 
    dplyr::mutate(dplyr::across(
      dplyr::where(is.numeric)
      , ~scales::percent(.x,accuracy=0.1)
    )) %>% 
    kableExtra::kbl(
      caption = "Predicted vs Reference Compositional Accuracy Metrics"
      , col.names = c("","value","interpretation")
    ) %>% 
    kableExtra::kable_styling(font_size = 12) %>% 
    kableExtra::column_spec(1:2, bold = TRUE)
  
  return(list(
    species_comp_df = species_comp_temp
    , kbl_html_comp_table = kbl_html_table_temp
    , sim_table_df = sim_table_temp
    , kbl_html_sim_table = kbl_html_sim_table_temp
  ))
}

# test that
similarity_tree_species_temp <- similarity_tree_class_fn(
  pred_trees = pred_trees_list$USFS_ManitouEF_N1
  , ref_trees = ref_trees_list$USFS_ManitouEF_N1
  , use_genus = F
)
similarity_tree_genus_temp <- similarity_tree_class_fn(
  pred_trees = pred_trees_list$USFS_ManitouEF_N1
  , ref_trees = ref_trees_list$USFS_ManitouEF_N1
  , use_genus = T
)

```

let's check out the proportional comparisons at the *species* level

```{r, results='asis'}
similarity_tree_species_temp$kbl_html_comp_table
# similarity_tree_species_temp$species_comp_df %>% dplyr::glimpse()
```

also, the site-wide *species* compositional accuracy metrics

```{r}
similarity_tree_species_temp$kbl_html_sim_table
```

let's check out the proportional comparisons at the *genus* level

```{r, results='asis'}
similarity_tree_genus_temp$kbl_html_comp_table
# similarity_tree_genus_temp$genus_comp_df %>% dplyr::glimpse()
```

also, the site-wide *genus* compositional accuracy metrics

```{r}
similarity_tree_genus_temp$kbl_html_sim_table
```

plot the proportion of total by *species*

```{r}
similarity_tree_species_temp$species_comp_df %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "reference")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "gray90"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_viridis_d(option = "turbo", begin=0.05, end = 0.95) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

plot the proportion of total by *genus*

```{r}
similarity_tree_species_temp$species_comp_df %>% 
  dplyr::group_by(species_genus) %>% 
  dplyr::summarise(dplyr::across(c(pred_pct,ref_pct),sum)) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(species_genus, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_genus
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "reference")
    , species_genus = forcats::fct_reorder(species_genus, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_genus)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_genus, fontface = "bold")
    # , color = "gray90"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_viridis_d(option = "magma", begin = 0.25) +
  ggplot2::labs(
    subtitle = "Genus Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

