# Penobscot Experimental Forest (Maine){#site_04}

Now that we have reviewed the [process and evaluation method](#site_01) and set up functions to standardize and streamline our data processing, let's do it for another site: USFS Manitou Experimental Forest ("N1" study site) 

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Field Data

Field data from this study site is from: 

* Kenefic, Laura S.; Allogio, Jeanette A.; Gerndt, Kathryn M.; Knapp, Rachel A.; Rogers, Nicole S.; Puhlick, Joshua J. 2025. Overstory tree and regeneration data from the "Silvicultural Effects on Composition, Structure, and Growth" study at Penobscot Experimental Forest. 3rd Edition. Fort Collins, CO: Forest Service Research Data Archive. [https://doi.org/10.2737/RDS-2012-0008-3](https://doi.org/10.2737/RDS-2012-0008-3)

>Data were collected on a series of permanent sample plots distributed across management units receiving various silvicultural treatments. Plot data are available in eight separate data sets. Overstory tree and sapling data include tree species, condition (e.g., merchantability status and cause of mortality), and diameter at breast height of living and dead trees, 1952 to 2023...Tree location data include the distance and azimuth from plot center of a subsample of trees, 2003 to 2023. Height and crown radius data include total height, height to crown base, and crown radius measurements for a subsample of live trees, 2003 to 2020 (heights) and 2003 to 2017 (radii)...Permanent sample plot location data include geospatial coordinates for permanent sample plots, 2018 and 2020...Future inventories are planned for this study indefinitely.

> 1/5th Acre Plots (were used to measure) Trees ≥ 4.5 inches DBH (with a) 52.7-foot plot radius

let's load in the plot data first here since we have distance and azimuth from plot center for some trees

```{r}
ft_to_m_temp <- 3.280839895
usfs_penobscotef_plots <- readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_PlotLocationsData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(
    ~ tolower(.x) %>% 
      stringr::str_squish() %>% 
      make.names() %>% 
      stringr::str_replace_all( "\\.{2,}", ".") %>% 
      stringr::str_remove( "\\.$") %>% 
      stringr::str_replace_all( "\\.", "_")
  ) %>% 
  dplyr::filter(
    geodoubt == "N"
    , active == 1
    , !is.na(utm_east)
    , !is.na(utm_north)
  ) %>% 
  dplyr::mutate(
    plot_id = 
      mu %>%
        stringr::str_squish() %>% 
        stringr::str_c("-", as.character(plot))
  ) %>% 
  # spatial that
  sf::st_as_sf(coords = c("utm_east", "utm_north"), crs = 26919, remove = F) %>% 
  dplyr::mutate(
    plot_area_ac = 1/5
    # 1 ha = 2.47105 acres
    , plot_area_ha = plot_area_ac/2.47105
    , plot_area_m2 = plot_area_ha*10000
    , plot_radius_meters = 52.7/ft_to_m_temp
  ) %>% 
  sf::st_buffer(.$plot_radius_meters) %>% 
  dplyr::select(
    plot_id, plot_area_m2, plot_radius_meters
    , latitude, longitude
    , utm_east, utm_north
    , mu,plot
  )
# huh?
# usfs_penobscotef_plots %>% dplyr::count(projection)
# usfs_penobscotef_plots %>% dplyr::glimpse()
# usfs_penobscotef_plots %>% sf::st_drop_geometry() %>% dplyr::count(mu,plot) %>% dplyr::filter(n>1)
# usfs_penobscotef_plots %>% sf::st_drop_geometry() %>% dplyr::count(plot_id) %>% dplyr::filter(n>1)
```

load in the trees

```{r}
usfs_penobscotef_trees <- 
  readr::read_csv(
    "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_TreeLocationData.csv"
    , progress = F
    , show_col_types = F
  ) %>% 
  dplyr::rename_with(
    ~ tolower(.x) %>% 
      stringr::str_squish() %>% 
      make.names() %>% 
      stringr::str_replace_all( "\\.{2,}", ".") %>% 
      stringr::str_remove( "\\.$") %>% 
      stringr::str_replace_all( "\\.", "_")
  ) %>% 
  dplyr::left_join(
    usfs_penobscotef_plots %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(plot_id,mu,plot,plot_utm_north=utm_north,plot_utm_east=utm_east)
    , by = dplyr::join_by(mu,plot)
  ) %>% 
  # convert from dist and azimuth to coords
  dplyr::mutate(
    # dist to m
    dist_m = dist/ft_to_m_temp
    # convert azimuth from degrees to radians
    , az_rad = az * pi / 180
    , tree_utm_east = plot_utm_east + (dist_m * sin(az_rad)),
    , tree_utm_north = plot_utm_north + (dist_m * cos(az_rad))
  ) %>% 
  # filter
  dplyr::filter(
    !is.na(tree_utm_east)
    , !is.na(tree_utm_north)
  ) %>% 
  # what if a tree has multiple locations? ... take the most recent
  dplyr::arrange(mu,plot,treenum,desc(year),desc(month),desc(inv)) %>% 
  dplyr::group_by(mu,plot,treenum) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(plot_id,mu,plot,treenum,tidyselect::starts_with("tree_utm_")) %>% 
  # spatial that
  sf::st_as_sf(coords = c("tree_utm_east", "tree_utm_north"), crs = sf::st_crs(usfs_penobscotef_plots), remove = F)

# read and attach tree information
usfs_penobscotef_trees <-
  usfs_penobscotef_trees %>% 
  dplyr::inner_join(
    readr::read_csv(
        "../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Data/PEF_CompartmentStudy_TreeData.csv"
        , progress = F
        , show_col_types = F
      ) %>% 
      dplyr::rename_with(
        ~ tolower(.x) %>% 
          stringr::str_squish() %>% 
          make.names() %>% 
          stringr::str_replace_all( "\\.{2,}", ".") %>% 
          stringr::str_remove( "\\.$") %>% 
          stringr::str_replace_all( "\\.", "_")
      )
    , by = dplyr::join_by(mu,plot,treenum)
    , relationship = "one-to-many"
  ) %>% 
  dplyr::rename(
    dbh_in = dbh
  ) %>% 
  # dplyr::select(dbh_in, tree_height_ft) %>% summary()
  dplyr::mutate(
    dbh_cm = dbh_in/0.394
    # , tree_height_m = tree_height_ft/3.281
  ) %>% 
  # dplyr::select(dbh_cm, tree_height_m) %>% summary()
  sf::st_set_geometry("geometry") %>% 
  dplyr::mutate(
    tree_x = sf::st_coordinates(geometry)[,1] #lon
    , tree_y = sf::st_coordinates(geometry)[,2] #lat
  ) %>% 
  # make a plot_id that is unique for all so we don't have to join by multiple columns
  # and for all sites we'll just use a plot_id
  dplyr::mutate(
    # plot_id = xxxxxx
    # also, each tree needs a treeID
    treeID = dplyr::row_number() %>% as.character()
    # # also, we need a species_symbol to match with the FIADatabase
    # , species_symbol = toupper(spp) %>% stringr::str_remove_all("\\s")
    
    , usfsspid = ifelse(
      usfsspid < 100
      , str_pad(usfsspid, width = 2, pad = "0")
      , as.character(usfsspid)
    )
  ) %>% 
  # clean
  dplyr::filter(
    !is.na(dbh_cm)
    & sf::st_is_valid(geometry)
    & dbh_cm >= 4.5/0.394 # measured trees >= 4.5 in
    # & !is.na(species_symbol)
    # & species_symbol!=""
  )
  
  
```

unfortunately, the researchers for this project chose to use a custom species code with a custom lookup table. this means that we have to do all of the work to tie the species in this research to a common, standardized data framework used to organize species level information

```{r}
#### now we need to get the species data........
## ....................................................
###                    :|

# what is this custom lookup table?
# oh, it's in a pdf???? so.great.
library(pdftools)
# read
species_lookup_temp <- 
  pdftools::pdf_text("../data/USFS_PenobscotEF/USFS_PenobscotEF_trees/Supplements/PEF_CompartmentStudy_SpeciesCodes.pdf") %>% 
  .[1] %>% 
  # get the unique lines of the pdf
  stringr::str_split("\n") %>% 
  purrr::pluck(1) %>% 
  # keep lines that begin with number
  purrr::keep(
    \(x) 
    # regular expression that matches a string starting with 
    # two, three, or four digits, followed by either a tab character or multiple spaces
    stringr::str_detect(stringr::str_trim(x), "^[0-9]{2,4}(\t| {2,})")
  ) %>% 
  # table that
  tibble::tibble() %>% 
  setNames(c("line")) %>% 
  dplyr::mutate(
    line = stringr::str_trim(line)
    , usfsspid = stringr::str_extract(line, "^[0-9]{2,4}?")
    # remaining, then next col
    , rest = stringr::str_remove(line, paste0("^",usfsspid)) %>% stringr::str_trim()
    , common_name = rest %>% 
      # everything from the beginning of a string up until (but not including) a tab or multiple spaces
      stringr::str_extract("^.*?(?=\\t| {2,})")
    # remaining, then next col
    , rest = stringr::str_remove(rest, paste0("^",common_name)) %>% stringr::str_trim()
    , sci_name = rest %>% 
      # everything from the beginning of a string up until (but not including) a tab or multiple spaces
      stringr::str_extract("^.*?(?=\\t| {2,})")
  ) %>% 
  #thats all we want
  dplyr::select(
    usfsspid, common_name, sci_name
  ) %>% 
  dplyr::mutate(
    dplyr::across(dplyr::where(is.character), stringr::str_squish)
    , genus = stringr::word(sci_name,1)
    , species = 
      stringr::word(sci_name,2) %>% 
      stringr::str_remove("spp.$") %>% 
      dplyr::na_if("")
  )
# species_lookup_temp %>% dplyr::glimpse()
# match with the actual FIA lookup
species_lookup_temp <- 
  species_lookup_temp %>% 
  dplyr::mutate(
    join_species = species %>% tolower() %>% stringr::str_squish()
    , join_genus = genus %>% tolower() %>% stringr::str_squish()
  ) %>% 
  dplyr::left_join(
    fia_species_lookup %>% 
      dplyr::mutate(
        join_species = species %>% tolower() %>% stringr::str_squish()
        , join_genus = genus %>% tolower() %>% stringr::str_squish()
        , sym_len = stringr::str_length(species_symbol)
      ) %>% 
      dplyr::select(spcd,species_symbol,join_genus,join_species,sym_len) %>% 
      dplyr::filter(!is.na(join_genus), !is.na(join_species)) %>% 
      dplyr::arrange(join_genus,join_species,sym_len) %>% 
      dplyr::group_by(join_genus,join_species) %>% 
      dplyr::slice(1) %>% 
      dplyr::select(-sym_len) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(join_genus,join_species)
    , relationship = "many-to-one"
  ) %>% 
  # for the missings...we'll manual it
  dplyr::mutate(
    species_symbol = dplyr::case_when(
      usfsspid == "33" ~ "POTR5" # only other aspen in list
      , usfsspid == "02" ~ "PIGL" # most common based on tree count of spruce
      , usfsspid == "36" ~ "QURU" # https://research.fs.usda.gov/treesearch/45567
      , usfsspid == "23" ~ "ULRU" # https://research.fs.usda.gov/treesearch/45567
      , T ~ species_symbol
    )
  ) %>% 
  dplyr::group_by(usfsspid) %>% 
  dplyr::summarise(species_symbol=dplyr::first(species_symbol)) %>% 
  dplyr::ungroup()
# species_lookup_temp

### adddd it
# for each tree, use only the most recent species
# ...on the assumption that the data collection process has improved over time
usfs_penobscotef_trees <-
  usfs_penobscotef_trees %>% 
  dplyr::select(-usfsspid) %>% 
  dplyr::inner_join(
    usfs_penobscotef_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::filter(
        !is.na(dplyr::na_if(usfsspid,""))
        , !is.na(year)
        , !is.na(month)
      ) %>% 
      dplyr::arrange(plot_id,treenum,desc(year),desc(month)) %>% 
      dplyr::group_by(plot_id,treenum) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup() %>% 
      dplyr::select(plot_id,treenum,usfsspid)
    , by = dplyr::join_by(plot_id,treenum)
    , relationship = "many-to-one"
  ) %>% 
  dplyr::left_join(species_lookup_temp, by = "usfsspid") %>% 
  dplyr::filter(
    !is.na(species_symbol)
    & species_symbol!=""
  ) %>% 
  dplyr::select(plot_id, treeID, tree_x, tree_y, species_symbol, dbh_cm, inv, month, year)

# huh?
usfs_penobscotef_trees %>% dplyr::glimpse()
# only keep plots we have trees for
usfs_penobscotef_plots <- usfs_penobscotef_plots %>% 
  dplyr::inner_join(
    usfs_penobscotef_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(plot_id)
    , by = dplyr::join_by(plot_id)
  )

# save to reference list
# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(data_site_nm)
ref_plots_list[[nm_temp]] <- usfs_penobscotef_plots
ref_trees_list[[nm_temp]] <- usfs_penobscotef_trees

```

let's check those field trees out to orient ourselves to the location

```{r}
usfs_penobscotef_trees %>% 
  dplyr::inner_join(
    usfs_penobscotef_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::arrange(plot_id,desc(year),desc(month)) %>% 
      dplyr::group_by(plot_id) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup() %>% 
      dplyr::select(plot_id,inv)
    , by = dplyr::join_by(plot_id,inv)
  ) %>% 
  mapview::mapview(
    zcol = "species_symbol"
    , layer.name="USFS Penobscot EF: field trees"
    , popup=F
  )
# usfs_penobscotef_trees %>%
#   sf::st_union() %>%
#   sf::st_bbox() %>%
#   sf::st_as_sfc() %>%
#   sf::st_centroid() %>%
#   sf::st_transform(crs=4326) %>%
#   sf::st_coordinates()
# # # 44.85681,-68.62513
```

what is the distribution of species?

```{r}
usfs_penobscotef_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::inner_join(
    usfs_penobscotef_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::arrange(plot_id,desc(year),desc(month)) %>% 
      dplyr::group_by(plot_id) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup() %>% 
      dplyr::select(plot_id,inv)
    , by = dplyr::join_by(plot_id,inv)
  ) %>% 
  dplyr::count(species_symbol) %>% 
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Point Cloud Data

we need to identify the most recent survey for the majority of the plots so that we can try to find the most chronologically proximal ALS data
 
```{r}
xxxxxxxxx
```


data was collected over the study site via a UAS-lidar flight at an altitude of 122 m (400 ft) and speed of 8.9 meters per second (20 miles per hour)

let's check out the raw point cloud data

```{r}
usfs_penobscotef_las_f_temp <- "c:/data/usfs/uas_lidar_n1/data/N1_400AGL_20MPH_TFOFF/"
usfs_penobscotef_las_ctg_temp <- lidR::readLAScatalog(usfs_penobscotef_las_f_temp)
usfs_penobscotef_las_ctg_temp
# usfs_penobscotef_las_ctg_temp@data %>% dplyr::glimpse()
```

let's check out the lidar data coverage compared to the field plot data

```{r}
usfs_penobscotef_las_ctg_temp@data$geometry %>% 
  sf::st_union() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(data = "UAS-lidar data extent") %>% 
mapview::mapview(
 zcol = "data"
  , layer.name="USFS Penobscot EF: UAS-lidar data"
  , popup=F
) +
mapview::mapview(
  usfs_penobscotef_trees
  , zcol = "species_symbol"
  , layer.name="USFS Penobscot EF: field trees"
  , popup=F
)
```

use `cloud2trees` to process the raw point cloud data to generate a CHM for use in tree detection

```{r}
################################################
# input/output data str
################################################
# outdir
c2t_output_dir_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(data_dir)
c2t_process_dir_temp <- file.path(c2t_output_dir_temp, "point_cloud_processing_delivery")
# timer file
c2t_tracking_fnm_temp <- file.path(c2t_process_dir_temp, "processed_tracking_data.csv")
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    c2t_process_dir = ifelse(data_dir==c2t_output_dir_temp, c2t_process_dir_temp, c2t_process_dir)
    , c2t_tracking_fnm = ifelse(data_dir==c2t_output_dir_temp, c2t_tracking_fnm_temp, c2t_tracking_fnm)
  )
# project_data_df %>% print(n=nrow(project_data_df))

# get point cloud density and raster resolutions
get_ptcld_density_chm_res_temp <- get_ptcld_density_chm_res_fn(usfs_penobscotef_las_ctg_temp)
# get_ptcld_density_chm_res_temp

# fnm
chm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("chm_",get_ptcld_density_chm_res_temp$chm_res_m,"m.tif") )
dtm_fnm_temp <- file.path(c2t_process_dir_temp, paste0("dtm_",get_ptcld_density_chm_res_temp$dtm_res_m,"m.tif") )
# save to tracking df
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    las_pts_m2 = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$pts_m2, las_pts_m2) # las_pts_m2
    , chm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$chm_res_m, chm_res_m) # chm_res_m
    , chm_fnm = ifelse(data_dir==c2t_output_dir_temp, chm_fnm_temp, chm_fnm) # chm_fnm
    , dtm_res_m = ifelse(data_dir==c2t_output_dir_temp, get_ptcld_density_chm_res_temp$dtm_res_m, dtm_res_m) # dtm_res_m
    , dtm_fnm = ifelse(data_dir==c2t_output_dir_temp, dtm_fnm_temp, dtm_fnm) # dtm_fnm
  )

# project_data_df %>% dplyr::filter(data_dir==c2t_output_dir_temp) %>% dplyr::glimpse()

##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists(chm_fnm_temp)
  || !file.exists(dtm_fnm_temp)
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir_temp
    , input_las_dir = usfs_penobscotef_las_f_temp
    , accuracy_level = 2
    , keep_intrmdt = F
    , dtm_res_m = get_ptcld_density_chm_res_temp$dtm_res_m # (1/sqrt(pts_m2_temp))*2
    , chm_res_m = get_ptcld_density_chm_res_temp$chm_res_m # 1/sqrt(pts_m2_temp)
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm_temp
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( dtm_fnm_temp )
  chm_temp <- terra::rast( chm_fnm_temp )
  
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

we'll set the minimum CHM resolution to `r min_chm_res` m to match the `cloud2trees` default. if the automatically-determined CHM resolution based on the point cloud density used in the `cloud2trees::cloud2raster()` was lower than this level, we'll aggregate the raster using `terra::aggregate()` and `terra::resample()` to get the exact target resolution. if the determined CHM resolution was higher than this level due to low point density, we will use the CHM as-is since this is already the finest resolution possible with the point cloud density.

let's aggregate the CHM for this site and put it into a list of CHM rasters that we can map over in all of the processing

```{r}
agg_chm_rast_temp <- agg_chm_rast_fn(
  chm = cloud2raster_ans$chm_rast
  , res = min_chm_res
  , dir = 
    project_data_df %>% 
    dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
    dplyr::pull(c2t_process_dir)
)
# agg_chm_rast_temp
# agg_chm_rast_temp %>% terra::clamp(lower = 1.37, values = F) %>% terra::plot()

# add to the chm list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(data_site_nm)
chm_list[[nm_temp]] <- agg_chm_rast_temp
# chm_list
```

let's check out the CHM

```{r}
# plot to check out the fine-resolution CHM raster
chm_list$USFS_PenobscotEF %>% 
  terra::plot(
    col = viridis::plasma(n=100)
    , axes=F
    , plg = list(
      title = "CHM (m)"
      , cex = 0.8
      , title.cex = 0.9
      , shrink = 0.6 
    )
  )
```

and convert the reference trees and plot data to the point cloud data projection

```{r}
ref_plots_list$USFS_PenobscotEF <- ref_plots_list$USFS_PenobscotEF %>% sf::st_transform(terra::crs(chm_list$USFS_PenobscotEF))
ref_trees_list$USFS_PenobscotEF <- ref_trees_list$USFS_PenobscotEF %>% 
  sf::st_transform(terra::crs(chm_list$USFS_PenobscotEF)) %>% 
  dplyr::mutate(
    tree_x = sf::st_coordinates(geometry)[,1] #lon
    , tree_y = sf::st_coordinates(geometry)[,2] #lat
  )
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(usfs_penobscotef_plots, usfs_penobscotef_trees)
gc()
```

## Tree Detection

`cloud2trees` implements a CHM-based individual tree detection method accessible via `cloud2trees::raster2trees()`. this method requires a window function used to set the search radius used to find local maxima in the CHM. These local maxima represent individual tree top points and are subsequently used to delineate individual tree crowns via a marker-controlled watershed segmentation

because the objective of this study is to evaluate a species/genera classification method, we just want the best ITD results we can get based on the field reference data. We are not interested in evaluating tree segmentation here, that has already been done across varied input data and forest types.

use our function to clip the CHM to the plot plus a buffer and run `cloud2trees::raster2trees()` for each window function. the return should be a spatial tree list unique by function.

get the individual tree predictions for each window function for the area within the field reference plots with a buffer

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, ws_pred_trees_fnm)
  )

if(!file.exists(fnm_temp)){
  usfs_penobscotef_ws_pred_trees <-
    1:nrow(ref_plots_list$USFS_PenobscotEF) %>%
    purrr::map(
      \(rn)
      ws_testtest_fn(
        chm = chm_list$USFS_PenobscotEF
        , buffer = (competition_radius_m+3) # since we only have one plot we're just going to go for it
        , plot_sf =
          ref_plots_list$USFS_PenobscotEF %>%
          dplyr::ungroup() %>%
          dplyr::slice(rn) %>%
          sf::st_union()
        # a named list of functions ex: cloud2trees::itd_ws_functions()
        , ws_fn_list = get_ws_fn_list_ans # cloud2trees::itd_ws_functions() #
      )
    ) %>%
    purrr::list_rbind(names_to = "plot_dta_idx")
  
  # add plot id
  usfs_penobscotef_ws_pred_trees <- 
    usfs_penobscotef_ws_pred_trees %>% 
    dplyr::left_join(
      ref_plots_list$USFS_PenobscotEF %>% 
        dplyr::mutate(plot_dta_idx=dplyr::row_number()) %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(plot_dta_idx,plot_id)
      , by = "plot_dta_idx"
    ) %>% 
    sf::st_set_geometry("geometry")
  
  # usfs_penobscotef_ws_pred_trees %>% sf::st_crs()
  # usfs_penobscotef_ws_pred_trees %>% dplyr::glimpse()
  # usfs_penobscotef_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_fn_nm)
  
  # get dbh
  usfs_penobscotef_ws_pred_trees_dbh_temp <- 
    usfs_penobscotef_ws_pred_trees %>% 
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number() %>% as.character()
    ) %>% 
    cloud2trees::trees_dbh()
  # usfs_penobscotef_ws_pred_trees_dbh_temp %>% dplyr::glimpse()
  # add to trees
  usfs_penobscotef_ws_pred_trees <-
    usfs_penobscotef_ws_pred_trees %>% 
    dplyr::mutate(
      join_id = dplyr::row_number() %>% as.character()
    ) %>% 
    dplyr::left_join(
      usfs_penobscotef_ws_pred_trees_dbh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::mutate(
          join_id = as.character(treeID)
        ) %>% 
        dplyr::select(
          join_id, tidyselect::starts_with("fia_est_dbh"), dbh_cm
        )
      , by = "join_id"
    )
  # usfs_penobscotef_ws_pred_trees %>% dplyr::glimpse()
  
  # # plot real quick
  # usfs_penobscotef_ws_pred_trees %>% 
  #   sf::st_drop_geometry() %>% 
  #   dplyr::slice_sample(n = 9999) %>% 
  #   ggplot2::ggplot(mapping = ggplot2::aes(x=tree_height_m,y=dbh_cm)) + 
  #   ggplot2::geom_point() +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_lower),color="gray") +
  #   ggplot2::geom_point(mapping = ggplot2::aes(y=fia_est_dbh_cm_upper),color="gray")

  # save
  sf::st_write(usfs_penobscotef_ws_pred_trees, dsn = fnm_temp, quiet = T, append = F)
}else{
  usfs_penobscotef_ws_pred_trees <- sf::st_read(fnm_temp, quiet = T)
}
usfs_penobscotef_ws_pred_trees <- usfs_penobscotef_ws_pred_trees %>% 
  dplyr::mutate(
    ws_type = 
      ws_fn_nm %>% 
      # removes all numeric digits
      stringr::str_remove_all("[0-9]+") %>% 
      stringr::str_replace_all("_"," ") %>% 
      stringr::str_squish() %>% 
      as.factor() %>% 
      forcats::fct_inorder(ordered = T)
    , treeID = as.character(treeID)
  )
# usfs_penobscotef_ws_pred_trees %>% dplyr::glimpse()
# usfs_penobscotef_ws_pred_trees %>% sf::st_drop_geometry() %>% dplyr::count(ws_type)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's check out one example field plot and a few different window function tree detection and crown delineation results

```{r, fig.height=10}
# plot
p_temp <- ref_plots_list$USFS_PenobscotEF %>% dplyr::slice(1)

if(as.numeric(sf::st_area(p_temp))>2000){
  p_temp <- p_temp %>% 
    sf::st_centroid() %>% 
    sf::st_buffer(sqrt(2000)/2) %>% 
    sf::st_bbox() %>% 
    sf::st_as_sfc() %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(plot_id = p_temp$plot_id[1])
}
# some trees
t_temp <- 
  usfs_penobscotef_ws_pred_trees %>% 
  dplyr::filter(plot_id==p_temp$plot_id[1]) %>% 
  sf::st_intersection(p_temp) %>% 
  dplyr::inner_join(
    usfs_penobscotef_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm,ws_type) %>% 
      dplyr::slice_sample(n = 3, by = ws_type) %>% 
      dplyr::select(ws_fn_nm)
    , by = "ws_fn_nm"
  )
chm_temp <- chm_list$USFS_PenobscotEF %>% 
  terra::crop(
    t_temp %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      terra::vect()
    , mask = T
  )
# terra::plot(chm_temp)
plt_temp <-
  chm_temp %>% 
  terra::clamp(lower = 1.37, values = F) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x=x,y=y,fill=f)
    , color = NA
    , lwd = 0
  ) +
  ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent", name = "CHM (m)") +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top")
plt_temp +
  ggplot2::geom_sf(
    data = p_temp %>% sf::st_transform(terra::crs(chm_temp))
    , fill = NA
    , color = "black"
    , lwd = 0.8
  ) + 
  ggplot2::geom_sf(
    data = t_temp
    , mapping = ggplot2::aes(color = ws_type)
    , fill = NA
    , lwd = 0.7
  ) +
  ggplot2::scale_color_viridis_d() +
  ggplot2::facet_wrap(facets = dplyr::vars(ws_fn_nm), ncol = 3) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold")
    , strip.background = ggplot2::element_rect(color = "black")
    , panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  ggplot2::guides(
    color = "none"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now we need a function to perform instance matching of the predictions and reference data. to keep things simple, since all the reference field data includes point locations, we'll perform matching based on a spatial intersection of these points with the predicted tree crowns. if multiple tree points fall within a single crown, we'll select the tree that is closest in terms of size based on whatever field data is available (height or DBH). One predicted tree will only be matched with one field reference tree and vice-versa.

```{r}
##################################################################
##################################################################
##################################################################
# plot_iter_instance_match_fn
# loop over plots and ws functions for instance match
##################################################################
##################################################################
##################################################################
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"ws_instance_match.csv")
if(!file.exists(fnm_temp)){
  usfs_penobscotef_ws_instance_match <- 
    plot_iter_instance_match_fn(
      ref_plots = ref_plots_list$USFS_PenobscotEF
      , ref_pts = ref_trees_list$USFS_PenobscotEF
      , pred_polys = usfs_penobscotef_ws_pred_trees
      , min_ht_m = 2
      , min_dbh_cm = NA
    )
  readr::write_csv(usfs_penobscotef_ws_instance_match, file = fnm_temp, append = F, progress = F)
}else{
  usfs_penobscotef_ws_instance_match <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# usfs_penobscotef_ws_instance_match %>% dplyr::glimpse()
##################################################################
##################################################################
##################################################################
# agg_ground_truth_match
# loop over plots and ws functions for accuracy eval
##################################################################
##################################################################
##################################################################
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"agg_ground_truth_match_ans.csv")
if(!file.exists(fnm_temp)){
  # ?purrr::pmap
  # A data frame is an important special case of .l. It will cause .f to be called once for each row.
  usfs_penobscotef_agg_ground_truth_match_ans <- 
    usfs_penobscotef_ws_instance_match %>% 
    dplyr::distinct(plot_id,ws_fn_nm) %>% 
    purrr::pmap(function(plot_id, ws_fn_nm) {
      # the unique row data
      d <- dplyr::tibble(plot_id = plot_id, ws_fn_nm = ws_fn_nm)
      # filter full set based on the data
      j <- 
        usfs_penobscotef_ws_instance_match %>% 
        dplyr::inner_join(
          d   
          , by = dplyr::join_by("plot_id","ws_fn_nm")
        )
      # agg_ground_truth_match
      d %>% 
        dplyr::bind_cols(agg_ground_truth_match(j))
      
    }) %>% 
    list_rbind()
  readr::write_csv(usfs_penobscotef_agg_ground_truth_match_ans, file = fnm_temp, append = F, progress = F)
}else{
  usfs_penobscotef_agg_ground_truth_match_ans <- readr::read_csv(file = fnm_temp, show_col_types = F, progress = F)
}
# huh?
# wfsi_eglinafb_agg_ground_truth_match_ans %>% dplyr::glimpse()
```

aggregate the plot level data to the window function to get the answer we've been searching for

```{r}
# usfs_penobscotef_agg_ground_truth_match_ans %>% 
#   dplyr::count(ws_fn_nm) %>% 
#   dplyr::count(n)
agg_temp <- 
  usfs_penobscotef_agg_ground_truth_match_ans %>% 
  dplyr::group_by(ws_fn_nm) %>% 
  dplyr::summarize(
    n_plots = dplyr::n()
    , dplyr::across(
      .cols = c(precision,recall,f_score,tidyselect::ends_with("_rate"))
      , list( 
        mean= ~ mean(.x,na.rm=T)
        , median= ~ median(.x,na.rm=T)
      )
    )
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::left_join(
    usfs_penobscotef_ws_pred_trees %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(ws_fn_nm, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::mutate(
    f_score_rank = dplyr::min_rank(-f_score_mean)
    , ws_fn_nm_fact = forcats::fct_reorder(ws_fn_nm, f_score_rank)
    # , dplyr::across(
    #   .cols = dplyr::where(is.numeric)
    #   , ~ round(.x,2) %>% as.character() %>% as.numeric()
    # )
  ) %>% 
  dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean))
# agg_temp %>% dplyr::glimpse()
```

let's check out the plot-level distribution of F-score for the top window functions (remember this particular data has one "plot", a stem map)

```{r}
usfs_penobscotef_agg_ground_truth_match_ans %>% 
  # add agg data
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::filter(f_score_rank<=12) %>% 
      dplyr::select(ws_fn_nm, ws_fn_nm_fact, ws_type)
    , by = "ws_fn_nm"
  ) %>% 
  dplyr::arrange(ws_fn_nm_fact) %>% 
  dplyr::mutate(ws_fn_nm_fact=forcats::fct_rev(ws_fn_nm_fact)) %>% 
  # dplyr::pull(ws_fn_nm) %>% levels()
  # View()
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = f_score, y = ws_fn_nm_fact, fill=ws_type)
  ) +
  ggplot2::geom_boxplot(
    width = 0.4
    , alpha=0.9
    , outliers = F
    , color = "black"
  ) +
  ggplot2::geom_point(color = "gray77") +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(x="F-score", y="") +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , legend.direction  = "horizontal"
  )
```

we can print the raw window function definition for those that achieved the highest F-score

```{r}
agg_temp %>% 
  dplyr::filter(f_score_rank<=7) %>% 
  dplyr::arrange(f_score_rank) %>% 
  dplyr::pull(ws_fn_nm) %>% 
  purrr::set_names() %>% 
  # purrr::map(\(nm) purrr::pluck(get_ws_fn_list_ans, nm))
  purrr::map(function(nm){
    fn <- purrr::pluck(get_ws_fn_list_ans, nm)
    dplyr::tibble(
      ws_fn_nm = nm
      , def = paste(format(fn), collapse = "\n")
    )
  }) %>%
  purrr::list_rbind() %>% 
  dplyr::inner_join(
    agg_temp %>% 
      dplyr::select(ws_fn_nm, f_score_mean, recall_mean, precision_mean) %>% 
      dplyr::mutate(
        dplyr::across(
          dplyr::where(is.numeric)
          , ~ scales::percent(.x,accuracy = 0.1)
        )
      )
  ) %>% 
  dplyr::relocate(def,.after = dplyr::everything()) %>% 
  kableExtra::kbl(
    caption = "WFSI Eglin Air Force Base top ITD window functions"
    , col.names = c(
      "function<br>name"
      , "F-score<br>(mean)"
      , "recall<br>(mean)"
      , "precision<br>(mean)"
      , "function<br>definition"
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling(font_size = 12, bootstrap_options = c("responsive", "condensed")) %>% # 
  # make r line breaks (\n) and use monospace font for code
  kableExtra::column_spec(5, monospace = T, extra_css = "white-space: pre;")
```

nice, we'll select the best non-constant window function for use in evaluation of the species accuracy. if we were to use the constant functions, even if they are the best performing, we would likely be overfitting to this specific data set. in addition to being more in-line with real-world tree height-to-crown relationships, the variable window functions are likely more broadly applicable

```{r}
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    best_ws_fn_nm = ifelse(
      tolower(data_site_nm)=="usfs_penobscotef"
      , agg_temp %>% 
        dplyr::filter(ws_type!="constant") %>% 
        dplyr::arrange(desc(f_score_mean),desc(f_score_median),desc(recall_mean),desc(precision_mean)) %>% 
        dplyr::slice(1) %>% 
        dplyr::pull(ws_fn_nm)
      , best_ws_fn_nm
    )
  )
# project_data_df %>% dplyr::glimpse()
```

full CHM data processing

```{r}
c2t_process_dir_temp <- 
  project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(c2t_process_dir)
fnm_temp <- file.path(c2t_process_dir_temp,"best_ws_pred_trees.gpkg")
project_data_df <- project_data_df %>% 
  dplyr::mutate(
    best_ws_pred_trees_fnm = ifelse(c2t_process_dir==c2t_process_dir_temp, fnm_temp, best_ws_pred_trees_fnm)
  )

###############################################################################################
###############################################################################################
## raster2trees that best ws fn
###############################################################################################
###############################################################################################
if(!file.exists(fnm_temp)){
  
  ws_temp <- project_data_df %>% 
    dplyr::filter(c2t_process_dir==c2t_process_dir_temp) %>% 
    dplyr::slice(1) %>% 
    dplyr::pull(best_ws_fn_nm)
  
  pred_trees_temp <- 
    usfs_penobscotef_ws_pred_trees %>% 
    dplyr::filter(ws_fn_nm==ws_temp) %>% 
    # this is because we did tree detection by plot
    dplyr::rename(xxxtreeID=treeID) %>% 
    dplyr::mutate(
      treeID = dplyr::row_number() %>% as.character()
    )
  
  # get rh
  pred_trees_temp_rh_temp <- 
    # have to do this by plot so that all trees are not included
    unique(pred_trees_temp$plot_dta_idx) %>% 
    purrr::map(
      \(idx)
      pred_trees_temp %>% 
        dplyr::filter(plot_dta_idx==idx) %>% 
        get_relative_height_fn(radius = competition_radius_m) 
    ) %>% 
    purrr::list_rbind()
  # pred_trees_temp_rh_temp %>% dplyr::glimpse()
  # pred_trees_temp_rh_temp %>% 
  #   ggplot2::ggplot() + 
  #   ggplot2::geom_point(ggplot2::aes(x = tree_height_m, y=rel_tree_height_m)) + 
  #   ggplot2::facet_wrap(facets=dplyr::vars(plot_id))
  
  # add to trees
  pred_trees_temp <- 
    pred_trees_temp %>% 
    dplyr::left_join(
      pred_trees_temp_rh_temp %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(
          plot_dta_idx,treeID,rel_tree_height_m
        )
      , by = dplyr::join_by(plot_dta_idx,treeID)
    )
  # pred_trees_temp %>% dplyr::glimpse()
  
  # save
  sf::st_write(pred_trees_temp, dsn = fnm_temp, quiet = T, append = F)
}else{
  pred_trees_temp <- sf::st_read(fnm_temp, quiet = T)
}
# pred_trees_temp %>% dplyr::glimpse()

###############################################################################################
###############################################################################################
## now we need to do the matching 
# again to only keep trees in plots for individual tree species comparisons
###############################################################################################
###############################################################################################

im_temp <- 
  plot_iter_instance_match_fn(
    ref_plots = ref_plots_list$USFS_PenobscotEF
    , ref_pts = ref_trees_list$USFS_PenobscotEF
    , pred_polys = pred_trees_temp
    , min_ht_m = 2
    , min_dbh_cm = NA
  )

# im_temp %>% dplyr::glimpse()

# add to the ref trees list
nm_temp <- project_data_df %>% 
  dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
  dplyr::pull(data_site_nm)
pred_ref_tree_match_list[[nm_temp]] <- im_temp

###############################################################################################
###############################################################################################
## save to the pred trees
###############################################################################################
###############################################################################################
# let's also save the plot-level ITD predictions for this ws

pred_trees_list[[nm_temp]] <-
  pred_trees_temp %>% 
  # which ws
  dplyr::filter(
    ws_fn_nm %in% (
      project_data_df %>% 
      dplyr::filter(tolower(data_site_nm)=="usfs_penobscotef") %>% 
      dplyr::pull(best_ws_fn_nm)
    )
  ) %>% 
  # filter for trees used in instance matching 
    # which included thresholds used for measuring field trees 
    # and plot spatial alignment
  dplyr::inner_join(
    pred_ref_tree_match_list[[nm_temp]] %>% 
      dplyr::select(plot_id,ws_fn_nm,pred_treeID) %>% 
      dplyr::filter(!is.na(pred_treeID)) %>% 
      dplyr::rename(treeID=pred_treeID)
    , by = dplyr::join_by(plot_id,ws_fn_nm,treeID)
  )
# pred_trees_list
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(list = ls()[grep("^usfs_penobscot",ls())])
remove(cloud2raster_ans)
gc()
```

## Species/Genus Prediction Method

for some context of the primary objective of this analysis, here is what we said in the `cloud2trees` manuscript (a draft of the manuscript which may be different than what was actually published): 

>However, taxonomic differentiation using airborne lidar alone remains one of the most technically challenging aspects of forest inventory workflows (White et al., 2016, 2025; Fassnacht et al., 2024). In practice, genus-level distinction is often more realistic when relying on remote sensing and supplemental spatial data ([Vauhkonen et al., 2014b](https://doi.org/10.1007/978-94-017-8663-8_7)). As a practical, computationally efficient approach, future research could evaluate predicting tree-level species or genus for potential integration into cloud2trees by leveraging supplemental spatial datasets. Testing could evaluate either using tree lists from FIA plots linked through a spatial overlay with TreeMap (Houtman et al., 2025) to train a taxonomic classification model based on structural and competition metrics shared between FIA and cloud2trees tree lists, or by proportionally assigning species using BigMap species-level proportional biomass raster data (Wilson et al., 2012, 2018). Incorporating finer-scale taxonomic classification than the currently implemented FIA forest type groups into the cloud2trees tree list will enable the integration of detailed biomass equations such as the National Scale Volume Biomass system (NSVB; Westfall et al., 2024) and other allometries that require species- or genera-level information (e.g., Réjou‐Méchain et al., 2017; Gonzalez‐Akre et al., 2022).

`cloud2trees` does not currently include functionality to attach species to it's tree list. there are a few paths forward for performing species-specific workflows with the `cloud2trees` tree list:

the proposed method which we are evaluating presently, will predict tree-level species using the TreeMap imputed FIA plot tree list to build a species classification model based on height and tree location to attach tree species to our tree list. The model will inherently account for the proportional distribution of different species. 

early testing of the proposed method used a  bayesian hierarchical model grouped by imputed FIA plot identifier that allows the intercept and the slope of height to vary by plot took quite a while to run for a relatively small area (~20-30 mins). as such, we are going to use a random forest (RF) classifier to enable more computationally efficient classification. `cloud2trees` already implements random forest models to model missing HMD and CBH values that were not successfully extracted from the point cloud and uses a model averaging method that subsamples the training data (to reduce model fit times), estimates the model several times with different subsamples, then model averages the predictions. Because the species is not a metric variable that can be averaged, we'll use the same method but employ a model majority voting system instead. we can do this with some of the `cloud2trees` built-in functionality.

ultimately, if the proposed species/genera classification method achieves decent accuracy results, it could be integrated into the `cloud2trees` framework as `cloud2trees::trees_species()`, for example. this function would take an input spatial tree list and attach the predicted species/genera

### Training Data

load in the TreeMap data and clip to our trees as in `cloud2trees::trees_dbh()`

```{r}
# test to get the treemap data for our aoi
treemap_data_loader_ans <-
  pred_trees_list$USFS_PenobscotEF %>% 
  dplyr::ungroup() %>% 
  sf::st_union() %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  treemap_data_loader()
```

Check out the TreeMap raster of forest types for our area of interest defined using the spatial extent of the predicted trees

```{r}
treemap_data_loader_ans$treemap_rast %>% terra::plot(axes=F)
```

here is the stability of those imputed TreeMap FIA plot id's

```{r}
treemap_data_loader_ans$treemap_prop_rast %>% terra::plot(col = viridis::mako(n=100), axes=F)
```

lets' review the spatial tree list we'll use for training data which comes from the TreeMap FIA imputed plots

what's in the FIA tree list using the TreeMap imputed plots filtered for our area of interest?

```{r}
treemap_data_loader_ans$treemap_trees_spatial_df %>% dplyr::glimpse()
```

species distribution?

```{r}
treemap_data_loader_ans$treemap_trees_spatial_df %>%
  sf::st_drop_geometry() %>%
  dplyr::count(species_symbol) %>%
  dplyr::arrange(desc(n)) %>% 
  dplyr::mutate(
    percent=n/sum(n)
    , percent = scales::percent(percent,accuracy=0.1)
    , n = scales::comma(n,accuracy = 1)
  ) %>%
  kableExtra::kbl(
    caption = "TreeMap Species Distribution over Study Area"
  ) %>% 
  kableExtra::kable_styling()
```

here is the spatial distribution of the species for a sample of the trees

```{r}
samp_temp <- treemap_data_loader_ans$treemap_trees_spatial_df %>%
  dplyr::slice_sample(n=2222)
samp_temp %>%
  ggplot2::ggplot(ggplot2::aes(x=x,y=y,color=species_symbol)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::scale_color_viridis_d(option = "turbo") +
  ggplot2::theme_void()
```

that is difficult to read, let's actually print the species symbol

```{r}
samp_temp %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = x, y = y, fill = species_symbol, color = species_symbol, label = species_symbol)
  ) +
  # ggplot2::geom_hline(yintercept = median(treemap_data_loader_ans$treemap_trees_spatial_df$y) , color = "black") + 
  # ggplot2::geom_vline(xintercept = median(treemap_data_loader_ans$treemap_trees_spatial_df$x), color = "black") + 
  ggplot2::geom_text(size = 2.5) +
  ggplot2::scale_color_viridis_d(option = "turbo") +
  ggplot2::labs(
    x = "x", y = "y"
  ) + 
  ggplot2::theme_void() + 
  ggplot2::theme(
    legend.position = "none"
  )
```

let's check out the `stability_weight` distribution which is unique for each raster cell in the data extent

```{r}
stab_temp <- 
  # stability_weight should be unique by cell
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::distinct(cell,stability_weight) %>% 
  # add on forest type
  dplyr::left_join(
    treemap_data_loader_ans$treemap_rast %>% 
      terra::as.data.frame(xy=T, cells = T)  %>% 
      dplyr::select(cell, ForTypName) %>% 
      dplyr::rename(f_type=ForTypName)
  ) %>% 
  dplyr::mutate(
    f_type = f_type %>% 
      forcats::fct_drop() %>% 
      forcats::fct_reorder(stability_weight,.fun = "median") %>% 
      as.ordered()
  )
stab_temp %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = f_type, x = stability_weight, fill = f_type)
  ) +
  ggplot2::geom_boxplot(
    width = 0.4
    , alpha=0.9
    , outliers = F
    , color = "black"
  ) +
  ggplot2::scale_fill_viridis_d(option = "magma",direction=-1,begin=0.1,end=0.5) +
  ggplot2::labs(y="") +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , legend.direction  = "horizontal"
  )
```

this essentially means that TreeMap is fairly confident that the study area includes `r paste0(levels(forcats::fct_rev(stab_temp$f_type))[1], collapse = " and ")`

lastly, we'll look at the relative height data for a sample of the data

```{r}
focal_cell_temp <- 
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::filter(
    stability_weight>=
      quantile(treemap_data_loader_ans$treemap_trees_spatial_df$stability_weight,probs=0.9)
  ) %>% 
  dplyr::count(cell) %>% 
  dplyr::slice_max(order_by = n, n = 10, with_ties = F) %>% 
  dplyr::slice_sample(n=1) %>% 
  dplyr::pull(cell)

aoi_temp <- 
  treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  dplyr::filter(cell == focal_cell_temp) %>% 
  sf::st_union() %>% 
  sf::st_centroid() %>% 
  sf::st_buffer(terra::res(treemap_data_loader_ans$treemap_rast)[1]) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc()

treemap_data_loader_ans$treemap_trees_spatial_df %>% 
  sf::st_intersection(aoi_temp) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(x,y,tree_height_m,rel_tree_height_m) %>% 
  # tidyr::pivot_longer(cols = -c(x,y)) %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(x=x,y=y)
  ) + 
  ggplot2::geom_point(mapping = ggplot2::aes(color = rel_tree_height_m), size = 4) +
  ggplot2::scale_color_distiller(palette = "Greys", direction = 1) +
  ggnewscale::new_scale_color() +
  ggplot2::geom_point(mapping = ggplot2::aes(color = tree_height_m), size = 1.8) +
  ggplot2::scale_color_distiller(palette = "Oranges", direction = 1) +
  ggplot2::theme_void()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Predict Species/Genus Application

let's implement a random forest model majority voting for species with the objective to reduce estimation time (compared to a Bayesian Softmax model)

tree species will be our dependent variable and the most simple model will use tree height and tree location (X and Y coordinates) as independent variables. The training FIA tree list data (identified from the TreeMap imputed plots of the area of interest) includes tree height and we'll use the centroid of the TreeMap 30m raster cell to represent the individual tree X and Y location (i.e. each tree within the same cell will have the same coordinate). Considering the combined tree height and spatial location via a full interaction model (i.e. `height*x*y`) allows the model to capture how the species-height relationship varies across space.

let's estimate the RF using data subsampling and model averaging for a computationally efficient way to get estimates over broad areas with potentially large tree lists. Model averaging can improve the robustness and accuracy of random forest models, especially when dealing with large datasets.

```{r}
# do it with our data
pred_trees_list$USFS_PenobscotEF <- 
  get_species_symbol_predictions(
    treemap_data_loader_ans = treemap_data_loader_ans
    , predict_df = pred_trees_list$USFS_PenobscotEF
  ) %>% 
  # for this project we'll just rename to maintain consistent across data sources
  dplyr::rename(species_symbol=predicted_species_symbol)

# pred_trees_list$USFS_PenobscotEF %>%
#   sf::st_drop_geometry() %>%
#   dplyr::count(species_symbol) %>%
#   dplyr::arrange(desc(n)) %>%
#   dplyr::mutate(pct=scales::percent(n/sum(n), accuracy=0.1))
```

let's check out the distribution of the predicted species

```{r}
pred_trees_list$USFS_PenobscotEF %>%
  dplyr::inner_join(
    ref_plots_list$USFS_PenobscotEF %>% 
      dplyr::slice(1) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(plot_id)
    , by = "plot_id"
  ) %>% 
  dplyr::arrange(desc(species_symbol)) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::ungroup() %>% 
  dplyr::count(species_symbol) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    pct = n/sum(n)
    , lab = ifelse(
      pct>0.0005
      , paste0(
        scales::percent(pct,accuracy=1) 
        ,"\n("
        , scales::comma(n,accuracy=1) 
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct, x = species_symbol)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = species_symbol)
    , width = 0.6
    , color = NA, alpha = 0.8
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab)
    , color = "black", size = 2.5, vjust = -0.2
  ) +
  ggplot2::scale_fill_viridis_d(option = "turbo") +
  ggplot2::scale_x_discrete(drop = F) +
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.15))
  ) +
  ggplot2::labs(
    x = "", y = "", fill = ""
    , subtitle = "Predicted Species"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 7, angle = 45, hjust = 1, vjust = 1)
    , axis.text.y = ggplot2::element_blank()
    , axis.ticks.y = ggplot2::element_blank()
  )
```

make sure the reference species data is clean

```{r}
# update the reference tree list
ref_trees_list$USFS_PenobscotEF <- 
  ref_trees_list$USFS_PenobscotEF %>% 
  clean_ref_species_fn(
    my_treemap_data_loader_ans = treemap_data_loader_ans
  )
```

what is the distribution of the cleaned reference field data species?

```{r}
ref_trees_list$USFS_PenobscotEF %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species_symbol) %>% 
  dplyr::arrange(desc(n))
```

## Species/Genus Accuracy Evaluation

### Tree-Level Accuracy

we're going to use the tree-level prediction and reference data for all TP (true positive) predictions to perform confusion matrix accuracy evaluation of the species predictions. 

using the confusion matrix evaluation we'll calculate the following metrics: 

* **Overall Accuracy**: Measures the global percentage of all correctly classified trees out of the entire dataset of true positive (TP) observations. Provides an overall summary metric but can mask performance on uncommon species.
* **User's Accuracy** (precision): Measures the reliability of specific predicted species. For example, if the method predicts a tree is a ponderosa pine, how often it is actually true.
* **Producer's Accuracy** (recall): Measures how well the model successfully aligns with what was measured on the ground. For example, out of all real ponderosa pine trees measured in the field, what percentage did the method correctly identify.

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's make a function to get the confusion matrix and those accuracy metrics

```{r}
# test that
confusion_matrix_species_temp <- confusion_matrix_tree_class_fn(
  pred_ref_match = pred_ref_tree_match_list$USFS_PenobscotEF
  , pred_trees = pred_trees_list$USFS_PenobscotEF
  , ref_trees = ref_trees_list$USFS_PenobscotEF
  , use_genus = F
)
confusion_matrix_genus_temp <- confusion_matrix_tree_class_fn(
  pred_ref_match = pred_ref_tree_match_list$USFS_PenobscotEF
  , pred_trees = pred_trees_list$USFS_PenobscotEF
  , ref_trees = ref_trees_list$USFS_PenobscotEF
  , use_genus = T
)
```

let's check out the confusion matrix at the *species* level

```{r, results='asis'}
confusion_matrix_species_temp$kbl_html_table
```

and the confusion matrix at the *genus* level

```{r, results='asis'}
confusion_matrix_genus_temp$kbl_html_table
```

### Site-Level Accuracy

we shift away from individual item matching and use the Bray-Curtis Dissimilarity Index. This metric compares the proportional compositions of each species predicted by the model against the reference proportions within a defined area or site.

```{r}
# test that
similarity_tree_species_temp <- similarity_tree_class_fn(
  pred_trees = pred_trees_list$USFS_PenobscotEF
  , ref_trees = ref_trees_list$USFS_PenobscotEF
  , use_genus = F
)
similarity_tree_genus_temp <- similarity_tree_class_fn(
  pred_trees = pred_trees_list$USFS_PenobscotEF
  , ref_trees = ref_trees_list$USFS_PenobscotEF
  , use_genus = T
)

```

let's check out the proportional comparisons at the *species* level

```{r, results='asis'}
similarity_tree_species_temp$kbl_html_comp_table
# similarity_tree_species_temp$species_comp_df %>% dplyr::glimpse()
```

also, the site-wide *species* compositional accuracy metrics

```{r}
similarity_tree_species_temp$kbl_html_sim_table
```

let's check out the proportional comparisons at the *genus* level

```{r, results='asis'}
similarity_tree_genus_temp$kbl_html_comp_table
# similarity_tree_genus_temp$genus_comp_df %>% dplyr::glimpse()
```

also, the site-wide *genus* compositional accuracy metrics

```{r}
similarity_tree_genus_temp$kbl_html_sim_table
```

plot the proportion of total by *species*

```{r}
similarity_tree_species_temp$species_comp_df %>% 
  dplyr::select(species_symbol, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_symbol
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "reference")
    , species_symbol = forcats::fct_reorder(species_symbol, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_symbol)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_symbol, fontface = "bold")
    # , color = "gray90"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_viridis_d(option = "turbo", begin=0.05, end = 0.95) +
  ggplot2::labs(
    subtitle = "Species Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

plot the proportion of total by *genus*

```{r}
similarity_tree_species_temp$species_comp_df %>% 
  dplyr::group_by(species_genus) %>% 
  dplyr::summarise(dplyr::across(c(pred_pct,ref_pct),sum)) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(species_genus, pred_pct, ref_pct) %>% 
  tidyr::pivot_longer(
    cols = c(pred_pct, ref_pct)
  ) %>% 
  dplyr::mutate(
    lab = ifelse(
      value>0.05
      , paste0(
        species_genus
        ,"\n("
        , scales::percent(value,accuracy=1)
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::mutate(
    name = ifelse(name == "pred_pct", "predicted", "reference")
    , species_genus = forcats::fct_reorder(species_genus, value, .fun = sum)
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = name, y = value, fill = species_genus)
  ) +
  ggplot2::geom_col(width = 0.5, color = NA, alpha = 0.9) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = species_genus, fontface = "bold")
    # , color = "gray90"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, breaks = scales::breaks_extended(n=9)) +
  ggplot2::scale_fill_viridis_d(option = "magma", begin = 0.25) +
  ggplot2::labs(
    subtitle = "Genus Composition Comparison"
    , x = ""
    , y = ""
    , fill = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(legend.position = "top")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

